### 2019-03-20 Rmd/02-plyranges.Rmd
inital commit, thanks @earowang!

# plyranges: a grammar of graphics for genomics data {#ch:plyranges}

```{r setup, include = FALSE}
library(tibble)
library(knitr)
opts_chunk$set(message = FALSE, 
               warning = FALSE, 
               comment = "#>", 
               fig.width = 5, 
               fig.height = 3, 
               fig.align = "center",
               fig.path = "./diagrams/")
# for peak figures
library(ggbio)

# for bam example
library(Biostrings)
library(plyranges)

# -- Data from the Human Epigenomics RoadMap consortium
# retrieving a BigWigFile from AnnotationHub

if (!dir.exists("./data")) dir.create("./data")

library(AnnotationHub)
q <- AnnotationHub()
# T-cell BW file
bw_file <- q[["AH33458"]]

# a BamFile from the H1 cell line
h1_bam <- "./data/GSM433167_BI.H3K27me3.bam"
if (!file.exists(h1_bam))
  download.file(
    url = "ftp://ftp.ncbi.nlm.nih.gov/geo/samples/GSM433nnn/GSM433167/suppl/GSM433167%5FBI%2EH3K27me3%2Ebam",
    destfile = h1_bam
  )

h1_bam_sorted <- "./data/GSM433167_BI.H3K27me3.sorted.bam"

# sort and index bam
if (!file.exists(h1_bam_sorted)) {
  Rsamtools::sortBam(h1_bam, "./data/GSM433167_BI.H3K27me3.sorted")
  Rsamtools::indexBam(h1_bam_sorted)
}

# array data for h1 cell line
# illumina annotations
array_info <- "data/GPL18952_HumanOmni25M-8v1-1_B.annotated.txt.gz"
if (!file.exists(array_info)) 
  download.file(
    url = "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GPL18952&format=file&file=GPL18952%5FHumanOmni25M%2D8v1%2D1%5FB%2Eannotated%2Etxt%2Egz",
    destfile = array_info
  )

snp_info <- "data/GSM1463263_JS-ESCH1.txt.gz"
if (!file.exists(snp_info))
  download.file(
    url = "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSM1463263&format=file&file=GSM1463263%5FJS%2DESCH1%2Etxt%2Egz",
    destfile = snp_info
  )

```
# Background

High-throughput genomics promises to unlock new disease therapies,
and strengthen our knowledge of basic biology. To deliver on those promises, 
scientists must derive a stream of knowledge from a deluge of data. Genomic 
data is challenging in both scale and complexity. Innovations in sequencing 
technology often outstrip our capacity to process the output. Beyond their common
association with genomic coordinates, genomic data are heterogeneous,
consisting of raw sequence read alignments, genomic feature
annotations like genes and exons, and summaries like coverage vectors,
ChIP-seq peak calls, variant calls, and per-feature read
counts. Genomic scientists need software tools to wrangle the
different types of data, process the data at scale, test hypotheses,
and generate new ones, all while focusing on the biology, not the
computation. For the tool developer, the challenge is to define ways
to model and operate on the data that align with the mental model of
scientists, and to provide an implementation that scales with their
ambition.

Several domain specific languages (DSLs) enable scientists to process and reason
about heterogeneous genomics data by expressing common
operations, such as range manipulation and overlap-based joins, using
the vocabulary of genomics. Their implementations either delegate
computations to a database, or operate over collections of files in
standard formats like BED. An example of the former is the Genome Query 
Language (GQL) and its distributed implementation GenAp which use a 
SQL-like syntax for fast retrieval of information of unprocessed 
sequencing data [@Kozanitis2014-va; @Kozanitis2016-bm]. Similarly, the Genometric 
Query Language (GMQL) implements a DSL for combining genomic datasets 
\cite{Kaitoua2017-pw}.  The command line application BEDtools develops an extensive 
algebra for performing arithmetic between two or more sets of genomic regions 
\cite{Quinlan2010-gc}. All of the aforementioned DSLs are designed to be evaluated
either at the command line or embedded in scripts for batch processing. They 
exist in a sparse ecosystem, mostly consisting of UNIX and database tools that
lack biological semantics and operate at the level of files and
database tables.

The Bioconductor/R packages `IRanges` and `GenomicRanges` 
[@r-core; @Lawrence2013-wg; @Huber2015-ei] define a DSL for 
analyzing genomics data with R, an interactive data analysis environment that 
encourages reproducibility and provides high-level abstractions for manipulating,
modelling and plotting data, through state of the art methods in
statistical computing. The packages define object-oriented (OO)
abstractions for representing genomic data and enable
interoperability by allowing users and developers to use these
abstractions in their own code and packages. Other genomic DSLs 
that are embedded in programming languages include pybedtools and valr 
[@Dale2011-js; @Kent2017], however these packages lack the interoperability
provided by the aforementioned Bioconductor packages and are not easily 
extended.

The Bioconductor infrastructure models the genomic data and operations from the
perspective of the power user, one who understands and wants to take
advantage of the subtle differences in data types. This design has
enabled the development of sophisticated tools, as evidenced by the
hundreds of packages depending on the framework. Unfortunately, the
myriad of data structures have overlapping purposes and important but
obscure differences in behavior that often confuse the typical end
user.

Recently, there has been a concerted, community effort to standardize
R data structures and workflows around the notion of tidy data
\cite{Wickham2014-jc}. A tidy dataset is defined as a tabular
data structure that has observations as rows and columns as variables,
and all measurements pertain to a single observational unit.
The tidy data pattern is useful because it allows 
us to see how the data relate to the design of an experiment
and the variables measured. The `dplyr` package \cite{Wickham2017-dplyr} 
defines an application programming interface (API) that maps notions from 
the general relational algebra to verbs that act on tidy data. These verbs
can be composed together on one or more tidy datasets with the pipe operator 
from the `magrittr` package \cite{R-magrittr}. Taken together these features 
enable a user to write human readable analysis workflows.

We have created a genomic DSL called `plyranges` that reformulates
notions from existing genomic algebras and embeds them in R as a
genomic extension of `dplyr`. By analogy, `plyranges` is to the
genomic algebra, as `dplyr` is to the relational algebra.  The
`plyranges` Bioconductor package implements the language on top of a
key subset of Bioconductor data structures and thus fully integrates
with the Bioconductor framework, gaining access to its scalable data
representations and sophisticated statistical methods.

# Results

## Genomic Relational Algebra

### Data Model

\begin{figure}
\includegraphics[width=\textwidth]{diagrams/GRanges}
\caption{An illustration of the GRanges data model for a
sample from an RNA-seq experiment. The core components of the data model
include a seqname column (representing the chromosome), a ranges column
which consists of start and end coordinates for a genomic region, and a
strand identifier (either positive, negative, or unstranded). Metadata
are included as columns to the right of the dotted line as annotations
(gene\_id) or range level covariates (score).}
\label{fig:GRanges} 
\end{figure}

The `plyranges` DSL is built on the core Bioconductor data structure
GRanges, which is a constrained table, with fixed columns
for the chromosome, start and end coordinates, and the strand, along
with an arbitrary set of additional columns, consisting of
measurements or metadata specific to the data type or experiment
(figure \ref{fig:GRanges}).  GRanges balances flexibility with formal
constraints, so that it is applicable to virtually any genomic
workflow, while also being semantically rich enough to support
high-level operations on genomic ranges. As a core data structure,
GRanges enables interoperability between `plyranges` and the rest of
Bioconductor. Adhering to a single data structure simplifies the API
and makes it easier to learn and understand, in part because
operations become endomorphic, i.e., they return the same type as
their input.

GRanges follow the intuitive tidy data pattern: it is a rectangular
table corresponding to a single biological context. Each row contains
a single observation and each column is a variable describing the
observations.  GRanges specializes the tidy pattern in that the
observations always pertain to some genomic feature, but it largely
remains compatible with the general relational operations defined by
`dplyr`. Thus, we define our algebra as an extension of the `dplyr`
algebra, and borrow its syntax conventions and design principles.

\begin{table}[!htbp]
\centering
\begin{tabular}{|l|l|p{6cm}|}
  \hline
  & Verb &  Description \\ 
  \hline
   & \textbf{\emph{summarize()}} & aggregate over column(s) \\ 
   Aggregate & \emph{disjoin\_ranges()} & aggregate column(s) over the union of end coordinates \\
   &  \emph{reduce\_ranges()} & aggregate column(s) by merging overlapping and neighboring ranges \\
   \hline
   &  \textbf{\emph{mutate()}} & modifies any column \\
   & \textbf{\emph{select()}} & select columns \\
  Modify (Unary) & \textbf{\emph{arrange()}} & sort by columns \\
   & \emph{stretch()} & extend range by fixed amount \\
   &  \emph{shift\_(direction)} & shift coordinates \\
   & \emph{flank\_(direction)} & generate flanking regions \\
   & \emph{\%intersection\% } & row-wise intersection \\
   & \emph{\%union\%} & row-wise union \\
   & \emph{compute\_coverage} & coverage over all ranges \\
  Modify (Binary) &  \emph{\%setdiff\%} & row-wise set difference \\
   & \emph{between()} & row-wise gap range \\
   & \emph{span()} & row-wise spanning range \\
   \hline
    & \emph{join\_overlap\_*()} & merge by overlapping ranges \\
    & \emph{join\_nearest} & merge by nearest neighbor ranges \\
    & \emph{join\_follow} & merge by following ranges \\
    Merge & \emph{join\_precedes} & merge by preceding ranges \\
    & \emph{union\_ranges} & range-wise union \\
    & \emph{intersect\_ranges} & range-wise intersect \\
    & \emph{setdiff\_ranges} & range-wise set difference \\
    & \emph{complement\_ranges} & range-wise set complement \\
  \hline
   & \emph{anchor\_direction()} & fix coordinates at direction \\
  Operate & \textbf{\emph{group\_by()}} & partition by column(s)  \\ 
   & \emph{group\_by\_overlaps()} & partition by overlaps \\
   \hline
   & \textbf{\emph{filter()}} & subset rows \\
  Restrict & \emph{filter\_by\_overlaps()} & subset by overlap \\
    & \emph{filter\_by\_non\_overlaps()} & subset by no overlap \\
   \hline
\end{tabular}
\caption{Overview of the \texttt{plyranges} grammar. The core verbs are
briefly described and categorized into one of the following higher level 
categories: aggregate, modify, merge, operate, or restrict. A verb is given bold text if
its origin is from the \texttt{dplyr} grammar.}\label{tab:grammar}
\end{table}

### Algebraic operations

The `plyranges` DSL defines an expressive algebra for performing
genomic operations with and between GRanges objects (see table
\ref{tab:grammar}). The grammar includes several classes of operation
that cover most use cases in genomics data analysis. There are
range arithmetic operators, such as for resizing ranges or finding their
intersection, and operators for merging, filtering and aggregating by
range-specific notions like overlap and proximity.

Arithmetic operations transform range coordinates, as defined by their
_start_, _end_ and _width_. The three dimensions are mutually
dependent and partially redundant, so direct manipulation of them is
problematic. For example, changing the _width_ column needs to change
either the _start_, _end_ or both to preserve integrity of the
object. We introduce the _anchor_ modifier to disambiguate these
adjustments. Supported anchor points include the start, end and
midpoint, as well as the 3' and 5' ends for strand-directed
ranges. For example, if we anchor the start, then setting the width
will adjust the end while leaving the start stationary.

The algebra also defines conveniences for relative coordinate
adjustments: _shift_ (unanchored adjustment to both start and end) and
_stretch_ (anchored adjustment of width). We can perform any relative
adjustment by some combination of those two operations.  The _stretch_
operation requires an anchor and assumes the midpoint by
default. Since _shift_ is unanchored, the user specifies a suffix for
indicating the direction: left/right or, for stranded features,
upstream/downstream. For example, _shift\_right_ shifts a range to the
right.

The _flank_ operation generates new ranges that are adjacent to
existing ones. This is useful, for example, when generating upstream
promoter regions for genes. Analogous to _shift_, a suffix indicates
the side of the input range to flank.

As with other genomic grammars, we define set operations that treat
ranges as sets of integers, including _intersect_, _union_,
_difference_, and _complement_. There are two sets of these: parallel
and merging. For example, the parallel intersection (_x %intersect% y_) finds the
intersecting range between _xi_ and _yi_ for _i_ in _1...n_, where _n_
is the length of both _x_ and _y_. In contrast, the merging
intersection (_intersect\_ranges(x, y)_) returns a new set of disjoint
ranges representing wherever there was overlap between a range in _x_
and a range in _y_. Finding the parallel union will fail when two ranges
have a gap, so we introduce a _span_ operator that takes the union
while filling any gap. The _complement_ operation is unique in that it
is unary. It finds the regions not covered by any of the ranges in a
single set. Closely related is the _between_ parallel operation, which
finds the gap separating _xi_ and _yi_. The binary operations are
callable from within arithmetic, restriction and aggregation
expressions.

\begin{figure}
\includegraphics[width=\textwidth]{diagrams/olap-fig}
\caption{Illustration of the three overlap join operators. Each join takes two 
GRanges objects, \textit{x} and \textit{y} as input. A 'Hits' object for the 
join is computed which consists of two components. The first component contains 
the indices of the ranges in \textit{x} that have been overlapped 
(the rectangles of \textit{x} that cross the orange lines). The second 
component consists of the indices of the ranges in \textit{y} that overlap 
the ranges in \textit{x}. In this case a range in \textit{y} overlaps the 
ranges in \textit{x} three times, so the index is repeated three times. The 
resulting 'Hits' object is used to modify \textit{x} by where it was 'hit' by
\textit{y} and merge all metadata columns from \textit{x} and \textit{y} 
based on the indices contained in the 'Hits' object. This procedure is applied 
generally in the \texttt{plyranges} DSL for both overlap and nearest neighbor
operations. The join semantics alter what is returned: \textbf{A}: for an 
\textbf{inner} join the \textit{x} ranges that are overlapped by \textit{y} 
are returned. The returned ranges also include the metadata from the \textit{y}
range that overlapped the three \textit{x} ranges. \textbf{B} An 
\textbf{intersect} join is identical to an inner join except that the 
intersection is taken between the overlapped \textit{x} ranges and the 
\textit{y} ranges. \textbf{C} For the \textbf{left} join all 
\textit{x} ranges are returned regardless of whether they are overlapped by 
\textit{y}. In this case the third range (rectangle with the asterisk next to it)
of the join would have missing values on metadata columns that came from \textit{y}.}
\label{fig:olaps-fig}
\end{figure}

To support merging, our algebra recasts finding overlaps or 
nearest neighbors between two
genomic regions as variants of the relational join operator. A join
acts on two GRanges objects: _x_ and _y_. The join operator
is relational in the sense that metadata from the _x_ and _y_
ranges are retained in the joined range.  All join operators in the
`plyranges` DSL generate a set of hits based on overlap or proximity
of ranges and use those hits to merge the two datasets in different
ways. There are four supported matching algorithms: _overlap_,
_nearest_, _precede_, and _follow_ (figure 
\ref{fig:olaps-fig}). We can further restrict the
matching by whether the query is completely _within_ the subject, and
adding the _directed_ suffix ensures that matching ranges have the
same direction (strand).

For merging based on the hits, we have three modes: _inner_,
_intersect_ and _left_. The _inner_ overlap join is similar to the
conventional inner join in that there is a row in the result for every
match. A major difference is that the matching is not by identity, so
we have to choose one of the ranges from each pair. We always choose
the left range. The _intersect_ join uses the intersection instead of
the left range. Finally, the overlap _left_ join is akin to left outer
join in Codd's relational algebra: it performs an overlap inner join
but also returns all _x_ ranges that are not hit by the _y_ ranges.

\begin{figure}
\includegraphics[width=\textwidth]{diagrams/code-comparison.pdf}
\caption{Idiomatic code examples for \texttt{plyranges} (\textbf{A}) and 
\texttt{GenomicRanges} (\textbf{B}) illustrating an overlap and aggregate 
operation that returns the same result. 
In each example, we have two BED files consisting of SNPs that are
genome-wide association study (GWAS) hits and reference exons. Each code block
counts for each SNP the number of distinct exons it overlaps. The 
\texttt{plyranges} code achieves this with an overlap join followed by partitioning
and aggregation. Strand is ignored by default here. 
The \texttt{GenomicRanges} code achieves this using the 'Hits' and 'List' 
classes and their methods.}
\label{fig:code}
\end{figure}

Since the GRanges object is a tabular data structure, our grammar includes
operators to filter, sort and aggregate by columns in a GRanges. These 
operations can be performed over partitions formed
using the *group_by* modifier. Together with our algebra for arithmetic and
merging, these operations conform to the semantics and syntax of the 
`dplyr` grammar. Consequently, `plyranges` code is generally more compact
than the equivalent `GenomicRanges` code (figure \ref{fig:code}).

## Developing workflows with `plyranges`

Here we provide illustrative examples of using the `plyranges` DSL to show
how our grammar could be integrated into genomic data workflows. As we
construct the workflows we show the data output intermittently to assist
the reader in understanding the pipeline steps. The workflows
highlight how interoperability with existing Bioconductor infrastructure, 
enables easy access to public datasets and methods for analysis and 
visualization.

### Peak Finding

In the workflow of ChIP-seq data analysis, we are interested in finding peaks 
from islands of coverage over chromosome. Here we will use `plyranges`
to call peaks from islands of coverage above 8 then plot the region
surrounding the tallest peak.

Using `plyranges` and the the Bioconductor package `AnnotationHub` \cite{R-ahub}
we can download and read BigWig files from ChIP-Seq experiments from 
the Human Epigenome Roadmap project \cite{Roadmap-Epigenomics-Consortium2015-pr}. 
Here we analyse a BigWig file corresponding to H3 lysine 27 trimethylation 
(H3K27Me3) of primary T CD8+ memory cells from peripheral blood, focussing
on coverage islands over chromosome 10.

First, we extract the genome information from the BigWig file and filter
to get the range for chromosome 10. This range will be used as a filter when 
reading the file.

```{r load-bw, cache = TRUE}
library(plyranges)
chr10_ranges <- bw_file %>% 
  get_genome_info() %>%
  filter(seqnames == "chr10")
```

Then we read the BigWig file only extracting scores if they overlap chromosome
10. We also add the genome build information to the resulting ranges. This
book-keeping is good practice as it ensures the integrity of any 
downstream operations such as finding overlaps.

```{r}
chr10_scores <- bw_file %>%
  read_bigwig(overlap_ranges = chr10_ranges) %>%
  set_genome_info(genome = "hg19")
chr10_scores
```

We then filter for regions with a coverage score greater than 8, and following
this reduce individual runs to ranges representing the islands of coverage.
This is achieved with the `reduce_ranges()` function, which allows a summary
to be computed over each island: in this case we take the maximum of 
the scores to find the coverage peaks over chromosome 10.

```{r}
all_peaks <- chr10_scores %>% 
  filter(score > 8) %>% 
  reduce_ranges(score = max(score))
all_peaks
```

Returning to the GRanges object containing normalized coverage scores, 
we filter to find the coordinates of the peak containing the maximum coverage 
score. We can then find a 5000 nt region centered around the maximum position 
by anchoring and modifying the width.

```{r max-score-region}
chr10_max_score_region <- chr10_scores %>%
  filter(score == max(score)) %>% 
  anchor_center() %>%
  mutate(width = 5000)
```

Finally, the overlap inner join is used to restrict the chromosome 10
coverage islands, to the islands that are contained in the 5000nt region that surrounds the max peak (figure \ref{fig:peak-viz}).

```{r peak_region}
peak_region <- chr10_scores %>%
  join_overlap_inner(chr10_max_score_region)
peak_region
```

```{r peak-viz, echo = FALSE, fig.cap = "The final result of the \\texttt{plyranges} operations to find a 5000nt region surrounding the peak of normalised coverage scores over chromosome 10, displayed as a density plot.", cache = TRUE}
ggbio::autoplot(peak_region, 
                aes(y = score.x), 
                geom = "blank") + 
  geom_area() +
  ylab("score")
```

### Computing Windowed Statistics

Another common operation in genomics data analysis is to compute data
summaries over genomic windows. In `plyranges` this can be achieved
via the `group_by_overlaps()` operator.  We bin and count and
find the average GC content of reads from a
H3K27Me3 ChIP-seq experiment by the Human Epigenome Roadmap
Consortium.

We can directly obtain the genome information from the header of the
BAM file: in this case the reads were aligned to the hg19 genome build
and there are no reads overlapping the mitochondrial genome.

```{r bins}
locations <- h1_bam_sorted %>%
  read_bam() %>% 
  get_genome_info() 
```

Next we only read in alignments that overlap the genomic locations
we are interested in and select the query sequence. Note that
the reading of the BAM file is deferred: only alignments that pass the filter 
are loaded into memory. We can add another column representing the GC proportion 
for each alignment using the `letterFrequency()` function from the `Biostrings`
package \cite{R-biostrings}. After computing the GC proportion as the score
column, we drop  all other columns in the GRanges object.

```{r, cache = TRUE}
alignments <- h1_bam_sorted %>% 
  read_bam() %>%
  filter_by_overlaps(locations) %>%
  select(seq) %>%
  mutate(
    score = as.numeric(letterFrequency(seq, "GC", as.prob = TRUE))
  ) %>% 
  select(score)
alignments
```

Finally, we create 10000nt tiles over the genome and compute the number
of reads and average GC content over all reads that fall within
each tile using an overlap join and merging endpoints.

```{r, cache = TRUE}
bins <- locations %>% 
  tile_ranges(width = 10000L)

alignments_summary <- bins %>%
  join_overlap_inner(alignments) %>% 
  disjoin_ranges(n = n(), avg_gc = mean(score))
alignments_summary
```
<!-- 
https://support.bioconductor.org/p/71601/ also useful example?
-->
### Quality Control Metrics

```{r, echo = FALSE, warning = FALSE, cache = TRUE}
# prepare a GRanges object from array data
library(readr)
# annotation information available here
# https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GPL18952
# chromosome zero are 'problem' probes and should be filtered
anno <- read_tsv(array_info,
                 trim_ws = TRUE,
                 col_types = c("cccc----")) %>%
  dplyr::filter(Chr != "0")

# genotypes for H1 cell line, data starts on line 10
# data obtained from https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM1463263
geno <- read_tsv(snp_info, skip = 10)

# rename columns
geno <- geno %>% 
  dplyr::select(name = `SNP Name`, 
         score = `GC Score`, 
         baf = `B Allele Freq`, 
         logR = `Log R Ratio`)

# merge everything
complete_vranges_df <- dplyr::inner_join(geno, anno, by = c("name" = "Name")) %>%
  dplyr::rename(alleles = Alleles, seqnames = Chr, start = MapInfo)

h1_snp_array <- as_granges(complete_vranges_df,
                           start = as.integer(start),
                           end = as.integer(start))
h1_snp_array <- h1_snp_array %>% 
  mutate(ref = stringr::str_replace(alleles, "\\[(.)/.*", "\\1"),
         alt = stringr::str_replace(alleles, ".*/(.)\\]", "\\1"))
```

We have created a GRanges object from genotyping performed on the
H1 cell line, consisting of approximately two million single nucleotide
polymorphisms (SNPs) and short insertion/deletions (indels). The GRanges object consists
of 7 columns, relating to the alleles of a SNP or indel, the B-allele frequency,
log relative intensity of the probes, GC content score over a probe, and the
name of the probe. We can use this information to compute the transition-transversion
ratio, a quality control metric, within each chromosome in GRanges object.

First we filter out the indels and mitochondrial variants. Then we
create a logical vector corresponding to whether there is a transition
event.

```{r, cache = TRUE}
h1_snp_array <- h1_snp_array %>%
  filter(!(ref %in% c("I", "D")), seqnames != "M") %>%
  mutate(transition = (ref %in% c("A", "G") & alt %in% c("G","A"))|
                      (ref %in% c("C","T") & alt %in% c("T", "C")))
```

We then compute the transition-transversion ratio over each chromosome using
`group_by()` in combination with `summarize()` (figure \ref{fig:titv-viz}).

```{r, cache = TRUE}
ti_tv_results <- h1_snp_array %>% 
  group_by(seqnames) %>%
  summarize(n_snps = n(),
            ti_tv = sum(transition) / sum(!transition)) 
ti_tv_results
```

```{r titv-viz, echo = FALSE, fig.cap = "The final result of computing quality control metrics over the SNP array data with \\texttt{plyranges}, displayed as a dot plot. Chromosomes are ordered by their estimated transition-transversion ratio. A white reference line is drawn at the expected ratio for a human exome."}
ti_tv_results %>% 
  as.data.frame() %>% 
  ggplot(aes(forcats::fct_reorder(seqnames, ti_tv), ti_tv)) +
  geom_point() + 
  geom_hline(yintercept = 3, colour = "white", size = 4) +
  labs(x = "Chromosome", y = "Transition Transversion Ratio") +
  coord_flip() 
```
<!-- 
Trivial example? Could do something with a smooth summarised over
genomic coordinates? 
-->

# Discussion

The design of `plyranges` adheres to well understood principles of
language and API design: cognitive consistency, cohesion,
endomorphism and expressiveness \cite{Green1996-qg}. 
To varying degrees, these principles also underlie the
design of `dplyr` and the Bioconductor infrastructure.

We have aimed for `plyranges` to have a simple and direct mapping to the
user's cognitive model, i.e., how the user thinks about the data. This
requires careful selection of the level of abstraction so that the
user can express workflows in the language of genomics. This motivates
the adoption of the tidy GRanges object as our central data
structure. The basic data.frame and `dplyr` tibble lack any notion of
genomic ranges and so could not easily support our genomic grammar,
with its specific verbs for range-oriented data manipulation. Another
example of cognitive consistency is how `plyranges` is insensitive to
direction/strand by default when, e.g., detecting
overlaps. `GenomicRanges` has the opposite behavior. We believe that
defaulting to purely spatial overlap is most intuitive to most users.

To further enable cognitive consistency, `plyranges` functions are cohesive. 
A function is defined to be cohesive if it performs a singular task without
producing any side-effects. Singular tasks can always be broken 
down further at lower levels of abstraction.
For example, to resize a range, the user needs to specify which
position (start, end, midpoint) should be invariant over the
transformation. The `resize()` function from the `GenomicRanges` package
has a `fix` argument that sets the anchor, so calling `resize()`
coalesces anchoring and width modification. The coupling at the
function call level is justified since the effect of setting the width
depends on the anchor. However, `plyranges` increases cohesion and
decouples the anchoring into its own function call.

Increasing cohesion simplifies the interface to each operation, makes
the meaning of arguments more intuitive, and relies on function names
as the primary means of expression, instead of a more complex mixture
of function and argument names. This results in the
user being able to conceptualize the `plyranges` DSL as a flat catalog of 
functions, without having to descend further into documentation to understand
a function's arguments. A flat function catalog also enhances API
discoverability, particularly through auto-completion in 
integrated developer environments (IDEs). One downside of pushing cohesion to 
this extreme is that function calls become coupled, and care is 
necessary to treat them as a group when modifying code.

Like `dplyr`, `plyranges` verbs are functional: they are free of side
effects and are generally endomorphic, meaning that when the input is a GRanges
object they return a GRanges object. This enables chaining of verbs
through syntax like the forward pipe operator from the `magrittr` package. 
This syntax has a direct cognitive mapping to natural language and the intuitive notion of
pipelines. The low-level object-oriented APIs of Bioconductor tend to
manipulate data via sub-replacement functions, like `start(gr) <-
x`. These ultimately produce the side effect of replacing a symbol
mapping in the current environment and thus are not amenable to so-called 
fluent syntax.

Expressiveness relates to the information content in code: the
programmer should be able to clarify intent without unnecessary
verbosity. For example, our overlap-based join operations are more
concise than the multiple steps necessary to achieve the same effect
in the original `GenomicRanges` API. In other cases, the `plyranges` API
increases verbosity for the sake of clarity and cohesion. Explicitly
calling _anchor()_ can require more typing, but the code is easier to
comprehend. Another example is the set of routines for importing
genomic annotations, including _read\_gff()_, _read\_bed()_, and
_read\_bam()_. Compared to the generic _import()_ in `rtracklayer`, the
explicit format-based naming in `plyranges` clarifies intent and the
type of data being returned. Similarly, every `plyranges` function that
computes with strand information indicates its intentions by including
suffixes such as _directed_, _upstream_ or _downstream_ in its name,
otherwise strand is ignored. The `GenomicRanges` API does not make this
distinction explicit in its function naming, instead relying on a
parameter that defaults to strand sensitivity, an arguably confusing
behavior.

The implementation of `plyranges` is built on top of 
Bioconductor infrastructure, meaning most functions are constructed
by composing generic functions from core Bioconductor packages. As a result, any 
Bioconductor packages that uses data structures that inherit from GRanges will
be able to use `plyranges`  for free. Another consequence of building on top of
Bioconductor generics is that the speed and memory usage of `plyranges` 
functions are similar to the highly optimized methods implemented in Bioconductor
for GRanges objects.

A caveat to constructing a compatible interface with `dplyr` is that
`plyranges` makes extensive use of non-standard evaluation in R via
the `rlang` package \cite{R-rlang}. Simply, this means that computations are
evaluated in the context of the GRanges objects. Both `dplyr` and
`plyranges` are based on the `rlang` language, because it allows for more
expressive code that is free of repeated references to the
container. Implicitly referencing the container is particularly
convenient when programming interactively.  Consequently, when
programming with `plyranges`, a user needs to generally understand
the `rlang` language and how to adapt their code accordingly. Users
familiar with the tidyverse should already have such knowledge.

# Conclusion

We have shown how to create expressive and 
reproducible genomic workflows using the `plyranges` DSL. By
realising that the GRanges data model is tidy we have highlighted how to 
implement a grammar for performing genomic arithmetic, aggregation, restriction 
and merging. Our examples show that `plyranges` code is succinct, human
readable and can take advantage of the interoperability provided by
the Bioconductor ecosystem and the R language.

We also note that the grammar elements and design principles we have described
are programming language agnostic and could be easily be implemented in another
language where genomic information could be represented as a tabular data
structure. We chose R because it is what we are familiar with and because
the aforementioned Bioconductor packages have implemented the GRanges data
structure.

We aim to continue developing the `plyranges` package and to extend it
for use with more complex data structures, such as the
SummarizedExperiment class, the core Bioconductor data structure for
representing experimental results (e.g., counts) from multiple sample
experiments in conjunction with feature and sample metadata. Although,
the SummarizedExperiment is not strictly tidy, it does consist of 
three tidy data structures that are related by feature and sample identifiers. 
Therefore, the grammar and design of the `plryanges` DSL is naturally 
extensible to the SummarizedExperiment.

As the `plyranges` interface encourages tidy data practices, it
integrates well with the grammar of graphics \cite{Wickham2016-gz}.  To
achieve responsive performance, interactive graphics rely on lazy data
access and computing patterns, so the deferred mechanisms within
`plyranges` should help support interactive genomics applications.

# Declarations

## Acknowledgements

We would like to thank Dr Matthew Ritchie at the Walter and Eliza Hall 
Institute and Dr Paul Harrison for their feedback on earlier drafts of
this work. We would also like to thank Lori Shepherd and Hèrve Pages
for the code review they performed and users who have submitted feedback
and pull requests.

## Availabliity of data and Materials

The BigWig file for the H3K27Me3 primary T CD8+ memory cells from peripheral 
blood ChIP-seq data from the Human Roadmap Epigenomics project  
was downloaded from the `AnnotationHub` package (2.13.1)
under accession AH33458 \cite{R-ahub, Roadmap-Epigenomics-Consortium2015-pr}. 
The BAM file corresponding to the H1 cell line ChIP-seq data is available at 
NCBI GEO under accession [GSM433167](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM433167) \cite{ncbi-geo, Roadmap-Epigenomics-Consortium2015-pr}. 
The SNP array data for the H1 cell line data is available at  NCBI GEO under accession
[GPL18952](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GPL18952) 
\cite{Roadmap-Epigenomics-Consortium2015-pr}.

The `plyranges` package is open source under an Artistic 2.0 license 
\cite{R-plyranges-z}. 
The  version (1.1.10) used in the paper is archived under DOI: [10.5281/zenodo.1469841](https://doi.org/10.5281/zenodo.1469841). The software 
can be obtained via the Bioconductor project website 
[https://bioconductor.org](https://bioconductor.org) or accessed via Github 
[https://github.com/sa-lee/plyranges](https://github.com/sa-lee/plyranges).

To produce the workflows as described in results section we used 
R version 3.5 with the development version of `plyranges` (1.1.10)   
and the `BioStrings` package (2.49.0) installed \cite{r-core, R-plyranges-z, R-biostrings}.

This article was written with `knitr` \cite{R-knitr} and 
the figures were made with `ggbio` \cite{R-ggbio}. All code required to 
reproduce this article is available at https://github.com/sa-lee/plyranges-paper.

## Funding

SL is funded by an Australian Government Research Training Program (RTP) 
scholarship and receives an additional top up scholarship from CSL. Funding 
bodies had no role in data collection, analysis, and interpretation.

## Authors Contributions

SL, DC, ML contributed ideas towards the software. SL wrote 
the manuscript and analysed data. SL and ML wrote the software. 
DC and ML supervised the research and revised the manuscript. 
All authors have read and approved the final version of the manuscript.

## Ethics approval and consent to participate

Not applicable to the current manuscript.

## Competing Interests

Not applicable to the current manuscript.

# References {#references .unnumbered}

### 2019-03-20 index.Rmd
inital commit, thanks @earowang!

---
title: 'TBD'
author: 'Stuart Lee'
degreetype: 'Doctor of Philosophy'
site: bookdown::bookdown_site
degrees: 'B.MatComSci., University of Adelaide; MSc.Bioinformatics. University of Melbourne'
bibliography: [bib/plyranges.bib]
link-citations: yes
---

```{r initial, echo = FALSE, cache = FALSE, results = 'hide'}
options("knitr.graphics.auto_pdf" = TRUE)
library(knitr)
opts_chunk$set(
  warning = FALSE, message = FALSE, echo = FALSE,
  fig.path = 'figure/', fig.align = 'center', fig.show = 'hold',
  cache = TRUE, external = TRUE, comment = "#>", dpi = 300,
  out.width = ifelse(is_html_output(), "100%", "\\textwidth"),
  cache.path = ifelse(is_html_output(), 'cache-gitbook/', 'cache-pdfbook/')
)
```

`r if (knitr::is_latex_output()) '<!--'`

# Welcome {-}

This is the website for my PhD thesis at Monash University (Australia), titled "TBD". 
It is a work in progress.

`r if (knitr::is_latex_output()) '-->'`

### 2019-03-20 _render.R

@@ -1,0 +1,47 @@
+options(warn = 1)
+library(cliapp)
+start_app(theme = simple_theme())
+
+cli_h1("Preprocessing")
+# spelling check
+cli_h2("Spelling check")
+rmd_files <- list.files(pattern = "*.Rmd", full.names = TRUE, recursive = TRUE)
+wordlist <- readLines("WORDLIST")
+spell_gb <- spelling::spell_check_files(rmd_files, wordlist, "en_GB")
+spell_us <- spelling::spell_check_files(rmd_files, wordlist, "en_US")
+spell_res <- intersect(spell_gb[[1]], spell_us[[1]])
+if (length(spell_res) > 0) {
+  print(subset(spell_gb, spell_gb[[1]] %in% spell_res))
+  stop("Can you please fix typos listed above first?", call. = FALSE)
+}
+
+# convert pdf to png for html output
+cli_h2("Coverting pdf to png")
+img_pdf <- list.files("img/", pattern = "*.pdf")
+for (i in img_pdf) {
+  file_pdf <- paste0("img/", i)
+  dest_pdf <- paste0("img/", sub("pdf$", "png", i))
+  magick::image_write(
+    magick::image_read(file_pdf, 300), dest_pdf, "png", 
+    density = 300
+  )
+}
+
+cli_h1("Compiling")
+
+if (Sys.getenv("RSTUDIO") != "1" && Sys.info()['sysname'] == "Darwin") {
+  Sys.setenv('RSTUDIO_PANDOC' = '/Applications/RStudio.app/Contents/MacOS/pandoc')
+}
+
+# provide default formats if necessary
+formats <- commandArgs(trailingOnly = TRUE)
+if (length(formats) == 0) {
+  formats <- "bookdown::pdfbook"
+}
+# render the book to all formats
+for (fmt in formats)
+  bookdown::render_book("index.Rmd", fmt, quiet = FALSE)
+
+gif_file <- list.files("figure", "*.gif", full.names = TRUE)
+invisible(file.copy(gif_file, "_thesis/figure/animate.gif"))
+cli_alert_success("You Rock!")

### 2019-10-21 Rmd/00-c-declaration.Rmd
restructuring

`r if (knitr::is_html_output()) '<!--'`

# Declaration {-}

I hereby declare that this thesis contains no material which has been accepted for the award of any other degree or diploma in any university or equivalent institution, and that, to the best of my knowledge and belief, this thesis contains no material previously published or written by another person, except where due reference is made in the text of the thesis.

\vspace*{2cm}\par\authorname

`r if (knitr::is_html_output()) '-->'`

### 2019-10-21 Rmd/00-e-preface.Rmd
restructuring

# Preface {-}

\clearpage\pagenumbering{arabic}\setcounter{page}{0}

### 2019-11-05 Rmd/00-b-abstract.Rmd
first draft of abstract

# Abstract {-}

Exploratory data analysis  (EDA) is a vital element of the modern 
statistical workflow - it is an analyst’s first pass at 
understanding their data; revealing all it’s messes and (possibly) uncovering hidden insights. It is an iterative process that can 
lead to new hypotheses that can be tested and formalised using 
modelling. However, as data grows in complexity and there is an 
increase in the number of observations, and variables EDA becomes 
increasingly difficult.  This is certainly true in bioinformatic 
data analysis, where there are constraints computation time and 
memory, as well as a lack of guidelines or tooling for performing 
EDA.

This thesis presents a set of tools for performing exploratory 
data analysis in the context of genomics and transcriptomics.  The first tool, “plryanges” provides a domain specific language for 
reasoning about range based genomics data that simplifies common 
operations such as overlaps and aggregations.  Next I extend 
“plyranges”  to develop “superintronic” which presents an 
approach for discovering interesting genomic regions within an 
experimental design. We apply this tool to explore intron signal 
in transcriptomic data. Finally, I turn my attention to focus on 
embedding methods currently used for visualisation of large data 
sets…

\newpage

### 2019-11-05 index.Rmd
add (potential) title

@@
 ---
-title: 'TBD'
+title: 'Exploratory data analysis for high-dimensional biological datasets'
 author: 'Stuart Lee'
@@
   warning = FALSE, message = FALSE, echo = FALSE,
-  fig.path = 'figure/', fig.align = 'center', fig.show = 'hold',
+  fig.path = 'img/', fig.align = 'center', fig.show = 'hold',
   cache = TRUE, external = TRUE, comment = "#>", dpi = 300,

### 2020-01-28 Rmd/04-fluentGenomics.Rmd
add in fluentGenomics

## Fluent genomics: integrating data with plyranges and tximeta {#ch:fluentGenomics}

```{r setup, include = FALSE}
library(fluentGenomics)
dir <- system.file("extdata", package="macrophage")

library(tximeta)
makeLinkedTxome(
  indexDir=file.path(dir, "gencode.v29_salmon_0.12.0"),
  source="Gencode",
  organism="Homo sapiens",
  release="29",
  genome="GRCh38",
  fasta="ftp://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_human/release_29/gencode.v29.transcripts.fa.gz",
  gtf=file.path(dir, "gencode.v29.annotation.gtf.gz"), # local version
  write=FALSE
)

knitr::opts_chunk$set(
  fig.align = "center"
)
```

(ref:workflow) An overview of the fluent genomics workflow. First, we *import*
data as a *SummarizedExperiment* object, which enables interoperability with
downstream analysis packages. Then we *model* our assay data, using the
existing Bioconductor packages *DESeq2* and *limma*. We take the results of our
models for each assay with respect to their genomic coordinates, and
*integrate* them. First, we compute the overlap between the results of each
assay, then aggregate over the combined genomic regions, and finally summarize
to compare enrichment for differentially expressed genes to non differentially
expressed genes. The final output can be used for downstream visualization or
further transformation.

(ref:maplot) Visualization of *DESeq2* results as an "MA plot". Genes that have
an adjusted *p-value* below 0.01 are colored red.

(ref:boxplot) A boxplot of maximum LFCs for DA peaks for DE genes compared to
non-DE genes where genes have at least one DA peak.

(ref:linechart) A line chart displaying how relative enrichment of DA peaks
change between DE genes compared to non-DE genes as the absolute DA LFC
threshold increases.

(ref:linechart2) A line chart displaying how gene and peak counts change as the
absolute DA LFC threshold increases. Lines are colored according to whether
they represent a gene that is DE or not. Note the x-axis is on a $\log_{10}$
scale.

# Introduction

In this workflow, we examine a subset of the RNA-seq and ATAC-seq data from
@alasoo, a study that involved treatment of macrophage cell lines from a number
of human donors with interferon gamma (IFNg), *Salmonella* infection, or both
treatments combined. @alasoo examined gene expression and chromatin
accessibility in a subset of 86 successfully differentiated induced pluripotent
stem cells (iPSC) lines, and compared baseline and response with respect to
chromatin accessibility and gene expression at specific quantitative trait loci
(QTL). The authors found that many of the stimulus-specific expression QTL were
already detectable as chromatin QTL in naive cells, and further hypothesize
about the nature and role of transcription factors implicated in the response
to stimulus.

We will perform a much simpler analysis than the one found in @alasoo, using
their publicly available RNA-seq and ATAC-seq data (ignoring the genotypes). We
will examine the effect of IFNg stimulation on gene expression and chromatin
accessibility, and look to see if there is an enrichment of differentially
accessible (DA) ATAC-seq peaks in the vicinity of differentially expressed (DE)
genes. This is plausible, as the transcriptomic response to IFNg stimulation
may be mediated through binding of regulatory proteins to accessible regions,
and this binding may increase the accessibility of those regions such that it
can be detected by ATAC-seq.

```{r workflow, fig.cap = "(ref:workflow)", fig.align='center', out.width="\\textwidth", echo = FALSE}
knitr::include_graphics("workflow.png")
```

Throughout the workflow (Figure \@ref(fig:workflow)), we will use existing
Bioconductor infrastructure to understand these datasets. In particular, we
will emphasize the use of the Bioconductor packages *plyranges* and *tximeta*.
The *plyranges* package fluently transforms data tied to genomic ranges using
operations like shifting, window construction, overlap detection, etc. It is
described by @Lee2019 and leverages underlying core Bioconductor infrastructure
[@granges; @bioc] and the *tidyverse* design principles @tidyverse.

The *tximeta* package described by @Love2019-tximeta is used to read RNA-seq
quantification data into R/Bioconductor, such that the transcript ranges and
their provenance are automatically attached to the object containing expression
values and differential expression results.

## Experimental Data

The data used in this workflow is available from two packages: the *macrophage*
Bioconductor ExperimentData package and from the workflow package
*fluentGenomics*.

The *macrophage* package contains RNA-seq quantification from 24 RNA-seq
samples, a subset of the RNA-seq samples generated and analyzed by @alasoo. The
paired-end reads were quantified using *Salmon* [@salmon], using the Gencode 29
human reference transcripts [@gencode]. For more details on quantification, and
the exact code used, consult the vignette of the
[macrophage](http://bioconductor.org/packages/macrophage) package. The package
also contains the `Snakemake` file that was used to distribute the *Salmon*
quantification jobs on a cluster [@snakemake].

The *fluentGenomics* package contains functionality to download and generate a
cached *SummarizedExperiment* object from the normalized ATAC-seq data provided
by @alasooZenodo. This object contains all 145 ATAC-seq samples across all
experimental conditions as analyzed by @alasoo. The data can be also be
downloaded directly from the
[Zenodo](https://zenodo.org/record/1188300#.XIAhXlNKjOQ) deposition.

The following code loads the path to the cached data file, or if it is not
present, will create the cache and generate a *SummarizedExperiment* using the
the *BiocFileCache* package [@bcfilecache].

```{r cache-path, eval = FALSE}
library(fluentGenomics)
path_to_se <- cache_atac_se()
```

We can then read the cached file and assign it to an object called `atac`.

```{r read-cache, eval = FALSE}
atac <- readRDS(path_to_se)
```

A precise description of how we obtained this *SummarizedExperiment* object can
be found in section \@ref(atac).

# Import Data as a *SummarizedExperiment* {#se}

## Using *tximeta* to import RNA-seq quantification data

First, we specify a directory `dir`, where the quantification files are stored.
You could simply specify this directory with:

```{r dir, eval=FALSE}
dir <- "/path/to/quant/files"
```

where the path is relative to your current R session. However, in this case we
have distributed the files in the *macrophage* package. The relevant directory
and associated files can be located using `system.file`.

```{r setdir}
dir <- system.file("extdata", package="macrophage")
```

Information about the experiment is contained in the `coldata.csv` file. We
leverage the *dplyr* and *readr* packages (as part of the *tidyverse*) to read
this file into R [@tidyverse]. We will see later that *plyranges* extends these
packages to accommodate genomic ranges.

```{r coldata-rna}
library(dplyr)
library(readr)
colfile <- file.path(dir, "coldata.csv")
coldata <- read_csv(colfile) %>%
  dplyr::select(
    names,
    id = sample_id,
    line = line_id,
    condition = condition_name
  ) %>%
  dplyr::mutate(
    files = file.path(dir, "quants", names, "quant.sf.gz"),
    line = factor(line),
    condition = relevel(factor(condition), "naive")
  )
coldata
```

After we have read the `coldata.csv` file, we select relevant columns from this
table, create a new column called `files`, and transform the existing `line`
and `condition` columns into factors.  In the case of `condition`, we specify
the "naive" cell line as the reference level.  The `files` column points to the
quantifications for each observation -- these files have been gzipped, but
would typically not have the 'gz' ending if used from *Salmon* directly. One
other thing to note is the use of the pipe operator,`%>%`, which can be read as
"then", i.e. first read the data, *then* select columns, *then* mutate them.

Now we have a table summarizing the experimental design and the locations of
the quantifications. The following lines of code do a lot of work for the
analyst: importing the RNA-seq quantification (dropping *inferential
replicates* in this case), locating the relevant reference transcriptome,
attaching the transcript ranges to the data, and fetching genome information.
Inferential replicates are especially useful for performing transcript-level
analysis, but here we will use a point estimate for the per-gene counts and
perform gene-level analysis.

The result is a *SummarizedExperiment* object.

```{r tximeta-run}
suppressPackageStartupMessages(library(SummarizedExperiment))
library(tximeta)
se <- tximeta(coldata, dropInfReps=TRUE)
se
```
<!-- should we describe the data structure in more detail, or use a figure -->

On a machine with a working internet connection, the above command works
without any extra steps, as the `tximeta` function obtains any necessary
metadata via FTP, unless it is already cached locally. The *tximeta* package
can also be used without an internet connection, in this case the linked
transcriptome can be created directly from a *Salmon* index and gtf.

```{r linkedtxome-ex, eval = FALSE}
makeLinkedTxome(
  indexDir=file.path(dir, "gencode.v29_salmon_0.12.0"),
  source="Gencode",
  organism="Homo sapiens",
  release="29",
  genome="GRCh38",
  fasta="ftp://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_human/release_29/gencode.v29.transcripts.fa.gz",
  gtf=file.path(dir, "gencode.v29.annotation.gtf.gz"), # local version
  write=FALSE
)
```

Because *tximeta* knows the correct reference transcriptome, we can ask
*tximeta* to summarize the transcript-level data to the gene level using the
methods of @Soneson2015.

```{r gse}
gse <- summarizeToGene(se)
```

One final note is that the `start` of positive strand genes and the `end` of
negative strand genes is now dictated by the genomic extent of the isoforms of
the gene (so the `start` and `end` of the reduced *GRanges*). Another
alternative would be to either operate on transcript abundance, and perform
differential analysis on transcript (and so avoid defining the TSS of a set of
isoforms), or to use gene-level summarized expression but to pick the most
representative TSS based on isoform expression.

## Importing ATAC-seq data as a *SummarizedExperiment* object {#atac}

The *SummarizedExperiment* object containing ATAC-seq peaks can be created from
the following tab-delimited files from @alasooZenodo:

* The sample metadata: `ATAC_sample_metadata.txt.gz` (<1M)
* The matrix of normalized read counts: `ATAC_cqn_matrix.txt.gz` (109M)
* The annotated peaks: `ATAC_peak_metadata.txt.gz` (5.6M)

To begin, we read in the sample metadata, following similar steps to those we
used to generate the `coldata` table for the RNA-seq experiment:

```{r coldata-atac, eval=FALSE}
atac_coldata <- read_tsv("ATAC_sample_metadata.txt.gz") %>%
  select(
    sample_id,
    donor,
    condition = condition_name
  ) %>%
  mutate(condition = relevel(factor(condition), "naive"))
```

The ATAC-seq counts have already been normalized with *cqn* [@Hansen2012] and
log2 transformed. Loading the *cqn*-normalized matrix of log2 transformed read
counts takes ~30 seconds and loads an object of ~370 Mb. We set the column
names so that the first column contains the rownames of the matrix, and the
remaining columns are the sample identities from the `atac_coldata` object.

```{r mat-atac, eval=FALSE}
atac_mat <- read_tsv("ATAC_cqn_matrix.txt.gz",
                     skip = 1,
                     col_names =c("rownames", atac_coldata[["sample_id"]]))
rownames <- atac_mat[["rownames"]]
atac_mat <- as.matrix(atac_mat[,-1])
rownames(atac_mat) <- rownames
```

We read in the peak metadata (locations in the genome), and convert it to a
*GRanges* object. The `as_granges()` function automatically converts the
*data.frame* into a *GRanges* object. From that result, we extract the peak_id
column and set the genome information to the build "GRCh38". We know this from
the [Zenodo entry](https://zenodo.org/record/1188300#.XJOFSlNKiL5).

```{r peaks-atac, eval=FALSE}
library(plyranges)
peaks_df <- read_tsv("ATAC_peak_metadata.txt.gz",
                     col_types = c("cidciicdc")
)

peaks_gr <- peaks_df %>%
  as_granges(seqnames = chr) %>%
  select(peak_id=gene_id) %>%
  set_genome_info(genome = "GRCh38")
```

Finally, we construct a *SummarizedExperiment* object.  We place the matrix
into the assays slot as a named list, the annotated peaks into the row-wise
ranges slot, and the sample metadata into the column-wise data slot:

```{r atac-se, eval=FALSE}
atac <- SummarizedExperiment(assays = list(cqndata=atac_mat),
                             rowRanges=peaks_gr,
                             colData=atac_coldata)
```

# Model assays

## RNA-seq differential gene expression analysis

We can easily run a differential expression analysis with *DESeq2* using the
following code chunks [@Love2014]. The design formula indicates that we want to
control for the donor baselines (`line`) and test for differences in gene
expression on the condition. For a more comprehensive discussion of DE
workflows in Bioconductor see @Love2016-f1000 and @Law2018-f1000.

```{r setup-deseq}
library(DESeq2)
dds <- DESeqDataSet(gse, ~line + condition)
# filter out lowly expressed genes
# at least 10 counts in at least 6 samples
keep <- rowSums(counts(dds) >= 10) >= 6
dds <- dds[keep,]
```

The model is fit with the following line of code:

```{r fit-model}
dds <- DESeq(dds)
```

Below we set the contrast on the condition variable, indicating we are
estimating the $\log_2$ fold change (LFC) of IFNg stimulated cell lines against
naive cell lines. We are interested in LFC greater than 1 at a nominal false
discovery rate (FDR) of 1%.

```{r results-DFrame}
res <- results(dds,
               contrast=c("condition","IFNg","naive"),
               lfcThreshold=1, alpha=0.01)
```

To see the results of the expression analysis, we can generate a summary table
and an MA plot:

```{r ma-plot, fig.cap="(ref:maplot)" }
summary(res)
DESeq2::plotMA(res, ylim=c(-10,10))
```

We now output the results as a *GRanges* object, and due to the conventions of
*plyranges*, we construct a new column called `gene_id` from the row names of
the results. Each row now contains the genomic region (`seqnames`, `start`,
`end`, `strand`) along with corresponding metadata columns (the `gene_id` and
the results of the test). Note that *tximeta* has correctly identified the
reference genome as "hg38", and this has also been added to the *GRanges* along
the results columns. This kind of book-keeping is vital once overlap operations
are performed to ensure that *plyranges* is not comparing across incompatible
genomes.

```{r results-GRanges}
suppressPackageStartupMessages(library(plyranges))
de_genes <- results(dds,
                    contrast=c("condition","IFNg","naive"),
                    lfcThreshold=1,
                    format="GRanges") %>%
  names_to_column("gene_id")
de_genes
```

From this, we can restrict the results to those that meet our FDR threshold and
select (and rename) the metadata columns we're interested in:

```{r de-genes}
de_genes <- de_genes %>%
  filter(padj < 0.01) %>%
  select(gene_id, de_log2FC = log2FoldChange, de_padj = padj)
```

We now wish to extract genes for which there is evidence that the LFC is *not*
large. We perform this test by specifying an LFC threshold and an alternative
hypothesis (`altHypothesis`) that the LFC is less than the threshold in
absolute value. To visualize the result of this test, you can run `results`
without `format="GRanges"`, and pass this object to `plotMA` as before. We
label these genes as `other_genes` and later as "non-DE genes", for comparison
with our `de_genes` set.

```{r not-de-genes}
other_genes <- results(dds,
                       contrast=c("condition","IFNg","naive"),
                       lfcThreshold=1,
                       altHypothesis="lessAbs",
                       format="GRanges") %>%
  filter(padj < 0.01) %>%
  names_to_column("gene_id") %>%
  dplyr::select(gene_id,
                de_log2FC = log2FoldChange,
                de_padj = padj)
```

## ATAC-seq peak differential abundance analysis

The following section describes the process we have used for generating a
*GRanges* object of differential peaks from the ATAC-seq data in @alasoo.

The code chunks for the remainder of this section are not run.

<!-- We can check the standard deviation over mean plot, to assess for any -->
<!-- systematic trends: -->

<!-- ```{r, eval = FALSE} --> <!-- library(ggplot2) --> <!-- mv <-
data.frame(rmu = rowMeans(assay(atac)), --> <!-- rvar = rowVars(assay(atac)))
--> <!-- ggplot(data =mv , aes(x = rmu, y = sqrt(rvar))) + --> <!-- geom_hex()
--> <!-- ``` -->

For assessing differential accessibility, we run *limma* [@Smyth2004], and
generate the a summary of LFCs and adjusted p-values for the peaks:

```{r limma, eval = FALSE}
library(limma)
design <- model.matrix(~donor + condition, colData(atac))
fit <- lmFit(assay(atac), design)
fit <- eBayes(fit)
idx <- which(colnames(fit$coefficients) == "conditionIFNg")
tt <- topTable(fit, coef=idx, sort.by="none", n=nrow(atac))
```

We now take the `rowRanges` of the *SummarizedExperiment* and attach the LFCs
and adjusted p-values from *limma*, so that we can consider the overlap with
differential expression. Note that we set the genome build to "hg38" and
restyle the chromosome information to use the "UCSC" style (e.g. "chr1",
"chr2", etc.). Again, we know the genome build from the Zenodo entry for the
ATAC-seq data.

```{r peaks-tidy, eval = FALSE}
atac_peaks <- rowRanges(atac) %>%
  remove_names() %>%
  mutate(
    da_log2FC = tt$logFC,
    da_padj = tt$adj.P.Val
  ) %>%
  set_genome_info(genome = "hg38")

seqlevelsStyle(atac_peaks) <- "UCSC"
```

The final *GRanges* object containing the DA peaks is included in the workflow
package and can be loaded as follows:

```{r load-peaks}
library(fluentGenomics)
peaks
```

# Integrate ranges

## Finding overlaps with *plyranges*

We have already used *plyranges* a number of times above, to `filter`,
`mutate`, and `select` on *GRanges* objects, as well as ensuring the correct
genome annotation and style has been used. The *plyranges* package provides a
grammar for performing transformations of genomic data [@Lee2019]. Computations
resulting from compositions of *plyranges* "verbs" are performed using
underlying, highly optimized range operations in the *GenomicRanges* package
[@granges].

For the overlap analysis, we filter the annotated peaks to have a nominal FDR
bound of 1%.

```{r filter-peaks}
da_peaks <- peaks %>%
  filter(da_padj < 0.01)
```

We now have *GRanges* objects that contain DE genes, genes without strong
signal of DE, and DA peaks. We are ready to answer the question: is there an
enrichment of DA ATAC-seq peaks in the vicinity of DE genes compared to genes
without sufficient DE signal?

## Down sampling non-differentially expressed genes

As *plyranges* is built on top of *dplyr*, it implements methods for many of
its verbs for *GRanges* objects. Here we can use `slice` to randomly sample the
rows of the `other_genes`. The `sample.int` function will generate random
samples of size equal to the number of DE-genes from the number of rows in
`other_genes`:

```{r slice-example}
size <- length(de_genes)
slice(other_genes, sample.int(n(), size))
```

We can repeat this many times to create many samples via `replicate`. By
replicating the sub-sampling multiple times, we minimize the variance on the
enrichment statistics induced by the sampling process.

```{r boot-set-01}
# set a seed for the results
set.seed(2019-08-02)
boot_genes <- replicate(10,
                        slice(other_genes, sample.int(n(), size)),
                        simplify = FALSE)
```

This creates a list of *GRanges* objects as a list, and we can bind these
together using the `bind_ranges` function. This function creates a new column
called "resample" on the result that identifies each of the input *GRanges*
objects:

```{r boot-set-02}
boot_genes <- bind_ranges(boot_genes, .id = "resample")
```

Similarly, we can then combine the `boot_genes` *GRanges*, with the DE
*GRanges* object. As the resample column was not present on the DE *GRanges*
object, this is given a missing value which we recode to a 0 using `mutate()`

```{r combine-results}
all_genes <- bind_ranges(
  de=de_genes,
  not_de = boot_genes,
  .id="origin"
) %>%
  mutate(
    origin = factor(origin, c("not_de", "de")),
    resample = ifelse(is.na(resample), 0L, as.integer(resample))
  )
all_genes
```

## Expanding genomic coordinates around the transcription start site

Now we would like to modify our gene ranges so they contain the 10 kilobases on
either side of their transcription start site (TSS). There are many ways one
could do this, but we prefer an approach via the anchoring methods in
*plyranges*. Because there is a mutual dependence between the start, end,
width, and strand of a *GRanges* object, we define anchors to fix one of
`start` and `end`, while modifying the `width`. As an example, to extract just
the TSS, we can anchor by the 5' end of the range and modify the width of the
range to equal 1.

```{r resize-01}
all_genes <- all_genes %>%
  anchor_5p() %>%
  mutate(width = 1)
```

Anchoring by the 5' end of a range will fix the `end` of negatively stranded
ranges, and fix the `start` of positively stranded ranges.

We can then repeat the same pattern but this time using `anchor_center()` to
tell *plyranges* that we are making the TSS the midpoint of a range that has
total width of 20kb, or 10kb both upstream and downstream of the TSS.

```{r resize-02}
all_genes <- all_genes %>%
  anchor_center() %>%
  mutate(width=2*1e4)
```

## Use overlap joins to find relative enrichment

We are now ready to compute overlaps between RNA-seq genes (our DE set and
bootstrap sets) and the ATAC-seq peaks. In *plyranges*, overlaps are defined as
joins between two *GRanges* objects: a _left_ and a _right_ *GRanges* object.
In an overlap join, a match is any range on the _left_ *GRanges* that is
overlapped by the _right_ *GRanges*. One powerful aspect of the overlap joins
is that the result maintains all (metadata) columns from each of the _left_ and
_right_ ranges which makes downstream summaries easy to compute.

To combine the DE genes with the DA peaks, we perform a left overlap join. This
returns to us the `all_genes` ranges (potentially with duplication), but with
the metadata columns from those overlapping DA peaks.  For any gene that has no
overlaps, the DA peak columns will have `NA`'s.

```{r olap-join}
genes_olap_peaks <- all_genes %>%
  join_overlap_left(da_peaks)
genes_olap_peaks
```

Now we can ask, how many DA peaks are near DE genes relative to "other" non-DE
genes?  A gene may appear more than once in `genes_olap_peaks`, because
multiple peaks may overlap a single gene, or because we have re-sampled the
same gene more than once, or a combination of these two cases.

For each gene (that is the combination of chromosome, the start, end, and
strand), and the "origin" (DE vs not-DE) we can compute the distinct number of
peaks for each gene and the maximum peak based on LFC. This is achieved via
`reduce_ranges_directed`, which allows an aggregation to result in a *GRanges*
object via merging neighboring genomic regions. The use of the directed suffix
indicates we're maintaining strand information. In this case, we are simply
merging ranges (genes) via the groups we mentioned above. We also have to
account for the number of resamples we have performed when counting if there
are any peaks, to ensure we do not double count the same peak:

```{r reduce-ex01}
gene_peak_max_lfc <- genes_olap_peaks %>%
  group_by(gene_id, origin)  %>%
  reduce_ranges_directed(
    peak_count = sum(!is.na(da_padj)) / n_distinct(resample),
    peak_max_lfc = max(abs(da_log2FC))
  )
```

We can then filter genes if they have any peaks and compare the peak fold
changes between non-DE and DE genes using a boxplot:

```{r boxplot, fig.cap = "(ref:boxplot)"}
library(ggplot2)
gene_peak_max_lfc %>%
  filter(peak_count > 0) %>%
  as.data.frame() %>%
  ggplot(aes(origin, peak_max_lfc)) +
  geom_boxplot()
```

In general, the DE genes have larger maximum DA fold changes relative to the
non-DE genes.

Next we examine how thresholds on the DA LFC modify the enrichment we observe
of DA peaks near DE or non-DE genes. First, we want to know how the number of
peaks within DE genes and non-DE genes change as we change threshold values on
the peak LFC. As an example, we could compute this by arbitrarily chosen LFC
thresholds of 1 or 2 as follows:

```{r summarize-ex01}
origin_peak_lfc <- genes_olap_peaks %>%
  group_by(origin) %>%
  summarize(
    peak_count = sum(!is.na(da_padj)) / n_distinct(resample),
    lfc1_peak_count =sum(abs(da_log2FC) > 1, na.rm=TRUE)/ n_distinct(resample),
    lfc2_peak_count = sum(abs(da_log2FC) > 2, na.rm=TRUE)/ n_distinct(resample)
  )
origin_peak_lfc
```

Here we see that DE genes tend to have more DA peaks near them, and that the
number of DA peaks decreases as we increase the DA LFC threshold (as expected).
We now show how to compute the ratio of peak counts from DE compared to non-DE
genes, so we can see how this ratio changes for various DA LFC thresholds.

For all variables except for the `origin` column we divide the first row's
values by the second row, which will be the enrichment of peaks in DE genes
compared to other genes. This requires us to reshape the summary table from
long form back to wide form using the *tidyr* package. First we pivot the
results of the `peak_count` columns into name-value pairs, then pivot again to
place values into the `origin` column. Then we create a new column with the
relative enrichment:

```{r pivot-enrich}
origin_peak_lfc %>%
  as.data.frame() %>%
  tidyr::pivot_longer(cols = -origin) %>%
  tidyr::pivot_wider(names_from = origin, values_from = value) %>%
  mutate(enrichment = de / not_de)
```

The above table shows that relative enrichment increases for a larger LFC
threshold.

Due to the one-to-many mappings of genes to peaks, it is unknown if we have the
same number of DE genes participating or less, as we increase the threshold on
the DA LFC. We can examine the number of genes with overlapping DA peaks at
various thresholds by grouping and aggregating twice. First, the number of
peaks that meet the thresholds are computed within each gene, origin, and
resample group. Second, within the origin column, we compute the total number
of peaks that meet the DA LFC threshold and the number of genes that have more
than zero peaks (again averaging over the number of resamples).

```{r reduce-summarize}
genes_olap_peaks %>%
  group_by(gene_id, origin, resample) %>%
  reduce_ranges_directed(
    lfc1 = sum(abs(da_log2FC) > 1, na.rm=TRUE),
    lfc2 = sum(abs(da_log2FC) > 2, na.rm=TRUE)
  ) %>%
  group_by(origin) %>%
  summarize(
    lfc1_gene_count = sum(lfc1 > 0) / n_distinct(resample),
    lfc1_peak_count = sum(lfc1) / n_distinct(resample),
    lfc2_gene_count = sum(lfc2 > 0) / n_distinct(resample),
    lfc2_peak_count = sum(lfc2) / n_distinct(resample)
  )
```

To do this for many thresholds is cumbersome and would create a lot of
duplicate code. Instead we create a single function called
`count_above_threshold` that accepts a variable and a vector of thresholds, and
computes the sum of the absolute value of the variable for each element in the
`thresholds` vector.

```{r count-fn}
count_if_above_threshold <- function(var, thresholds) {
  lapply(thresholds, function(.) sum(abs(var) > ., na.rm = TRUE))
}
```

The above function will compute the counts for any arbitrary threshold, so we
can apply it over possible LFC thresholds of interest. We choose a grid of one
hundred thresholds based on the range of absolute LFC values in the `da_peaks`
*GRanges* object:

```{r thresholds}
thresholds <- da_peaks %>%
  mutate(abs_lfc = abs(da_log2FC)) %>%
  with(
    seq(min(abs_lfc), max(abs_lfc), length.out = 100)
  )
```

The peak counts for each threshold are computed as a new list-column called
`value`. First, the *GRanges* object has been grouped by the gene, origin, and
the number of resamples columns. Then we aggregate over those columns, so each
row will contain the peak counts for all of the thresholds for a gene, origin,
and resample. We also maintain another list-column that contains the threshold
values.

```{r reduce-ex02}
genes_peak_all_thresholds <- genes_olap_peaks %>%
  group_by(gene_id, origin, resample) %>%
  reduce_ranges_directed(
    value = count_if_above_threshold(da_log2FC, thresholds),
    threshold = list(thresholds)
  )
genes_peak_all_thresholds
```

Now we can expand these list-columns into a long *GRanges* object using the
`expand_ranges()` function. This function will unlist the `value` and
`threshold` columns and lengthen the resulting *GRanges* object.  To compute
the peak and gene counts for each threshold, we apply the same summarization as
before:

```{r expand-summarize}
origin_peak_all_thresholds <- genes_peak_all_thresholds %>%
  expand_ranges() %>%
  group_by(origin, threshold) %>%
  summarize(
    gene_count = sum(value > 0) / n_distinct(resample),
    peak_count = sum(value) / n_distinct(resample)
  )
origin_peak_all_thresholds
```

Again we can compute the relative enrichment in LFCs in the same manner as
before, by pivoting the results to long form then back to wide form to compute
the enrichment. We visualize the peak enrichment changes of DE genes relative
to other genes as a line chart:

```{r line-chart, fig.cap = "(ref:linechart)"}
origin_threshold_counts <- origin_peak_all_thresholds %>%
  as.data.frame() %>%
  tidyr::pivot_longer(cols = -c(origin, threshold),
                      names_to = c("type", "var"),
                      names_sep = "_",
                      values_to = "count") %>%
  select(-var)

origin_threshold_counts %>%
  filter(type == "peak") %>%
  tidyr::pivot_wider(names_from = origin, values_from = count) %>%
  mutate(enrichment =  de / not_de) %>%
  ggplot(aes(x = threshold, y = enrichment)) +
  geom_line() +
  labs(x = "logFC threshold", y = "Relative Enrichment")
```

We computed the sum of DA peaks near the DE genes, for increasing LFC
thresholds on the accessibility change. As we increased the threshold, the
number of total peaks went down (likewise the mean number of DA peaks per
gene). It is also likely the number of DE genes with a DA peak nearby with such
a large change went down. We can investigate this with a plot that summarizes
many of the aspects underlying the enrichment plot above.

```{r line-chart2, fig.cap = "(ref:linechart2)"}
origin_threshold_counts %>%
  ggplot(aes(x = threshold,
             y = count + 1,
             color = origin,
             linetype = type)) +
  geom_line() +
  scale_y_log10()
```

# Discussion

We have shown that by using *plyranges* and *tximeta* (with support of
Bioconductor and *tidyverse* ecosystems) we can fluently iterate through the
biological data science workflow: from import, through to modeling, and data
integration.

There are several further steps that would be interesting to perform in this
analysis; for example, we could modify window size around the TSS to see how it
affects enrichment, and vary the FDR cut-offs for both the DE gene and DA peak
sets. We could also have computed variance in addition to the mean of the
bootstrap set, and so drawn an interval around the enrichment line.

Finally, our workflow illustrates the benefits of using appropriate data
abstractions provided by Bioconductor such as the *SummarizedExperiment* and
*GRanges*. These abstractions provide users with a mental model of their
experimental data and are the building blocks for constructing the modular and
iterative analyses we have shown here. Consequently, we have been able to
interoperate many decoupled R packages (from both Bioconductor and the
tidyverse) to construct a seamless end-to-end workflow that is far too
specialized for a single monolithic tool.

# Software Availability

The workflow materials, including this article can be fully reproduced
following the instructions found at the Github repository
[sa-lee/fluentGenomics](https://github.com/sa-lee/fluentGenomics). Moreover,
the development version of the workflow and all downstream dependencies can be
installed using the `BiocManager` package by running:

```{r, eval = FALSE}
# development version from Github
BiocManager::install("sa-lee/fluentGenomics")
# version available from Bioconductor
BiocManager::install("fluentGenomics")
```

This article and the analyses were performed with R [@baser] using the
*rmarkdown* [@rmarkdown], and *knitr* [@knitr; @xie2015] packages.

## Session Info

```{r}
sessionInfo()
```

## Author Contributions

All authors contributed to the writing and development of the workflow.

## Competing interests

The authors declare that they have no competing interests.

## Funding

SL is supported by an Australian Government Research Training Program (RTP)
scholarship with a top up scholarship from CSL Limited.

MIL's contribution is supported by NIH grant R01 HG009937.

__I confirm that the funders had no role in study design, data collection and
analysis, decision to publish, or preparation of the manuscript.__

## Acknowledgements

The authors would like to thank all participants of the Bioconductor 2019 and
BiocAsia 2019 conferences who attended and provided feedback on early versions
of this workflow paper.

# References

### 2020-01-30 scripts/superintronic.R
data for superintronic chapter

@@ -1,0 +1,232 @@
+## ---- superintronic-setup
+library(plyranges)
+library(BiocParallel)
+library(superintronic)
+library(ggplot2)
+library(here)
+theme_set(theme_bw())
+
+# Plotting helper functions
+set_plot_rng <- function(cvg, target, design, alpha) {
+  if (is(cvg, "GRangesList")) {
+    stopifnot(is(design, "DataFrame"))
+    cvg_rng <- as(lapply(cvg, function(x) superintronic::join_parts(x, target)),
+                  "GRangesList")
+    md <- S4Vectors::DataFrame(
+      Sample = S4Vectors::Rle(names(cvg_rng), lengths(cvg_rng)),
+      CellLine = S4Vectors::Rle(design$CellLine, lengths(cvg_rng)),
+      Kit = S4Vectors::Rle(as.factor(design$Kit), lengths(cvg_rng))
+    )
+
+    cvg_rng <- unlist(cvg_rng, use.names = FALSE)
+    S4Vectors::mcols(cvg_rng) <- cbind(S4Vectors::mcols(cvg_rng), md)
+
+    cvg_rng <- plyranges::mutate(cvg_rng,
+                                 log_score = log2(score + 0.5),
+                                 strand = BiocGenerics::strand(target),
+                                 var = paste0("Kit: ", Kit, " Cellline: ", CellLine))
+  }
+
+  if (is(cvg, "GRanges")) {
+    cvg_rng <- plyranges::filter_by_overlaps(cvg, target)
+    cvg_rng <- plyranges::mutate(cvg_rng,
+                                 strand = BiocGenerics::strand(target),
+                                 var = paste0("Kit: ", Kit, " Cellline: ", CellLine))
+  }
+
+  if (alpha) {
+    cvg_rng <- plyranges::disjoin_ranges_directed(group_by(cvg_rng, var),
+                                                  log_score = BiocGenerics::mean(log_score),
+                                                  feature_type = unlist(BiocGenerics::unique(feature_type)))
+
+    cvg_rng <- plyranges::mutate(cvg_rng,
+                                 alpha = dplyr::case_when(
+                                   score > 3 & feature == "intron" ~ 1,
+                                   score <= 3 & feature == "intron"~ 0.5,
+                                   TRUE ~ 1
+                                 ))
+  }
+
+  return(cvg_rng)
+}
+
+
+# pretty track plot
[…]

### 2020-01-30 tic.R
add travis, this will fail for the moment

@@ -1,0 +1,21 @@
+get_stage("before_install") %>%
+  add_code_step(update.packages(ask = FALSE))
+
+get_stage("install") %>%
+  add_code_step(remotes::install_deps(dependencies = TRUE))
+
+get_stage("deploy") %>%
+  add_code_step(bookdown::render_book('index.Rmd', 'bookdown::gitbook'))
+
+if (Sys.getenv("id_rsa") != "" && !ci()$is_tag()) {
+  # Other example criteria:
+  # - `inherits(ci(), "TravisCI")`: Only for Travis CI
+  # - `Sys.getenv("BUILD_PKGDOWN") != ""`: If the env var "BUILD_PKGDOWN" is set
+  # - `Sys.getenv("TRAVIS_EVENT_TYPE") == "cron"`: Only for Travis cron jobs
+
+  get_stage("before_deploy") %>%
+    add_step(step_setup_ssh())
+
+  get_stage("deploy") %>%
+    add_step(step_push_deploy(path = "_book", branch = "gh-pages"))
+}

### 2020-01-31 scripts/fluentGenomics.R
generate scripts

@@ -1,0 +1,380 @@
+## ----setup, include = FALSE------------------------
+library(fluentGenomics)
+dir <- system.file("extdata", package="macrophage")
+
+library(tximeta)
+makeLinkedTxome(
+  indexDir=file.path(dir, "gencode.v29_salmon_0.12.0"),
+  source="Gencode",
+  organism="Homo sapiens",
+  release="29",
+  genome="GRCh38",
+  fasta="ftp://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_human/release_29/gencode.v29.transcripts.fa.gz",
+  gtf=file.path(dir, "gencode.v29.annotation.gtf.gz"), # local version
+  write=FALSE
+)
+
+knitr::opts_chunk$set(
+  fig.align = "center"
+)
+
+
+## ----workflow, fig.cap = "(ref:workflow)", fig.align='center', out.width="\\textwidth", echo = FALSE----
+knitr::include_graphics("workflow.png")
+
+
+## ----cache-path, eval = FALSE----------------------
+## library(fluentGenomics)
+## path_to_se <- cache_atac_se()
+
+
+## ----read-cache, eval = FALSE----------------------
+## atac <- readRDS(path_to_se)
+
+
+## ----dir, eval=FALSE-------------------------------
+## dir <- "/path/to/quant/files"
+
+
+## ----setdir----------------------------------------
+dir <- system.file("extdata", package="macrophage")
+
+
+## ----coldata-rna-----------------------------------
+library(dplyr)
+library(readr)
+colfile <- file.path(dir, "coldata.csv")
+coldata <- read_csv(colfile) %>%
+  dplyr::select(
+    names,
+    id = sample_id,
+    line = line_id,
+    condition = condition_name
+  ) %>%
+  dplyr::mutate(
+    files = file.path(dir, "quants", names, "quant.sf.gz"),
+    line = factor(line),
+    condition = relevel(factor(condition), "naive")
+  )
+coldata
+
+
+## ----tximeta-run-----------------------------------
+suppressPackageStartupMessages(library(SummarizedExperiment))
+library(tximeta)
+se <- tximeta(coldata, dropInfReps=TRUE)
+se
+
+
+## ----linkedtxome-ex, eval = FALSE------------------
+## makeLinkedTxome(
+##   indexDir=file.path(dir, "gencode.v29_salmon_0.12.0"),
+##   source="Gencode",
+##   organism="Homo sapiens",
+##   release="29",
[…]

### 2020-01-31 scripts/plyranges.R

@@ -1,0 +1,189 @@
+## ----setup, include = FALSE------------------------
+library(tibble)
+library(knitr)
+opts_chunk$set(message = FALSE,
+               warning = FALSE,
+               comment = "#>",
+               fig.width = 5,
+               fig.height = 3,
+               fig.align = "center",
+               fig.path = "./diagrams/")
+# for peak figures
+library(ggbio)
+
+# for bam example
+library(Biostrings)
+library(plyranges)
+
+# -- Data from the Human Epigenomics RoadMap consortium
+# retrieving a BigWigFile from AnnotationHub
+
+if (!dir.exists("./data")) dir.create("./data")
+
+library(AnnotationHub)
+q <- AnnotationHub()
+# T-cell BW file
+bw_file <- q[["AH33458"]]
+
+# a BamFile from the H1 cell line
+h1_bam <- "./data/GSM433167_BI.H3K27me3.bam"
+if (!file.exists(h1_bam))
+  download.file(
+    url = "ftp://ftp.ncbi.nlm.nih.gov/geo/samples/GSM433nnn/GSM433167/suppl/GSM433167%5FBI%2EH3K27me3%2Ebam",
+    destfile = h1_bam
+  )
+
+h1_bam_sorted <- "./data/GSM433167_BI.H3K27me3.sorted.bam"
+
+# sort and index bam
+if (!file.exists(h1_bam_sorted)) {
+  Rsamtools::sortBam(h1_bam, "./data/GSM433167_BI.H3K27me3.sorted")
+  Rsamtools::indexBam(h1_bam_sorted)
+}
+
+# array data for h1 cell line
+# illumina annotations
+array_info <- "data/GPL18952_HumanOmni25M-8v1-1_B.annotated.txt.gz"
+if (!file.exists(array_info))
+  download.file(
+    url = "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GPL18952&format=file&file=GPL18952%5FHumanOmni25M%2D8v1%2D1%5FB%2Eannotated%2Etxt%2Egz",
+    destfile = array_info
+  )
+
+snp_info <- "data/GSM1463263_JS-ESCH1.txt.gz"
+if (!file.exists(snp_info))
+  download.file(
+    url = "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSM1463263&format=file&file=GSM1463263%5FJS%2DESCH1%2Etxt%2Egz",
+    destfile = snp_info
+  )
[…]

### 2020-02-21 data-raw/plyranges-ex-bam.R
add cache scripts for plyranges examples

@@ -1,0 +1,31 @@
+# download and prepare BAM file example
+
+library(BiocFileCache)
+bfc <- BiocFileCache(cache = here::here("data"))
+
+# ftp link on GEO
+url <- "ftp://ftp.ncbi.nlm.nih.gov/geo/samples/GSM433nnn/GSM433167/suppl/GSM433167%5FBI%2EH3K27me3%2Ebam"
+fpath <- "GSM433167_BI.H3K27me3.bam"
+
+res <- bfcquery(bfc, fpath)
+
+if (bfccount(res) == 0) {
+  download.file(url = url, destfile = fpath)
+  ans <- bfcadd(bfc, rname = fpath, action = "move" )
+} else {
+  ans <- bfcrpath(bfc, fpath)
+}
+
+
+sortedBam <- "GSM433167_BI.H3K27me3.sorted.bam"
+res <- bfcquery(bfc, sortedBam)
+
+if (bfccount(res) == 0) {
+  Rsamtools::sortBam(ans, sub("\\.bam", "", sortedBam))
+  Rsamtools::indexBam(sortedBam)
+  indexBai <- paste0(sortedBam, ".bai")
+  bfcadd(bfc, rname = indexBai, action = "move", rtype = "local")
+  ans <- bfcadd(bfc, rname = sortedBam, action = "move", rtype = "local")
+} else {
+  ans <- bfcrpath(bfc, sortedBam)
+}

### 2020-02-21 data-raw/plyranges-ex-snparray.R

@@ -1,0 +1,59 @@
+# Prepare GRanges object for SNP array resource
+library(readr)
+library(BiocFileCache)
+library(plyranges)
+
+bfc <- BiocFileCache(cache = here::here("data"))
+
+rname <- "h1_snp_array"
+res <- bfcquery(bfc, rname)
+
+# if the resource doesn't exist start downloading files
+if (bfccount(res) == 0) {
+  array_info <- here::here("data-raw", "GPL18952_HumanOmni25M-8v1-1_B.annotated.txt.gz")
+  if (!file.exists(array_info))
+    download.file(
+      url = "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GPL18952&format=file&file=GPL18952%5FHumanOmni25M%2D8v1%2D1%5FB%2Eannotated%2Etxt%2Egz",
+      destfile = array_info
+    )
+  snp_info <- here::here("data-raw", "GSM1463263_JS-ESCH1.txt.gz")
+  if (!file.exists(snp_info))
+    download.file(
+      url = "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSM1463263&format=file&file=GSM1463263%5FJS%2DESCH1%2Etxt%2Egz",
+      destfile = snp_info
+    )
+  # prepare a GRanges object from array data
+  # annotation information available here
+  # https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GPL18952
+  # chromosome zero are 'problem' probes and should be filtered
+  anno <- read_tsv(array_info,
+                   trim_ws = TRUE,
+                   col_types = c("cccc----")) %>%
+    dplyr::filter(Chr != "0")
+
+  # genotypes for H1 cell line, data starts on line 10
+  # data obtained from https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM1463263
+  geno <- read_tsv(snp_info, skip = 10)
+
+  # rename columns
+  geno <- geno %>%
+    dplyr::select(name = `SNP Name`,
+                  score = `GC Score`,
+                  baf = `B Allele Freq`,
+                  logR = `Log R Ratio`)
+
+  # merge everything
+  complete_vranges_df <- dplyr::inner_join(geno, anno, by = c("name" = "Name")) %>%
+    dplyr::rename(alleles = Alleles, seqnames = Chr, start = MapInfo)
+
+  h1_snp_array <- as_granges(complete_vranges_df,
[…]

### 2020-02-21 Rmd/02-plyranges.Rmd
plyranges chapter edits

@@
 
-```{r setup, include = FALSE}
-library(tibble)
-library(knitr)
-opts_chunk$set(message = FALSE, 
-               warning = FALSE, 
-               comment = "#>", 
-               fig.width = 5, 
-               fig.height = 3, 
-               fig.align = "center",
-               fig.path = "./diagrams/")
-# for peak figures
-library(ggbio)
+```{r plyranges-init, echo = FALSE, cache = FALSE, include = FALSE}
+knitr::read_chunk(here::here("scripts", "plyranges.R"))
+```
 
-# for bam example
-library(Biostrings)
-library(plyranges)
+```{r plyranges-setup, message = FALSE, echo = FALSE}
+```
 
-# -- Data from the Human Epigenomics RoadMap consortium
-# retrieving a BigWigFile from AnnotationHub
-
-if (!dir.exists("./data")) dir.create("./data")
-
-library(AnnotationHub)
-q <- AnnotationHub()
-# T-cell BW file
-bw_file <- q[["AH33458"]]
-
-# a BamFile from the H1 cell line
-h1_bam <- "./data/GSM433167_BI.H3K27me3.bam"
-if (!file.exists(h1_bam))
-  download.file(
-    url = "ftp://ftp.ncbi.nlm.nih.gov/geo/samples/GSM433nnn/GSM433167/suppl/GSM433167%5FBI%2EH3K27me3%2Ebam",
-    destfile = h1_bam
-  )
-
-h1_bam_sorted <- "./data/GSM433167_BI.H3K27me3.sorted.bam"
-
-# sort and index bam
-if (!file.exists(h1_bam_sorted)) {
-  Rsamtools::sortBam(h1_bam, "./data/GSM433167_BI.H3K27me3.sorted")
-  Rsamtools::indexBam(h1_bam_sorted)
-}
-
-# array data for h1 cell line
-# illumina annotations
-array_info <- "data/GPL18952_HumanOmni25M-8v1-1_B.annotated.txt.gz"
-if (!file.exists(array_info)) 
-  download.file(
-    url = "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GPL18952&format=file&file=GPL18952%5FHumanOmni25M%2D8v1%2D1%5FB%2Eannotated%2Etxt%2Egz",
-    destfile = array_info
-  )
-
-snp_info <- "data/GSM1463263_JS-ESCH1.txt.gz"
-if (!file.exists(snp_info))
-  download.file(
-    url = "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSM1463263&format=file&file=GSM1463263%5FJS%2DESCH1%2Etxt%2Egz",
-    destfile = snp_info
-  )
-
-```
-# Background
+## Background
 
@@
 Query Language (GMQL) implements a DSL for combining genomic datasets 
-\cite{Kaitoua2017-pw}.  The command line application BEDtools develops an extensive 
+[@Kaitoua2017-pw].  The command line application BEDtools develops an extensive 
 algebra for performing arithmetic between two or more sets of genomic regions 
-\cite{Quinlan2010-gc}. All of the aforementioned DSLs are designed to be evaluated
+[@Quinlan2010-gc]. All of the aforementioned DSLs are designed to be evaluated
 either at the command line or embedded in scripts for batch processing. They 
@@
 R data structures and workflows around the notion of tidy data
-\cite{Wickham2014-jc}. A tidy dataset is defined as a tabular
+[@Wickham2014-jc]. A tidy dataset is defined as a tabular
 data structure that has observations as rows and columns as variables,
@@
 us to see how the data relate to the design of an experiment
-and the variables measured. The `dplyr` package \cite{Wickham2017-dplyr} 
+and the variables measured. The `dplyr` package [@Wickham2017-dplyr] 
 defines an application programming interface (API) that maps notions from 
@@
 can be composed together on one or more tidy datasets with the pipe operator 
-from the `magrittr` package \cite{R-magrittr}. Taken together these features 
+from the `magrittr` package [@R-magrittr]. Taken together these features 
 enable a user to write human readable analysis workflows.
@@
 
-# Results
+## Results
 
-## Genomic Relational Algebra
+### Genomic Relational Algebra
 
-### Data Model
+#### Data Model
 
-\begin{figure}
-\includegraphics[width=\textwidth]{diagrams/GRanges}
-\caption{An illustration of the GRanges data model for a
-sample from an RNA-seq experiment. The core components of the data model
-include a seqname column (representing the chromosome), a ranges column
-which consists of start and end coordinates for a genomic region, and a
-strand identifier (either positive, negative, or unstranded). Metadata
-are included as columns to the right of the dotted line as annotations
-(gene\_id) or range level covariates (score).}
-\label{fig:GRanges} 
-\end{figure}
+(ref:GRanges-cap): An illustration of the GRanges data model for a sample from an RNA-seq experiment.
+The core components of the data model include a seqname column (representing the chromosome), a ranges column which consists of start and end coordinates for a genomic region, and a strand identifier (either positive, negative, or unstranded).
+Metadata are included as columns to the right of the dotted line as annotations (gene\_id) or range level covariates (score).
+
+```{r GRanges, fig.cap = "(ref:GRanges-cap)", echo = FALSE}
+knitr::include_graphics(here::here("img", "GRanges.png"))
+```
 
@@
 measurements or metadata specific to the data type or experiment
-(figure \ref{fig:GRanges}).  GRanges balances flexibility with formal
+(figure \@ref(fig:GRanges)).  GRanges balances flexibility with formal
 constraints, so that it is applicable to virtually any genomic
@@
 
-### Algebraic operations
+#### Algebraic operations
 
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{diagrams/olap-fig}
-\caption{Illustration of the three overlap join operators. Each join takes two 
-GRanges objects, \textit{x} and \textit{y} as input. A 'Hits' object for the 
-join is computed which consists of two components. The first component contains 
-the indices of the ranges in \textit{x} that have been overlapped 
-(the rectangles of \textit{x} that cross the orange lines). The second 
-component consists of the indices of the ranges in \textit{y} that overlap 
-the ranges in \textit{x}. In this case a range in \textit{y} overlaps the 
-ranges in \textit{x} three times, so the index is repeated three times. The 
-resulting 'Hits' object is used to modify \textit{x} by where it was 'hit' by
-\textit{y} and merge all metadata columns from \textit{x} and \textit{y} 
-based on the indices contained in the 'Hits' object. This procedure is applied 
-generally in the \texttt{plyranges} DSL for both overlap and nearest neighbor
-operations. The join semantics alter what is returned: \textbf{A}: for an 
-\textbf{inner} join the \textit{x} ranges that are overlapped by \textit{y} 
-are returned. The returned ranges also include the metadata from the \textit{y}
-range that overlapped the three \textit{x} ranges. \textbf{B} An 
-\textbf{intersect} join is identical to an inner join except that the 
-intersection is taken between the overlapped \textit{x} ranges and the 
-\textit{y} ranges. \textbf{C} For the \textbf{left} join all 
-\textit{x} ranges are returned regardless of whether they are overlapped by 
-\textit{y}. In this case the third range (rectangle with the asterisk next to it)
-of the join would have missing values on metadata columns that came from \textit{y}.}
-\label{fig:olaps-fig}
-\end{figure}
+```{r olaps-fig, fig.cap = "(ref:olaps-cap)", echo = FALSE}
+knitr::include_graphics(here::here("img", "olap-fig.png"))
+```
+
+(ref:olaps-cap): Illustration of the three overlap join operators.
+Each join takes two  GRanges objects, \textit{x} and \textit{y} as input.
+A 'Hits' object for the  join is computed which consists of two components.
+The first component contains  the indices of the ranges in \textit{x} that have been overlapped  (the rectangles of \textit{x} that cross the orange lines).
+The second  component consists of the indices of the ranges in \textit{y} that overlap the ranges in \textit{x}.
+In this case a range in \textit{y} overlaps the ranges in \textit{x} three times, so the index is repeated three times.
+The resulting 'Hits' object is used to modify \textit{x} by where it was 'hit' by \textit{y} and merge all metadata columns from \textit{x} and \textit{y} based on the indices contained in the 'Hits' object.
+This procedure is applied generally in the \texttt{plyranges} DSL for both overlap and nearest neighbor operations.
+The join semantics alter what is returned: \textbf{A}: for an \textbf{inner} join the \textit{x} ranges that are overlapped by \textit{y} are returned.
+The returned ranges also include the metadata from the \textit{y} range that overlapped the three \textit{x} ranges.
+\textbf{B} An \textbf{intersect} join is identical to an inner join except that the intersection is taken between the overlapped \textit{x} ranges and the \textit{y} ranges.
+\textbf{C} For the \textbf{left} join all \textit{x} ranges are returned regardless of whether they are overlapped by \textit{y}.
+In this case the third range (rectangle with the asterisk next to it) of the join would have missing values on metadata columns that came from \textit{y}.
 
@@
 _nearest_, _precede_, and _follow_ (figure 
-\ref{fig:olaps-fig}). We can further restrict the
+\@ref(fig:olaps-fig)). We can further restrict the
 matching by whether the query is completely _within_ the subject, and
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{diagrams/code-comparison.pdf}
-\caption{Idiomatic code examples for \texttt{plyranges} (\textbf{A}) and 
-\texttt{GenomicRanges} (\textbf{B}) illustrating an overlap and aggregate 
-operation that returns the same result. 
-In each example, we have two BED files consisting of SNPs that are
-genome-wide association study (GWAS) hits and reference exons. Each code block
-counts for each SNP the number of distinct exons it overlaps. The 
-\texttt{plyranges} code achieves this with an overlap join followed by partitioning
-and aggregation. Strand is ignored by default here. 
-The \texttt{GenomicRanges} code achieves this using the 'Hits' and 'List' 
-classes and their methods.}
-\label{fig:code}
-\end{figure}
+```{r code-comparsion, fig.cap = "(ref:code-comparison-cap)", echo = FALSE}
+knitr::include_graphics(here::here("img", "code-comparison.png"))
+```
+
+(ref:code-comparison-cap): Idiomatic code examples for \texttt{plyranges} (\textbf{A}) and \texttt{GenomicRanges} (\textbf{B}) illustrating an overlap and aggregate operation that returns the same result.
+In each example, we have two BED files consisting of SNPs that are genome-wide association study (GWAS) hits and reference exons.
+Each code block counts for each SNP the number of distinct exons it overlaps.
+The \texttt{plyranges} code achieves this with an overlap join followed by partitioning and aggregation.
+Strand is ignored by default here.
+The \texttt{GenomicRanges} code achieves this using the 'Hits' and 'List' classes and their methods.
 
@@
 `dplyr` grammar. Consequently, `plyranges` code is generally more compact
-than the equivalent `GenomicRanges` code (figure \ref{fig:code}).
+than the equivalent `GenomicRanges` code (figure \@ref(fig:code-comparison)).
 
-## Developing workflows with `plyranges`
+### Developing workflows with `plyranges`
 
@@
 
-### Peak Finding
+#### Peak Finding
 
@@
 
-Using `plyranges` and the the Bioconductor package `AnnotationHub` \cite{R-ahub}
+Using `plyranges` and the the Bioconductor package `AnnotationHub` [@R-ahub]
 we can download and read BigWig files from ChIP-Seq experiments from 
-the Human Epigenome Roadmap project \cite{Roadmap-Epigenomics-Consortium2015-pr}. 
+the Human Epigenome Roadmap project [@Roadmap-Epigenomics-Consortium2015-pr]. 
 Here we analyse a BigWig file corresponding to H3 lysine 27 trimethylation 
@@
 
-```{r load-bw, cache = TRUE}
-library(plyranges)
-chr10_ranges <- bw_file %>% 
-  get_genome_info() %>%
-  filter(seqnames == "chr10")
+```{r load-bw, echo = TRUE}
 ```
@@
 
-```{r}
-chr10_scores <- bw_file %>%
-  read_bigwig(overlap_ranges = chr10_ranges) %>%
-  set_genome_info(genome = "hg19")
-chr10_scores
+```{r chr10-scores-read, echo = TRUE}
 ```
@@
 
-```{r}
+```{r chr10-scores-reduce, echo = TRUE}
 all_peaks <- chr10_scores %>% 
@@
 ```{r max-score-region}
-chr10_max_score_region <- chr10_scores %>%
-  filter(score == max(score)) %>% 
-  anchor_center() %>%
-  mutate(width = 5000)
 ```
@@
 Finally, the overlap inner join is used to restrict the chromosome 10
-coverage islands, to the islands that are contained in the 5000nt region that surrounds the max peak (figure \ref{fig:peak-viz}).
+coverage islands, to the islands that are contained in the 5000nt region that surrounds the max peak (figure \@ref(fig:peak-viz)).
 
 ```{r peak_region}
-peak_region <- chr10_scores %>%
-  join_overlap_inner(chr10_max_score_region)
-peak_region
 ```
 
-```{r peak-viz, echo = FALSE, fig.cap = "The final result of the \\texttt{plyranges} operations to find a 5000nt region surrounding the peak of normalised coverage scores over chromosome 10, displayed as a density plot.", cache = TRUE}
-ggbio::autoplot(peak_region, 
-                aes(y = score.x), 
-                geom = "blank") + 
-  geom_area() +
-  ylab("score")
+(ref:peak-viz-cap): The final result of the `plyranges` operations to find a 5000nt region surrounding the peak of normalised coverage scores over chromosome 10, displayed as a density plot.
+
+```{r peak-viz, fig.width = 7, out.width = "70%", fig.cap = "(ref:peak-viz-cap)"}
 ```
 
-### Computing Windowed Statistics
+#### Computing Windowed Statistics
 
@@
 for each alignment using the `letterFrequency()` function from the `Biostrings`
-package \cite{R-biostrings}. After computing the GC proportion as the score
+package [@R-biostrings]. After computing the GC proportion as the score
 column, we drop  all other columns in the GRanges object.
 
-```{r, cache = TRUE}
-alignments <- h1_bam_sorted %>% 
-  read_bam() %>%
-  filter_by_overlaps(locations) %>%
-  select(seq) %>%
-  mutate(
-    score = as.numeric(letterFrequency(seq, "GC", as.prob = TRUE))
-  ) %>% 
-  select(score)
-alignments
+```{r read-bam, echo = TRUE}
 ```
@@
 
-```{r, cache = TRUE}
-bins <- locations %>% 
-  tile_ranges(width = 10000L)
+```{r read-bam-summary, echo = TRUE}
+```
 
-alignments_summary <- bins %>%
-  join_overlap_inner(alignments) %>% 
-  disjoin_ranges(n = n(), avg_gc = mean(score))
-alignments_summary
-```
-<!-- 
-https://support.bioconductor.org/p/71601/ also useful example?
--->
-### Quality Control Metrics
+#### Quality Control Metrics
 
-```{r, echo = FALSE, warning = FALSE, cache = TRUE}
-# prepare a GRanges object from array data
-library(readr)
-# annotation information available here
-# https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GPL18952
-# chromosome zero are 'problem' probes and should be filtered
-anno <- read_tsv(array_info,
-                 trim_ws = TRUE,
-                 col_types = c("cccc----")) %>%
-  dplyr::filter(Chr != "0")
-
-# genotypes for H1 cell line, data starts on line 10
-# data obtained from https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM1463263
-geno <- read_tsv(snp_info, skip = 10)
-
-# rename columns
-geno <- geno %>% 
-  dplyr::select(name = `SNP Name`, 
-         score = `GC Score`, 
-         baf = `B Allele Freq`, 
-         logR = `Log R Ratio`)
-
-# merge everything
-complete_vranges_df <- dplyr::inner_join(geno, anno, by = c("name" = "Name")) %>%
-  dplyr::rename(alleles = Alleles, seqnames = Chr, start = MapInfo)
-
-h1_snp_array <- as_granges(complete_vranges_df,
-                           start = as.integer(start),
-                           end = as.integer(start))
-h1_snp_array <- h1_snp_array %>% 
-  mutate(ref = stringr::str_replace(alleles, "\\[(.)/.*", "\\1"),
-         alt = stringr::str_replace(alleles, ".*/(.)\\]", "\\1"))
+```{r load-array, cache = TRUE}
 ```
@@
 
-```{r, cache = TRUE}
-h1_snp_array <- h1_snp_array %>%
-  filter(!(ref %in% c("I", "D")), seqnames != "M") %>%
-  mutate(transition = (ref %in% c("A", "G") & alt %in% c("G","A"))|
-                      (ref %in% c("C","T") & alt %in% c("T", "C")))
+```{r process-array, cache = TRUE, echo = TRUE}
 ```
@@
 We then compute the transition-transversion ratio over each chromosome using
-`group_by()` in combination with `summarize()` (figure \ref{fig:titv-viz}).
+`group_by()` in combination with `summarize()` (figure \@ref(fig:titv-viz)).
 
-```{r, cache = TRUE}
-ti_tv_results <- h1_snp_array %>% 
-  group_by(seqnames) %>%
-  summarize(n_snps = n(),
-            ti_tv = sum(transition) / sum(!transition)) 
-ti_tv_results
+```{r titv, cache = TRUE, echo = TRUE}
 ```
 
-```{r titv-viz, echo = FALSE, fig.cap = "The final result of computing quality control metrics over the SNP array data with \\texttt{plyranges}, displayed as a dot plot.
+```{r titv-viz, echo = FALSE, fig.cap = "(ref:titv-viz-cap)"}
+```
+
+(ref:titv-viz-cap): The final result of computing quality control metrics over the SNP array data with `plyranges`, displayed as a dot plot.
 Chromosomes are ordered by their estimated transition-transversion ratio.
-A white reference line is drawn at the expected ratio for a human exome."}
-ti_tv_results %>% 
-  as.data.frame() %>% 
-  ggplot(aes(forcats::fct_reorder(seqnames, ti_tv), ti_tv)) +
-  geom_point() + 
-  geom_hline(yintercept = 3, colour = "white", size = 4) +
-  labs(x = "Chromosome", y = "Transition Transversion Ratio") +
-  coord_flip() 
-```
-<!-- 
-Trivial example? Could do something with a smooth summarised over
-genomic coordinates? 
--->
+A white reference line is drawn at the expected ratio for a human exome."
 
-# Discussion
+## Discussion
 
@@
 language and API design: cognitive consistency, cohesion,
-endomorphism and expressiveness \cite{Green1996-qg}. 
+endomorphism and expressiveness [@Green1996-qg]. 
 To varying degrees, these principles also underlie the
@@
 `plyranges` makes extensive use of non-standard evaluation in R via
-the `rlang` package \cite{R-rlang}. Simply, this means that computations are
+the `rlang` package [@R-rlang]. Simply, this means that computations are
 evaluated in the context of the GRanges objects. Both `dplyr` and
@@
 
-# Conclusion
+## Conclusion
 
@@
 As the `plyranges` interface encourages tidy data practices, it
-integrates well with the grammar of graphics \cite{Wickham2016-gz}.  To
+integrates well with the grammar of graphics [@Wickham2016-gz].  To
 achieve responsive performance, interactive graphics rely on lazy data
@@
 `plyranges` should help support interactive genomics applications.
-
-# Declarations
 
@@
 
-## Availabliity of data and Materials
+## Availability of Data and Materials
 
@@
 was downloaded from the `AnnotationHub` package (2.13.1)
-under accession AH33458 \cite{R-ahub, Roadmap-Epigenomics-Consortium2015-pr}. 
+under accession AH33458 [@R-ahub; @Roadmap-Epigenomics-Consortium2015-pr]. 
 The BAM file corresponding to the H1 cell line ChIP-seq data is available at 
-NCBI GEO under accession [GSM433167](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM433167) \cite{ncbi-geo, Roadmap-Epigenomics-Consortium2015-pr}. 
+NCBI GEO under accession [GSM433167](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM433167) [@ncbi-geo; @Roadmap-Epigenomics-Consortium2015-pr]. 
 The SNP array data for the H1 cell line data is available at  NCBI GEO under accession
 [GPL18952](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GPL18952) 
-\cite{Roadmap-Epigenomics-Consortium2015-pr}.
+[@Roadmap-Epigenomics-Consortium2015-pr].
 
 The `plyranges` package is open source under an Artistic 2.0 license 
-\cite{R-plyranges-z}. 
-The  version (1.1.10) used in the paper is archived under DOI: [10.5281/zenodo.1469841](https://doi.org/10.5281/zenodo.1469841). The software 
+[@R-plyranges-z]. The software 
 can be obtained via the Bioconductor project website 
@@
 [https://github.com/sa-lee/plyranges](https://github.com/sa-lee/plyranges).
-
-To produce the workflows as described in results section we used 
-R version 3.5 with the development version of `plyranges` (1.1.10)   
-and the `BioStrings` package (2.49.0) installed \cite{r-core, R-plyranges-z, R-biostrings}.
-
-This article was written with `knitr` \cite{R-knitr} and 
-the figures were made with `ggbio` \cite{R-ggbio}. All code required to 
-reproduce this article is available at https://github.com/sa-lee/plyranges-paper.
-
-## Funding
-
-SL is funded by an Australian Government Research Training Program (RTP) 
-scholarship and receives an additional top up scholarship from CSL. Funding 
-bodies had no role in data collection, analysis, and interpretation.
-
-## Authors Contributions
-
-SL, DC, ML contributed ideas towards the software. SL wrote 
-the manuscript and analysed data. SL and ML wrote the software. 
-DC and ML supervised the research and revised the manuscript. 
-All authors have read and approved the final version of the manuscript.
-
-## Ethics approval and consent to participate
-
-Not applicable to the current manuscript.
-
-## Competing Interests
-
-Not applicable to the current manuscript.
-
-# References {#references .unnumbered}

### 2020-02-21 index.Rmd
update path to bib
plyranges chapter edits

@@
 degrees: 'B.MatComSci., University of Adelaide; MSc.Bioinformatics. University of Melbourne'
-bibliography: [bib/plyranges.bib]
+bibliography: [bib/plyranges.bib, bib/fluentGenomics.bib]
 link-citations: yes
@@
   warning = FALSE, message = FALSE, echo = FALSE,
-  fig.path = 'img/', fig.align = 'center', fig.show = 'hold',
+  fig.path = here::here('img'), fig.align = 'center', fig.show = 'hold',
   cache = TRUE, external = TRUE, comment = "#>", dpi = 300,
   out.width = ifelse(is_html_output(), "100%", "\\textwidth"),
-  cache.path = ifelse(is_html_output(), 'cache-gitbook/', 'cache-pdfbook/')
+  cache.path = ifelse(is_html_output(), 
+                      here::here('cache-gitbook'), 
+                      here::here('cache-pdfbook')
+  )
 )
@@
 
-This is the website for my PhD thesis at Monash University (Australia), titled "TBD". 
+This is the website for my PhD thesis at Monash University (Australia), titled "Fluent statistical computing interfaces for biological data analysis". 
 It is a work in progress.

### 2020-02-21 scripts/plyranges.R

@@ -1,79 +1,32 @@
-## ----setup, include = FALSE------------------------
-library(tibble)
-library(knitr)
-opts_chunk$set(message = FALSE,
-               warning = FALSE,
-               comment = "#>",
-               fig.width = 5,
-               fig.height = 3,
-               fig.align = "center",
-               fig.path = "./diagrams/")
-# for peak figures
+## ----plyranges-setup, include = FALSE------------------------
+library(BiocFileCache)
 library(ggbio)
-
-# for bam example
 library(Biostrings)
 library(plyranges)
-
-# -- Data from the Human Epigenomics RoadMap consortium
-# retrieving a BigWigFile from AnnotationHub
-
-if (!dir.exists("./data")) dir.create("./data")
-
 library(AnnotationHub)
-q <- AnnotationHub()
-# T-cell BW file
-bw_file <- q[["AH33458"]]
+# bam and snp array resources are stored in BiocFileCache
+bfc <- BiocFileCache(cache = here::here("data"))
+h1_bam_sorted <- bfcrpath(bfc, "GSM433167_BI.H3K27me3.sorted.bam")
+h1_bam_sorted_index <- bfcrpath(bfc, "GSM433167_BI.H3K27me3.sorted.bam.bai")
+h1_snp_array_path <- bfcrpath(bfc, "h1_snp_array") # this is an rds file
+# AnnotationHub resource for T-cell BigWig file
+bw_file <- AnnotationHub()[["AH33458"]]
 
-# a BamFile from the H1 cell line
-h1_bam <- "./data/GSM433167_BI.H3K27me3.bam"
-if (!file.exists(h1_bam))
-  download.file(
-    url = "ftp://ftp.ncbi.nlm.nih.gov/geo/samples/GSM433nnn/GSM433167/suppl/GSM433167%5FBI%2EH3K27me3%2Ebam",
-    destfile = h1_bam
-  )
-
-h1_bam_sorted <- "./data/GSM433167_BI.H3K27me3.sorted.bam"
-
-# sort and index bam
-if (!file.exists(h1_bam_sorted)) {
-  Rsamtools::sortBam(h1_bam, "./data/GSM433167_BI.H3K27me3.sorted")
-  Rsamtools::indexBam(h1_bam_sorted)
-}
-
-# array data for h1 cell line
-# illumina annotations
-array_info <- "data/GPL18952_HumanOmni25M-8v1-1_B.annotated.txt.gz"
-if (!file.exists(array_info))
-  download.file(
[…]

### 2020-02-21 Rmd/04-intron.Rmd
reorder chapters

# Tidy coverage analysis uncovers intron retention {#ch:intron}

# Introduction

## Data

We will use data from 12 samples from the human cell line mixture RNA-seq
data. There are three biological replicates across two library preparation kits,
a polyA kit and a total RNA kit. The BAM files have been previously aligned
with subread using the hg20 reference genome.

```{r prepare-design, echo= FALSE}
design <- readRDS(here("data", "superintronic", "design.rds"))
```

## `superintronic` analysis

A superintronic analysis consists of four steps:
1. Preparing the annotation and extract exonic/intronic parts
2. Computing coverage over regions of interest
3. Summarising coverage over regions of interest
4. Visualising the results

Steps 1 and 2. are performed in the `scripts/run_coverage.R`

### Preparing GFF/GTF files

```{r gff}
library(superintronic)
parts <- readRDS(here::here("data", "complete-annotation.rds"))
parts
```

These can be interrogated as desired, but for our purposes we will keep
genes that are protein coding, on main chromosomes but not mitochondrial 
genome, and do not overlap any other genes and have more than one exon.

```{r default-filter, echo = FALSE}
parts_sub <- readRDS(here::here("data", "filtered-annotation.rds"))
parts_sub
```

## Coverage over intronic/exonic regions

Now all the ingredients are in place to merge the coverage scores
to our prepared annotation with `join_parts()`. This again
returns a GRanges object, restricted to the intersection of the 
coverage ranges with intron/exon ranges. Additional columns are added
corresponding to the properties of the intron/exon.

```{r, load-features, echo = FALSE }
cvg_over_features <- readRDS(here::here("data", "parts-coverage.rds"))
```

Now we will compute summaries for each cell line and kit to detect IR 
genes. This consists of three steps

1. Computing the average log-coverage over all exon/intron features:

```{r rangenostics-01}
# compute intron/exon features
cvg_over_features <- cvg_over_features %>% 
  mutate(log_score = log2(score + 0.5))

cvg_by_features <- cvg_over_features %>% 
  group_by(Kit, CellLine, gene_id, feature_type)
E_vals <- cvg_by_features %>% 
  summarise(mn = Hmisc::wtd.mean(log_score, width),
            sd = sqrt(Hmisc::wtd.var(log_score, width)),
            raw_mn = Hmisc::wtd.mean(score, width),
            score = score,
            n_bases = width,
            start = min(start),
            end = max(end),
            seqnames = unlist(unique(seqnames)),
            strand = unlist(unique(feature_strand))) %>% 
  makeGRangesFromDataFrame(keep.extra.columns = TRUE)
```

2. For exon features computing the mean and variance of average log-coverage 
   over all genes (lumpiness).

```{r rangenostics-02}
lumpy <- E_vals %>% 
  filter(sd > 0, feature_type == "exon") %>% 
  group_by(Kit, CellLine) %>% 
  summarise(
    E_mu = mean(mn),
    E_sd = mean(sd),
    E_raw = mean(raw_mn)
)
lumpy 
mcols(E_vals) <- cbind(mcols(E_vals),
                       lumpy[match(paste0(E_vals$Kit, E_vals$CellLine), 
                                   paste0(lumpy$Kit, lumpy$CellLine)), -c(1,2)])
E_vals <- E_vals %>% 
  mutate(
    n_bases_E_raw = sum(n_bases[score > E_raw]),
  ) 
```

3. Compute the average number of bases with a 
   log-coverage score above the E values defined above. We have also computed
   a smooth between over the intron/exon feature standard deviation against
   the average for visualisation purposes.

```{r rangenostics-03}
rango <- E_vals %>% 
  plyranges::select(-n_bases, -score, -seqnames, -start, -end, -width, -strand, 
         .drop_ranges = TRUE) %>% 
  dplyr::as_tibble() %>% 
  group_by(Kit, CellLine, feature_type) %>% 
  tidyr::nest() %>% 
  mutate(
    smooth = lapply(data, function(x) {
      mgcv::gam(sd ~ s(mn), data = x)
      }),
    augment = lapply(smooth, broom::augment),
  ) %>% 
  tidyr::unnest(data, augment) %>% 
  dplyr::select(-dplyr::ends_with("1"), -.sigma)
```

4. Plot the results.

We get very similar results to a `voom` plot in a standard limma analysis. We
have also overlaid on the E_mu and E_sigma values as potential cut-offs for
each feature.

```{r voom-like-plot}
# rangenostics feature_average vs feature_sd coloured by n of bases above
# average raw exon coverage
library(ggplot2)
voom_plot <- function(.x, .y) {
  p <- ggplot(data = .x, aes(x = mn, y = sd)) +
    geom_point() +
    geom_line(aes(y = .fitted), colour = "blue") +
    geom_vline(aes(xintercept = E_mu), 
               data  = dplyr::distinct(.x, feature_type, E_mu)) +
    geom_hline(aes(yintercept = E_sd),
               data = dplyr::distinct(.x, feature_type, E_sd)) +
    facet_wrap(~feature_type) +
    labs(
         subtitle = paste(.y$Kit, .y$CellLine),
         x = "mean log-coverage",
         y = "sd log-covreage"
    )
  fname <- here::here("img", 
                      paste0(.y$Kit, "-", .y$CellLine, 
                             "voom-like-plot.png")
  )
  ggsave(fname, p)
}
rango %>% 
  group_by(Kit, CellLine) %>% 
  dplyr::group_walk(voom_plot)
```

We also plot pairs of summary values for each feature in a scatter plot matrix
(using a hexbin scatter plot). Again we get similar results to the plots
produced by a DE analysis from InDeX.

```{r rangenostics-pairs}
all_features <- tidyr::gather(rango, "key", "value", 
                              -Kit, -CellLine, -gene_id, -feature_type, 
                              -E_mu, -E_sd, -E_raw) %>% 
  mutate(var = paste0(feature_type, "_", key)) %>% 
  plyranges::select(-feature_type, -key) %>% 
  arrange(gene_id) %>% 
  tidyr::spread(var, value)
pairs_plot <- function(.x, .y) {
  sub <- plyranges::select(.x, 
                           exon_mn, 
                           intron_mn, 
                           intron_sd, 
                           bases_above = intron_n_bases_E_raw)
  fname <- here::here("img", 
                      paste0(.y$Kit, "-", .y$CellLine, 
                             "-pairs-plot.png")
  )

  p <- GGally::ggpairs(sub, columns = 1:4, lower = list(continuous = hex))
  ggsave(fname, p)
}
all_features %>% 
  group_by(Kit, CellLine) %>% 
  dplyr::group_walk(pairs_plot)

```

5. Decide on cut-off over each feature. It appears that choosing genes
with high intron feature standard deviations, and with number of bases above
E_mu in the 99th percentile discriminates intron retention features well.

```{r rangenositcs-hits}
hits <- all_features %>% 
  group_by(Kit, CellLine) %>% 
  filter(exon_mn > E_mu, 
         intron_sd > E_sd, 
         intron_n_bases_E_raw > quantile(intron_n_bases_E_raw, 0.99))

dplyr::count(hits)
```

6. We can also look at overlaps between sets using an UpSet plot:

```{r, echo  = FALSE}
sets <- hits %>% 
  dplyr::distinct(gene_id) %>%
  arrange(gene_id) %>% 
  mutate(value = 1L, var = paste(CellLine, Kit)) %>% 
  ungroup() %>%  
  plyranges::select(gene_id, var, value) %>% 
  tidyr::spread(var, value) %>% 
  dplyr::mutate_at(2:5, .funs = ~ dplyr::if_else(is.na(.), 0L, .)) 
sets <- as.data.frame(sets)
library(UpSetR)
pdf(here::here("img", "olaps-superintronic-hits.pdf"))
upset(sets)
dev.off()
```

### 2020-02-21 Rmd/03-fluentGenomics.Rmd
update fluentGenomics chapter

--- 2020-01-28 Rmd/04-fluentGenomics.Rmd
@@
-## Fluent genomics: integrating data with plyranges and tximeta {#ch:fluentGenomics}
+# Fluent genomics: integrating data with plyranges and tximeta {#ch:fluentGenomics}
+
+```{r fluentGenomics-init, echo = FALSE, cache = FALSE, include = FALSE}
+knitr::read_chunk(here::here("scripts", "fluentGenomics.R"))
+```
 
 ```{r setup, include = FALSE}
-library(fluentGenomics)
-dir <- system.file("extdata", package="macrophage")
-
-library(tximeta)
-makeLinkedTxome(
-  indexDir=file.path(dir, "gencode.v29_salmon_0.12.0"),
-  source="Gencode",
-  organism="Homo sapiens",
-  release="29",
-  genome="GRCh38",
-  fasta="ftp://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_human/release_29/gencode.v29.transcripts.fa.gz",
-  gtf=file.path(dir, "gencode.v29.annotation.gtf.gz"), # local version
-  write=FALSE
-)
-
-knitr::opts_chunk$set(
-  fig.align = "center"
-)
 ```
@@
 
-# Introduction
+## Introduction
 
@@
 
-```{r workflow, fig.cap = "(ref:workflow)", fig.align='center', out.width="\\textwidth", echo = FALSE}
-knitr::include_graphics("workflow.png")
+```{r workflow, fig.cap = "(ref:workflow)"}
+knitr::include_graphics(here::here("img", "workflow.png"))
 ```
@@
 
-## Experimental Data
+### Experimental Data
 
@@
 
-A precise description of how we obtained this *SummarizedExperiment* object can
-be found in section \@ref(atac).
+A precise description of how we obtained this *SummarizedExperiment* object can be found in section \@ref(atac).
 
-# Import Data as a *SummarizedExperiment* {#se}
+## Import Data as a *SummarizedExperiment* {#se}
 
-## Using *tximeta* to import RNA-seq quantification data
+### Using *tximeta* to import RNA-seq quantification data
 
@@
 
-```{r setdir}
-dir <- system.file("extdata", package="macrophage")
+```{r setdir, echo = TRUE}
 ```
@@
 
-```{r coldata-rna}
+```{r coldata-rna, echo = TRUE}
 library(dplyr)
@@
 
-```{r tximeta-run}
-suppressPackageStartupMessages(library(SummarizedExperiment))
-library(tximeta)
-se <- tximeta(coldata, dropInfReps=TRUE)
-se
+```{r tximeta-run, echo = TRUE}
 ```
@@
 
-```{r gse}
-gse <- summarizeToGene(se)
+```{r gse, echo = TRUE}
 ```
@@
 
-## Importing ATAC-seq data as a *SummarizedExperiment* object {#atac}
+### Importing ATAC-seq data as a *SummarizedExperiment* object {#atac}
 
@@
 
-# Model assays
+## Model assays
 
-## RNA-seq differential gene expression analysis
+### RNA-seq differential gene expression analysis
 
@@
 
-```{r setup-deseq}
-library(DESeq2)
-dds <- DESeqDataSet(gse, ~line + condition)
-# filter out lowly expressed genes
-# at least 10 counts in at least 6 samples
-keep <- rowSums(counts(dds) >= 10) >= 6
-dds <- dds[keep,]
+```{r setup-deseq, echo = TRUE}
 ```
@@
 
-```{r fit-model}
-dds <- DESeq(dds)
+```{r fit-model, echo = TRUE}
 ```
@@
 
-```{r results-DFrame}
-res <- results(dds,
-               contrast=c("condition","IFNg","naive"),
-               lfcThreshold=1, alpha=0.01)
+```{r results-DFrame, echo = TRUE}
 ```
@@
 
-```{r ma-plot, fig.cap="(ref:maplot)" }
-summary(res)
-DESeq2::plotMA(res, ylim=c(-10,10))
+```{r ma-plot, fig.cap="(ref:maplot)", echo = TRUE}
 ```
@@
 
-```{r results-GRanges}
-suppressPackageStartupMessages(library(plyranges))
-de_genes <- results(dds,
-                    contrast=c("condition","IFNg","naive"),
-                    lfcThreshold=1,
-                    format="GRanges") %>%
-  names_to_column("gene_id")
-de_genes
+```{r results-GRanges, echo = TRUE}
 ```
@@
 
-```{r de-genes}
-de_genes <- de_genes %>%
-  filter(padj < 0.01) %>%
-  select(gene_id, de_log2FC = log2FoldChange, de_padj = padj)
+```{r de-genes, echo = TRUE}
 ```
@@
 
-```{r not-de-genes}
-other_genes <- results(dds,
-                       contrast=c("condition","IFNg","naive"),
-                       lfcThreshold=1,
-                       altHypothesis="lessAbs",
-                       format="GRanges") %>%
-  filter(padj < 0.01) %>%
-  names_to_column("gene_id") %>%
-  dplyr::select(gene_id,
-                de_log2FC = log2FoldChange,
-                de_padj = padj)
+```{r not-de-genes, echo = TRUE}
 ```
 
-## ATAC-seq peak differential abundance analysis
+### ATAC-seq peak differential abundance analysis
 
@@
 
-<!-- We can check the standard deviation over mean plot, to assess for any -->
-<!-- systematic trends: -->
-
-<!-- ```{r, eval = FALSE} --> <!-- library(ggplot2) --> <!-- mv <-
-data.frame(rmu = rowMeans(assay(atac)), --> <!-- rvar = rowVars(assay(atac)))
---> <!-- ggplot(data =mv , aes(x = rmu, y = sqrt(rvar))) + --> <!-- geom_hex()
---> <!-- ``` -->
-
 For assessing differential accessibility, we run *limma* [@Smyth2004], and
@@
 
-```{r limma, eval = FALSE}
+```{r limma, eval = FALSE, echo = TRUE}
 library(limma)
@@
 
-```{r peaks-tidy, eval = FALSE}
+```{r peaks-tidy, eval = FALSE, echo = TRUE}
 atac_peaks <- rowRanges(atac) %>%
@@
 
-```{r load-peaks}
-library(fluentGenomics)
-peaks
+```{r load-peaks, echo = TRUE}
 ```
 
-# Integrate ranges
+## Integrate ranges
 
-## Finding overlaps with *plyranges*
+### Finding overlaps with *plyranges*
 
@@
 
-```{r filter-peaks}
-da_peaks <- peaks %>%
-  filter(da_padj < 0.01)
+```{r filter-peaks, echo = TRUE}
 ```
@@
 
-## Down sampling non-differentially expressed genes
+### Down sampling non-differentially expressed genes
 
@@
 
-```{r slice-example}
-size <- length(de_genes)
-slice(other_genes, sample.int(n(), size))
+```{r slice-example, echo = TRUE}
 ```
@@
 
-```{r boot-set-01}
-# set a seed for the results
-set.seed(2019-08-02)
-boot_genes <- replicate(10,
-                        slice(other_genes, sample.int(n(), size)),
-                        simplify = FALSE)
+```{r boot-set-01, echo = TRUE}
 ```
@@
 
-```{r boot-set-02}
-boot_genes <- bind_ranges(boot_genes, .id = "resample")
+```{r boot-set-02, echo = TRUE}
 ```
@@
 
-```{r combine-results}
-all_genes <- bind_ranges(
-  de=de_genes,
-  not_de = boot_genes,
-  .id="origin"
-) %>%
-  mutate(
-    origin = factor(origin, c("not_de", "de")),
-    resample = ifelse(is.na(resample), 0L, as.integer(resample))
-  )
-all_genes
+```{r combine-results, echo = TRUE}
 ```
 
-## Expanding genomic coordinates around the transcription start site
+### Expanding genomic coordinates around the transcription start site
 
@@
 
-```{r resize-01}
-all_genes <- all_genes %>%
-  anchor_5p() %>%
-  mutate(width = 1)
+```{r resize-01, echo = TRUE}
 ```
@@
 
-```{r resize-02}
-all_genes <- all_genes %>%
-  anchor_center() %>%
-  mutate(width=2*1e4)
+```{r resize-02, echo = TRUE}
 ```
 
-## Use overlap joins to find relative enrichment
+### Use overlap joins to find relative enrichment
 
@@
 
-```{r olap-join}
-genes_olap_peaks <- all_genes %>%
-  join_overlap_left(da_peaks)
-genes_olap_peaks
+```{r olap-join, echo = FALSE}
 ```
@@
 
-```{r reduce-ex01}
-gene_peak_max_lfc <- genes_olap_peaks %>%
-  group_by(gene_id, origin)  %>%
-  reduce_ranges_directed(
-    peak_count = sum(!is.na(da_padj)) / n_distinct(resample),
-    peak_max_lfc = max(abs(da_log2FC))
-  )
+```{r reduce-ex01, echo = FALSE}
 ```
@@
 
-```{r boxplot, fig.cap = "(ref:boxplot)"}
-library(ggplot2)
-gene_peak_max_lfc %>%
-  filter(peak_count > 0) %>%
-  as.data.frame() %>%
-  ggplot(aes(origin, peak_max_lfc)) +
-  geom_boxplot()
+```{r boxplot, fig.cap = "(ref:boxplot)", echo = TRUE}
 ```
@@
 
-```{r summarize-ex01}
-origin_peak_lfc <- genes_olap_peaks %>%
-  group_by(origin) %>%
-  summarize(
-    peak_count = sum(!is.na(da_padj)) / n_distinct(resample),
-    lfc1_peak_count =sum(abs(da_log2FC) > 1, na.rm=TRUE)/ n_distinct(resample),
-    lfc2_peak_count = sum(abs(da_log2FC) > 2, na.rm=TRUE)/ n_distinct(resample)
-  )
-origin_peak_lfc
+```{r summarize-ex01, echo = TRUE}
 ```
@@
 
-```{r pivot-enrich}
-origin_peak_lfc %>%
-  as.data.frame() %>%
-  tidyr::pivot_longer(cols = -origin) %>%
-  tidyr::pivot_wider(names_from = origin, values_from = value) %>%
-  mutate(enrichment = de / not_de)
+```{r pivot-enrich, echo = TRUE}
 ```
@@
 
-```{r reduce-summarize}
-genes_olap_peaks %>%
-  group_by(gene_id, origin, resample) %>%
-  reduce_ranges_directed(
-    lfc1 = sum(abs(da_log2FC) > 1, na.rm=TRUE),
-    lfc2 = sum(abs(da_log2FC) > 2, na.rm=TRUE)
-  ) %>%
-  group_by(origin) %>%
-  summarize(
-    lfc1_gene_count = sum(lfc1 > 0) / n_distinct(resample),
-    lfc1_peak_count = sum(lfc1) / n_distinct(resample),
-    lfc2_gene_count = sum(lfc2 > 0) / n_distinct(resample),
-    lfc2_peak_count = sum(lfc2) / n_distinct(resample)
-  )
+```{r reduce-summarize, echo = TRUE}
 ```
@@
 
-```{r count-fn}
-count_if_above_threshold <- function(var, thresholds) {
-  lapply(thresholds, function(.) sum(abs(var) > ., na.rm = TRUE))
-}
+```{r count-fn, echo = TRUE}
 ```
@@
 
-```{r thresholds}
-thresholds <- da_peaks %>%
-  mutate(abs_lfc = abs(da_log2FC)) %>%
-  with(
-    seq(min(abs_lfc), max(abs_lfc), length.out = 100)
-  )
+```{r thresholds, echo = TRUE}
 ```
@@
 
-```{r reduce-ex02}
-genes_peak_all_thresholds <- genes_olap_peaks %>%
-  group_by(gene_id, origin, resample) %>%
-  reduce_ranges_directed(
-    value = count_if_above_threshold(da_log2FC, thresholds),
-    threshold = list(thresholds)
-  )
-genes_peak_all_thresholds
+```{r reduce-ex02, echo = TRUE}
 ```
@@
 
-```{r expand-summarize}
-origin_peak_all_thresholds <- genes_peak_all_thresholds %>%
-  expand_ranges() %>%
-  group_by(origin, threshold) %>%
-  summarize(
-    gene_count = sum(value > 0) / n_distinct(resample),
-    peak_count = sum(value) / n_distinct(resample)
-  )
-origin_peak_all_thresholds
+```{r expand-summarize, echo = TRUE}
 ```
@@
 
-```{r line-chart, fig.cap = "(ref:linechart)"}
-origin_threshold_counts <- origin_peak_all_thresholds %>%
-  as.data.frame() %>%
-  tidyr::pivot_longer(cols = -c(origin, threshold),
-                      names_to = c("type", "var"),
-                      names_sep = "_",
-                      values_to = "count") %>%
-  select(-var)
-
-origin_threshold_counts %>%
-  filter(type == "peak") %>%
-  tidyr::pivot_wider(names_from = origin, values_from = count) %>%
-  mutate(enrichment =  de / not_de) %>%
-  ggplot(aes(x = threshold, y = enrichment)) +
-  geom_line() +
-  labs(x = "logFC threshold", y = "Relative Enrichment")
+```{r line-chart, fig.cap = "(ref:linechart)", echo = TRUE}
 ```
@@
 
-```{r line-chart2, fig.cap = "(ref:linechart2)"}
-origin_threshold_counts %>%
-  ggplot(aes(x = threshold,
-             y = count + 1,
-             color = origin,
-             linetype = type)) +
-  geom_line() +
-  scale_y_log10()
+```{r line-chart2, fig.cap = "(ref:linechart2)", echo = TRUE}
 ```
 
-# Discussion
+## Discussion
 
@@
 
-# Software Availability
+## Software Availability
 
-The workflow materials, including this article can be fully reproduced
+The workflow materials can be fully reproduced
 following the instructions found at the Github repository
@@
 
-This article and the analyses were performed with R [@baser] using the
-*rmarkdown* [@rmarkdown], and *knitr* [@knitr; @xie2015] packages.
-
-## Session Info
-
-```{r}
-sessionInfo()
-```
-
-## Author Contributions
-
-All authors contributed to the writing and development of the workflow.
-
-## Competing interests
-
-The authors declare that they have no competing interests.
-
-## Funding
-
-SL is supported by an Australian Government Research Training Program (RTP)
-scholarship with a top up scholarship from CSL Limited.
-
-MIL's contribution is supported by NIH grant R01 HG009937.
-
-__I confirm that the funders had no role in study design, data collection and
-analysis, decision to publish, or preparation of the manuscript.__
-
 ## Acknowledgements
@@
 of this workflow paper.
-
-# References

### 2020-02-21 scripts/fluentGenomics.R

@@ -14,27 +14,6 @@ makeLinkedTxome(
   write=FALSE
 )
 
-knitr::opts_chunk$set(
-  fig.align = "center"
-)
-
-
-## ----workflow, fig.cap = "(ref:workflow)", fig.align='center', out.width="\\textwidth", echo = FALSE----
-knitr::include_graphics("workflow.png")
-
-
-## ----cache-path, eval = FALSE----------------------
-## library(fluentGenomics)
-## path_to_se <- cache_atac_se()
-
-
-## ----read-cache, eval = FALSE----------------------
-## atac <- readRDS(path_to_se)
-
-
-## ----dir, eval=FALSE-------------------------------
-## dir <- "/path/to/quant/files"
-
 
 ## ----setdir----------------------------------------
 dir <- system.file("extdata", package="macrophage")
@@ -66,58 +45,8 @@ se <- tximeta(coldata, dropInfReps=TRUE)
 se
 
 
-## ----linkedtxome-ex, eval = FALSE------------------
-## makeLinkedTxome(
-##   indexDir=file.path(dir, "gencode.v29_salmon_0.12.0"),
-##   source="Gencode",
-##   organism="Homo sapiens",
-##   release="29",
-##   genome="GRCh38",
-##   fasta="ftp://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_human/release_29/gencode.v29.transcripts.fa.gz",
-##   gtf=file.path(dir, "gencode.v29.annotation.gtf.gz"), # local version
-##   write=FALSE
-## )
-
-
 ## ----gse-------------------------------------------
 gse <- summarizeToGene(se)
-
-
-## ----coldata-atac, eval=FALSE----------------------
-## atac_coldata <- read_tsv("ATAC_sample_metadata.txt.gz") %>%
-##   select(
-##     sample_id,
-##     donor,
-##     condition = condition_name
-##   ) %>%
-##   mutate(condition = relevel(factor(condition), "naive"))
-
-
-## ----mat-atac, eval=FALSE--------------------------
-## atac_mat <- read_tsv("ATAC_cqn_matrix.txt.gz",
-##                      skip = 1,
-##                      col_names =c("rownames", atac_coldata[["sample_id"]]))
-## rownames <- atac_mat[["rownames"]]
-## atac_mat <- as.matrix(atac_mat[,-1])
-## rownames(atac_mat) <- rownames
-
-
-## ----peaks-atac, eval=FALSE------------------------
-## library(plyranges)
[…]

### 2020-02-21 Rmd/01-intro.Rmd
renaming

# Introduction {#ch:intro}

overview

describe main topic - research questions
chapters 1, 2, 3, 4 (your contributions to the joint work)
scope

### 2020-02-21 Rmd/05-embeddings.Rmd
renaming

# Spurious embeddings from non-linear dimensionality reduction methods and how to find them {#ch:tsne}

### 2020-02-21 Rmd/00-b-abstract.Rmd
first build of chapters

@@
 This thesis presents a set of tools for performing exploratory 
-data analysis in the context of genomics and transcriptomics.  The first tool, “plryanges” provides a domain specific language for 
+data analysis in the context of genomics and transcriptomics.  The first tool, “plyranges” provides a domain specific language for 
 reasoning about range based genomics data that simplifies common 

### 2020-02-21 Rmd/03-fluentGenomics.Rmd
first build of chapters

@@
 
-(ref:workflow) An overview of the fluent genomics workflow. First, we *import*
-data as a *SummarizedExperiment* object, which enables interoperability with
-downstream analysis packages. Then we *model* our assay data, using the
-existing Bioconductor packages *DESeq2* and *limma*. We take the results of our
-models for each assay with respect to their genomic coordinates, and
-*integrate* them. First, we compute the overlap between the results of each
-assay, then aggregate over the combined genomic regions, and finally summarize
-to compare enrichment for differentially expressed genes to non differentially
-expressed genes. The final output can be used for downstream visualization or
-further transformation.
+(ref:workflow) An overview of the fluent genomics workflow.
+First, we *import* data as a *SummarizedExperiment* object, which enables interoperability with downstream analysis packages.
+Then we *model* our assay data, using the existing Bioconductor packages *DESeq2* and *limma*.
+We take the results of our models for each assay with respect to their genomic coordinates, and *integrate* them.
+First, we compute the overlap between the results of each assay, then aggregate over the combined genomic regions, and finally summarize to compare enrichment for differentially expressed genes to non differentially expressed genes.
+The final output can be used for downstream visualization or further transformation.
 
-(ref:maplot) Visualization of *DESeq2* results as an "MA plot". Genes that have
-an adjusted *p-value* below 0.01 are colored red.
+(ref:maplot) Visualization of *DESeq2* results as an "MA plot". Genes that have an adjusted *p-value* below 0.01 are colored red.
 
-(ref:boxplot) A boxplot of maximum LFCs for DA peaks for DE genes compared to
-non-DE genes where genes have at least one DA peak.
+(ref:boxplot) A boxplot of maximum LFCs for DA peaks for DE genes compared to non-DE genes where genes have at least one DA peak.
 
-(ref:linechart) A line chart displaying how relative enrichment of DA peaks
-change between DE genes compared to non-DE genes as the absolute DA LFC
-threshold increases.
+(ref:linechart) A line chart displaying how relative enrichment of DA peaks change between DE genes compared to non-DE genes as the absolute DA LFC threshold increases.
 
-(ref:linechart2) A line chart displaying how gene and peak counts change as the
-absolute DA LFC threshold increases. Lines are colored according to whether
-they represent a gene that is DE or not. Note the x-axis is on a $\log_{10}$
-scale.
+(ref:linechart2) A line chart displaying how gene and peak counts change as the absolute DA LFC threshold increases.
+Lines are colored according to whether they represent a gene that is DE or not.
+Note the x-axis is on a $\log_{10}$ scale.
 
@@
 
-```{r cache-path, eval = FALSE}
+```{r cache-path, eval = FALSE, echo = TRUE}
 library(fluentGenomics)
@@
 
-```{r read-cache, eval = FALSE}
+```{r read-cache, eval = FALSE, echo = TRUE}
 atac <- readRDS(path_to_se)
@@
 
-```{r linkedtxome-ex, eval = FALSE}
+```{r linkedtxome-ex, eval = FALSE, echo = TRUE}
 makeLinkedTxome(
@@
 
-```{r, eval = FALSE}
+```{r, eval = FALSE, echo = TRUE}
 # development version from Github

### 2020-02-21 Rmd/04-intron.Rmd
first build of chapters

@@
 
-# Introduction
+<!-- # Introduction -->
 
-## Data
+<!-- ## Data -->
 
-We will use data from 12 samples from the human cell line mixture RNA-seq
-data. There are three biological replicates across two library preparation kits,
-a polyA kit and a total RNA kit. The BAM files have been previously aligned
-with subread using the hg20 reference genome.
+<!-- We will use data from 12 samples from the human cell line mixture RNA-seq -->
+<!-- data. There are three biological replicates across two library preparation kits, -->
+<!-- a polyA kit and a total RNA kit. The BAM files have been previously aligned -->
+<!-- with subread using the hg20 reference genome.  -->
 
-```{r prepare-design, echo= FALSE}
-design <- readRDS(here("data", "superintronic", "design.rds"))
-```
+<!-- ```{r prepare-design, echo= FALSE} -->
+<!-- design <- readRDS(here("data", "superintronic", "design.rds")) -->
+<!-- ``` -->
 
-## `superintronic` analysis
+<!-- ## `superintronic` analysis -->
 
-A superintronic analysis consists of four steps:
-1. Preparing the annotation and extract exonic/intronic parts
-2. Computing coverage over regions of interest
-3. Summarising coverage over regions of interest
-4. Visualising the results
+<!-- A superintronic analysis consists of four steps: -->
+<!-- 1. Preparing the annotation and extract exonic/intronic parts -->
+<!-- 2. Computing coverage over regions of interest -->
+<!-- 3. Summarising coverage over regions of interest -->
+<!-- 4. Visualising the results -->
 
-Steps 1 and 2. are performed in the `scripts/run_coverage.R`
+<!-- Steps 1 and 2. are performed in the `scripts/run_coverage.R` -->
 
-### Preparing GFF/GTF files
+<!-- ### Preparing GFF/GTF files -->
 
-```{r gff}
-library(superintronic)
-parts <- readRDS(here::here("data", "complete-annotation.rds"))
-parts
-```
+<!-- ```{r gff} -->
+<!-- library(superintronic) -->
+<!-- parts <- readRDS(here::here("data", "complete-annotation.rds")) -->
+<!-- parts -->
+<!-- ``` -->
 
-These can be interrogated as desired, but for our purposes we will keep
-genes that are protein coding, on main chromosomes but not mitochondrial 
-genome, and do not overlap any other genes and have more than one exon.
+<!-- These can be interrogated as desired, but for our purposes we will keep -->
+<!-- genes that are protein coding, on main chromosomes but not mitochondrial  -->
+<!-- genome, and do not overlap any other genes and have more than one exon. -->
 
-```{r default-filter, echo = FALSE}
-parts_sub <- readRDS(here::here("data", "filtered-annotation.rds"))
-parts_sub
-```
+<!-- ```{r default-filter, echo = FALSE} -->
+<!-- parts_sub <- readRDS(here::here("data", "filtered-annotation.rds")) -->
+<!-- parts_sub -->
+<!-- ``` -->
 
-## Coverage over intronic/exonic regions
+<!-- ## Coverage over intronic/exonic regions -->
 
-Now all the ingredients are in place to merge the coverage scores
-to our prepared annotation with `join_parts()`. This again
-returns a GRanges object, restricted to the intersection of the 
-coverage ranges with intron/exon ranges. Additional columns are added
-corresponding to the properties of the intron/exon.
+<!-- Now all the ingredients are in place to merge the coverage scores -->
+<!-- to our prepared annotation with `join_parts()`. This again -->
+<!-- returns a GRanges object, restricted to the intersection of the  -->
+<!-- coverage ranges with intron/exon ranges. Additional columns are added -->
+<!-- corresponding to the properties of the intron/exon. -->
 
-```{r, load-features, echo = FALSE }
-cvg_over_features <- readRDS(here::here("data", "parts-coverage.rds"))
-```
+<!-- ```{r, load-features, echo = FALSE } -->
+<!-- cvg_over_features <- readRDS(here::here("data", "parts-coverage.rds")) -->
+<!-- ``` -->
 
-Now we will compute summaries for each cell line and kit to detect IR 
-genes. This consists of three steps
+<!-- Now we will compute summaries for each cell line and kit to detect IR  -->
+<!-- genes. This consists of three steps -->
 
-1. Computing the average log-coverage over all exon/intron features:
+<!-- 1. Computing the average log-coverage over all exon/intron features: -->
 
-```{r rangenostics-01}
-# compute intron/exon features
-cvg_over_features <- cvg_over_features %>% 
-  mutate(log_score = log2(score + 0.5))
+<!-- ```{r rangenostics-01} -->
+<!-- # compute intron/exon features -->
+<!-- cvg_over_features <- cvg_over_features %>%  -->
+<!--   mutate(log_score = log2(score + 0.5))   -->
 
-cvg_by_features <- cvg_over_features %>% 
-  group_by(Kit, CellLine, gene_id, feature_type)
-E_vals <- cvg_by_features %>% 
-  summarise(mn = Hmisc::wtd.mean(log_score, width),
-            sd = sqrt(Hmisc::wtd.var(log_score, width)),
-            raw_mn = Hmisc::wtd.mean(score, width),
-            score = score,
-            n_bases = width,
-            start = min(start),
-            end = max(end),
-            seqnames = unlist(unique(seqnames)),
-            strand = unlist(unique(feature_strand))) %>% 
-  makeGRangesFromDataFrame(keep.extra.columns = TRUE)
-```
+<!-- cvg_by_features <- cvg_over_features %>%  -->
+<!--   group_by(Kit, CellLine, gene_id, feature_type) -->
+<!-- E_vals <- cvg_by_features %>%  -->
+<!--   summarise(mn = Hmisc::wtd.mean(log_score, width), -->
+<!--             sd = sqrt(Hmisc::wtd.var(log_score, width)), -->
+<!--             raw_mn = Hmisc::wtd.mean(score, width), -->
+<!--             score = score, -->
+<!--             n_bases = width, -->
+<!--             start = min(start), -->
+<!--             end = max(end), -->
+<!--             seqnames = unlist(unique(seqnames)), -->
+<!--             strand = unlist(unique(feature_strand))) %>%  -->
+<!--   makeGRangesFromDataFrame(keep.extra.columns = TRUE) -->
+<!-- ``` -->
 
-2. For exon features computing the mean and variance of average log-coverage 
-   over all genes (lumpiness).
+<!-- 2. For exon features computing the mean and variance of average log-coverage  -->
+<!--    over all genes (lumpiness). -->
 
-```{r rangenostics-02}
-lumpy <- E_vals %>% 
-  filter(sd > 0, feature_type == "exon") %>% 
-  group_by(Kit, CellLine) %>% 
-  summarise(
-    E_mu = mean(mn),
-    E_sd = mean(sd),
-    E_raw = mean(raw_mn)
-)
-lumpy 
-mcols(E_vals) <- cbind(mcols(E_vals),
-                       lumpy[match(paste0(E_vals$Kit, E_vals$CellLine), 
-                                   paste0(lumpy$Kit, lumpy$CellLine)), -c(1,2)])
-E_vals <- E_vals %>% 
-  mutate(
-    n_bases_E_raw = sum(n_bases[score > E_raw]),
-  ) 
-```
+<!-- ```{r rangenostics-02} -->
+<!-- lumpy <- E_vals %>%  -->
+<!--   filter(sd > 0, feature_type == "exon") %>%  -->
+<!--   group_by(Kit, CellLine) %>%  -->
+<!--   summarise( -->
+<!--     E_mu = mean(mn), -->
+<!--     E_sd = mean(sd), -->
+<!--     E_raw = mean(raw_mn) -->
+<!-- ) -->
+<!-- lumpy  -->
+<!-- mcols(E_vals) <- cbind(mcols(E_vals), -->
+<!--                        lumpy[match(paste0(E_vals$Kit, E_vals$CellLine),  -->
+<!--                                    paste0(lumpy$Kit, lumpy$CellLine)), -c(1,2)]) -->
+<!-- E_vals <- E_vals %>%  -->
+<!--   mutate( -->
+<!--     n_bases_E_raw = sum(n_bases[score > E_raw]), -->
+<!--   )  -->
+<!-- ``` -->
 
-3. Compute the average number of bases with a 
-   log-coverage score above the E values defined above. We have also computed
-   a smooth between over the intron/exon feature standard deviation against
-   the average for visualisation purposes.
+<!-- 3. Compute the average number of bases with a  -->
+<!--    log-coverage score above the E values defined above. We have also computed -->
+<!--    a smooth between over the intron/exon feature standard deviation against -->
+<!--    the average for visualisation purposes. -->
 
-```{r rangenostics-03}
-rango <- E_vals %>% 
-  plyranges::select(-n_bases, -score, -seqnames, -start, -end, -width, -strand, 
-         .drop_ranges = TRUE) %>% 
-  dplyr::as_tibble() %>% 
-  group_by(Kit, CellLine, feature_type) %>% 
-  tidyr::nest() %>% 
-  mutate(
-    smooth = lapply(data, function(x) {
-      mgcv::gam(sd ~ s(mn), data = x)
-      }),
-    augment = lapply(smooth, broom::augment),
-  ) %>% 
-  tidyr::unnest(data, augment) %>% 
-  dplyr::select(-dplyr::ends_with("1"), -.sigma)
-```
+<!-- ```{r rangenostics-03} -->
+<!-- rango <- E_vals %>%  -->
+<!--   plyranges::select(-n_bases, -score, -seqnames, -start, -end, -width, -strand,  -->
+<!--          .drop_ranges = TRUE) %>%  -->
+<!--   dplyr::as_tibble() %>%  -->
+<!--   group_by(Kit, CellLine, feature_type) %>%  -->
+<!--   tidyr::nest() %>%  -->
+<!--   mutate( -->
+<!--     smooth = lapply(data, function(x) { -->
+<!--       mgcv::gam(sd ~ s(mn), data = x) -->
+<!--       }), -->
+<!--     augment = lapply(smooth, broom::augment), -->
+<!--   ) %>%  -->
+<!--   tidyr::unnest(data, augment) %>%  -->
+<!--   dplyr::select(-dplyr::ends_with("1"), -.sigma) -->
+<!-- ``` -->
 
-4. Plot the results.
+<!-- 4. Plot the results. -->
 
-We get very similar results to a `voom` plot in a standard limma analysis. We
-have also overlaid on the E_mu and E_sigma values as potential cut-offs for
-each feature.
+<!-- We get very similar results to a `voom` plot in a standard limma analysis. We -->
+<!-- have also overlaid on the E_mu and E_sigma values as potential cut-offs for -->
+<!-- each feature.  -->
 
-```{r voom-like-plot}
-# rangenostics feature_average vs feature_sd coloured by n of bases above
-# average raw exon coverage
-library(ggplot2)
-voom_plot <- function(.x, .y) {
-  p <- ggplot(data = .x, aes(x = mn, y = sd)) +
-    geom_point() +
-    geom_line(aes(y = .fitted), colour = "blue") +
-    geom_vline(aes(xintercept = E_mu), 
-               data  = dplyr::distinct(.x, feature_type, E_mu)) +
-    geom_hline(aes(yintercept = E_sd),
-               data = dplyr::distinct(.x, feature_type, E_sd)) +
-    facet_wrap(~feature_type) +
-    labs(
-         subtitle = paste(.y$Kit, .y$CellLine),
-         x = "mean log-coverage",
-         y = "sd log-covreage"
-    )
-  fname <- here::here("img", 
-                      paste0(.y$Kit, "-", .y$CellLine, 
-                             "voom-like-plot.png")
-  )
-  ggsave(fname, p)
-}
-rango %>% 
-  group_by(Kit, CellLine) %>% 
-  dplyr::group_walk(voom_plot)
-```
+<!-- ```{r voom-like-plot} -->
+<!-- # rangenostics feature_average vs feature_sd coloured by n of bases above -->
+<!-- # average raw exon coverage -->
+<!-- library(ggplot2) -->
+<!-- voom_plot <- function(.x, .y) { -->
+<!--   p <- ggplot(data = .x, aes(x = mn, y = sd)) + -->
+<!--     geom_point() + -->
+<!--     geom_line(aes(y = .fitted), colour = "blue") + -->
+<!--     geom_vline(aes(xintercept = E_mu),  -->
+<!--                data  = dplyr::distinct(.x, feature_type, E_mu)) + -->
+<!--     geom_hline(aes(yintercept = E_sd), -->
+<!--                data = dplyr::distinct(.x, feature_type, E_sd)) + -->
+<!--     facet_wrap(~feature_type) + -->
+<!--     labs( -->
+<!--          subtitle = paste(.y$Kit, .y$CellLine), -->
+<!--          x = "mean log-coverage", -->
+<!--          y = "sd log-covreage" -->
+<!--     ) -->
+<!--   fname <- here::here("img",  -->
+<!--                       paste0(.y$Kit, "-", .y$CellLine,  -->
+<!--                              "voom-like-plot.png") -->
+<!--   ) -->
+<!--   ggsave(fname, p) -->
+<!-- } -->
+<!-- rango %>%  -->
+<!--   group_by(Kit, CellLine) %>%  -->
+<!--   dplyr::group_walk(voom_plot) -->
+<!-- ``` -->
 
-We also plot pairs of summary values for each feature in a scatter plot matrix
-(using a hexbin scatter plot). Again we get similar results to the plots
-produced by a DE analysis from InDeX.
+<!-- We also plot pairs of summary values for each feature in a scatter plot matrix -->
+<!-- (using a hexbin scatter plot). Again we get similar results to the plots -->
+<!-- produced by a DE analysis from InDeX. -->
 
-```{r rangenostics-pairs}
-all_features <- tidyr::gather(rango, "key", "value", 
-                              -Kit, -CellLine, -gene_id, -feature_type, 
-                              -E_mu, -E_sd, -E_raw) %>% 
-  mutate(var = paste0(feature_type, "_", key)) %>% 
-  plyranges::select(-feature_type, -key) %>% 
-  arrange(gene_id) %>% 
-  tidyr::spread(var, value)
-pairs_plot <- function(.x, .y) {
-  sub <- plyranges::select(.x, 
-                           exon_mn, 
-                           intron_mn, 
-                           intron_sd, 
-                           bases_above = intron_n_bases_E_raw)
-  fname <- here::here("img", 
-                      paste0(.y$Kit, "-", .y$CellLine, 
-                             "-pairs-plot.png")
-  )
+<!-- ```{r rangenostics-pairs} -->
+<!-- all_features <- tidyr::gather(rango, "key", "value",  -->
+<!--                               -Kit, -CellLine, -gene_id, -feature_type,  -->
+<!--                               -E_mu, -E_sd, -E_raw) %>%  -->
+<!--   mutate(var = paste0(feature_type, "_", key)) %>%  -->
+<!--   plyranges::select(-feature_type, -key) %>%  -->
+<!--   arrange(gene_id) %>%  -->
+<!--   tidyr::spread(var, value) -->
+<!-- pairs_plot <- function(.x, .y) { -->
+<!--   sub <- plyranges::select(.x,  -->
+<!--                            exon_mn,  -->
+<!--                            intron_mn,  -->
+<!--                            intron_sd,  -->
+<!--                            bases_above = intron_n_bases_E_raw) -->
+<!--   fname <- here::here("img",  -->
+<!--                       paste0(.y$Kit, "-", .y$CellLine,  -->
+<!--                              "-pairs-plot.png") -->
+<!--   ) -->
 
-  p <- GGally::ggpairs(sub, columns = 1:4, lower = list(continuous = hex))
-  ggsave(fname, p)
-}
-all_features %>% 
-  group_by(Kit, CellLine) %>% 
-  dplyr::group_walk(pairs_plot)
+<!--   p <- GGally::ggpairs(sub, columns = 1:4, lower = list(continuous = hex)) -->
+<!--   ggsave(fname, p) -->
+<!-- } -->
+<!-- all_features %>%  -->
+<!--   group_by(Kit, CellLine) %>%  -->
+<!--   dplyr::group_walk(pairs_plot) -->
 
-```
+<!-- ``` -->
 
-5. Decide on cut-off over each feature. It appears that choosing genes
-with high intron feature standard deviations, and with number of bases above
-E_mu in the 99th percentile discriminates intron retention features well.
+<!-- 5. Decide on cut-off over each feature. It appears that choosing genes -->
+<!-- with high intron feature standard deviations, and with number of bases above -->
+<!-- E_mu in the 99th percentile discriminates intron retention features well. -->
 
-```{r rangenositcs-hits}
-hits <- all_features %>% 
-  group_by(Kit, CellLine) %>% 
-  filter(exon_mn > E_mu, 
-         intron_sd > E_sd, 
-         intron_n_bases_E_raw > quantile(intron_n_bases_E_raw, 0.99))
+<!-- ```{r rangenositcs-hits} -->
+<!-- hits <- all_features %>%  -->
+<!--   group_by(Kit, CellLine) %>%  -->
+<!--   filter(exon_mn > E_mu,  -->
+<!--          intron_sd > E_sd,  -->
+<!--          intron_n_bases_E_raw > quantile(intron_n_bases_E_raw, 0.99))   -->
 
-dplyr::count(hits)
-```
+<!-- dplyr::count(hits) -->
+<!-- ``` -->
 
-6. We can also look at overlaps between sets using an UpSet plot:
+<!-- 6. We can also look at overlaps between sets using an UpSet plot: -->
 
-```{r, echo  = FALSE}
-sets <- hits %>% 
-  dplyr::distinct(gene_id) %>%
-  arrange(gene_id) %>% 
-  mutate(value = 1L, var = paste(CellLine, Kit)) %>% 
-  ungroup() %>%  
-  plyranges::select(gene_id, var, value) %>% 
-  tidyr::spread(var, value) %>% 
-  dplyr::mutate_at(2:5, .funs = ~ dplyr::if_else(is.na(.), 0L, .)) 
-sets <- as.data.frame(sets)
-library(UpSetR)
-pdf(here::here("img", "olaps-superintronic-hits.pdf"))
-upset(sets)
-dev.off()
-```
+<!-- ```{r, echo  = FALSE} -->
+<!-- sets <- hits %>%  -->
+<!--   dplyr::distinct(gene_id) %>% -->
+<!--   arrange(gene_id) %>%  -->
+<!--   mutate(value = 1L, var = paste(CellLine, Kit)) %>%  -->
+<!--   ungroup() %>%   -->
+<!--   plyranges::select(gene_id, var, value) %>%  -->
+<!--   tidyr::spread(var, value) %>%  -->
+<!--   dplyr::mutate_at(2:5, .funs = ~ dplyr::if_else(is.na(.), 0L, .))  -->
+<!-- sets <- as.data.frame(sets) -->
+<!-- library(UpSetR) -->
+<!-- pdf(here::here("img", "olaps-superintronic-hits.pdf")) -->
+<!-- upset(sets) -->
+<!-- dev.off() -->
+<!-- ``` -->

### 2020-02-21 index.Rmd
first build of chapters

@@
   warning = FALSE, message = FALSE, echo = FALSE,
-  fig.path = here::here('img'), fig.align = 'center', fig.show = 'hold',
+  fig.path = here::here('img/'), fig.align = 'center', fig.show = 'hold',
   cache = TRUE, external = TRUE, comment = "#>", dpi = 300,
@@
   cache.path = ifelse(is_html_output(), 
-                      here::here('cache-gitbook'), 
-                      here::here('cache-pdfbook')
+                      here::here('cache-gitbook/'), 
+                      here::here('cache-pdfbook/')
   )

### 2020-02-21 _render.R

@@ -22,7 +22,7 @@ for (i in img_pdf) {
   file_pdf <- paste0("img/", i)
   dest_pdf <- paste0("img/", sub("pdf$", "png", i))
   magick::image_write(
-    magick::image_read(file_pdf, 300), dest_pdf, "png", 
+    magick::image_read(file_pdf, 300), dest_pdf, "png",
     density = 300
   )
 }

### 2020-02-21 scripts/fluentGenomics.R

@@ -20,8 +20,8 @@ dir <- system.file("extdata", package="macrophage")
 
 
 ## ----coldata-rna-----------------------------------
-library(dplyr)
 library(readr)
+library(dplyr)
 colfile <- file.path(dir, "coldata.csv")
 coldata <- read_csv(colfile) %>%
   dplyr::select(
@@ -86,7 +86,11 @@ de_genes
 ## ----de-genes--------------------------------------
 de_genes <- de_genes %>%
   filter(padj < 0.01) %>%
-  select(gene_id, de_log2FC = log2FoldChange, de_padj = padj)
+  dplyr::select(
+    gene_id,
+    de_log2FC = log2FoldChange,
+    de_padj = padj
+  )
 
 
 ## ----not-de-genes----------------------------------
@@ -97,9 +101,11 @@ other_genes <- results(dds,
                        format="GRanges") %>%
   filter(padj < 0.01) %>%
   names_to_column("gene_id") %>%
-  dplyr::select(gene_id,
-                de_log2FC = log2FoldChange,
-                de_padj = padj)
+  dplyr::select(
+    gene_id,
+    de_log2FC = log2FoldChange,
+    de_padj = padj
+  )
 
 
 ## ----load-peaks------------------------------------
@@ -114,14 +120,14 @@ da_peaks <- peaks %>%
 
 ## ----slice-example---------------------------------
 size <- length(de_genes)
-slice(other_genes, sample.int(n(), size))
+slice(other_genes, sample.int(plyranges::n(), size))
 
 
 ## ----boot-set-01-----------------------------------
 # set a seed for the results
 set.seed(2019-08-02)
 boot_genes <- replicate(10,
-                        slice(other_genes, sample.int(n(), size)),
+                        slice(other_genes, sample.int(plyranges::n(), size)),
                         simplify = FALSE)
 
 
@@ -164,7 +170,7 @@ genes_olap_peaks
 gene_peak_max_lfc <- genes_olap_peaks %>%
   group_by(gene_id, origin)  %>%
   reduce_ranges_directed(
-    peak_count = sum(!is.na(da_padj)) / n_distinct(resample),
+    peak_count = sum(!is.na(da_padj)) / plyranges::n_distinct(resample),
     peak_max_lfc = max(abs(da_log2FC))
   )
 
[…]

### 2020-02-21 Rmd/02-plyranges.Rmd
first build of chapters
rebuild

@@
-# plyranges: a grammar of graphics for genomics data {#ch:plyranges}
+# plyranges: a grammar of data transformation for genomics {#ch:plyranges}
 
@@
 
-(ref:GRanges-cap): An illustration of the GRanges data model for a sample from an RNA-seq experiment.
+(ref:GRanges-cap) An illustration of the GRanges data model for a sample from an RNA-seq experiment.
 The core components of the data model include a seqname column (representing the chromosome), a ranges column which consists of start and end coordinates for a genomic region, and a strand identifier (either positive, negative, or unstranded).
@@
 
-(ref:olaps-cap): Illustration of the three overlap join operators.
+(ref:olaps-cap) Illustration of the three overlap join operators.
 Each join takes two  GRanges objects, \textit{x} and \textit{y} as input.
@@
 
-```{r code-comparsion, fig.cap = "(ref:code-comparison-cap)", echo = FALSE}
+```{r code-comparison, fig.cap = "(ref:code-comparison-cap)", echo = FALSE}
 knitr::include_graphics(here::here("img", "code-comparison.png"))
@@
 
-(ref:code-comparison-cap): Idiomatic code examples for \texttt{plyranges} (\textbf{A}) and \texttt{GenomicRanges} (\textbf{B}) illustrating an overlap and aggregate operation that returns the same result.
+(ref:code-comparison-cap) Idiomatic code examples for \texttt{plyranges} (\textbf{A}) and \texttt{GenomicRanges} (\textbf{B}) illustrating an overlap and aggregate operation that returns the same result.
 In each example, we have two BED files consisting of SNPs that are genome-wide association study (GWAS) hits and reference exons.
@@
 
-(ref:peak-viz-cap): The final result of the `plyranges` operations to find a 5000nt region surrounding the peak of normalised coverage scores over chromosome 10, displayed as a density plot.
+(ref:peak-viz-cap) The final result of the `plyranges` operations to find a 5000nt region surrounding the peak of normalised coverage scores over chromosome 10, displayed as a density plot.
 
@@
 
-(ref:titv-viz-cap): The final result of computing quality control metrics over the SNP array data with `plyranges`, displayed as a dot plot.
+(ref:titv-viz-cap) The final result of computing quality control metrics over the SNP array data with `plyranges`, displayed as a dot plot.
 Chromosomes are ordered by their estimated transition-transversion ratio.

### 2020-02-21 index.Rmd
modify templates
rebuild

@@
 ---
-title: 'Exploratory data analysis for high-dimensional biological datasets'
+title: 'Fluent statistical computing interfaces for biological data analysis'
 author: 'Stuart Lee'
@@
 degrees: 'B.MatComSci., University of Adelaide; MSc.Bioinformatics. University of Melbourne'
-bibliography: [bib/plyranges.bib, bib/fluentGenomics.bib]
+bibliography: [bib/plyranges.bib, bib/fluentGenomics.bib, bib/thesisrefs.bib]
 link-citations: yes

### 2020-02-21 Rmd/03-fluentGenomics.Rmd
rebuild
update output

@@
-# Fluent genomics: integrating data with plyranges and tximeta {#ch:fluentGenomics}
+# Fluent genomics with _plyranges_ and _tximeta_ {#ch:fluentGenomics}
 
@@
 ```
-
-(ref:workflow) An overview of the fluent genomics workflow.
-First, we *import* data as a *SummarizedExperiment* object, which enables interoperability with downstream analysis packages.
-Then we *model* our assay data, using the existing Bioconductor packages *DESeq2* and *limma*.
-We take the results of our models for each assay with respect to their genomic coordinates, and *integrate* them.
-First, we compute the overlap between the results of each assay, then aggregate over the combined genomic regions, and finally summarize to compare enrichment for differentially expressed genes to non differentially expressed genes.
-The final output can be used for downstream visualization or further transformation.
-
-(ref:maplot) Visualization of *DESeq2* results as an "MA plot". Genes that have an adjusted *p-value* below 0.01 are colored red.
-
-(ref:boxplot) A boxplot of maximum LFCs for DA peaks for DE genes compared to non-DE genes where genes have at least one DA peak.
-
-(ref:linechart) A line chart displaying how relative enrichment of DA peaks change between DE genes compared to non-DE genes as the absolute DA LFC threshold increases.
-
-(ref:linechart2) A line chart displaying how gene and peak counts change as the absolute DA LFC threshold increases.
-Lines are colored according to whether they represent a gene that is DE or not.
-Note the x-axis is on a $\log_{10}$ scale.
 
@@
 
+(ref:workflow) An overview of the fluent genomics workflow.
+First, we *import* data as a *SummarizedExperiment* object, which enables interoperability with downstream analysis packages.
+Then we *model* our assay data, using the existing Bioconductor packages `DESeq2` and `limma`.
+We take the results of our models for each assay with respect to their genomic coordinates, and *integrate* them.
+First, we compute the overlap between the results of each assay, then aggregate over the combined genomic regions, and finally summarize to compare enrichment for differentially expressed genes to non differentially expressed genes.
+The final output can be used for downstream visualization or further transformation.
+
 Throughout the workflow (Figure \@ref(fig:workflow)), we will use existing
 Bioconductor infrastructure to understand these datasets. In particular, we
-will emphasize the use of the Bioconductor packages *plyranges* and *tximeta*.
-The *plyranges* package fluently transforms data tied to genomic ranges using
+will emphasize the use of the Bioconductor packages `plyranges` and `tximeta`.
+The `plyranges` package fluently transforms data tied to genomic ranges using
 operations like shifting, window construction, overlap detection, etc. It is
 described by @Lee2019 and leverages underlying core Bioconductor infrastructure
-[@granges; @bioc] and the *tidyverse* design principles @tidyverse.
+[@granges; @bioc] and the `tidyverse` design principles @tidyverse.
 
-The *tximeta* package described by @Love2019-tximeta is used to read RNA-seq
+The `tximeta` package described by @Love2019-tximeta is used to read RNA-seq
 quantification data into R/Bioconductor, such that the transcript ranges and
@@
 
-The data used in this workflow is available from two packages: the *macrophage*
+The data used in this workflow is available from two packages: the `macrophage`
 Bioconductor ExperimentData package and from the workflow package
-*fluentGenomics*.
+`fluentGenomics`.
 
-The *macrophage* package contains RNA-seq quantification from 24 RNA-seq
+The `macrophage` package contains RNA-seq quantification from 24 RNA-seq
 samples, a subset of the RNA-seq samples generated and analyzed by @alasoo. The
@@
 
-The *fluentGenomics* package contains functionality to download and generate a
+The `fluentGenomics` package contains functionality to download and generate a
 cached *SummarizedExperiment* object from the normalized ATAC-seq data provided
@@
 present, will create the cache and generate a *SummarizedExperiment* using the
-the *BiocFileCache* package [@bcfilecache].
+the `BiocFileCache` package [@bcfilecache].
 
@@
 
-### Using *tximeta* to import RNA-seq quantification data
+### Using `tximeta` to import RNA-seq quantification data
 
@@
 
-```{r dir, eval=FALSE}
+```{r dir, eval=FALSE, echo = TRUE}
 dir <- "/path/to/quant/files"
@@
 where the path is relative to your current R session. However, in this case we
-have distributed the files in the *macrophage* package. The relevant directory
+have distributed the files in the `macrophage` package. The relevant directory
 and associated files can be located using `system.file`.
@@
 Information about the experiment is contained in the `coldata.csv` file. We
-leverage the *dplyr* and *readr* packages (as part of the *tidyverse*) to read
-this file into R [@tidyverse]. We will see later that *plyranges* extends these
+leverage the `dplyr` and `readr` packages (as part of the `tidyverse`) to read
+this file into R [@tidyverse]. We will see later that `plyranges` extends these
 packages to accommodate genomic ranges.
@@
 without any extra steps, as the `tximeta` function obtains any necessary
-metadata via FTP, unless it is already cached locally. The *tximeta* package
+metadata via FTP, unless it is already cached locally. The `tximeta` package
 can also be used without an internet connection, in this case the linked
@@
 
-Because *tximeta* knows the correct reference transcriptome, we can ask
-*tximeta* to summarize the transcript-level data to the gene level using the
+Because `tximeta` knows the correct reference transcriptome, we can ask
+`tximeta` to summarize the transcript-level data to the gene level using the
 methods of @Soneson2015.
@@
 
-```{r coldata-atac, eval=FALSE}
+```{r coldata-atac, eval=FALSE, echo = TRUE}
 atac_coldata <- read_tsv("ATAC_sample_metadata.txt.gz") %>%
@@
 
-```{r mat-atac, eval=FALSE}
+```{r mat-atac, eval=FALSE, echo = TRUE}
 atac_mat <- read_tsv("ATAC_cqn_matrix.txt.gz",
@@
 
-```{r peaks-atac, eval=FALSE}
+```{r peaks-atac, eval=FALSE, echo = TRUE}
 library(plyranges)
@@
 
-```{r atac-se, eval=FALSE}
+```{r atac-se, eval=FALSE, echo = TRUE}
 atac <- SummarizedExperiment(assays = list(cqndata=atac_mat),
@@
 
-We can easily run a differential expression analysis with *DESeq2* using the
+We can easily run a differential expression analysis with `DESeq2` using the
 following code chunks [@Love2014]. The design formula indicates that we want to
@@
 
+(ref:maplot) Visualization of `DESeq2` results as an "MA plot". Genes that have an adjusted *p-value* below 0.01 are colored red.
+
 We now output the results as a *GRanges* object, and due to the conventions of
-*plyranges*, we construct a new column called `gene_id` from the row names of
+`plyranges`, we construct a new column called `gene_id` from the row names of
 the results. Each row now contains the genomic region (`seqnames`, `start`,
 `end`, `strand`) along with corresponding metadata columns (the `gene_id` and
-the results of the test). Note that *tximeta* has correctly identified the
+the results of the test). Note that `tximeta` has correctly identified the
 reference genome as "hg38", and this has also been added to the *GRanges* along
 the results columns. This kind of book-keeping is vital once overlap operations
-are performed to ensure that *plyranges* is not comparing across incompatible
+are performed to ensure that `plyranges` is not comparing across incompatible
 genomes.
@@
 
-For assessing differential accessibility, we run *limma* [@Smyth2004], and
+For assessing differential accessibility, we run `limma` [@Smyth2004], and
 generate the a summary of LFCs and adjusted p-values for the peaks:
@@
 We now take the `rowRanges` of the *SummarizedExperiment* and attach the LFCs
-and adjusted p-values from *limma*, so that we can consider the overlap with
+and adjusted p-values from `limma`, so that we can consider the overlap with
 differential expression. Note that we set the genome build to "hg38" and
@@
 
-### Finding overlaps with *plyranges*
+### Finding overlaps with `plyranges`
 
-We have already used *plyranges* a number of times above, to `filter`,
+We have already used `plyranges` a number of times above, to `filter`,
 `mutate`, and `select` on *GRanges* objects, as well as ensuring the correct
-genome annotation and style has been used. The *plyranges* package provides a
+genome annotation and style has been used. The `plyranges` package provides a
 grammar for performing transformations of genomic data [@Lee2019]. Computations
-resulting from compositions of *plyranges* "verbs" are performed using
-underlying, highly optimized range operations in the *GenomicRanges* package
+resulting from compositions of `plyranges` "verbs" are performed using
+underlying, highly optimized range operations in the `GenomicRanges` package
 [@granges].
@@
 
-As *plyranges* is built on top of *dplyr*, it implements methods for many of
+As `plyranges` is built on top of *dplyr*, it implements methods for many of
 its verbs for *GRanges* objects. Here we can use `slice` to randomly sample the
@@
 could do this, but we prefer an approach via the anchoring methods in
-*plyranges*. Because there is a mutual dependence between the start, end,
+`plyranges`. Because there is a mutual dependence between the start, end,
 width, and strand of a *GRanges* object, we define anchors to fix one of
@@
 We can then repeat the same pattern but this time using `anchor_center()` to
-tell *plyranges* that we are making the TSS the midpoint of a range that has
+tell `plyranges` that we are making the TSS the midpoint of a range that has
 total width of 20kb, or 10kb both upstream and downstream of the TSS.
@@
 We are now ready to compute overlaps between RNA-seq genes (our DE set and
-bootstrap sets) and the ATAC-seq peaks. In *plyranges*, overlaps are defined as
+bootstrap sets) and the ATAC-seq peaks. In `plyranges`, overlaps are defined as
 joins between two *GRanges* objects: a _left_ and a _right_ *GRanges* object.
@@
 ```
+
+(ref:boxplot) A boxplot of maximum LFCs for DA peaks for DE genes compared to non-DE genes where genes have at least one DA peak.
 
@@
 
+(ref:linechart) A line chart displaying how relative enrichment of DA peaks change between DE genes compared to non-DE genes as the absolute DA LFC threshold increases.
+
 We computed the sum of DA peaks near the DE genes, for increasing LFC
@@
 
+(ref:linechart2) A line chart displaying how gene and peak counts change as the absolute DA LFC threshold increases.
+Lines are colored according to whether they represent a gene that is DE or not.
+Note the x-axis is on a $log_{10}$ scale.
+
 ## Discussion
 
-We have shown that by using *plyranges* and *tximeta* (with support of
-Bioconductor and *tidyverse* ecosystems) we can fluently iterate through the
+We have shown that by using `plyranges` and `tximeta` (with support of
+Bioconductor and `tidyverse` ecosystems) we can fluently iterate through the
 biological data science workflow: from import, through to modeling, and data
@@
 
-The authors would like to thank all participants of the Bioconductor 2019 and
-BiocAsia 2019 conferences who attended and provided feedback on early versions
-of this workflow paper.
+We would like to thank all participants of the Bioconductor 2019 and
+BiocAsia 2019 conferences who attended and provided feedback on early versions of this workflow paper.

### 2020-02-21 index.Rmd
update output

@@
 library(knitr)
+
 opts_chunk$set(
@@
 )
+
 ```

### 2020-02-21 _render.R

@@ -36,7 +36,7 @@ if (Sys.getenv("RSTUDIO") != "1" && Sys.info()['sysname'] == "Darwin") {
 # provide default formats if necessary
 formats <- commandArgs(trailingOnly = TRUE)
 if (length(formats) == 0) {
-  formats <- "bookdown::pdfbook"
+  formats <- "bookdown::pdf_book"
 }
 # render the book to all formats
 for (fmt in formats)

### 2020-02-21 Rmd/03-fluentGenomics.Rmd
don't cache by default, rebuild output

@@
   genome="GRCh38",
-  fasta="ftp://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_human/release_29/gencode.v29.transcripts.fa.gz",
+  fasta="gencode.v29.transcripts.fa.gz", # ftp link to fasta file
   gtf=file.path(dir, "gencode.v29.annotation.gtf.gz"), # local version
@@
 ```{r mat-atac, eval=FALSE, echo = TRUE}
-atac_mat <- read_tsv("ATAC_cqn_matrix.txt.gz",
+atac_mat <- read_tsv(
+  "ATAC_cqn_matrix.txt.gz",
   skip = 1,
-                     col_names =c("rownames", atac_coldata[["sample_id"]]))
+  col_names = c("rownames", atac_coldata[["sample_id"]])
+ )
 rownames <- atac_mat[["rownames"]]
@@
 ```{r atac-se, eval=FALSE, echo = TRUE}
-atac <- SummarizedExperiment(assays = list(cqndata=atac_mat),
-                             rowRanges=peaks_gr,
-                             colData=atac_coldata)
+atac <- SummarizedExperiment(
+  assays = list(cqndata=atac_mat),
+  rowRanges = peaks_gr,
+  colData = atac_coldata
+)
 ```

### 2020-02-21 index.Rmd
don't cache by default, rebuild output

@@
   fig.path = here::here('img/'), fig.align = 'center', fig.show = 'hold',
-  cache = TRUE, external = TRUE, comment = "#>", dpi = 300,
+  external = TRUE, comment = "#>", dpi = 300,
   out.width = ifelse(is_html_output(), "100%", "\\textwidth"),

### 2020-02-21 _render.R

@@ -21,6 +21,9 @@ img_pdf <- list.files("img/", pattern = "*.pdf")
 for (i in img_pdf) {
   file_pdf <- paste0("img/", i)
   dest_pdf <- paste0("img/", sub("pdf$", "png", i))
+  if (file.exists(dest_pdf)) {
+    break
+  }
   magick::image_write(
     magick::image_read(file_pdf, 300), dest_pdf, "png",
     density = 300
@@ -36,12 +39,11 @@ if (Sys.getenv("RSTUDIO") != "1" && Sys.info()['sysname'] == "Darwin") {
 # provide default formats if necessary
 formats <- commandArgs(trailingOnly = TRUE)
 if (length(formats) == 0) {
-  formats <- "bookdown::pdf_book"
+  formats <- c("bookdown::pdf_book", "bookdown::gitbook")
 }
 # render the book to all formats
-for (fmt in formats)
+for (fmt in formats) {
   bookdown::render_book("index.Rmd", fmt, quiet = FALSE)
+}
 
-gif_file <- list.files("figure", "*.gif", full.names = TRUE)
-invisible(file.copy(gif_file, "_thesis/figure/animate.gif"))
 cli_alert_success("You Rock!")

### 2020-02-21 scripts/plyranges.R

@@ -60,7 +60,7 @@ locations <- bam %>%
   get_genome_info()
 
 
-## ----read-bam, cache = TRUE---------------------------------
+## ----read-bam------------------------------------------
 alignments <- bam %>%
   filter_by_overlaps(locations) %>%
   select(seq) %>%

### 2020-02-23 Rmd/00-c-declaration.Rmd
add declaration

@@
 
-\vspace*{2cm}\par\authorname
+This thesis includes 3 publications, two of which have been published and one 
+which has not yet been submitted. As the core theme of my thesis is the development of software
+interfaces for biological data analysis, and given the collaborative nature
+of statistical computing and bioinformatics research; all of the included
+papers in this thesis reflect and acknowledge the contributions of my 
+co-authors, including my  primary supervisors Professor Dianne Cook (DC) in the 
+Department of Econometrics and Business Statistics at Monash University
+and Associate Professor Matthew Ritchie in the Epigenetics and Development 
+Division at the Walter and Eliza Hall Institute. The following table details
+the publications and my and my fellow co-authors contributions:
+
+\begin{table}
+\centering\footnotesize\tabcolsep=0.12cm
+\begin{tabular}{|p{1cm}|p{2cm}|p{1.5cm}|p{3.5cm}|p{3.5cm}|p{1.5cm}|p{1cm}}
+\hline
+\RaggedRight\textbf{Thesis Chapter}  & \RaggedRight\textbf{Publication Title}  & \RaggedRight\textbf{Status (published, in press, accepted or returned for revision)}  & \RaggedRight\textbf{Nature and} {\%} \RaggedRight\textbf{of student contribution} & \RaggedRight\textbf{Co-author name(s) Nature and} {\%} \RaggedRight\textbf{of Co-author’s contribution} & \RaggedRight\textbf{Co-author(s), Monash student Y/N}  & \RaggedRight\textbf{Reference} \\ \hline
+\@ref(ch:plyranges) & plyranges: a grammar of genomic data transformation & Published  & 70\%.
+Concept, software development, data analysis, and manuscript writing & (1) Dianne Cook, Concept and manuscript revision 10\% (2) Michael Lawrence, Concept and software development  20\% & \multicolumn{1}{c|}{N} & @Lee2019 \\
+\hline
+\@ref(ch:fluentGenomics) 3 & Fluent genomics with _plyranges_ and _tximeta_ & Published (awaiting peer review) & 60\%.
+Concept, software development, data analysis and manuscript writing  & (1) Michael Lawrence, manuscript feedback and editing 5\% (2) Michael I Love, Concept, data analysis and manuscript writing 35\% & \multicolumn{1}{c|}{N} & @Lee2020-ie \\
+\hline
+\@ref(ch:tsne) & Casting multiple shadows: high-dimensional interactive data visualisation with tours and embeddings & To be submitted & 80\%.
+Concept, software development, and manuscript writing & (1) Dianne Cook, Concept and manuscript writing 15\% \multicolumn{1}{c|}{N} & Not Applicable \\
+\hline
+\end{tabular}
+\end{table}
+
+\FloatBarrier
+
+Note that chapter \@ref(ch:intron), reflects my  contribution (in the form of a 
+software package and data analysis) to the submitted manuscript entitled 
+_Covering all your bases: incorporating intron signal from RNA-seq data_
+[@Lee2019-mf] and is joint work under the supervision of Dr Charity Law
+and Associate Professor Matthew Ritchie.
+
+\textbf{Student name:} Stuart Andrew Lee
+
+\textbf{Student signature:}
+
+```{r signature, out.width = "20%", fig.align = "left", eval = file.exists("img/signature.png")}
+knitr::include_graphics("img/signature.png")
+```
 

### 2020-02-23 Rmd/05-embeddings.Rmd
add outline of liminal paper

@@
-# Spurious embeddings from non-linear dimensionality reduction methods and how to find them {#ch:tsne}
+# Casting multiple shadows: high-dimensional interactive data visualisation with tours and embeddings {#ch:tsne}
+
+<!--
+- review of embedding algorithms, interactive graphics, tours
+- technology
+- design
+- examples/benchmarks
+  - Kevin moon
+  - pdfsense
+  - single cell
+  - quick draw
+https://jlmelville.github.io/smallvis/init.html#other_initialization_options
+-->
+
+## Introduction
+
+High dimensional data is increasingly prevalent in the natural sciences and
+beyond but presents a challenge to the analyst in terms of both data cleaning /
+preprocessing and visualisation. Methods to embed data from a high-dimesional
+space into a low-dimensional one now form a core step of the data analysis
+workflow where they are used to ascertain hidden structure and denoise data for
+downstream analysis (thereby nullifying the 'curse of dimensionality').
+
+Choosing an appropriate embedding presents a challenge to the analyst. How does
+an analyst know whether the embedding has captured the underlying topology and
+geometry of the high dimensional space? The answer depends on the analyst's
+workflow.  @Brehmer2014-hk characterised two main workflow steps that an
+analyst performs when using embedding techniques: dimension reduction and
+cluster orientation. The first relates to dimension reduction achieved by using
+an embedding method, here an analyst wants to characterise and map meaning onto
+the embedded form, for example identifying batch effects from a high throughput
+sequencing experiment, or identifying a gradient or trajectory along the
+embedded form @Nguyen2019-yh. The second relates to using embeddings as part of
+a clustering workflow. Here analysts are interested in identifying and naming
+clusters and verifying them by either applying known labels or colouring by
+variables that are a-priori known to distinguish clusters. Both of these
+workflow steps rely on the embedding being 'faithful' or the original high
+dimensional dataset, and become much more difficult when there is no underlying
+ground truth.
+
+Embedding methods can be classified into two broad groups: linear and
+non-linear methods. Linear methods perform a linear transformation of the data;
+one example is principal components analysis (PCA) which performs an
+eigendecomposition of the estimated sample covariance matrix. The eigenvalues
+are sorted in decreasing order and represent the variance explained by each
+component (eigenvector). A common approach to deciding on the number of
+principal components to retain is to plot the proportion of variance explained
+by each component and choose a cut-off.
+
+Non-linear methods generally perform preprocessing on the high-dimensional data
+such as generating a neighborhood graph and perform transformations on the
+preprocessed form. We restrict our attention to three methods that are commonly
+used in high-throughput biology: t-distributed stochastic neighbor embedding
+(t-SNE), uniform manifold alignment and projection (UMAP), and potential of
+heat-diffusion for affinity-based transition embedding (PHATE). The t-SNE
+algorithm estimates the similarity of (Euclidean) distances of points in a high
+dimensional space using a Gaussian distribution and then estimates a
+configuration in the low dimensional embedding space by modelling similarities
+using a t-distribution with 1 degree of freedom. The resulting configuration is
+the one that minimizes the Kulback-Leibler divergence between the two
+distributions. A recent theoretical contribution by @Linderman2019-dq proved
+that t-SNE can recover spherical and well separated cluster shapes, and
+proposed new approaches for tuning the optimisation parameters. It is a known
+problem that t-SNE can have trouble recovering global structure and that
+configurations can be highly dependent on how the algorithm is initialised and
+parameterised [@Wattenberg2016-ji; @Kobak2019-lm]. UMAP is a method that is
+related to LargeVis [@Tang2016-oz], and like t-SNE acts on the k-nearest
+neighbor graph. Its main differences are that it uses a different cost function
+(cross entropy) which is optimized using stochastic gradient descent and
+defines a different kernel for similiarites in the low dimensional space. Due
+to it's computational speed it's possible to generate UMAP embeddings in more
+than three dimensions. Finally, PHATE, inspired by diffusion maps, is based on
+estimating an affinity matrix via a distance matrix and k-nearest neighbors
+graph. The algorithm denoises estimated distances in high dimensional space via
+transforming the affinity matrix into a Markov transition probaility matrix and
+diffusing this matrix over a fixed number of time steps. Then the diffused
+probabilities are transformed once more to construct a distance matrix, and
+multidimensional scaling is performed to construct a 2d embedding for
+visualization.
+
+As part of a visualization workflow, it's important to consider the perception
+and interpretation of embedding methods as well. @Sedlmair2013-pn showed that
+2D scatter plots were mostly suffucient for detecting class seperation, however
+they noted that often multiple embeddings were required. For the task of
+cluster identification, @Lewis2012-ai showed experimentally that novice users
+of non-linear embedding techniques were more likely to consider clusters of
+points on a 2d scatter plot to be the result of a spurious embedding compared
+to advanced users who were aware of the inner workings of the embedding
+algorithm.
+
+A complimentary approach for visualizing structure in high dimensional data is
+the tour. A tour is a sequence of projections of a high dimensional dataset
+onto a low-dimensional orthonormal basis matrix, that is represented as a
+dynamic visualization. The sequence of generated bases are interpolated to form
+the tour path, allowing a user to explore the subspace of projections. A grand
+tour corresponds to choosing new bases at random, and can give an overview of
+the structure in the data. Instead of picking projections at random, a guided
+tour can be used to generate a sequence 'interesting' projections as quantified
+by an index function. Given the dynamic nature of the tour, user interaction is
+important for controlling and exploring the visualisation: the tour has been
+used previously by @Wickham2015-cx as tool for exploring statistical model fits
+and by @Buja1996-fk for exploring factorial experimental designs.
+
+While there has been much work on the algorithmic details of the aformentioned
+embedding methods, there has been relatively few tools designed to assist users
+to interact with these techniques and assist them in making comparsions between
+embeddings and performing the aforementioned cluster orientation tasks. Several
+interactive interfaces have been proposed for evaluating or using embedding
+techniques: the Sleepwalk interface provides a click and highlight
+visualisation for colouring points in an embedding according to their distance
+in the original high-dimensional space [@Ovchinnikova2019-gf]. The work by
+@Pezzotti2017-cz provides a user guided and modified form of the t-SNE
+algorithm, that allows users to modified optimisation parameters in real-time.
+Similarly, the embedding projector is a web interface to running UMAP, t-SNE or
+PCA live in the browser and provides interactions to color points, and
+highlights nearest neighbors [@Smilkov2016-hp].
+
+There is no one-size fits all: finding an appropriate embedding for a given
+dataset is a difficult and somewhat poorly defined problem. For non-linear
+methods, there are a lot of parameters to explore that can have an effeft on
+the resulting visualisation and interpretation. Interfaces for evaluating
+embeddings require interaction but should also be able to be incorporated into
+an analysts workflow. We propose a more pragmatic workflow inspired by
+incorporating interactive graphics and tours with embeddings that allows users
+to see a global overview of their high dimensional data and assists them with
+cluster orientation tasks. This workflow is incorporated into an R package
+called `liminal` (Available: https://github.com/sa-lee/liminal).
+
+## Design
+
+We propose using tours as part of an analyst's workflow in in performing
+dimensionality reduction tasks. We have made extensive use of ensemble
+graphics, that is aligning related plots alongside each other to provide
+context. As we will see in the case studies, this allows analysts to quickly
+compare views from embedding methods and allows them to see how the embedding
+method alters the global structure of their data. Using ensembles allows the
+use of interaction techniques, that allow analysts to perform cluster
+orientation tasks via linking multiple views. This approach allows our
+interface, to achieve the three principles for interactive high-dimensional
+data visualisation outlined by @Buja1996-fk: finding gestalt, posing queries,
+and making comparisons.
+
+### Finding Gestalt: focus and context
+
+To investigate latent structure and the shape of a high dimensional dataset, a
+tour can be run without the use of an external embedding. It is often useful to
+first run principal components on the input as an intial dimension reduction
+step, and then tour a subset of those components instead, i.e. by selecting
+them from a scree plot. The default tour layout is a scatter plot with an axis
+layout displaying the magnitude and direction of each basis vector. Since the
+tour is dynamic, it is often useful to be able to pause and highlight a
+particular view. In our interface, brushing will pause the tour path, allowing
+users to identify 'interesting' projections. The domain of the axis scales from
+running a tour is called the half range, and is computed by rescaling the input
+data onto hyperdimensional unit cube. We bind the half range to a mouse wheel
+event, allowing a user to pan and zoom on the tour view dynamically. This is
+useful for peeling back dense clumps of points to reveal structure.
+
+### Posing Queries: multiple views, many contexts
+
+We combine the tour view in a side by side layout with a scatter
+plot view from an embedding algorithm. The views are linked; analysts can brush a group of points in the embedding view, and see its shape in the tour view.
+This can be used for either manually identifying clusers, or verifying the location of clusters relative to other groupings. Consequently, simple linked
+brushing can alleviate a downfall of non-linear methods such as UMAP or t-SNE
+that break hierachial structure of clusterings. Aside from simple
+linked brushing, we can also pose more complex queries via interrogating
+the neighbourhood graph. Using a linked neighbourhood brush, we can 
+visually investigate the nearest neighbour relationships in the 
+high-dimensional space via brushing in the embedding view. The user can select
+the number of nearest neighbours directly, and modify the distance metric 
+used for determining the neighbours. This allows users to interrogate the stability of clusters generated in the embedding view. Multiple brushes
+can be used to pose queries in either the tour view or the embedding view;
+interface controls allow these brushes to combine using logical operators
+such as 'and', 'or', or 'not'. The use of linked brushing goes beyond
+single point highlights or colorings, allowing analysts to get a more holistic
+view of the effect of an embedding algorithm.
+
+### Making comparsions
+
+TODO
+
+## Implementation
+
+The `liminal` software is implemented as an open source R package @baseR. All
+visualisations produced by the package were constructed using the `vegawidget`
+package, which is an R interface to the Vega-Lite grammar of interactive
+graphics. Ensemble graphics for arranging tour and embedding views are
+generated using the `shiny` packages. We have made extensive use of the `tourr`
+package for constructing tour paths.
+
+### Tours as a streaming data problem
+
+### Linking views via spatial brushes
+
+## Case Studies
+
+### Case Study 1: Fake trees
+
+### Case Study 2: well-clustered smallish single cell
+
+### Case Study 3: largeish single cell data
+
+### Case Study 4: non-spherical clusters, pdfsense
+
+## Discussion

### 2020-02-23 Rmd/00-c-declaration.Rmd
fixing up declarations section

--- 2019-10-21 Rmd/00-c-declaration.Rmd
@@
+---
+output:
+  pdf_document: default
+  html_document: default
+---
 `r if (knitr::is_html_output()) '<!--'`
@@
 
-\vspace*{2cm}\par\authorname
+This thesis includes 3 publications, two of which have been published and one 
+which has not yet been submitted [@Lee2019; @Lee2020-ie] . As the core theme of my thesis is the development of software
+interfaces for biological data analysis, and given the collaborative nature
+of statistical computing and bioinformatics research; all of the included
+papers in this thesis reflect and acknowledge the contributions of my 
+co-authors, including my  primary supervisors Professor Dianne Cook (DC) in the 
+Department of Econometrics and Business Statistics at Monash University
+and Associate Professor Matthew Ritchie in the Epigenetics and Development 
+Division at the Walter and Eliza Hall Institute. The following table details
+the publications and my and my fellow co-authors contributions:
+
+\begin{table}
+\centering\footnotesize\tabcolsep=0.12cm
+\begin{tabular}{|p{1cm}|p{2cm}|p{1.5cm}|p{3.5cm}|p{3.5cm}|p{1.5cm}|}
+\hline
+\RaggedRight\textbf{Thesis Chapter}  & \RaggedRight\textbf{Publication Title}  & \RaggedRight\textbf{Status (published, in press, accepted or returned for revision)}  & \RaggedRight\textbf{Nature and} {\%} \RaggedRight\textbf{of student contribution} & \RaggedRight\textbf{Co-author name(s) Nature and} {\%} \RaggedRight\textbf{of Co-author’s contribution} & \RaggedRight\textbf{Co-author(s), Monash student Y/N} \\ \hline
+2 & plyranges: a grammar of genomic data transformation & Published  & 70\%.
+Concept, software development, data analysis, and manuscript writing & (1) Dianne Cook, Concept and manuscript revision 10\% (2) Michael Lawrence, Concept and software development  20\% & \multicolumn{1}{c|}{N}  \\
+\hline
+3 & Fluent genomics with \textit{plyranges} and \textit{tximeta} & Published (awaiting peer review) & 60\%.
+Concept, software development, data analysis and manuscript writing  & (1) Michael Lawrence, manuscript feedback and editing 5\% (2) Michael I Love, Concept, data analysis and manuscript writing 35\% & \multicolumn{1}{c|}{N} \\
+\hline
+5 & Casting multiple shadows: high-dimensional interactive data visualisation with tours and embeddings & To be submitted & 80\%.
+Concept, software development, and manuscript writing & (1) Dianne Cook, Concept and manuscript writing 15\% & \multicolumn{1}{c|}{N} \\
+\hline
+\end{tabular}
+\end{table}
+
+\FloatBarrier
+
+Note that chapter \@ref(ch:intron), reflects my  contribution (in the form of a 
+software package and data analysis) to the submitted manuscript entitled 
+_Covering all your bases: incorporating intron signal from RNA-seq data_
+[@Lee2019-mf] and is joint work under the supervision of Dr Charity Law
+and Associate Professor Matthew Ritchie.
+
+\textbf{Student name:} Stuart Andrew Lee
+
+\textbf{Student signature:}
+
+```{r signature, out.width = "20%", fig.align = "left", eval = file.exists("img/signature.png")}
+knitr::include_graphics("img/signature.png")
+```
 

### 2020-02-23 Rmd/05-embeddings.Rmd
fixing up declarations section

@@
 beyond but presents a challenge to the analyst in terms of both data cleaning /
-preprocessing and visualisation. Methods to embed data from a high-dimesional
+pre-processing and visualisation. Methods to embed data from a high-dimensional
 space into a low-dimensional one now form a core step of the data analysis
-workflow where they are used to ascertain hidden structure and denoise data for
+workflow where they are used to ascertain hidden structure and de-noise data for
 downstream analysis (thereby nullifying the 'curse of dimensionality').
@@
 
-Non-linear methods generally perform preprocessing on the high-dimensional data
+Non-linear methods generally perform pre-processing on the high-dimensional data
 such as generating a neighborhood graph and perform transformations on the
-preprocessed form. We restrict our attention to three methods that are commonly
+pre-processed form. We restrict our attention to three methods that are commonly
 used in high-throughput biology: t-distributed stochastic neighbor embedding
@@
 using a t-distribution with 1 degree of freedom. The resulting configuration is
-the one that minimizes the Kulback-Leibler divergence between the two
+the one that minimizes the Kullback-Leibler divergence between the two
 distributions. A recent theoretical contribution by @Linderman2019-dq proved
@@
 configurations can be highly dependent on how the algorithm is initialised and
-parameterised [@Wattenberg2016-ji; @Kobak2019-lm]. UMAP is a method that is
+parameterized [@Wattenberg2016-ji; @Kobak2019-lm]. UMAP is a method that is
 related to LargeVis [@Tang2016-oz], and like t-SNE acts on the k-nearest
@@
 (cross entropy) which is optimized using stochastic gradient descent and
-defines a different kernel for similiarites in the low dimensional space. Due
+defines a different kernel for similarities in the low dimensional space. Due
 to it's computational speed it's possible to generate UMAP embeddings in more
@@
 estimating an affinity matrix via a distance matrix and k-nearest neighbors
-graph. The algorithm denoises estimated distances in high dimensional space via
-transforming the affinity matrix into a Markov transition probaility matrix and
+graph. The algorithm de-noises estimated distances in high dimensional space via
+transforming the affinity matrix into a Markov transition probability matrix and
 diffusing this matrix over a fixed number of time steps. Then the diffused
@@
 and interpretation of embedding methods as well. @Sedlmair2013-pn showed that
-2D scatter plots were mostly suffucient for detecting class seperation, however
+2D scatter plots were mostly sufficient for detecting class separation, however
 they noted that often multiple embeddings were required. For the task of
@@
 
-While there has been much work on the algorithmic details of the aformentioned
+While there has been much work on the algorithmic details of the aforementioned
 embedding methods, there has been relatively few tools designed to assist users
-to interact with these techniques and assist them in making comparsions between
+to interact with these techniques and assist them in making comparisons between
 embeddings and performing the aforementioned cluster orientation tasks. Several
@@
 dataset is a difficult and somewhat poorly defined problem. For non-linear
-methods, there are a lot of parameters to explore that can have an effeft on
+methods, there are a lot of parameters to explore that can have an effect on
 the resulting visualisation and interpretation. Interfaces for evaluating
@@
 tour can be run without the use of an external embedding. It is often useful to
-first run principal components on the input as an intial dimension reduction
+first run principal components on the input as an initial dimension reduction
 step, and then tour a subset of those components instead, i.e. by selecting
@@
 running a tour is called the half range, and is computed by rescaling the input
-data onto hyperdimensional unit cube. We bind the half range to a mouse wheel
+data onto hyper-dimensional unit cube. We bind the half range to a mouse wheel
 event, allowing a user to pan and zoom on the tour view dynamically. This is
@@
 plot view from an embedding algorithm. The views are linked; analysts can brush a group of points in the embedding view, and see its shape in the tour view.
-This can be used for either manually identifying clusers, or verifying the location of clusters relative to other groupings. Consequently, simple linked
+This can be used for either manually identifying clusters, or verifying the location of clusters relative to other groupings. Consequently, simple linked
 brushing can alleviate a downfall of non-linear methods such as UMAP or t-SNE
-that break hierachial structure of clusterings. Aside from simple
+that break hierarchical structure of clusters. Aside from simple
 linked brushing, we can also pose more complex queries via interrogating
@@
 such as 'and', 'or', or 'not'. The use of linked brushing goes beyond
-single point highlights or colorings, allowing analysts to get a more holistic
+simple color highlights, allowing analysts to get a more holistic
 view of the effect of an embedding algorithm.
 
-### Making comparsions
+### Making comparisons
 
@@
 
-### Case Study 3: largeish single cell data
+### Case Study 3: largish single cell data
 

### 2020-02-23 Rmd/02-plyranges.Rmd
don't cache by default, rebuild output
rebuild output
fix space
update chunks
add abstracts

@@
 # plyranges: a grammar of data transformation for genomics {#ch:plyranges}
+
+There is a cognitive load placed on users in learning a data abstraction from the Bioconductor project and understanding its appropriate use.
+Users must navigate these abstractions to perform a genomic analysis task, when a single data abstraction, a GRanges object will suffice.
+By recognizing that the GRanges class follows 'tidy' data principles, we create a grammar of genomic data transformation, defining verbs for performing actions on and between genomic interval data and providing a way of performing common data analysis tasks through a coherent interface to existing Bioconductor infrastructure.
+We implement this grammar as a Bioconductor/R package called `plyranges`.
 
@@
 ```{r chr10-scores-reduce, echo = TRUE}
-all_peaks <- chr10_scores %>% 
-  filter(score > 8) %>% 
-  reduce_ranges(score = max(score))
-all_peaks
 ```
@@
 
-```{r bins}
-locations <- h1_bam_sorted %>%
-  read_bam() %>% 
-  get_genome_info() 
+```{r bins, echo = TRUE}
 ```
@@
 
-```{r load-array, cache = TRUE}
+```{r load-array}
 ```
@@
 
-```{r process-array, cache = TRUE, echo = TRUE}
+```{r process-array, echo = TRUE}
 ```
@@
 
-```{r titv, cache = TRUE, echo = TRUE}
+```{r titv, echo = TRUE}
 ```

### 2020-02-23 Rmd/03-fluentGenomics.Rmd
rebuild output
add abstracts

@@
 ```
+
+We construct a simple workflow for fluent genomics data analysis using the R/Bioconductor ecosystem.
+This involves three core steps: import the data into an appropriate abstraction, model the data with respect to the biological questions of interest, and integrate the results with respect to their underlying genomic coordinates.
+Here we show how to implement these steps to integrate published RNA-seq and ATAC-seq experiments on macrophage cell lines.
+Using tximeta, we import RNA-seq transcript quantifications into an analysis-ready data structure, called the SummarizedExperiment, that contains the ranges of the reference transcripts and metadata on their provenance.
+Using SummarizedExperiments to represent the ATAC-seq and RNA-seq data, we model differentially accessible (DA) chromatin peaks and differentially expressed (DE) genes with existing Bioconductor packages.
+Using plyranges we then integrate the results to see if there is an enrichment of DA peaks near DE genes by finding overlaps and aggregating over log-fold change thresholds.
+The combination of these packages and their integration with the Bioconductor ecosystem provide a coherent framework for analysts to iteratively and reproducibly explore their biological data.
 
@@
 library(plyranges)
-peaks_df <- read_tsv("ATAC_peak_metadata.txt.gz",
+peaks_df <- read_tsv(
+  "ATAC_peak_metadata.txt.gz",
   col_types = c("cidciicdc")

### 2020-02-23 Rmd/01-intro.Rmd
delete abstract; rewrite intro

@@
 
-overview
+Exploratory data analysis (EDA) is a vital element of the modern 
+statistical workflow - it is an analyst’s first pass at 
+understanding their data; revealing all it’s messes and (possibly) 
+uncovering hidden insights [@Tukey1977-pn; @Grolemund2017-uc]. 
+It is an iterative process invovling both computation and visualization 
+that can generate new hypotheses that can be tested and formalised using 
+stastical modelling. Moreover, as datasets grow in complexity and become
+increasingly heterogenous and multidimensional, 
+the use of EDA becomes vital to ensure the integrity and quality of analysis 
+outputs.  This is certainly true in high-throughput biological 
+data analysis, where there are constraints on computation time and 
+memory, leading to the situation that as data becomes larger,
+EDA becomes increasingly difficult.
 
-describe main topic - research questions
-chapters 1, 2, 3, 4 (your contributions to the joint work)
-scope
+The approach taken by the suite of software packages collectively known as the *tidyverse* 
+is an attempt to formalise aspects of the EDA process in the R programming language
+under a single semantic known as *tidy data* [@baser; @tidyverse; @Wickham2014-jc]. 
+Simply put, a *tidy data* set is a rectangular table where each row of the table corresponds to an 
+observation and each column corresponds to a variable. There is a surprisingly
+large amount of utility that can be achieved with this definition. By having
+each column representing a variable,  variables can be mapped
+to graphical aesthetics  of visualizations. This enables the grammar of graphics as implemented
+by *ggplot2* [@Wickham2016-gz; @Wilkinson2005-kq]. User interfaces
+as implemented by *tidyverse* and in particular the *dplyr* package are *fluent*; 
+they form a domain specific language (DSL) that gives users a mental model 
+for performing and composing common data transformation tasks 
+[@Wickham2017-dplyr; @FowlerFluent].
+
+It is unclear whether the *fluent* interfaces as implemented using the
+*tidy data* framework can be more generally applied and useful
+in fields such as high-throughput biology where domain specific semantics as
+implemented through more complex data structures are required. This is particularly
+true in the Bioconductor ecosystem, where much thought has gone into the
+design of data structures that enable interoperability between different tools 
+[@Huber2015-ei]. One contribution of this thesis is the implementation
+of a *fluent* interface for genomics data that is expressive enough to
+enable EDA and be used as a building block for more software.
+
+## A grammar for genomic data analysis
+
+## Integration and representation of genomic data structures
+
+## Visual analytics for dimension reduction and clustering workflows
+
+## Summary
+
+This thesis presents a set of tools for performing exploratory 
+data analysis in the context of genomics and transcriptomics.  The first tool, “plyranges” provides a domain specific language for 
+reasoning about range based genomics data that simplifies common 
+operations such as overlaps and aggregations.  Next I extend 
+“plyranges”  to develop “superintronic” which presents an 
+approach for discovering interesting genomic regions within an 
+experimental design. We apply this tool to explore intron signal 
+in transcriptomic data. Finally, I turn my attention to focus on 
+embedding methods currently used for visualisation of large data 
+sets…

### 2020-02-24 Rmd/01-intro.Rmd
update toc; rebuild

@@
 uncovering hidden insights [@Tukey1977-pn; @Grolemund2017-uc]. 
-It is an iterative process invovling both computation and visualization 
+It is an iterative process involving both computation and visualization 
 that can generate new hypotheses that can be tested and formalised using 
-stastical modelling. Moreover, as datasets grow in complexity and become
-increasingly heterogenous and multidimensional, 
+statistical modelling. Moreover, as datasets grow in complexity and become
+increasingly heterogeneous and multidimensional, 
 the use of EDA becomes vital to ensure the integrity and quality of analysis 

### 2020-02-24 Rmd/06-conclusion.Rmd
update toc; rebuild

# Conclusion  {#ch:conclusion}

<!-- remind the reader of the research problem and purpose and how they were addressed -->
<!-- briefly summarise what has been covered in the paper -->
<!-- make some kind of holistic assessment/judgement/ claim that pertains to the whole project (i.e., more than a descriptive summary) -->
<!-- assess the value/relevance/ implications of the key findings in light of existing studies and literature -->
<!-- ‘speak’ to the Introduction -->
<!-- outline implications of the study (for theory, practice, further research) -->
<!-- comment on the findings that failed to support or only partially support the hypothesis or research questions directing the study -->
<!-- refer to the limitations of the studies that may affect the validity or the generalisability of results -->
<!-- make recommendations for further research -->
<!-- make claims for new knowledge/ contribution to knowledge. -->

## Findings and limitations

## Software development and impact

## Further research

### 2020-02-24 Rmd/05-embeddings.Rmd
update toc; rebuild
add abstract

@@
 -->
+
+There has been a rapid uptake in the use of non-linear dimensionality reduction (NLDR) methods such as t-distributed stochastic neighbour embedding (t-SNE) in the natural sciences as part of cluster orientation and dimension reduction workflows.
+The appropriate use of these methods is made difficult by their complex parameterisations and the multitude of decisions required to balance the preservation of
+local and global structure in the resulting visualisation.
+We present a visual analytics framework for the pragmatic usage of NLDR methods by combining them with a technique called the tour.
+A tour is a sequence of interpolated linear projections of multivariate data onto a lower dimensional space.
+The sequence is displayed as a dynamic visualisation, allowing a user to see the shadows the high-dimensional data casts in a lower dimensional view.
+By linking the tour to a view
+obtained from an NLDR method, we can preserve global structure and through user interactions like spatial linked brushing observe where
+the NLDR view may be misleading. We display show several case 
+studies from single cell genomics, that our approach is useful for
+cluster orientation tasks. 
+The implementation of our framework is available as an R package called `liminal` available at https://github.com/sa-lee/liminal.
 
@@
 
-### Case Study 2: well-clustered smallish single cell
+### Case Study 2: Clustering PBMC 10x single cell RNA-seq data
 
-### Case Study 3: largish single cell data
-
-### Case Study 4: non-spherical clusters, pdfsense
+### Case Study 3: Single cell mouse retina data
 

### 2020-03-13 Rmd/01-intro.Rmd
a bit of restructuring

--- 2020-02-21 Rmd/01-intro.Rmd
@@
 
-overview
+Exploratory data analysis (EDA) is a vital element of the modern 
+statistical workflow - it is an analyst’s first pass at 
+understanding their data; revealing all it’s messes and 
+uncovering hidden insights [@Tukey1977-pn; @Grolemund2017-uc]. 
+It is an iterative process involving computation and visualization,
+leading to new hypotheses that can be tested and formalised using 
+statistical modelling. As datasets grow in complexity and become
+increasingly heterogeneous and multidimensional, 
+the use of EDA becomes vital to ensure the integrity and quality of analysis 
+outputs. This is certainly true in high-throughput biological 
+data analysis, where constraints on computation time and memory, in addition
+to the analyst's time, results in EDA becoming increasingly difficult and 
+neglected.
 
-describe main topic - research questions
-chapters 1, 2, 3, 4 (your contributions to the joint work)
-scope
+## A grammar for genomic data analysis
+
+The approach taken by the suite of software packages collectively known as the 
+*tidyverse*  is an attempt to formalise aspects of the EDA process in the R
+programming language under a single semantic known as *tidy data* 
+[@baser; @tidyverse; @Wickham2014-jc]. Simply put, a *tidy data* set is a 
+rectangular table where each row of the table corresponds to an  observation, 
+each column corresponds to a variable and each cell a value. There is a 
+surprisingly large amount of utility that can be achieved with this definition. 
+By having each column representing a variable, variables in the data can be mapped
+to graphical aesthetics of plots. This paradigm enables the grammar of graphics 
+as implemented by *ggplot2* [@Wickham2016-gz; @Wilkinson2005-kq]. User interfaces
+as implemented by *tidyverse*, and in particular the *dplyr* package, are *fluent*; 
+they form a domain specific language (DSL) that gives users a mental model 
+for performing and composing common data transformation tasks 
+[@Wickham2017-dplyr; @FowlerFluent].
+
+It is unclear whether the *fluent* interfaces as implemented using the
+*tidy data* framework can be more generally applied and useful
+in fields such as high-throughput biology where domain specific semantics are 
+required. This is particularly true in the *Bioconductor* ecosystem, where much 
+thought has gone into the design of data structures that enable interoperability
+between different tools, biological assays and analysis goals [@Huber2015-ei].
+
+Chapter \@ref(ch:plyranges) shows that the *tidy data* semantic is applicable
+to range-based genomics data and develops a *fluent* interface to transforming
+it called *plyranges*. The software provides a framework to an assist
+an analyst to compose queries on genomics datasets. This chapter has been 
+published in Genome Biology [@Lee2019].
+
+## Integration and representation of genomic data structures
+
+## Visual analytics for dimension reduction and clustering workflows

### 2020-03-13 Rmd/01-intro.Rmd
words etc.

@@
 outputs. This is certainly true in high-throughput biological 
-data analysis, where constraints on computation time and memory, in addition
+data sceince, where constraints on computation time and memory, in addition
 to the analyst's time, results in EDA becoming increasingly difficult and 
 neglected.
+
+This thesis develops tools and frameworks to perform EDA as part of the
+biological data science workflow...
 

### 2020-03-16 Rmd/01-intro.Rmd
short dot points
more dot points

@@
 
-## Visual analytics for dimension reduction and clustering workflows
+- talking about chapters 3 and 4 here
+- having the grammar doesn't impair interoperability, in fact we show it
+can be used to glue two different approches
+- the long form representation can be taken even further... 
+- recasting EDA problems as search, we want to find regions measured along the
+genome that display a characteristic of interest, and then link that to raw 
+data plots.
+
+## Visual analytics for dimension reduction workflows
+
+- talking about chapter 5
+- taking a step back in the last chapter, have been focused on data wrangling
+but am now looking the integration of vis and models
+- focussed on a common part of EDA as applied to transcriptomics, dimension
+reduction workflows
+- especially non-linear dr as that has gotten extremely popular
+- combining with the tour, we can grasp the trade-off between global and
+local views of the data
+- user interaction provides ability to dissect the quality of a DR, and
+perform cluster oriented tasks

### 2020-03-19 Rmd/05-embeddings.Rmd
rebuild
add examples in...

@@
 obtained from an NLDR method, we can preserve global structure and through user interactions like spatial linked brushing observe where
-the NLDR view may be misleading. We display show several case 
-studies from single cell genomics, that our approach is useful for
+the NLDR view may be misleading. We display several case 
+studies from single cell genomics, that shows our approach is useful for
 cluster orientation tasks. 
@@
 
-TODO
+**distortion**: points that are far away in one view, will show neighbours that are close together in the opposing view.
+
+**diffusion**: points that are close in one view, will show neighbours that are far apart in the opposing view.
 
@@
 
+- **concatenated views:** see local and global structures
+- **user interactions:** verify cluster shapes and locations
+- **quality checks:** interrogate an embedding with reference
+to the underlying data
+
 ### Tours as a streaming data problem
 
-### Linking views via spatial brushes
+- don't need to realise the entire sequence
+- instead generate new bases according to a fixed frame rate
+- allows user interactions to play/pause on the current view
+
+### Linking views via brushes
 
@@
 
-### Case Study 1: Fake trees
+### Case Study 1: Exploring tree structured data
+
+(ref:fake-trees) Example from PHATE
+
+```{r fake-trees, echo = FALSE, out.width = "50%", fig.cap="(ref:fake-trees)"}
+knitr::include_graphics(here::here("img", "tree.png"))
+```
+
+#### Using t-SNE
+
+- there's a lot of parameters to tweak
+- emphasises locality, distance between clusters of points can be misleading
+- size of clusters can be misleading 
+- may require a few different runs to capture topology of data
+- if there are clusters it will find them
 
@@
 ## Discussion
+
+## Acknowledgements

### 2020-03-23 Rmd/05-embeddings.Rmd
update bib

@@
   - Kevin moon
-  - pdfsense
   - single cell
-  - quick draw
 https://jlmelville.github.io/smallvis/init.html#other_initialization_options

### 2020-03-23 Rmd/05-embeddings.Rmd
adjusting posing queries para

@@
 
-We combine the tour view in a side by side layout with a scatter
-plot view from an embedding algorithm. The views are linked; analysts can brush a group of points in the embedding view, and see its shape in the tour view.
-This can be used for either manually identifying clusters, or verifying the location of clusters relative to other groupings. Consequently, simple linked
-brushing can alleviate a downfall of non-linear methods such as UMAP or t-SNE
-that break hierarchical structure of clusters. Aside from simple
+We have combined the tour view in a side by side layout with 
+a scatter plot view that represents the output of an embedding algorithm. 
+These views are linked; analysts can brush regions or
+highlight collections of points in either view. Linked highlighting can
+be performed when points have been previously labeled according to some
+discrete structure, i.e. cluster labels are available. This is achieved
+via the analyst clicking on groups in the legend, which causes unclicked
+groupings to have their points become less opaque. Consequently, 
+simple linked highlighting can alleviate a known downfall of methods such as 
+UMAP or t-SNE: that is distances between clusters are misleading. By
+highlighting corresponding clusters in the tour view, the analyst can
+see the relationship between clusters, and therefore obtain a more accurate 
+representation of the topology of their data.
+
+Simple linked brushing is achieved via mouse-click and drag movements. By
+default, when brushing occurs in the tour view, the current projection is 
+paused and corresponding points in the embedding view are highlighted. 
+Likewise, when brushing occurs in the embedding view, corresponding points
+in the tour view are highlighted. In this case, an analyst can use brushing
+for manually identifying clusters and verifying cluster locations and shapes:
+brushing in the embedding view gives analysts a sense of the shape and 
+proximity of cluster in high-dimensional space.
+
+### Making comparisons: think global, act local
+
+Aside from simple
 linked brushing, we can also pose more complex queries via interrogating
@@
 view of the effect of an embedding algorithm.
-
-### Making comparisons
 

### 2020-03-23 Rmd/05-embeddings.Rmd
fleshing out para

@@
 
-A complimentary approach for visualizing structure in high dimensional data is
+A complementary approach for visualizing structure in high dimensional data is
 the tour. A tour is a sequence of projections of a high dimensional dataset
@@
 
-### Making comparisons: think global, act local
+### Making comparisons: revising embeddings
 
-Aside from simple
-linked brushing, we can also pose more complex queries via interrogating
-the neighbourhood graph. Using a linked neighbourhood brush, we can 
+As mentioned previously, when using any DR method, we are assuming
+the embedding is representative of the high-dimensional dataset it was
+computed from. Defining what it means for embedding to be 'representative` or
+'faithful' to high-dimensional data is ill-posed and depends on the underlying
+taks an analyst is trying to achieve. At the very minimum, we are interested
+in distortions and diffusions of the high-dimensional data. Distortions
+occur when points that are near each other in the embedding view are 
+far from each other in the original dataset. This implies that the embedding 
+is not continuous. Diffusions occur when points are far from each other in the embedding view are near in the original data.  Whether, points are near or far 
+is reliant on the distance metric used; distortions and diffusions can
+be thought of as the preservation of distances or the nearest neighbors
+graphs between the high-dimensional space and the embedding space. As distances
+can be noisy in high-dimensions, ranks can be used instead as has been 
+proposed by ... Identifying distortions and diffusions allows an analyst
+to investigate the quality of their embedding and revise them iteratively. We
+propose this can be done visually using our side-by-side tour and embedding
+views. Some quality checks, such as finding distortions can be identified using simple linked brushing, however we can also interrogate them via spatial
+brushes and brush compostions.
+
+Using a linked neighbourhood brush, we can 
 visually investigate the nearest neighbour relationships in the 
@@
 view of the effect of an embedding algorithm.
-
-**distortion**: points that are far away in one view, will show neighbours that are close together in the opposing view.
-
-**diffusion**: points that are close in one view, will show neighbours that are far apart in the opposing view.
 

### 2020-03-23 Rmd/05-embeddings.Rmd
add stuff from talk

@@
 
+ Look up nearest neighbours graph from points that lie in a brushing region.
+
+Highlight the corresponding neighbours using colour or transparency in the linked view.
+
+- The $k$ nearest neighbours graph can be pre-computed quickly, for either $X$, $Y$ or both.
+- Instead of using the neighbour indices, we could use the neighbour distances instead.
+- Composition of multiple brushes could be used to show where there are matches/mismatches between nearest neighbour graphs.
+
 Using a linked neighbourhood brush, we can 

### 2020-03-24 Rmd/01-intro.Rmd
typos

@@
 outputs. This is certainly true in high-throughput biological 
-data sceince, where constraints on computation time and memory, in addition
+data science, where constraints on computation time and memory, in addition
 to the analyst's time, results in EDA becoming increasingly difficult and 
@@
 - having the grammar doesn't impair interoperability, in fact we show it
-can be used to glue two different approches
+can be used to glue two different approaches
 - the long form representation can be taken even further... 

### 2020-03-24 Rmd/05-embeddings.Rmd
typos

@@
 discrete structure, i.e. cluster labels are available. This is achieved
-via the analyst clicking on groups in the legend, which causes unclicked
-groupings to have their points become less opaque. Consequently, 
+via the analyst clicking on groups in the legend, which causes unselected groupings to have their points become less opaque. Consequently, 
 simple linked highlighting can alleviate a known downfall of methods such as 
@@
 'faithful' to high-dimensional data is ill-posed and depends on the underlying
-taks an analyst is trying to achieve. At the very minimum, we are interested
+task an analyst is trying to achieve. At the very minimum, we are interested
 in distortions and diffusions of the high-dimensional data. Distortions
@@
 views. Some quality checks, such as finding distortions can be identified using simple linked brushing, however we can also interrogate them via spatial
-brushes and brush compostions.
+brushes and brush composition.
 

### 2020-03-24 Rmd/05-embeddings.Rmd
update text

@@
 
-The `liminal` software is implemented as an open source R package @baseR. All
-visualisations produced by the package were constructed using the `vegawidget`
-package, which is an R interface to the Vega-Lite grammar of interactive
-graphics. Ensemble graphics for arranging tour and embedding views are
-generated using the `shiny` packages. We have made extensive use of the `tourr`
-package for constructing tour paths.
-
-- **concatenated views:** see local and global structures
-- **user interactions:** verify cluster shapes and locations
-- **quality checks:** interrogate an embedding with reference
-to the underlying data
+We have implemnted the above design as an open source R package 
+called `liminal` [@baseR]. The package allows analysts to constuct
+concatenated visualisations, via the
+Vega-Lite grammar of interactive graphics (using `vegawidget` package)
+and provides an interface for constructing brushes and manipulating tour paths
+using the `shiny` and `tourr` packages. `liminal` also provides a stand-alone
+interface to the tour, in addition to the linked scatterplot interfaces 
+discussed above. Below we display the `liminal` API, and our approach to
+generating tour paths and user interactions.
 
@@
 
-### Case Study 1: Exploring tree structured data
+### Case Study 1: Exploring tree structured data with tours and t-SNE
 

### 2020-03-24 Rmd/05-embeddings.Rmd
spelling

@@
 
-We have implemnted the above design as an open source R package 
-called `liminal` [@baseR]. The package allows analysts to constuct
+We have implemented the above design as an open source R package 
+called `liminal` [@baseR]. The package allows analysts to construct
 concatenated visualisations, via the

### 2020-04-15 Rmd/01-intro.Rmd
add paragraph

@@
 
-This thesis develops tools and frameworks to perform EDA as part of the
-biological data science workflow...
+This thesis focuses on three core aspects of EDA as part of a biological data
+science workflow: wrangling, integration and visualisation (in particular we 
+focus on the exploration of genomics and transcriptomics data).
+It is organised in the same manner that you would iterate back and
+forth between these tasks  during an analysis. To begin you wrangle your data into a structure 
+that you can understand and use; section \@ref(sec:tidygranges) introudces a
+grammar-based framework for transforming genomics data that is described in 
+chapter \@ref(ch:plyranges). Next, you integrate data and model outputs
+over genomic regions to gain biological insight; \@ref(sec:ir) introduces a
+framework for incorporating and discovering 'interesting' genomic regions,
+described further in chapters \@ref(ch:fluentGenomics) and \@ref(ch:intron).
+Third, you cycle between model-based representations and visualisations
+of your data; section {#sec:va} introduces a visual analytics approach for
+understanding non-linear dimension reduction techniques and is described in
+detail in chapter \@ref(ch:tsne). Lastly, you take a step back and look at
+the implications and limitations of your work; in chapter \@ref(ch:conclusion)
+I describe the impacts of my work and plans for future research.
 
-## A grammar for genomic data analysis
+## A grammar for genomic data analysis {#sec:tidygranges}
 
@@
 
-## Integration and representation of genomic data structures
+## Integration and representation of genomic data structures {#sec:ir}
 

### 2020-04-15 Rmd/01-intro.Rmd
add citation

@@
 It is organised in the same manner that you would iterate back and
-forth between these tasks  during an analysis. To begin you wrangle your data into a structure 
-that you can understand and use; section \@ref(sec:tidygranges) introudces a
-grammar-based framework for transforming genomics data that is described in 
-chapter \@ref(ch:plyranges). Next, you integrate data and model outputs
-over genomic regions to gain biological insight; \@ref(sec:ir) introduces a
-framework for incorporating and discovering 'interesting' genomic regions,
-described further in chapters \@ref(ch:fluentGenomics) and \@ref(ch:intron).
-Third, you cycle between model-based representations and visualisations
-of your data; section {#sec:va} introduces a visual analytics approach for
-understanding non-linear dimension reduction techniques and is described in
+forth between these tasks  during an analysis. To begin you wrangle your data 
+into a structure that you can understand and use; section \@ref(sec:tidygranges)
+introudces a grammar-based framework for transforming genomics data that is 
+described in chapter \@ref(ch:plyranges). Next, you integrate data and model 
+outputs over genomic regions to gain biological insight; \@ref(sec:ir) 
+introduces a framework for incorporating and discovering 'interesting' genomic
+regions, described further in chapters \@ref(ch:fluentGenomics) and 
+\@ref(ch:intron). Third, you cycle between model-based representations and 
+visualisations of your data; section {#sec:va} introduces a visual analytics 
+approach for understanding non-linear dimension reduction techniques and is described in
 detail in chapter \@ref(ch:tsne). Lastly, you take a step back and look at
 the implications and limitations of your work; in chapter \@ref(ch:conclusion)
-I describe the impacts of my work and plans for future research.
+I describe the outpus of my thesis and plans for future research.
 
@@
 
-- talking about chapters 3 and 4 here
-- having the grammar doesn't impair interoperability, in fact we show it
-can be used to glue two different approaches
-- the long form representation can be taken even further... 
-- recasting EDA problems as search, we want to find regions measured along the
-genome that display a characteristic of interest, and then link that to raw 
-data plots.
+It is rare that a biological data analysis will involve
+a single measurement assay or that only one aspect of a measurement assay
+will be of interest to the biological question under study. While there are 
+many approaches to integrating data sets from multiple assays using multivariate
+statistical techniques [@Meng2016-pz], in chapter \@ref(ch:fluentGenomics)
+we describe a simpler end-to-end workflow for integrating 
+results along the genome using *plyranges*. This workflow also shows that
+our grammar based approach does not impair interoperability between
+the *tidyverse* and *Bioconductor* approaches, and in fact they work seamlessly
+together.
 
-## Visual analytics for dimension reduction workflows
+Moreover, in chapter \@ref(ch:intron) we explore the limits of the *tidy data*
+semantic by extending *plyranges* to analyse coverage estimated on RNA-seq
+data by developing a new software tool called *superintronic*. We show that the
+long-form tidy representation is an effective way of combining the experimental
+design and reference annotations into a single genomic data structure for
+exploration. We use *superintronic* to develop a framework for discovering
+interesting regions of coverage and apply our approach to integrating intron
+signal from RNA-seq data.
+
+## Visual analytics for dimension reduction workflows {#sec:va}
 

### 2020-04-16 Rmd/01-intro.Rmd
update intro

@@
 
-This thesis focuses on three core aspects of EDA as part of a biological data
-science workflow: wrangling, integration and visualisation (in particular we 
-focus on the exploration of genomics and transcriptomics data).
-It is organised in the same manner that you would iterate back and
+This thesis focuses on core aspects of EDA as part of a biological data
+science workflow: wrangling, integration and visualisation, with a focus 
+on applications to genomics and transcriptomics. It is organised in 
+the same manner that you would iterate back and
 forth between these tasks during an analysis. To begin you wrangle your data 
 into a structure that you can understand and use; section \@ref(sec:tidygranges)
-introudces a grammar-based framework for transforming genomics data that is 
+introduces a grammar-based framework for transforming genomics data that is 
 described in chapter \@ref(ch:plyranges). Next, you integrate data and model 
@@
 \@ref(ch:intron). Third, you cycle between model-based representations and 
-visualisations of your data; section {#sec:va} introduces a visual analytics 
-approach for understanding non-linear dimension reduction techniques and is described in
-detail in chapter \@ref(ch:tsne). Lastly, you take a step back and look at
-the implications and limitations of your work; in chapter \@ref(ch:conclusion)
-I describe the outpus of my thesis and plans for future research.
+visualisations of your data; section \@ref(sec:va) introduces a visual analytics 
+approach for understanding non-linear dimension reduction techniques and is 
+described in detail in chapter \@ref(ch:tsne). Lastly, you take a step back and 
+look at the implications and limitations of your work; in chapter 
+\@ref(ch:conclusion) I describe the outputs of my thesis and plans for future 
+research.
 
@@
 
-- talking about chapter 5
-- taking a step back in the last chapter, have been focused on data wrangling
-but am now looking the integration of vis and models
-- focussed on a common part of EDA as applied to transcriptomics, dimension
-reduction workflows
-- especially non-linear dr as that has gotten extremely popular
-- combining with the tour, we can grasp the trade-off between global and
-local views of the data
-- user interaction provides ability to dissect the quality of a DR, and
-perform cluster oriented tasks
+In the final chapter, we move away from data wrangling and towards the 
+integration of visualisation with model-based summaries of high-dimensional
+data sets. We focus on a common tool for EDA (especially applied to single-cell
+transcriptomics): dimension reduction (DR). We consider the incorporation of 
+interactive and dynamic graphics to assist analysts in using DR techniques
+for cluster orientation tasks. In particular, we advocate for the use of tours
+[@Cook1995-bi] alongside a DR visualisation to highlight potential pitfalls and 
+distortions of obtained from a DR method. This approach acknowledges that
+there is no 'one' best visualisation or DR for a high-dimensional dataset,
+and we often want to have an understanding of both the global and local structure
+within our data.
+
+Chapter \@ref(ch:liminal) introduces
+a software package called *liminal* for constructing these views and
+a user interaction framework for identifying distortions. We present several
+case studies for using our approach to diagnose the quality of 
+results obtained via t-distributed stochastic neighbour embeddings (t-SNE)

### 2020-04-16 Rmd/01-intro.Rmd
update references

@@
 
-Chapter \@ref(ch:liminal) introduces
+Chapter \@ref(ch:tsne) introduces
 a software package called *liminal* for constructing these views and

### 2020-06-03 Rmd/04-intron.Rmd
update toc; rebuild
start adding in previous content

@@
-# Tidy coverage analysis uncovers intron retention {#ch:intron}
+# Tidy coverage analysis and visualisation with superintronic {#ch:intron}
 
-<!-- # Introduction -->
+## Introduction
+
+## Methods
+
+### Motivation: intron signal in RNA-seq data
+
+Read coverage of intronic and exonic regions were calculated for poly(A) RNA and Total RNA HCC827 human cell lines using our \textit{superintronic} package, via the \textit{Rsamtools} package \cite{rsamtools,GenomicRanges}.
+Genes of interest were restricted to protein coding genes on reference chromosomes, and we removed any genes that overlapped another to simplify the analysis and reduce coverage ambiguity.  
+Genes were then further filtered if they were not expressed in the poly(A) RNA protocol (requiring at least 3 reads overlapping intronic and exonic regions). 
+A total of 3,262 genes were examined and categorised as short, regular or long (roughly 1,087 genes in each category) by splitting the length of each gene into three bins by tertiles.
+
+Using associated BAM files and GENCODE v27 annotation GTF, \textit{superintronic} summarised the number of bases covered at a given coverage score for each gene and sample.
+For exonic regions and intronic regions, coverage scores were transformed to log$_2$-scale using an offset of 0.5, and then normalised by dividing by each gene and each sample's maximum log-coverage score.
+Normalised log-coverage scores, or {\it relative log-coverage} (relative to each gene's maximum coverage), were divided into 20 windows along the length of each gene using the \textit{GenomicRanges} package \cite{GenomicRanges}.
+To summarise coverage patterns across genes, \textit{plyranges} \cite{Lee2019plyranges} was applied to relative log-coverage scores by intersecting it with the positional windows.
+Each window's mean coverage score (mean relative log-coverage) was calculated for each gene.
+Based on the position of the windows, they were further summarised across genes by taking its mean. % (mean of mean relative log-coverage).  
+The summary values were calculated separately for genes categorised as short, regular and long to represent general coverage trends along the gene body.
+
+## Data
+
+\textit{Human cell lines} of lung adenocarcinoma HCC827 and NCI-H1975 were cultured on three
+separate occasions by Holik {\it et al.} (2017) \cite{Holik:NAR:2017} giving three pseudo biological replicates. 
+RNA was extracted from each pseudo biological replicate and split into two and prepared as poly(A) RNA and Total RNA libraries.
+Raw sequencing reads were downloaded from the Gene Expression Omnibus (GEO) \citep{GEO:2013} under accession number GSE64098.
+12 libraries were examined for this dataset.
+
+FASTQ files containing raw sequencing reads were aligned to the human \textit{hg38} or mouse \textit{mm10}
+% or {\color{blue}{zebrafish \textit{GRCz11}}} 
+genome using \textit{subjunc} \citep{Liao:NAR:2013} with default parameters in the \textit{Rsubread} software package \citep{Rsubread:2019}. 
+GENCODE's main \textit{Comprehensive gene annotation} file in GTF format was downloaded from {https://www.gencodegenes.org} for human (Release 27) and mouse (Release M12). 
+Using ``gene types'' (rather than ``transcript types'') from Gencode, the annotation files were simplified by taking the union of two or more overlapping exons from transcripts of the same gene.
+The adjustment provides a simplification of genomic positions on each strand, such that each position is classified as belonging to ``exon'', ``intron'' or otherwise outside of an annotated gene.
+Three resultant annotation files were saved in standard annotation format (SAF) -- exon annotation, intron annotation (region between exons), and genebody annotation (region spanning first to last exon).
+Our Supplementary Materials available at \href{https://github.com/charitylaw/Intron-reads}{https://github.com/charitylaw/Intron-reads} contain the scripts to process annotation files, together with other data analyses and supplementary figures.
+
+## Results
+
+### Representation of coverage estimation
+
+### Integration of external gene annotations
+
+### Discovery of regions of interest via 'ranglers'
+
+### A superintronic workflow for uncovering intron retention
+
+Considering IR also as another possible source of intron reads we propose a new method using our {\it superintronic} software to explore intron signal directly from aligned sequencing data with the assumption that most intron reads do not point to IR but pre-mRNA instead.
+{\it Superintronic} is an R package that is available to download and install at \href{https://github.com/sa-lee/superintronic}{https://github.com/sa-lee/superintronic}. 
+It extends the \textit{plyranges} Bioconductor package \citep{Lee2019plyranges} for genomics data analysis to develop a simple and modular interface for performing exploratory  genomics data analysis via coverage estimation.
+Each aspect of the \textit{superintronic} data analysis workflow as it has been applied for exploring intron signal is outlined in Figure~\ref{fig:workflows}b.
+
+Our software records the per base coverage over intron and exon regions of each gene, with the option of storing these per sample or summarised over variables in the experimental design such as by biological group or by RNA library preparation.
+Coverage scores are normalised using a log$_2$-transformation with an offset of 0.5 to get log-coverage values for which intron and exon summary statistics are constructed for each gene (described below).
+Within \textit{superintronic}, a suite of visualization tools to construct coverage plots for genes with intron and exon structures and scatter plots are provided.
+
+\subsection{\textit{Superintronic} finds genes with IR-like coverage profiles in human cell lines}
+
+\begin{figure*}[!tpb]
+\centerline{\includegraphics[width=\textwidth]{coverage-si-hits.pdf}}
+\caption{{\it Superintronic} selects genes with IR-like coverage profiles in poly(A) RNA HCC827 cell lines.
+Genes (a) PSMB7, (b) EIF2S3, and (c) ARGLU1 are highlighted out of 43 genes selected.
+Whilst the analysis was carried out on poly(A) RNA libraries, coverage is shown for both poly(A) RNA HCC827 (top) and Total RNA HCC827 (bottom) samples to ensure that results are not an artifact of RNA library preparation.
+Coverage is oriented from 5' to 3', with exon regions coloured green and intron regions coloured orange.
+}\label{fig:ir}
+\end{figure*}
+
+Using \textit{superintronic}, poly(A) RNA HCC827 cell lines were examined for genes with IR  after selecting genes in the hg38 reference that were protein coding, did not overlap any other gene and were placed on the main contigs -- a total of 6,606 genes.
+These genes were then split into intron and exon regions and intersected with the coverage of each sample. 
+Per gene intron and exon summary statistics, mean and standard deviation, were computed on log-coverage values. 
+We selected genes enriched for IR-like coverage profiles by looking for ``expressed'' genes, where for a substantial number of intron bases its coverage is much higher than other intron features within the same gene while having similar expression levels to the exon features.
+To do this we used the following thresholds -- genes had an average exon log-coverage of greater than 2 (corresponding to the mean of average exon log-coverage values across all genes), a standard deviation of intron log-coverage greater than 1.5 (corresponding to the mean of intron standard deviation values across all genes), and genes with a large number of intron bases with log-coverage greater than 2 (top 1\% of genes).
+The thresholds were chosen after examining distributions of the summary statistics (Supplementary Figure 4).
+Forty-three genes met these criteria, where a manual check of coverage profiles revealed that 36 genes indeed appear IR-like (see `Analyses' page in Supplementary Materials).
+We highlight 3 of these genes in Figure~\ref{fig:ir}.
+The coverage of remaining 7 genes appear to be more pre-mRNA-like, with large variation in intron coverage, where at its peak it is expressed similarly to exon features.
+
+We use poly(A) RNA samples in our analysis for consistency with previous studies on IR.
+In theory, Total RNA samples may be more appropriate for this exercise since it is less biased towards the 3' end, where high 3' intron coverage as a result of 3' bias in poly(A) RNA libraries can be mistaken as a 3' retained intron.
+For this reason, we review both the coverage of poly(A) RNA and Total RNA samples in  our \textcolor{red}{S}upplementary \textcolor{red}{M}aterials for all selected genes to ensure that retained introns found in poly(A) RNA samples are not an artifact of RNA library preparation.
+
+If differentially retained introns are of interest, one could simply run \textit{superintronic} on two conditions separately and compare lists of IR-like genes between groups.
+We found that differences between the lists of IR-like genes were concordant with \textit{index} results for genes DE by intron counts, as expected (see `Analyses' page in Supplementary Materials).
+For example, 14 genes were uniquely selected as IR-like in Total RNA HCC827 cell line when compared to Total RNA NCI-H1975 cell line using \textit{superintronic}, where 13 of those genes were also found to be DE in intron counts by \textit{index}, and in the expected direction.
+Similarly, 19 genes were uniquely detected as IR-like in Total RNA NCI-H1975 by \textit{superintronic}, 16 of which were also detected as DE in intron counts by \textit{index}.
+Additionally, 12 genes which were found to be IR-like in both cell lines appear to have varying intron expression according to \textit{index} since it is found to be DE in intron counts.
+
+\textcolor{red}{
+Whilst genes that are uniquely IR-like were concordant with directional changes in intron counts, we did not find that IR-like genes played a large part in \textit{index} gene categorisation.
+This was mostly expected since thousands of genes were detected as significantly DE by \textit{index} and a relatively small number of genes were detected as IR-like by \textit{superintronic}.
+Specifically, IR-like genes did not overlap with one particular \textit{index} category (see `Analyses' page in Supplementary Materials).
+Of genes that are classified as intron+ or intron-, only 3 out of 1,006 genes were also detected by \textit{superintronic} as IR-like.
+In the ``mixed'' category, 2 out of 54 genes were also detected as IR-like.
+}
+
+## Discussion
+
+## Conclusion
 

### 2020-07-07 Rmd/01-intro.Rmd
rejig intro

@@
 on applications to genomics and transcriptomics. It is organised in 
-the same manner that you would iterate back and
-forth between these tasks during an analysis. To begin you wrangle your data 
-into a structure that you can understand and use; section \@ref(sec:tidygranges)
+according to the aforementioned workflow steps. To begin we discuss wrangling
+biological data using a coherent represntation and programming interface; 
+section \@ref(sec:tidygranges)
 introduces a grammar-based framework for transforming genomics data that is 
-described in chapter \@ref(ch:plyranges). Next, you integrate data and model 
-outputs over genomic regions to gain biological insight; \@ref(sec:ir) 
+described in chapter \@ref(ch:plyranges). Next, we look at integrating data and 
+model outputs over genomic regions to gain biological insight; \@ref(sec:ir) 
 introduces a framework for incorporating and discovering 'interesting' genomic
 regions, described further in chapters \@ref(ch:fluentGenomics) and 
-\@ref(ch:intron). Third, you cycle between model-based representations and 
-visualisations of your data; section \@ref(sec:va) introduces a visual analytics 
-approach for understanding non-linear dimension reduction techniques and is 
-described in detail in chapter \@ref(ch:tsne). Lastly, you take a step back and 
-look at the implications and limitations of your work; in chapter 
+\@ref(ch:intron). Third, we consider the challenges in visualising  
+high dimensional data; section \@ref(sec:va) introduces a visual analytics 
+approach for understanding non-linear dimension reduction techniques described
+in chapter \@ref(ch:tsne). Lastly, in chapter 
 \@ref(ch:conclusion) I describe the outputs of my thesis and plans for future 
-research.
+developments.
 
@@
 an analyst to compose queries on genomics datasets. This chapter has been 
-published in Genome Biology [@Lee2019].
+published in @Lee2019.
 
@@
 statistical techniques [@Meng2016-pz], in chapter \@ref(ch:fluentGenomics)
-we describe a simpler end-to-end workflow for integrating 
+we describe a simple end-to-end workflow for integrating 
 results along the genome using *plyranges*. This workflow also shows that
@@
 the *tidyverse* and *Bioconductor* approaches, and in fact they work seamlessly
-together.
+together. This chapter has been published in @Lee2020-ie.
 
@@
 interesting regions of coverage and apply our approach to integrating intron
-signal from RNA-seq data.
+signal from RNA-seq data. This chapter is based on the preprint @Lee2019-mf.
 
@@
 a user interaction framework for identifying distortions. We present several
-case studies for using our approach to diagnose the quality of 
-results obtained via t-distributed stochastic neighbour embeddings (t-SNE)
+case studies using data sets from both single-cell transciptomics and 
+machine learning  for using our approach to diagnose the quality of 
+results obtained via popular non-linear DR methods
+like t-distributed stochastic neighbour embeddings (t-SNE).

### 2020-07-08 Rmd/06-conclusion.Rmd
first pass at intro to conculsion

@@
 # Conclusion  {#ch:conclusion}
+
+In this thesis, I have designed tools to explore workflow steps
+that are integral to modern biological data science. In particular, I have
+implemented software that facilitates the wrangling, integration, and 
+visualisation of high-throughput biological data in a principled and pragmatic
+manner. Chapter \@ref(ch:plyranges), "plyranges: a grammar of genomic data
+transformation" developed a new domain specific language for genomics data.
+Chapter \@ref(ch:fluentGenomics), "Fluent genomics with plyranges and tximeta"
+described techniques integrating data along the genome, and emphasised
+the importance of interoperability between tools. Chapter \@ref(ch:intron),
+"Exploratory coverage analysis with superintronic and plyranges", tackled data 
+integration from a different angle by looking at multiple summaries
+of variables measured along the genome to find regions of biological interest. 
+Chapter \@ref(ch:tsne), "Casting multiple shadows: high-dimensional 
+interactive data visualisation with tours and embeddings", explored pragmatic
+approaches to high dimensional data visualisation in light of the rise of 
+popular non-linear embedding methods.
+
+The chapters
 

### 2020-07-08 Rmd/06-conclusion.Rmd
add first draft of conclusion

--- 2020-02-24 Rmd/06-conclusion.Rmd
@@
+```{r bioc-stats, include = FALSE}
+library(dplyr)
+library(ggplot2)
+dl_stats <- BiocPkgTools::biocDownloadStats() %>%
+  filter(Package == "plyranges")
+
+plyranges_stats <- dl_stats %>% 
+  summarise(no_ip = sum(Nb_of_distinct_IPs), no_downloads = sum(Nb_of_downloads))
+
+```
+
 # Conclusion  {#ch:conclusion}
 
-<!-- remind the reader of the research problem and purpose and how they were addressed -->
-<!-- briefly summarise what has been covered in the paper -->
-<!-- make some kind of holistic assessment/judgement/ claim that pertains to the whole project (i.e., more than a descriptive summary) -->
-<!-- assess the value/relevance/ implications of the key findings in light of existing studies and literature -->
-<!-- ‘speak’ to the Introduction -->
-<!-- outline implications of the study (for theory, practice, further research) -->
-<!-- comment on the findings that failed to support or only partially support the hypothesis or research questions directing the study -->
-<!-- refer to the limitations of the studies that may affect the validity or the generalisability of results -->
-<!-- make recommendations for further research -->
-<!-- make claims for new knowledge/ contribution to knowledge. -->
+In this thesis, I have designed tools to explore workflow steps
+that are integral to modern biological data science. In particular, I have
+implemented software that facilitates the wrangling, integration, and 
+visualisation of high-throughput biological data in a principled and pragmatic
+manner.  A significant amount of my work has been devoted to the development of 
+open source R/Bioconductor packages: **plyranges**, **superintronic**
+and **liminal**. I have emphasised how coherent software packages 
+are tools for thought; they enable analysts to reason about their data and 
+models through the composition of workflows.
 
-## Findings and limitations
+Chapter \@ref(ch:plyranges), "plyranges: a grammar of genomic data
+transformation" developed a new domain specific language for genomics data analysis.
+Chapter \@ref(ch:fluentGenomics), "Fluent genomics with plyranges and tximeta"
+described techniques integrating data along the genome, and emphasised
+the importance of interoperability between analysis tools. Chapter 
+\@ref(ch:intron), "Exploratory coverage analysis with superintronic and 
+plyranges", tackled data  integration from a different angle by looking at 
+multiple summaries of variables measured along the genome to putative regions of 
+intron retention.  Chapter \@ref(ch:tsne), "Casting multiple shadows: 
+high-dimensional  interactive data visualisation with tours and embeddings", 
+explored pragmatic approaches to high dimensional data visualisation in light of
+the rise of popular non-linear embedding methods. 
+To finish, I will dicuss the implications of this work and provide 
+suggestions for further research.
 
-## Software development and impact
+## Findings and Impact
 
-## Further research
+The early chapters of this thesis explored the *tidy data* semantic and its
+extension to range based genomics data. This culminated in the development
+of the **plyranges** package and consequent explorations of the applicability
+of its interface and use of the *tidy data* concept in chapters 
+\@ref(ch:fluentGenomics) and \@ref(ch:intron). The package has been relatively 
+successful: it has been downloaded `r plyranges_stats$no_downloads` times from 
+`r plyranges_stats$no_ip` unique IP addresses since its availability on 
+Bioconductor. I have also had the privilege of teaching workshops on **plyranges**
+at Bioconductor conferences which also led to the development of the work in 
+chapter\@ref(ch:fluentGenomics) and refined the development of the API. A 
+broader impact of the work, has been the discussions around the
+concepts of fluent interfaces and tidy data within the Bioconductor community,
+which has led to several developments currently in place that are exploring
+different approaches for fluent interfaces for other types of 'omics data.
+
+The **superintronic** software described in chapter \@ref(ch:intron) has
+been used in @Lee2019-mf to disentangle and view intron signal in RNA-seq
+data. Here, we again show the strengths of providing a long-form representations
+of genomics data (in this case coverage vectors). By leveraging **plyranges**
+we were then able to create a set of data descriptors that we could link back
+to the raw data to discover genes thought to be associated with a real biological 
+signal. An interesting extension to this work would be applying it to single
+cell and long-read based transcriptomics data, where scalability and much
+larger design matrices would be come an issue.
+
+Finally, the **liminal** software aims to provide a more holistic approach to
+analysis tasks requiring the use of dimensionality reduction algorithms. We
+showed how to incorporate interactive graphics and tours to identify problems
+with embeddings. Based on the case studies provided I believe that the methods
+used in **liminal** could be broadly applicable many high dimensional datasets.
+
+## Further Research
+
+### Scalable grammars for non-rectangular data structures
+
+A limitation of the grammar as we have implemented it in 
+**plyranges** is lack of scalability and computational speed for data sets
+that do not fit in memory. We attempted several techniques for perfoming 
+delayed operations over range-based data, however a more general approach
+that allows for data stored on the cloud or in scientific data formats
+like HDF5 that leverages existing Bioconductor frameworks would be useful.
+We showed in chapters \@ref(ch:fluentGenomics) ad \@ref(ch:intron) that an a
+nalyst is able to do some very complex data transformations and resampling 
+procedures via casting results into *GRanges* object. However, it is unclear 
+whether the semantics of our grammar can be extended to data that can not be 
+easily reshaped into long form tidy representations. Moreover, further work
+is required to explore the design space of grammars for data 
+transformations and grammars for graphics when the data large, multifaceted
+and non-rectangular.
+
+### Extending tour displays for large $n$ data sets
+
+We showed in chapter \@ref(ch:liminal) that tours provide a global overview
+that can be used as tool for exploring model fits. An issue that arises is
+how to scale the tour as the number of observations increases. There are
+latencies in sending data from the back end to the visualisation client
+that causes lag during animation. One could also question whether point
+based displays are appropriate in this case, and it would be worth exploring 
+the usability of animations based on binning the projections. Moreover, 
+when the number of observations are large, the points in the projections are 
+concentrated in the centre of the tour display obscuring interesting aspects
+of the data. This is mitigated via having
+the ability to zoom, but further research into transforming the projections
+to avoid crowding would be valuable. An added complexity to changes in visual
+displays are thinking about the design of user interactions, several promising
+avenues based on section tours should be explored [@Laa2020-wr; @Laa2020-fy].

### 2020-07-09 Rmd/04-intron.Rmd
rejig chapter 4
rejigging sections for chapter

@@
-# Tidy coverage analysis and visualisation with superintronic {#ch:intron}
+# Exploratory coverage analysis with superintronic and plyranges {#ch:intron}
 
@@
 
+### Motivation: intron signal in RNA-seq data
+
 ## Methods
 
-### Motivation: intron signal in RNA-seq data
+### Representation of coverage estimation
+
+### Integration of external gene annotations
+
+### Discovery of regions of interest via 'data descriptors'
+
+## A workflow for uncovering intron retention
 
@@
 The summary values were calculated separately for genes categorised as short, regular and long to represent general coverage trends along the gene body.
-
-## Data
 
@@
 Our Supplementary Materials available at \href{https://github.com/charitylaw/Intron-reads}{https://github.com/charitylaw/Intron-reads} contain the scripts to process annotation files, together with other data analyses and supplementary figures.
-
-## Results
-
-### Representation of coverage estimation
-
-### Integration of external gene annotations
-
-### Discovery of regions of interest via 'ranglers'
-
-### A superintronic workflow for uncovering intron retention
 
@@
 We highlight 3 of these genes in Figure~\ref{fig:ir}.
-The coverage of remaining 7 genes appear to be more pre-mRNA-like, with large variation in intron coverage, where at its peak it is expressed similarly to exon features.
+The coverage of remaining 7 genes appear to be more pre-mRNA-like, with large 
+variation in intron coverage, where at its peak it is expressed similarly to exon features.
 

### 2020-07-13 Rmd/04-intron.Rmd
shuffling text
add abstract

@@
 # Exploratory coverage analysis with superintronic and plyranges {#ch:intron}
+
+Here we consider a tidy-data approach for exploring
+estimated coverage from RNA-seq data. We establish a simple 
+framework, for aggregating accross experimental design, and annotated genomic regions to discover 'interesting' coverage trace plots. We highlight how this framework can be used to
+develop data descriptions that find putative genes with
+intron retention. Our framework is implemented in a software
+package called **superintronic**, available at https://github.com/sa-lee/superintronic.
 
@@
 ## Methods
+
+Considering IR also as another possible source of intron reads we propose a new method using our {\it superintronic} software to explore intron signal directly from aligned sequencing data with the assumption that most intron reads do not point to IR but pre-mRNA instead.
+{\it Superintronic} is an R package that is available to download and install at \href{https://github.com/sa-lee/superintronic}{https://github.com/sa-lee/superintronic}. 
+It extends the \textit{plyranges} Bioconductor package \citep{Lee2019plyranges} for genomics data analysis to develop a simple and modular interface for performing exploratory  genomics data analysis via coverage estimation.
+Each aspect of the \textit{superintronic} data analysis workflow as it has been applied for exploring intron signal is outlined in Figure~\ref{fig:workflows}b.
+
+Our software records the per base coverage over intron and exon regions of each gene, with the option of storing these per sample or summarised over variables in the experimental design such as by biological group or by RNA library preparation.
+Coverage scores are normalised using a log$_2$-transformation with an offset of 0.5 to get log-coverage values for which intron and exon summary statistics are constructed for each gene (described below).
+Within \textit{superintronic}, a suite of visualization tools to construct coverage plots for genes with intron and exon structures and scatter plots are provided.
 
@@
 Our Supplementary Materials available at \href{https://github.com/charitylaw/Intron-reads}{https://github.com/charitylaw/Intron-reads} contain the scripts to process annotation files, together with other data analyses and supplementary figures.
-
-Considering IR also as another possible source of intron reads we propose a new method using our {\it superintronic} software to explore intron signal directly from aligned sequencing data with the assumption that most intron reads do not point to IR but pre-mRNA instead.
-{\it Superintronic} is an R package that is available to download and install at \href{https://github.com/sa-lee/superintronic}{https://github.com/sa-lee/superintronic}. 
-It extends the \textit{plyranges} Bioconductor package \citep{Lee2019plyranges} for genomics data analysis to develop a simple and modular interface for performing exploratory  genomics data analysis via coverage estimation.
-Each aspect of the \textit{superintronic} data analysis workflow as it has been applied for exploring intron signal is outlined in Figure~\ref{fig:workflows}b.
-
-Our software records the per base coverage over intron and exon regions of each gene, with the option of storing these per sample or summarised over variables in the experimental design such as by biological group or by RNA library preparation.
-Coverage scores are normalised using a log$_2$-transformation with an offset of 0.5 to get log-coverage values for which intron and exon summary statistics are constructed for each gene (described below).
-Within \textit{superintronic}, a suite of visualization tools to construct coverage plots for genes with intron and exon structures and scatter plots are provided.
 

### 2020-07-13 Rmd/04-intron.Rmd
abstract and intro

@@
+```{r setup}
+library(superintronic)
+
+```
+
 # Exploratory coverage analysis with superintronic and plyranges {#ch:intron}
@@
 ### Motivation: intron signal in RNA-seq data
+
+In @Lee2019-mf, we showed that there is evidence that a major
+source of intron reads in RNA-seq datasets is pre-mRNA, and sought data analysis techniques to unravel different aspects of intron signal. In light of this,
+we made the assumption that most intron reads do not necessarily point to
+intron retention (IR) events, and developed a workflow based on combining
+multiple summary statistics, "data descriptors", to find coverage traces that appear to have IR-like events by collapsing coverage scores over a design matrix alongside the exonic and intronic parts of a gene.
+To do this we developed a new R package
+called **superintronic** that provides tooling for exploratory 
+coverage analysis by extending and integrating our previous software package,
+**plyranges** [@Lee2019]. We provide a workflow below
+using a zebrafish RNA-seq dataset for developing data descriptors to find IR like coverage traces.
+
+## Exploratory coverage analysis
+
+Although there are many flexible and powerful software tools, like BEDTools or Deeptools, for exploring and estimating coverage from common genomic data formats [@Quinlan2010-gc; @Ramirez2014-sh], it is advantegous to have tooling
+that is tightly coupled to a statistical computing language such as R.
+This allows interoperability between other software packages for data wrangling, visualisation and modelling within the ecosystem of the language that may not
+be possible with a single command line tool. There are also gains in reproduciblity as analysts do not have move between multiple software suites to explore their processed transcriptomics data.
+
+Here we emphasise the importance
+of actually looking at coverage traces to find biological events of interest, rather than only relying on numerical summaries of the data, and by facetting these traces over different combinations of the experimental design and with their biological context such as gene annotations.
+Of course, due to the sheer size of most reference genome annotaions 
+it would take an extremely long time for an analyst to look at every 
+single possible region where there is some interesting signal within the
+coverage trace. Because visualisation does not scale, we need a method to help
+us search the possible space of coverage traces and provide diagnostics for
+identifying traces with interesting biology. We have taken inspiration from
+visualisation literature; in particular the idea of scatterplot diagnostics ("scagnostics") for summarising the space of all possible 2D scatterplots
+to a small number of descriptors of each scatterplots properties such as 
+density or monotonicity [@Friedman2002-yx;@Wilkinson2005-xt].  Similarly,
+searching for unusual time series via estimating descriptors such as seasonality or autocorrelation, and visualising those descriptors instead [@Hyndman2015-sc].
 

### 2020-07-13 scripts/liminal-trees.R
remove old stuff

@@ -1,57 +1,0 @@
-## ----init-----------------------------------------------------------
-library(liminal)
-dim(fake_trees)
-
-
-## ----pc-view--------------------------------------------------------
-pcs  <- prcomp(fake_trees[, -ncol(fake_trees)])
-# var explained
-summary(pcs)$importance[3,1:12]
-
-## ----pca-xy---------------------------------------------------------
-fake_trees <- dplyr::bind_cols(fake_trees, as.data.frame(pcs$x))
-limn_xy(fake_trees, x = PC1, y = PC2, color = branches)
-
-
-## ----tsne-xy--------------------------------------------------------
-set.seed(2099)
-tsne <- Rtsne::Rtsne(dplyr::select(fake_trees, dplyr::starts_with("dim")))
-
-tsne_df <- data.frame(tsneX = tsne$Y[,1],
-                      tsneY = tsne$Y[,2],
-                      branches = fake_trees$branches)
-
-limn_xy(tsne_df, x = tsneX, y = tsneY, color = branches)
-
-# now together
-limn_tour_xylink(
-  dplyr::select(fake_trees, PC1:PC12, branches),
-  tsne_df,
-  x_color = branches,
-  y_color = branches
-)
-
-
-
-## ----pca-init--------------------------------------------------------
-tsne_v2 <- Rtsne::Rtsne(
-  dplyr::select(fake_trees, dplyr::starts_with("dim")),
-  Y_init = clamp_sd(as.matrix(dplyr::select(fake_trees, PC1,PC2)), sd = 1e-4),
-  perplexity = nrow(fake_trees) / 100,
-  eta = nrow(fake_trees) / 48,
-  max_iter = 3000
-)
-
-tsne_v2_df <- data.frame(
-  tsneX = tsne_v2$Y[,1], 
-  tsneY = tsne_v2$Y[,2], 
-  branches = fake_trees$branches
-)
-
-limn_tour_xylink(
-  dplyr::select(fake_trees, PC1:PC12, branches),
-  tsne_v2_df,
-  x_color = branches,
-  y_color = branches
-)
-

### 2020-07-13 scripts/superintronic.R

@@ -1,232 +1,0 @@
-## ---- superintronic-setup
-library(plyranges)
-library(BiocParallel)
-library(superintronic)
-library(ggplot2)
-library(here)
-theme_set(theme_bw())
-
-# Plotting helper functions
-set_plot_rng <- function(cvg, target, design, alpha) {
-  if (is(cvg, "GRangesList")) {
-    stopifnot(is(design, "DataFrame"))
-    cvg_rng <- as(lapply(cvg, function(x) superintronic::join_parts(x, target)),
-                  "GRangesList")
-    md <- S4Vectors::DataFrame(
-      Sample = S4Vectors::Rle(names(cvg_rng), lengths(cvg_rng)),
-      CellLine = S4Vectors::Rle(design$CellLine, lengths(cvg_rng)),
-      Kit = S4Vectors::Rle(as.factor(design$Kit), lengths(cvg_rng))
-    )
-
-    cvg_rng <- unlist(cvg_rng, use.names = FALSE)
-    S4Vectors::mcols(cvg_rng) <- cbind(S4Vectors::mcols(cvg_rng), md)
-
-    cvg_rng <- plyranges::mutate(cvg_rng,
-                                 log_score = log2(score + 0.5),
-                                 strand = BiocGenerics::strand(target),
-                                 var = paste0("Kit: ", Kit, " Cellline: ", CellLine))
-  }
-
-  if (is(cvg, "GRanges")) {
-    cvg_rng <- plyranges::filter_by_overlaps(cvg, target)
-    cvg_rng <- plyranges::mutate(cvg_rng,
-                                 strand = BiocGenerics::strand(target),
-                                 var = paste0("Kit: ", Kit, " Cellline: ", CellLine))
-  }
-
-  if (alpha) {
-    cvg_rng <- plyranges::disjoin_ranges_directed(group_by(cvg_rng, var),
-                                                  log_score = BiocGenerics::mean(log_score),
-                                                  feature_type = unlist(BiocGenerics::unique(feature_type)))
-
-    cvg_rng <- plyranges::mutate(cvg_rng,
-                                 alpha = dplyr::case_when(
-                                   score > 3 & feature == "intron" ~ 1,
-                                   score <= 3 & feature == "intron"~ 0.5,
-                                   TRUE ~ 1
-                                 ))
-  }
-
-  return(cvg_rng)
-}
-
-
-# pretty track plot
[…]

### 2020-07-13 Rmd/04-intron.Rmd
add next section

@@
 
-### Motivation: intron signal in RNA-seq data
+In high-throughput sequencing data sets, coverage is the estimated number of
+reads that overlap a single position of the reference genome, and is important for assessing sequencing data quality and used in many different aspects of omic's analysis such peak-calling in ChIP-seq or varinat calling in DNA-seq [@Sims2014-qk].
+Here we emphasise looking at coverage traces to find biological events of interest, rather than only relying on numerical summaries of the data.
+By facetting these traces over combinations of the experimental design and with their biological context such as gene annotations, we can gain an
+insight into biological signal under study. 
+Of course, due to the sheer size of most reference genome annotaions 
+it would take an extremely long time for an analyst to look at every 
+single possible region where there is some interesting signal within the
+coverage trace. Because visualisation does not scale, we need to search the possible space of coverage traces and provide diagnostics for
+identifying traces with interesting biology. We have taken inspiration from
+visualisation literature; in particular the idea of scatterplot diagnostics ("scagnostics") for summarising the space of all possible 2D scatterplots
+to a small number of descriptors of each scatterplots properties such as 
+density or monotonicity [@Friedman2002-yx;@Wilkinson2005-xt].  Similarly,
+searching for unusual time series via estimating descriptors such as seasonality or autocorrelation, and visualising those descriptors instead [@Hyndman2015-sc].
+
+Although there are many flexible and powerful software tools, like BEDTools or DeepTools, for exploring and estimating coverage from common genomic data formats [@Quinlan2010-gc; @Ramirez2014-sh], it is advantegous to have tooling
+that is tightly coupled to a statistical computing language such as R.
+This allows interoperability between other software packages for data wrangling, visualisation and modelling within the ecosystem of the language that may not
+be possible with a single command line tool. There are also gains in reproduciblity as analysts do not have to move between multiple software suites to explore their processed transcriptomics data.
 
@@
 multiple summary statistics, "data descriptors", to find coverage traces that appear to have IR-like events by collapsing coverage scores over a design matrix alongside the exonic and intronic parts of a gene.
-To do this we developed a new R package
-called **superintronic** that provides tooling for exploratory 
+To do this we developed a new R package called **superintronic** that provides tooling for exploratory
 coverage analysis by extending and integrating our previous software package,
-**plyranges** [@Lee2019]. We provide a workflow below
-using a zebrafish RNA-seq dataset for developing data descriptors to find IR like coverage traces.
+**plyranges** [@Lee2019].
 
-## Exploratory coverage analysis
-
-Although there are many flexible and powerful software tools, like BEDTools or Deeptools, for exploring and estimating coverage from common genomic data formats [@Quinlan2010-gc; @Ramirez2014-sh], it is advantegous to have tooling
-that is tightly coupled to a statistical computing language such as R.
-This allows interoperability between other software packages for data wrangling, visualisation and modelling within the ecosystem of the language that may not
-be possible with a single command line tool. There are also gains in reproduciblity as analysts do not have move between multiple software suites to explore their processed transcriptomics data.
-
-Here we emphasise the importance
-of actually looking at coverage traces to find biological events of interest, rather than only relying on numerical summaries of the data, and by facetting these traces over different combinations of the experimental design and with their biological context such as gene annotations.
-Of course, due to the sheer size of most reference genome annotaions 
-it would take an extremely long time for an analyst to look at every 
-single possible region where there is some interesting signal within the
-coverage trace. Because visualisation does not scale, we need a method to help
-us search the possible space of coverage traces and provide diagnostics for
-identifying traces with interesting biology. We have taken inspiration from
-visualisation literature; in particular the idea of scatterplot diagnostics ("scagnostics") for summarising the space of all possible 2D scatterplots
-to a small number of descriptors of each scatterplots properties such as 
-density or monotonicity [@Friedman2002-yx;@Wilkinson2005-xt].  Similarly,
-searching for unusual time series via estimating descriptors such as seasonality or autocorrelation, and visualising those descriptors instead [@Hyndman2015-sc].
+In this paper, we describe methodology for establishing data descriptors by turning coverage vectors into long form tidy data using **superintronic**
+and **plyranges**. We provide a workflow below
+using a zebrafish RNA-seq dataset for developing data descriptors to find IR like coverage traces within genes known to have minor class splicing events.
 
@@
 
-Considering IR also as another possible source of intron reads we propose a new method using our {\it superintronic} software to explore intron signal directly from aligned sequencing data with the assumption that most intron reads do not point to IR but pre-mRNA instead.
-{\it Superintronic} is an R package that is available to download and install at \href{https://github.com/sa-lee/superintronic}{https://github.com/sa-lee/superintronic}. 
-It extends the \textit{plyranges} Bioconductor package \citep{Lee2019plyranges} for genomics data analysis to develop a simple and modular interface for performing exploratory  genomics data analysis via coverage estimation.
-Each aspect of the \textit{superintronic} data analysis workflow as it has been applied for exploring intron signal is outlined in Figure~\ref{fig:workflows}b.
+**superintronic** is an R package available on Github at https://github.com/sa-lee/superintronic and used for estimating, representing,
+and visualising per-base coverage scores, that can then be 
+flexibly summarised over factors within an experimental design and collapsed
+over regions of the genome using our companion package **plyranges**. The
+aspects of this combined workflow are summarised in figure \@ref(fig:si-fig).
 
-Our software records the per base coverage over intron and exon regions of each gene, with the option of storing these per sample or summarised over variables in the experimental design such as by biological group or by RNA library preparation.
-Coverage scores are normalised using a log$_2$-transformation with an offset of 0.5 to get log-coverage values for which intron and exon summary statistics are constructed for each gene (described below).
-Within \textit{superintronic}, a suite of visualization tools to construct coverage plots for genes with intron and exon structures and scatter plots are provided.
+(ref:si-fig): An overview of the **superintronic** and **plyranges** workflow.
+
+```{r si-fig, fig.cap = "(ref:si-fig)", echo = FALSE}
+knitr::include_graphics(here::here("img", "olap-fig.png"))
+```
 
 ### Representation of coverage estimation
+
+The per base coverage score is estimated directly from one or more BAM files
+that represent the units within the experimental design,
+along with an optional experimental design table that is returns 
+a long-form tidy **GRanges** data structure. The coverage estimation is
+computed via the **Rsamtools** package, and
+users have the ability to estimate coverage in parallel and drop regions
+in the genome where there is no coverage [@rsamtools; @GenomicRanges]. This
+representation is tidy, since each row of the resulting **GRanges** data
+structure corresponds to position(s) within a given sample with a given 
+coverage score alongside any variables such as biological group. While the long-form representation repeats the same information
+for a sample within the design, the size of the resulting **GRanges** in
+memory can be compressed using run-length encoding for any categorical 
+variable. This representation allows us to easily transform the coverage scores
+and integrate annotations using the **plyranges** grammar, and visualise traces using **ggplot2** [@Wickham2016-gz].
 
@@
 
+Coverage scores are normalised using a log$_2$-transformation with an offset of 0.5 to get log-coverage values for which intron and exon summary statistics are constructed for each gene (described below).
+
 ## A workflow for uncovering intron retention
 
-Read coverage of intronic and exonic regions were calculated for poly(A) RNA and Total RNA HCC827 human cell lines using our \textit{superintronic} package, via the \textit{Rsamtools} package \cite{rsamtools,GenomicRanges}.
+Read coverage of intronic and exonic regions were calculated for poly(A) RNA and Total RNA HCC827 human cell lines using our \textit{superintronic} package,.
 Genes of interest were restricted to protein coding genes on reference chromosomes, and we removed any genes that overlapped another to simplify the analysis and reduce coverage ambiguity.  

### 2020-07-13 Rmd/04-intron.Rmd
finish section

@@
 ```{r si-fig, fig.cap = "(ref:si-fig)", echo = FALSE}
-knitr::include_graphics(here::here("img", "olap-fig.png"))
+knitr::include_graphics(here::here("img", "si-wflow.png"))
 ```
@@
 
-### Integration of external gene annotations
+### Integration of external annotations
+
+External reference annotations, perhaps transcripts or exons, 
+can be coereced to **GRanges** objects, are  incorporated into the coverage **GRanges** by taking the intersection of
+the annotation with the **GRanges** using an overlap intersect join from the **plyranges** software. The resulting intersection will now contain the
+per base coverage that are overlapped the genomic features in the annotation,
+along side any metadata about the features themselves. Since our main workflow
+interest is in discovering coverage traces with IR profiles, **superintronic**
+provides some syntatic sugar for unravelling gene annotations into their exonic
+and intronic parts, and intersecting them with a coverage **GRanges**.
 
@@
 
-Coverage scores are normalised using a log$_2$-transformation with an offset of 0.5 to get log-coverage values for which intron and exon summary statistics are constructed for each gene (described below).
+Once the coverage **GRanges** has reference genomic features, data 
+descriptors can be computed via collecting summary statistics across
+the factors of the experimental design and features of interest. This can be
+achieved using **plyranges** directly by first grouping across variables
+of interest, computing descriptors defined by **superintronic** and then
+pivoting the results into a wide form table for additional processing or
+visualisation. There are many descriptors defined by **superintronic**
+that are weighted statistics (as we have to account for the number of bases
+covered, or the width of the range) of the coverage score, such as the mean 
+and standard deviation. There are also descriptors that can be used to find
+the number of times the coverage trace is above a certain number of bases
+or score. To find coverage traces that have unusual descriptors, 
+the descriptors can be visualised directly as a scatterplot matrix. After that
+thresholds can be applied to filter the genomic features that had extreme
+descriptors on the coverage **GRanges**, and the traces can be visualised.
 
-## A workflow for uncovering intron retention
+## Results
+
+### A workflow for uncovering intron retention in zebrafish
 

### 2020-07-13 Rmd/04-intron.Rmd
partial spelling

@@
 estimated coverage from RNA-seq data. We establish a simple 
-framework, for aggregating accross experimental design, and annotated genomic regions to discover 'interesting' coverage trace plots. We highlight how this framework can be used to
+framework, for aggregating across experimental design, and annotated genomic regions to discover 'interesting' coverage trace plots. We highlight how this framework can be used to
 develop data descriptions that find putative genes with
@@
 In high-throughput sequencing data sets, coverage is the estimated number of
-reads that overlap a single position of the reference genome, and is important for assessing sequencing data quality and used in many different aspects of omic's analysis such peak-calling in ChIP-seq or varinat calling in DNA-seq [@Sims2014-qk].
+reads that overlap a single position of the reference genome, and is important for assessing sequencing data quality and used in many different aspects of omics analysis such peak-calling in ChIP-seq or variant calling in DNA-seq [@Sims2014-qk].
 Here we emphasise looking at coverage traces to find biological events of interest, rather than only relying on numerical summaries of the data.
-By facetting these traces over combinations of the experimental design and with their biological context such as gene annotations, we can gain an
+By faceting these traces over combinations of the experimental design and with their biological context such as gene annotations, we can gain an
 insight into biological signal under study. 
-Of course, due to the sheer size of most reference genome annotaions 
+Of course, due to the sheer size of most reference genome annotations 
 it would take an extremely long time for an analyst to look at every 
@@
 identifying traces with interesting biology. We have taken inspiration from
-visualisation literature; in particular the idea of scatterplot diagnostics ("scagnostics") for summarising the space of all possible 2D scatterplots
-to a small number of descriptors of each scatterplots properties such as 
+visualisation literature; in particular the idea of scatter plot diagnostics ("scagnostics") for summarising the space of all possible 2D scatter plots
+to a small number of descriptors of each scatter plots properties such as 
 density or monotonicity [@Friedman2002-yx;@Wilkinson2005-xt].  Similarly,
@@
 
-Although there are many flexible and powerful software tools, like BEDTools or DeepTools, for exploring and estimating coverage from common genomic data formats [@Quinlan2010-gc; @Ramirez2014-sh], it is advantegous to have tooling
+Although there are many flexible and powerful software tools, like BEDTools or DeepTools, for exploring and estimating coverage from common genomic data formats [@Quinlan2010-gc; @Ramirez2014-sh], it is advantageous to have tooling
 that is tightly coupled to a statistical computing language such as R.
 This allows interoperability between other software packages for data wrangling, visualisation and modelling within the ecosystem of the language that may not
-be possible with a single command line tool. There are also gains in reproduciblity as analysts do not have to move between multiple software suites to explore their processed transcriptomics data.
+be possible with a single command line tool. There are also gains in reproducibility as analysts do not have to move between multiple software suites to explore their processed transcriptomics data.
 
@@
 
-**superintronic** is an R package available on Github at https://github.com/sa-lee/superintronic and used for estimating, representing,
+**superintronic** is an R package available on GitHub at https://github.com/sa-lee/superintronic and used for estimating, representing,
 and visualising per-base coverage scores, that can then be 
@@
 External reference annotations, perhaps transcripts or exons, 
-can be coereced to **GRanges** objects, are  incorporated into the coverage **GRanges** by taking the intersection of
+can be coerced to **GRanges** objects, are  incorporated into the coverage **GRanges** by taking the intersection of
 the annotation with the **GRanges** using an overlap intersect join from the **plyranges** software. The resulting intersection will now contain the
@@
 interest is in discovering coverage traces with IR profiles, **superintronic**
-provides some syntatic sugar for unravelling gene annotations into their exonic
+provides some syntactic sugar for unravelling gene annotations into their exonic
 and intronic parts, and intersecting them with a coverage **GRanges**.
@@
 or score. To find coverage traces that have unusual descriptors, 
-the descriptors can be visualised directly as a scatterplot matrix. After that
+the descriptors can be visualised directly as a scatter plot matrix. After that
 thresholds can be applied to filter the genomic features that had extreme
@@
 Genes (a) PSMB7, (b) EIF2S3, and (c) ARGLU1 are highlighted out of 43 genes selected.
-Whilst the analysis was carried out on poly(A) RNA libraries, coverage is shown for both poly(A) RNA HCC827 (top) and Total RNA HCC827 (bottom) samples to ensure that results are not an artifact of RNA library preparation.
+Whilst the analysis was carried out on poly(A) RNA libraries, coverage is shown for both poly(A) RNA HCC827 (top) and Total RNA HCC827 (bottom) samples to ensure that results are not an artefact of RNA library preparation.
 Coverage is oriented from 5' to 3', with exon regions coloured green and intron regions coloured orange.
@@
 
-Using \textit{superintronic}, poly(A) RNA HCC827 cell lines were examined for genes with IR  after selecting genes in the hg38 reference that were protein coding, did not overlap any other gene and were placed on the main contigs -- a total of 6,606 genes.
+Using \textit{superintronic}, poly(A) RNA HCC827 cell lines were examined for genes with IR  after selecting genes in the hg38 reference that were protein coding, did not overlap any other gene and were placed on the main configs -- a total of 6,606 genes.
 These genes were then split into intron and exon regions and intersected with the coverage of each sample. 

### 2020-07-14 Rmd/04-intron.Rmd
add workflow image

@@
 (ref:si-fig): An overview of the **superintronic** and **plyranges** workflow.
+Coverage is estimated directly using the design matrix that contains a source column pointing to the locations of BAM files.
+The long-form representation is output as a **GRanges** object, and contains columsns that were part of the design.
+Additional annotations are add with join functions, here we show the particular case of expanding the coverage **GRanges** to include exonic and intronic parts of a gene.
+This object can be further anlaysed using **plyranges** and our data descriptors approach, and then descriptors can be visualised as a scatter plot matrix with the `GGally` and `ggplot2` packages [@ggally; @Wickham2016-gz].
+Coverage traces can be directly generated with **superintronic** and collapsed over parts of the design matrix to identify differences between groups.
 
 ```{r si-fig, fig.cap = "(ref:si-fig)", echo = FALSE}
-knitr::include_graphics(here::here("img", "si-wflow.png"))
+knitr::include_graphics(here::here("img", "si-workflowflow.png"))
 ```

### 2020-07-15 Rmd/04-intron.Rmd
add intro para

@@
 ### A workflow for uncovering intron retention in zebrafish
+
+In the study of gene regulation, there is much interest in uncovering the
+effects of aberrant minor class splicing (called U12 splicing) on the transcriptome.
+Minor class splicing is a regulatory process where a class of 
+introns (called U12 introns in this case but there exist other classes in eukaryotic organisms) are removed from pre-mRNA prior to gene expression.
+The removal of these introns is catalysed by small ribonucleoproteins (snRNPs)
+which identify key motifs and branch sites in the intron to begin splicing.
+Here we describe an exploratory workflow for finding coverage traces
+with evidence for intron retention using a knockout experiment in zebrafish.
 

### 2020-07-15 Rmd/04-intron.Rmd
add results section

--- 2020-07-14 Rmd/04-intron.Rmd
@@
 
-```{r si-fig, fig.cap = "(ref:si-fig)", echo = FALSE}
+```{r si-fig, fig.cap = "(ref:si-fig)", out.width = "80%", echo = FALSE}
 knitr::include_graphics(here::here("img", "si-workflowflow.png"))
@@
 
-## Results
+## A workflow for uncovering intron retention in  a zebrafish experiment
 
-### A workflow for uncovering intron retention in zebrafish
+In the study of gene regulation, there is much interest in uncovering the
+effects of aberrant minor class splicing (called U12 splicing) on the transcriptome.
+Minor class splicing is a regulatory process where a class of 
+introns (called U12 introns in this case but there exist other classes in eukaryotic organisms such as U2 introns) are removed from pre-mRNA prior to gene expression.
+The removal of these introns is catalysed by small ribonucleoproteins (snRNPs)
+which identify key motifs and branch sites in the intron to begin splicing.
+Here we describe an exploratory workflow for finding coverage traces
+with evidence for intron retention using RNA-seq data from a knockout experiment in zebrafish obtained from the Heath Lab at WEHI.
 
-Read coverage of intronic and exonic regions were calculated for poly(A) RNA and Total RNA HCC827 human cell lines using our \textit{superintronic} package,.
-Genes of interest were restricted to protein coding genes on reference chromosomes, and we removed any genes that overlapped another to simplify the analysis and reduce coverage ambiguity.  
-Genes were then further filtered if they were not expressed in the poly(A) RNA protocol (requiring at least 3 reads overlapping intronic and exonic regions). 
-A total of 3,262 genes were examined and categorised as short, regular or long (roughly 1,087 genes in each category) by splitting the length of each gene into three bins by tertiles.
+The data consist of 11 zebrafish samples from single-end polyA enriched RNA-seq libraries pooled from zebrafish larvae.
+The experimental factors looked at combinations of genotype (whether the gene rnpc3 has been knocked out or not) and line (whether the zebrafish larvae have the caliban *cal* or mutant caliban *zm-cal* phenotype).
+Within each combination there are three biological replicates, except for the combination *zm-cal* and wildtype rnpc3 which
+had two replicates.
 
-Using associated BAM files and GENCODE v27 annotation GTF, \textit{superintronic} summarised the number of bases covered at a given coverage score for each gene and sample.
-For exonic regions and intronic regions, coverage scores were transformed to log$_2$-scale using an offset of 0.5, and then normalised by dividing by each gene and each sample's maximum log-coverage score.
-Normalised log-coverage scores, or {\it relative log-coverage} (relative to each gene's maximum coverage), were divided into 20 windows along the length of each gene using the \textit{GenomicRanges} package \cite{GenomicRanges}.
-To summarise coverage patterns across genes, \textit{plyranges} \cite{Lee2019plyranges} was applied to relative log-coverage scores by intersecting it with the positional windows.
-Each window's mean coverage score (mean relative log-coverage) was calculated for each gene.
-Based on the position of the windows, they were further summarised across genes by taking its mean. % (mean of mean relative log-coverage).  
-The summary values were calculated separately for genes categorised as short, regular and long to represent general coverage trends along the gene body.
+FASTQ files were aligned to the GRCz11 reference genome using subjunc with
+the default parameters called from Rsubread to produce BAM files for each
+sample. The coverage was then estimated directly from the set of BAM files
+using **superintronic** into the long form *GRanges* representation we described above.
 
-\textit{Human cell lines} of lung adenocarcinoma HCC827 and NCI-H1975 were cultured on three
-separate occasions by Holik {\it et al.} (2017) \cite{Holik:NAR:2017} giving three pseudo biological replicates. 
-RNA was extracted from each pseudo biological replicate and split into two and prepared as poly(A) RNA and Total RNA libraries.
-Raw sequencing reads were downloaded from the Gene Expression Omnibus (GEO) \citep{GEO:2013} under accession number GSE64098.
-12 libraries were examined for this dataset.
+The gene annotation files were
+obtained as GFF files from RefSeq and used that to construct the exonic
+and intronic parts of each gene as **GRanges** object using *superintronic*. 
+We further filtered genes that had a single exon or genes that were overlapping
+others in the annotation and that were not on the main contigs of the reference genome (i.e.
+exlcuding mitochondria) to simplify our analysis and reduce any coverage ambiguity.
+This left 18,270 genes available for
+computing data descriptors on in order to detecting IR-like coverage traces.
+Accross each combination of genotype and line, we first log$_2$ transformed the coverage score with an offset of one, and then intersected the coverage *GRanges* with exonic and intronic features of each gene.
+For each gene,
+the weighted mean and standard deviation of the log-transformed coverage score were computed over all intron and exon parts within the groupings of 
+genotype and line. We also computed the bases above descriptor for intron
+features. It refers to the total number of bases within an intron that has a score above the overall average exon coverage score. The scatter plot 
+matrix view of these descriptors is shown in figure \@ref(fig:si-splom)
+for a single biological group in the experiment. Using these views and
+by summarising over the data descriptors, we came up with thresholds
+for finding genes that have IR-like traces within each biological group.
 
-FASTQ files containing raw sequencing reads were aligned to the human \textit{hg38} or mouse \textit{mm10}
-% or {\color{blue}{zebrafish \textit{GRCz11}}} 
-genome using \textit{subjunc} \citep{Liao:NAR:2013} with default parameters in the \textit{Rsubread} software package \citep{Rsubread:2019}. 
-GENCODE's main \textit{Comprehensive gene annotation} file in GTF format was downloaded from {https://www.gencodegenes.org} for human (Release 27) and mouse (Release M12). 
-Using ``gene types'' (rather than ``transcript types'') from Gencode, the annotation files were simplified by taking the union of two or more overlapping exons from transcripts of the same gene.
-The adjustment provides a simplification of genomic positions on each strand, such that each position is classified as belonging to ``exon'', ``intron'' or otherwise outside of an annotated gene.
-Three resultant annotation files were saved in standard annotation format (SAF) -- exon annotation, intron annotation (region between exons), and genebody annotation (region spanning first to last exon).
-Our Supplementary Materials available at \href{https://github.com/charitylaw/Intron-reads}{https://github.com/charitylaw/Intron-reads} contain the scripts to process annotation files, together with other data analyses and supplementary figures.
+(ref:si-splom): A hexbin scatter plot matrix of the data descriptors estimated for the *cal* rnpc3 knock out zebrafish line.
+To identify coverage traces with IR like events, we want a set of descriptors that will find genes with the following characteristics: the gene is "expressed" that has a large number of intron bases relative to the coverage of other intron features, and has  relatively stable coverage within exon features.
+To do this we looked at the  descriptor, as well as computing the the mean and standard deviation of both exon and intron features.
 
-\subsection{\textit{Superintronic} finds genes with IR-like coverage profiles in human cell lines}
+```{r si-splom, fig.cap = "(ref:si-splom)", out.width="80%"}
+knitr::include_graphics(here::here("img", "Cal-Rnpc3_KO-pairs-plot.png"))
+```
 
-\begin{figure*}[!tpb]
-\centerline{\includegraphics[width=\textwidth]{coverage-si-hits.pdf}}
-\caption{{\it Superintronic} selects genes with IR-like coverage profiles in poly(A) RNA HCC827 cell lines.
-Genes (a) PSMB7, (b) EIF2S3, and (c) ARGLU1 are highlighted out of 43 genes selected.
-Whilst the analysis was carried out on poly(A) RNA libraries, coverage is shown for both poly(A) RNA HCC827 (top) and Total RNA HCC827 (bottom) samples to ensure that results are not an artefact of RNA library preparation.
-Coverage is oriented from 5' to 3', with exon regions coloured green and intron regions coloured orange.
-}\label{fig:ir}
-\end{figure*}
+We decided to select genes with the following thresholds:  genes have an average exon log-coverage greater than the mean of average exon log-coverage values accross all genes, have an average intron log-coverage greater than the average standard deviation of exon log-coverage values over all genes, and
+the standard deviation of intron log-coverage values is twice the standard
+deviation of exon log-coverage values. That is, we are selecting genes that
+are expressed but  have large average intron expression that is more variable than the gene's exon expression. This results in a total of 86 genes selected
+to link back to their underlying coverage traces, with the overlaps shown
+in the UpSet plot in figure \@ref(fig:si-upset).
+The procedure produces gene coverage traces with known minor class splice sites affected by the knockout procedure such as ccdc43 and nat15 (figures \@ref(fig:si-cvg-ccdc43) and \@ref(fig:si-cvg-nat15)), as well as some that appear to have U12 intron retention like events such as mapk3 or tspan31 (figures \@ref(fig:si-cvg-mapk3) and \@ref(fig:si-cvg-tspan31)) that affect other parts of the gene.
 
-Using \textit{superintronic}, poly(A) RNA HCC827 cell lines were examined for genes with IR  after selecting genes in the hg38 reference that were protein coding, did not overlap any other gene and were placed on the main configs -- a total of 6,606 genes.
-These genes were then split into intron and exon regions and intersected with the coverage of each sample. 
-Per gene intron and exon summary statistics, mean and standard deviation, were computed on log-coverage values. 
-We selected genes enriched for IR-like coverage profiles by looking for ``expressed'' genes, where for a substantial number of intron bases its coverage is much higher than other intron features within the same gene while having similar expression levels to the exon features.
-To do this we used the following thresholds -- genes had an average exon log-coverage of greater than 2 (corresponding to the mean of average exon log-coverage values across all genes), a standard deviation of intron log-coverage greater than 1.5 (corresponding to the mean of intron standard deviation values across all genes), and genes with a large number of intron bases with log-coverage greater than 2 (top 1\% of genes).
-The thresholds were chosen after examining distributions of the summary statistics (Supplementary Figure 4).
-Forty-three genes met these criteria, where a manual check of coverage profiles revealed that 36 genes indeed appear IR-like (see `Analyses' page in Supplementary Materials).
-We highlight 3 of these genes in Figure~\ref{fig:ir}.
-The coverage of remaining 7 genes appear to be more pre-mRNA-like, with large 
-variation in intron coverage, where at its peak it is expressed similarly to exon features.
+(ref:si-upset): Gene overlaps found between each combination of genotype and line using the thresholds defined in the text.
+Our procedure mostly finds genes with IR like profiles across all groups (28 shared between all four) or that is unique to a single group, since we do not consider looking at differential IR and run our thresholding seperately for each group of replicates.
+The rnpc3 knockout lines
 
-We use poly(A) RNA samples in our analysis for consistency with previous studies on IR.
-In theory, Total RNA samples may be more appropriate for this exercise since it is less biased towards the 3' end, where high 3' intron coverage as a result of 3' bias in poly(A) RNA libraries can be mistaken as a 3' retained intron.
-For this reason, we review both the coverage of poly(A) RNA and Total RNA samples in  our \textcolor{red}{S}upplementary \textcolor{red}{M}aterials for all selected genes to ensure that retained introns found in poly(A) RNA samples are not an artifact of RNA library preparation.
+```{r si-upset, echo=FALSE, fig.cap = "(ref:si-upset)", out.width="80%"}
+knitr::include_graphics(here::here("img", "si-upset.pdf"))
+```
 
-If differentially retained introns are of interest, one could simply run \textit{superintronic} on two conditions separately and compare lists of IR-like genes between groups.
-We found that differences between the lists of IR-like genes were concordant with \textit{index} results for genes DE by intron counts, as expected (see `Analyses' page in Supplementary Materials).
-For example, 14 genes were uniquely selected as IR-like in Total RNA HCC827 cell line when compared to Total RNA NCI-H1975 cell line using \textit{superintronic}, where 13 of those genes were also found to be DE in intron counts by \textit{index}, and in the expected direction.
-Similarly, 19 genes were uniquely detected as IR-like in Total RNA NCI-H1975 by \textit{superintronic}, 16 of which were also detected as DE in intron counts by \textit{index}.
-Additionally, 12 genes which were found to be IR-like in both cell lines appear to have varying intron expression according to \textit{index} since it is found to be DE in intron counts.
+(ref:si-cvg-ccdc43): Coverage traces facetted by biological group for the zebrafish knockout experiment, oriented from the 5' to 3' end of the gene.
+Orange areas refer to intronic parts of the gene, while dark green areas refer to the exonic part.
+The ccdc43 gene is known to have enhanced U12 intron retention in the caliban phenotype and increase retention when rnpc3 is knocked out as can be seen directly from the intron located at around 31.603Mb.
 
-\textcolor{red}{
-Whilst genes that are uniquely IR-like were concordant with directional changes in intron counts, we did not find that IR-like genes played a large part in \textit{index} gene categorisation.
-This was mostly expected since thousands of genes were detected as significantly DE by \textit{index} and a relatively small number of genes were detected as IR-like by \textit{superintronic}.
-Specifically, IR-like genes did not overlap with one particular \textit{index} category (see `Analyses' page in Supplementary Materials).
-Of genes that are classified as intron+ or intron-, only 3 out of 1,006 genes were also detected by \textit{superintronic} as IR-like.
-In the ``mixed'' category, 2 out of 54 genes were also detected as IR-like.
-}
+```{r si-cvg-ccdc43, echo=FALSE, fig.cap = "(ref:si-cvg-ccdc43)", out.width="50%"}
+knitr::include_graphics(here::here("img", "ccdc43.pdf"))
+```
+
+(ref:si-cvg-nat15): The gene nat15 exhibits another example of U12 intron retention (located at around 27.93Mb) in the rnpc3 knockout groups.
+
+```{r si-cvg-nat15,  echo=FALSE, fig.cap = "(ref:si-cvg-nat15)", out.width="50%"}
+knitr::include_graphics(here::here("img", "nat15.pdf"))
+```
+
+(ref:si-cvg-mapk3): The gene mapk3 appears to have intron retention close to the start of the gene that have different impacts downstream, which result in different IR profiles between the knockout and wildtype groups.
+
+```{r si-cvg-mapk3,  echo=FALSE, fig.cap = "(ref:si-cvg-mapk3)", out.width="50%"}
+knitr::include_graphics(here::here("img", "mapk3.pdf"))
+```
+
+(ref:si-cvg-tspan31): The gene tspan31 has a potential cryptic splice site
+within the intron retained in the knockout groups.
+
+```{r si-cvg-tspan31, echo=FALSE, fig.cap = "(ref:si-cvg-tspan31)", out.width="50%"}
+knitr::include_graphics(here::here("img", "tspan31.pdf"))
+```
 
@@
 
-## Conclusion
+## Acknowledgements
 
-<!-- ## Data -->
-
-<!-- We will use data from 12 samples from the human cell line mixture RNA-seq -->
-<!-- data. There are three biological replicates across two library preparation kits, -->
-<!-- a polyA kit and a total RNA kit. The BAM files have been previously aligned -->
-<!-- with subread using the hg20 reference genome.  -->
-
-<!-- ```{r prepare-design, echo= FALSE} -->
-<!-- design <- readRDS(here("data", "superintronic", "design.rds")) -->
-<!-- ``` -->
-
-<!-- ## `superintronic` analysis -->
-
-<!-- A superintronic analysis consists of four steps: -->
-<!-- 1. Preparing the annotation and extract exonic/intronic parts -->
-<!-- 2. Computing coverage over regions of interest -->
-<!-- 3. Summarising coverage over regions of interest -->
-<!-- 4. Visualising the results -->
-
-<!-- Steps 1 and 2. are performed in the `scripts/run_coverage.R` -->
-
-<!-- ### Preparing GFF/GTF files -->
-
-<!-- ```{r gff} -->
-<!-- library(superintronic) -->
-<!-- parts <- readRDS(here::here("data", "complete-annotation.rds")) -->
-<!-- parts -->
-<!-- ``` -->
-
-<!-- These can be interrogated as desired, but for our purposes we will keep -->
-<!-- genes that are protein coding, on main chromosomes but not mitochondrial  -->
-<!-- genome, and do not overlap any other genes and have more than one exon. -->
-
-<!-- ```{r default-filter, echo = FALSE} -->
-<!-- parts_sub <- readRDS(here::here("data", "filtered-annotation.rds")) -->
-<!-- parts_sub -->
-<!-- ``` -->
-
-<!-- ## Coverage over intronic/exonic regions -->
-
-<!-- Now all the ingredients are in place to merge the coverage scores -->
-<!-- to our prepared annotation with `join_parts()`. This again -->
-<!-- returns a GRanges object, restricted to the intersection of the  -->
-<!-- coverage ranges with intron/exon ranges. Additional columns are added -->
-<!-- corresponding to the properties of the intron/exon. -->
-
-<!-- ```{r, load-features, echo = FALSE } -->
-<!-- cvg_over_features <- readRDS(here::here("data", "parts-coverage.rds")) -->
-<!-- ``` -->
-
-<!-- Now we will compute summaries for each cell line and kit to detect IR  -->
-<!-- genes. This consists of three steps -->
-
-<!-- 1. Computing the average log-coverage over all exon/intron features: -->
-
-<!-- ```{r rangenostics-01} -->
-<!-- # compute intron/exon features -->
-<!-- cvg_over_features <- cvg_over_features %>%  -->
-<!--   mutate(log_score = log2(score + 0.5))   -->
-
-<!-- cvg_by_features <- cvg_over_features %>%  -->
-<!--   group_by(Kit, CellLine, gene_id, feature_type) -->
-<!-- E_vals <- cvg_by_features %>%  -->
-<!--   summarise(mn = Hmisc::wtd.mean(log_score, width), -->
-<!--             sd = sqrt(Hmisc::wtd.var(log_score, width)), -->
-<!--             raw_mn = Hmisc::wtd.mean(score, width), -->
-<!--             score = score, -->
-<!--             n_bases = width, -->
-<!--             start = min(start), -->
-<!--             end = max(end), -->
-<!--             seqnames = unlist(unique(seqnames)), -->
-<!--             strand = unlist(unique(feature_strand))) %>%  -->
-<!--   makeGRangesFromDataFrame(keep.extra.columns = TRUE) -->
-<!-- ``` -->
-
-<!-- 2. For exon features computing the mean and variance of average log-coverage  -->
-<!--    over all genes (lumpiness). -->
-
-<!-- ```{r rangenostics-02} -->
-<!-- lumpy <- E_vals %>%  -->
-<!--   filter(sd > 0, feature_type == "exon") %>%  -->
-<!--   group_by(Kit, CellLine) %>%  -->
-<!--   summarise( -->
-<!--     E_mu = mean(mn), -->
-<!--     E_sd = mean(sd), -->
-<!--     E_raw = mean(raw_mn) -->
-<!-- ) -->
-<!-- lumpy  -->
-<!-- mcols(E_vals) <- cbind(mcols(E_vals), -->
-<!--                        lumpy[match(paste0(E_vals$Kit, E_vals$CellLine),  -->
-<!--                                    paste0(lumpy$Kit, lumpy$CellLine)), -c(1,2)]) -->
-<!-- E_vals <- E_vals %>%  -->
-<!--   mutate( -->
-<!--     n_bases_E_raw = sum(n_bases[score > E_raw]), -->
-<!--   )  -->
-<!-- ``` -->
-
-<!-- 3. Compute the average number of bases with a  -->
-<!--    log-coverage score above the E values defined above. We have also computed -->
-<!--    a smooth between over the intron/exon feature standard deviation against -->
-<!--    the average for visualisation purposes. -->
-
-<!-- ```{r rangenostics-03} -->
-<!-- rango <- E_vals %>%  -->
-<!--   plyranges::select(-n_bases, -score, -seqnames, -start, -end, -width, -strand,  -->
-<!--          .drop_ranges = TRUE) %>%  -->
-<!--   dplyr::as_tibble() %>%  -->
-<!--   group_by(Kit, CellLine, feature_type) %>%  -->
-<!--   tidyr::nest() %>%  -->
-<!--   mutate( -->
-<!--     smooth = lapply(data, function(x) { -->
-<!--       mgcv::gam(sd ~ s(mn), data = x) -->
-<!--       }), -->
-<!--     augment = lapply(smooth, broom::augment), -->
-<!--   ) %>%  -->
-<!--   tidyr::unnest(data, augment) %>%  -->
-<!--   dplyr::select(-dplyr::ends_with("1"), -.sigma) -->
-<!-- ``` -->
-
-<!-- 4. Plot the results. -->
-
-<!-- We get very similar results to a `voom` plot in a standard limma analysis. We -->
-<!-- have also overlaid on the E_mu and E_sigma values as potential cut-offs for -->
-<!-- each feature.  -->
-
-<!-- ```{r voom-like-plot} -->
-<!-- # rangenostics feature_average vs feature_sd coloured by n of bases above -->
-<!-- # average raw exon coverage -->
-<!-- library(ggplot2) -->
-<!-- voom_plot <- function(.x, .y) { -->
-<!--   p <- ggplot(data = .x, aes(x = mn, y = sd)) + -->
-<!--     geom_point() + -->
-<!--     geom_line(aes(y = .fitted), colour = "blue") + -->
-<!--     geom_vline(aes(xintercept = E_mu),  -->
-<!--                data  = dplyr::distinct(.x, feature_type, E_mu)) + -->
-<!--     geom_hline(aes(yintercept = E_sd), -->
-<!--                data = dplyr::distinct(.x, feature_type, E_sd)) + -->
-<!--     facet_wrap(~feature_type) + -->
-<!--     labs( -->
-<!--          subtitle = paste(.y$Kit, .y$CellLine), -->
-<!--          x = "mean log-coverage", -->
-<!--          y = "sd log-covreage" -->
-<!--     ) -->
-<!--   fname <- here::here("img",  -->
-<!--                       paste0(.y$Kit, "-", .y$CellLine,  -->
-<!--                              "voom-like-plot.png") -->
-<!--   ) -->
-<!--   ggsave(fname, p) -->
-<!-- } -->
-<!-- rango %>%  -->
-<!--   group_by(Kit, CellLine) %>%  -->
-<!--   dplyr::group_walk(voom_plot) -->
-<!-- ``` -->
-
-<!-- We also plot pairs of summary values for each feature in a scatter plot matrix -->
-<!-- (using a hexbin scatter plot). Again we get similar results to the plots -->
-<!-- produced by a DE analysis from InDeX. -->
-
-<!-- ```{r rangenostics-pairs} -->
-<!-- all_features <- tidyr::gather(rango, "key", "value",  -->
-<!--                               -Kit, -CellLine, -gene_id, -feature_type,  -->
-<!--                               -E_mu, -E_sd, -E_raw) %>%  -->
-<!--   mutate(var = paste0(feature_type, "_", key)) %>%  -->
-<!--   plyranges::select(-feature_type, -key) %>%  -->
-<!--   arrange(gene_id) %>%  -->
-<!--   tidyr::spread(var, value) -->
-<!-- pairs_plot <- function(.x, .y) { -->
-<!--   sub <- plyranges::select(.x,  -->
-<!--                            exon_mn,  -->
-<!--                            intron_mn,  -->
-<!--                            intron_sd,  -->
-<!--                            bases_above = intron_n_bases_E_raw) -->
-<!--   fname <- here::here("img",  -->
-<!--                       paste0(.y$Kit, "-", .y$CellLine,  -->
-<!--                              "-pairs-plot.png") -->
-<!--   ) -->
-
-<!--   p <- GGally::ggpairs(sub, columns = 1:4, lower = list(continuous = hex)) -->
-<!--   ggsave(fname, p) -->
-<!-- } -->
-<!-- all_features %>%  -->
-<!--   group_by(Kit, CellLine) %>%  -->
-<!--   dplyr::group_walk(pairs_plot) -->
-
-<!-- ``` -->
-
-<!-- 5. Decide on cut-off over each feature. It appears that choosing genes -->
-<!-- with high intron feature standard deviations, and with number of bases above -->
-<!-- E_mu in the 99th percentile discriminates intron retention features well. -->
-
-<!-- ```{r rangenositcs-hits} -->
-<!-- hits <- all_features %>%  -->
-<!--   group_by(Kit, CellLine) %>%  -->
-<!--   filter(exon_mn > E_mu,  -->
-<!--          intron_sd > E_sd,  -->
-<!--          intron_n_bases_E_raw > quantile(intron_n_bases_E_raw, 0.99))   -->
-
-<!-- dplyr::count(hits) -->
-<!-- ``` -->
-
-<!-- 6. We can also look at overlaps between sets using an UpSet plot: -->
-
-<!-- ```{r, echo  = FALSE} -->
-<!-- sets <- hits %>%  -->
-<!--   dplyr::distinct(gene_id) %>% -->
-<!--   arrange(gene_id) %>%  -->
-<!--   mutate(value = 1L, var = paste(CellLine, Kit)) %>%  -->
-<!--   ungroup() %>%   -->
-<!--   plyranges::select(gene_id, var, value) %>%  -->
-<!--   tidyr::spread(var, value) %>%  -->
-<!--   dplyr::mutate_at(2:5, .funs = ~ dplyr::if_else(is.na(.), 0L, .))  -->
-<!-- sets <- as.data.frame(sets) -->
-<!-- library(UpSetR) -->
-<!-- pdf(here::here("img", "olaps-superintronic-hits.pdf")) -->
-<!-- upset(sets) -->
-<!-- dev.off() -->
-<!-- ``` -->
+We would like to thank Alex Garnham, Stephen Mieruszynski and Joan Heath
+for providing the zebrafish data and helping interpret the results from the workflow.

### 2020-07-15 Rmd/04-intron.Rmd
first draft chapter 4

@@
-```{r setup}
-library(superintronic)
-
-```
-
 # Exploratory coverage analysis with superintronic and plyranges {#ch:intron}
@@
 
-Although there are many flexible and powerful software tools, like BEDTools or DeepTools, for exploring and estimating coverage from common genomic data formats [@Quinlan2010-gc; @Ramirez2014-sh], it is advantageous to have tooling
+Although there are many flexible and powerful software tools, like BEDtools or DeepTools, for exploring and estimating coverage from common genomic data formats [@Quinlan2010-gc; @Ramirez2014-sh], it is advantageous to have tooling
 that is tightly coupled to a statistical computing language such as R.
@@
 Coverage is estimated directly using the design matrix that contains a source column pointing to the locations of BAM files.
-The long-form representation is output as a **GRanges** object, and contains columsns that were part of the design.
+The long-form representation is output as a **GRanges** object, and contains columns that were part of the design.
 Additional annotations are add with join functions, here we show the particular case of expanding the coverage **GRanges** to include exonic and intronic parts of a gene.
-This object can be further anlaysed using **plyranges** and our data descriptors approach, and then descriptors can be visualised as a scatter plot matrix with the `GGally` and `ggplot2` packages [@ggally; @Wickham2016-gz].
+This object can be further analysed using **plyranges** and our data descriptors approach, and then descriptors can be visualised as a scatter plot matrix with the `GGally` and `ggplot2` packages [@ggally; @Wickham2016-gz].
 Coverage traces can be directly generated with **superintronic** and collapsed over parts of the design matrix to identify differences between groups.
@@
 ```{r si-fig, fig.cap = "(ref:si-fig)", out.width = "80%", echo = FALSE}
-knitr::include_graphics(here::here("img", "si-workflowflow.png"))
+knitr::include_graphics(here::here("img", "si-workflow.png"))
 ```
@@
 The removal of these introns is catalysed by small ribonucleoproteins (snRNPs)
-which identify key motifs and branch sites in the intron to begin splicing.
+which identify key motifs and branch sites in the intron to begin splicing [@Markmiller2014-kg].
 Here we describe an exploratory workflow for finding coverage traces
@@
 The experimental factors looked at combinations of genotype (whether the gene rnpc3 has been knocked out or not) and line (whether the zebrafish larvae have the caliban *cal* or mutant caliban *zm-cal* phenotype).
-Within each combination there are three biological replicates, except for the combination *zm-cal* and wildtype rnpc3 which
+Within each combination there are three biological replicates, except for the combination *zm-cal* and wild-type rnpc3 which
 had two replicates.
@@
 others in the annotation and that were not on the main contigs of the reference genome (i.e.
-exlcuding mitochondria) to simplify our analysis and reduce any coverage ambiguity.
+excluding mitochondria) to simplify our analysis and reduce any coverage ambiguity.
 This left 18,270 genes available for
 computing data descriptors on in order to detecting IR-like coverage traces.
-Accross each combination of genotype and line, we first log$_2$ transformed the coverage score with an offset of one, and then intersected the coverage *GRanges* with exonic and intronic features of each gene.
+Across each combination of genotype and line, we first log$_2$ transformed the coverage score with an offset of one, and then intersected the coverage *GRanges* with exonic and intronic features of each gene.
 For each gene,
@@
 
-We decided to select genes with the following thresholds:  genes have an average exon log-coverage greater than the mean of average exon log-coverage values accross all genes, have an average intron log-coverage greater than the average standard deviation of exon log-coverage values over all genes, and
+We decided to select genes with the following thresholds:  genes have an average exon log-coverage greater than the mean of average exon log-coverage values across all genes, have an average intron log-coverage greater than the average standard deviation of exon log-coverage values over all genes, and
 the standard deviation of intron log-coverage values is twice the standard
@@
 to link back to their underlying coverage traces, with the overlaps shown
-in the UpSet plot in figure \@ref(fig:si-upset).
+in the UpSet plot in figure \@ref(fig:si-upset) [@Lex2014-bh].
 The procedure produces gene coverage traces with known minor class splice sites affected by the knockout procedure such as ccdc43 and nat15 (figures \@ref(fig:si-cvg-ccdc43) and \@ref(fig:si-cvg-nat15)), as well as some that appear to have U12 intron retention like events such as mapk3 or tspan31 (figures \@ref(fig:si-cvg-mapk3) and \@ref(fig:si-cvg-tspan31)) that affect other parts of the gene.
@@
 (ref:si-upset): Gene overlaps found between each combination of genotype and line using the thresholds defined in the text.
-Our procedure mostly finds genes with IR like profiles across all groups (28 shared between all four) or that is unique to a single group, since we do not consider looking at differential IR and run our thresholding seperately for each group of replicates.
+Our procedure mostly finds genes with IR like profiles across all groups (28 shared between all four) or that is unique to a single group, since we do not consider looking at differential IR and run our thresholding separately for each group of replicates.
 The rnpc3 knockout lines
@@
 
-(ref:si-cvg-ccdc43): Coverage traces facetted by biological group for the zebrafish knockout experiment, oriented from the 5' to 3' end of the gene.
+(ref:si-cvg-ccdc43): Coverage traces faceted by biological group for the zebrafish knockout experiment, oriented from the 5' to 3' end of the gene.
 Orange areas refer to intronic parts of the gene, while dark green areas refer to the exonic part.
@@
 
-(ref:si-cvg-mapk3): The gene mapk3 appears to have intron retention close to the start of the gene that have different impacts downstream, which result in different IR profiles between the knockout and wildtype groups.
+(ref:si-cvg-mapk3): The gene mapk3 appears to have intron retention close to the start of the gene that have different impacts downstream, which result in different IR profiles between the knockout and wild-type groups.
 
@@
 
+We have shown how to coverage can be represented in the tidy data framework
+and integrated with experimental metadata and reference annotations. This
+framework allowed us to build data descriptions that are simple aggregations of
+various aspects of genomic features over factors within a designed experiment
+and link those descriptions to their underlying coverage traces.
+
+Our zebrafish workflow shows that our approach using **superintronic** 
+and **plyranges** is able to uncover interesting biological signals in a
+purely data-driven manner. We did not include additional information
+that could have been useful when deriving our selected genes, such as
+sequence motifs for U12 class of introns, or exploit the experimental
+design to find differential IR like profiles. However, if that was of interest,
+one could look at the overlaps, like we did in figure \@ref(fig:si-upset),
+or combine our data descriptors with external estimates using **limma** 
+[@Ritchie2015], like our proposed index method in @Lee2019-mf.
+
+Although the example we have explored has related to finding coverage
+traces with IR-like events, the workflow of building and then visualising
+data descriptors could be generalised to other types of omics analyses,
+and to use more sophisticated methods for identifying 
+thresholds of 'interesting' traces. Our approach would also greatly benefit
+from interactive graphics that dynamically link say a gene description to
+it's underlying coverage trace, for rapid exploration. This is left for 
+future work.
+
 ## Acknowledgements
 
-We would like to thank Alex Garnham, Stephen Mieruszynski and Joan Heath
+We would like to thank Alexandra Garnham, Stephen Mieruszynski and Joan Heath
 for providing the zebrafish data and helping interpret the results from the workflow.

### 2020-07-15 Rmd/04-intron.Rmd
add extra references

@@
 the default parameters called from Rsubread to produce BAM files for each
-sample. The coverage was then estimated directly from the set of BAM files
+sample [@Liao2013; @Liao2019-qn]. The coverage was then estimated directly from the set of BAM files
 using **superintronic** into the long form *GRanges* representation we described above.

### 2020-07-15 Rmd/01-intro.Rmd
spelling

@@
 according to the aforementioned workflow steps. To begin we discuss wrangling
-biological data using a coherent represntation and programming interface; 
+biological data using a coherent representation and programming interface; 
 section \@ref(sec:tidygranges)
@@
 a user interaction framework for identifying distortions. We present several
-case studies using data sets from both single-cell transciptomics and 
+case studies using data sets from both single-cell transcriptomics and 
 machine learning  for using our approach to diagnose the quality of 

### 2020-07-15 Rmd/06-conclusion.Rmd
spelling

@@
 the rise of popular non-linear embedding methods. 
-To finish, I will dicuss the implications of this work and provide 
+To finish, I will discuss the implications of this work and provide 
 suggestions for further research.
@@
 which has led to several developments currently in place that are exploring
-different approaches for fluent interfaces for other types of 'omics data.
+different approaches for fluent interfaces for other types of omics data.
 
@@
 **plyranges** is lack of scalability and computational speed for data sets
-that do not fit in memory. We attempted several techniques for perfoming 
+that do not fit in memory. We attempted several techniques for performing 
 delayed operations over range-based data, however a more general approach
 that allows for data stored on the cloud or in scientific data formats
-like HDF5 that leverages existing Bioconductor frameworks would be useful.
+like HDF5 that leverage existing Bioconductor frameworks would be useful.
 We showed in chapters \@ref(ch:fluentGenomics) ad \@ref(ch:intron) that an a
-nalyst is able to do some very complex data transformations and resampling 
+analyst is able to do some very complex data transformations and re-sampling 
 procedures via casting results into *GRanges* object. However, it is unclear 

### 2020-07-16 Rmd/04-intron.Rmd
fix reference and delete old stuff
update bib

--- 2020-07-15 Rmd/04-intron.Rmd
@@
 
-Although there are many flexible and powerful software tools, like BEDtools or DeepTools, for exploring and estimating coverage from common genomic data formats [@Quinlan2010-gc; @Ramirez2014-sh], it is advantageous to have tooling
+Although there are many flexible and powerful software tools, like BEDtools or deepTools, for exploring and estimating coverage from common genomic data formats [@Quinlan2010-gc; @Ramirez2014-sh], it is advantageous to have tooling
 that is tightly coupled to a statistical computing language such as R.
@@
 the default parameters called from Rsubread to produce BAM files for each
-sample. The coverage was then estimated directly from the set of BAM files
+sample [@Liao2013-eb; @Liao2019-qn]. The coverage was then estimated directly from the set of BAM files
 using **superintronic** into the long form *GRanges* representation we described above.
@@
 ```{r si-upset, echo=FALSE, fig.cap = "(ref:si-upset)", out.width="80%"}
-knitr::include_graphics(here::here("img", "si-upset.pdf"))
+knitr::include_graphics(here::here("img", "si-olaps.png"))
 ```
@@
 or combine our data descriptors with external estimates using **limma** 
-[@Ritchie2015], like our proposed index method in @Lee2019-mf.
+[@Ritchie2015-mj], like our proposed index method in @Lee2019-mf.
 

### 2020-07-16 Rmd/00-e-preface.Rmd
add preface and rejig intro

@@
 
+Chapter \@ref(ch:plyranges) has been published in _Genome Biology_. It
+won the ACEMS Business Analytics Prize for Best Paper in 2019. Chapter \@ref(ch:fluentGenomics) has been published in _F1000 Research_.  Chapter
+\@ref(ch:intron) is based on my software contributions to the submitted
+article @Lee2019-mf. Chapter \@ref(ch:embedding) has not been submitted yet.
+
+## Open Research
+
+This thesis has been written using R Markdown with the **bookdown**
+package [@Xie2016-oa; @Xie2017-ke]. All materials required to build the
+thesis from source can be obtained from https://github.com/sa-lee/thesis. An
+online version of the thesis is available at https://sa-lee.github.io/thesis.
+
 \clearpage\pagenumbering{arabic}\setcounter{page}{0}

### 2020-07-16 Rmd/01-intro.Rmd
add preface and rejig intro

@@
 science workflow: wrangling, integration and visualisation, with a focus 
-on applications to genomics and transcriptomics. It is organised in 
-according to the aforementioned workflow steps. To begin we discuss wrangling
+on applications to genomics and transcriptomics. To begin we discuss wrangling
 biological data using a coherent representation and programming interface; 
@@
 introduces a grammar-based framework for transforming genomics data that is 
-described in chapter \@ref(ch:plyranges). Next, we look at integrating data and 
-model outputs over genomic regions to gain biological insight; \@ref(sec:ir) 
-introduces a framework for incorporating and discovering 'interesting' genomic
-regions, described further in chapters \@ref(ch:fluentGenomics) and 
-\@ref(ch:intron). Third, we consider the challenges in visualising  
-high dimensional data; section \@ref(sec:va) introduces a visual analytics 
-approach for understanding non-linear dimension reduction techniques described
-in chapter \@ref(ch:tsne). Lastly, in chapter 
+described in chapter \@ref(ch:plyranges). We then look at integrating data and 
+model outputs over genomic regions to gain biological insight; \@ref(sec:intron) 
+introduces a framework for incorporating genomic regions over multiple assays, 
+described in chapter \@ref(ch:fluentGenomics), while \@ref(sec:coverage)
+discusses finding 'interesting' genomic regions via multiple summaries of a 
+single assay, described in \@ref(ch:intron). Next, we consider the challenges in 
+visualising  high dimensional data; section \@ref(sec:va) introduces an 
+interactive visualisation approach for understanding non-linear dimension 
+reduction  techniques described in chapter \@ref(ch:tsne). Lastly, in chapter 
 \@ref(ch:conclusion) I describe the outputs of my thesis and plans for future 
@@
 surprisingly large amount of utility that can be achieved with this definition. 
-By having each column representing a variable, variables in the data can be mapped
-to graphical aesthetics of plots. This paradigm enables the grammar of graphics 
-as implemented by *ggplot2* [@Wickham2016-gz; @Wilkinson2005-kq]. User interfaces
-as implemented by *tidyverse*, and in particular the *dplyr* package, are *fluent*; 
-they form a domain specific language (DSL) that gives users a mental model 
-for performing and composing common data transformation tasks 
+By having each column representing a variable, variables in the data can be 
+mapped to graphical aesthetics of plots.
+This paradigm enables the grammar of graphics as implemented by *ggplot2* [@Wickham2016-gz; @Wilkinson2005-kq].
+User interfaces as implemented by *tidyverse*, and in particular the *dplyr* package, are *fluent*;  they form a domain specific language (DSL) that gives users a mental model for performing and composing common data transformation tasks
 [@Wickham2017-dplyr; @FowlerFluent].
@@
 an analyst to compose queries on genomics datasets. This chapter has been 
-published in @Lee2019.
+published as @Lee2019.
 
-## Integration and representation of genomic data structures {#sec:ir}
+## Integration of genomic data structures {#sec:integrate}
 
@@
 the *tidyverse* and *Bioconductor* approaches, and in fact they work seamlessly
-together. This chapter has been published in @Lee2020-ie.
+together. This chapter has been published as @Lee2020-ie.
 
-Moreover, in chapter \@ref(ch:intron) we explore the limits of the *tidy data*
+## Representation of genomic data structures{#sec:coverage}
+
+In chapter \@ref(ch:intron) we explore the limits of the *tidy data*
 semantic by extending *plyranges* to analyse coverage estimated on RNA-seq
@@
 interesting regions of coverage and apply our approach to integrating intron
-signal from RNA-seq data. This chapter is based on the preprint @Lee2019-mf.
+signal from RNA-seq data. This chapter is based on my software and analysis
+contributions to the preprint @Lee2019-mf.
 
-## Visual analytics for dimension reduction workflows {#sec:va}
+## Interactive visualisation for high-dimensional data {#sec:va}
 
@@
 data sets. We focus on a common tool for EDA (especially applied to single-cell
-transcriptomics): dimension reduction (DR). We consider the incorporation of 
-interactive and dynamic graphics to assist analysts in using DR techniques
+transcriptomics): non-linear dimension reduction (NLDR). We consider the incorporation of 
+interactive and dynamic graphics to assist analysts in using NLDR techniques
 for cluster orientation tasks. In particular, we advocate for the use of tours
-[@Cook1995-bi] alongside a DR visualisation to highlight potential pitfalls and 
-distortions of obtained from a DR method. This approach acknowledges that
-there is no 'one' best visualisation or DR for a high-dimensional dataset,
-and we often want to have an understanding of both the global and local structure
-within our data.
+[@Cook1995-bi] alongside an NLDR visualisation to highlight potential pitfalls 
+and distortions of obtained from an NLDR method. This approach acknowledges that
+there is no 'one' best visualisation or dimension reduction for a 
+high-dimensional dataset, and we often want to have an understanding of both the global and local structure within our data.
 
-Chapter \@ref(ch:tsne) introduces
-a software package called *liminal* for constructing these views and
-a user interaction framework for identifying distortions. We present several
-case studies using data sets from both single-cell transcriptomics and 
-machine learning  for using our approach to diagnose the quality of 
-results obtained via popular non-linear DR methods
+Chapter \@ref(ch:tsne) introduces a software package called *liminal* for 
+constructing these views and a user interaction framework for identifying 
+distortions. We present several case studies using data sets from both 
+single-cell transcriptomics and  machine learning  for using our approach to diagnose the quality of  results obtained via popular NLDR methods
 like t-distributed stochastic neighbour embeddings (t-SNE).

### 2020-07-16 Rmd/01-intro.Rmd
update text

@@
 described in chapter \@ref(ch:plyranges). We then look at integrating data and 
-model outputs over genomic regions to gain biological insight; \@ref(sec:intron) 
-introduces a framework for incorporating genomic regions over multiple assays, 
-described in chapter \@ref(ch:fluentGenomics), while \@ref(sec:coverage)
-discusses finding 'interesting' genomic regions via multiple summaries of a 
-single assay, described in \@ref(ch:intron). Next, we consider the challenges in 
-visualising  high dimensional data; section \@ref(sec:va) introduces an 
-interactive visualisation approach for understanding non-linear dimension 
-reduction  techniques described in chapter \@ref(ch:tsne). Lastly, in chapter 
-\@ref(ch:conclusion) I describe the outputs of my thesis and plans for future 
-developments.
+model outputs over genomic regions to gain biological insight; 
+\@ref(sec:integrate) introduces a framework for incorporating genomic regions 
+over multiple assays, described in chapter \@ref(ch:fluentGenomics), while 
+\@ref(sec:coverage) discusses finding 'interesting' genomic regions via 
+multiple summaries of a single assay, described in \@ref(ch:intron). Next, we 
+consider the challenges in  visualising  high dimensional data; section 
+\@ref(sec:va) introduces an  interactive visualisation approach for 
+understanding non-linear dimension reduction  techniques described in chapter 
+\@ref(ch:tsne). Lastly, in chapter \@ref(ch:conclusion) I describe the outputs 
+of my thesis and plans for future developments.
 

### 2020-07-16 Rmd/03-fluentGenomics.Rmd
update toc; rebuild
update text

@@
 ```{r fluentGenomics-init, echo = FALSE, cache = FALSE, include = FALSE}
+knitr::opts_chunk$set(cache = TRUE)
 knitr::read_chunk(here::here("scripts", "fluentGenomics.R"))
@@
 
-## Acknowledgements
+## Acknowledgements {-}
 

### 2020-07-16 _render.R

@@ -1,4 +1,6 @@
 options(warn = 1)
+library(bookdown)
+options(bookdown.render.file_scope = FALSE)
 library(cliapp)
 start_app(theme = simple_theme())
 
@@ -43,7 +45,7 @@ if (length(formats) == 0) {
 }
 # render the book to all formats
 for (fmt in formats) {
-  bookdown::render_book("index.Rmd", fmt, quiet = FALSE)
+  render_book("index.Rmd", fmt, quiet = FALSE)
 }
 
 cli_alert_success("You Rock!")

### 2020-07-16 Rmd/04-intron.Rmd
fix citations

@@
 users have the ability to estimate coverage in parallel and drop regions
-in the genome where there is no coverage [@rsamtools; @GenomicRanges]. This
+in the genome where there is no coverage [@rsamtools; @granges]. This
 representation is tidy, since each row of the resulting **GRanges** data

### 2020-07-16 Rmd/06-conclusion.Rmd
fix reference
rejig conculsion

@@
-```{r bioc-stats, include = FALSE}
+```{r bioc-stats, include = FALSE, cache = TRUE}
 library(dplyr)
@@
 plyranges_stats <- dl_stats %>% 
-  summarise(no_ip = sum(Nb_of_distinct_IPs), no_downloads = sum(Nb_of_downloads))
+  summarise(
+    no_ip = sum(Nb_of_distinct_IPs), 
+    no_downloads = sum(Nb_of_downloads)
+  )
 
@@
 visualisation of high-throughput biological data in a principled and pragmatic
-manner.  A significant amount of my work has been devoted to the development of 
-open source R/Bioconductor packages: **plyranges**, **superintronic**
-and **liminal**. I have emphasised how coherent software packages 
-are tools for thought; they enable analysts to reason about their data and 
-models through the composition of workflows.
+manner.  The early chapters of this thesis explored the *tidy data* semantic and 
+its extension to range based genomics data. This culminated in the development
+of "plyranges: a grammar of genomic data transformation" in chapter 
+\@ref(ch:plyranges), which developed a new domain specific language for genomics
+data analysis. The applicability of the **plyranges** interface and use of the 
+*tidy data* concept were further interrogated in chapter \@ref(ch:fluentGenomics),
+"Fluent genomics with plyranges and tximeta", which described techniques 
+integrating data along the genome, and emphasised the importance of 
+interoperability between analysis tools. Similarly, chapter  \@ref(ch:intron), 
+"Exploratory coverage analysis with superintronic and 
+plyranges", tackled data integration from a different angle by looking at 
+multiple summaries of variables measured along the genome to find putative 
+regions of intron retention. In the final part of the thesis I moved towards 
+visualisation issues as they related to working with high-dimensional data
+common in biological data science.
+Chapter \@ref(ch:tsne), "Casting multiple shadows:  high-dimensional interactive data visualisation with tours and embeddings",  explored pragmatic approaches to high dimensional data
+visualisation in light of the rise of popular non-linear embedding methods.
 
-Chapter \@ref(ch:plyranges), "plyranges: a grammar of genomic data
-transformation" developed a new domain specific language for genomics data analysis.
-Chapter \@ref(ch:fluentGenomics), "Fluent genomics with plyranges and tximeta"
-described techniques integrating data along the genome, and emphasised
-the importance of interoperability between analysis tools. Chapter 
-\@ref(ch:intron), "Exploratory coverage analysis with superintronic and 
-plyranges", tackled data  integration from a different angle by looking at 
-multiple summaries of variables measured along the genome to putative regions of 
-intron retention.  Chapter \@ref(ch:tsne), "Casting multiple shadows: 
-high-dimensional  interactive data visualisation with tours and embeddings", 
-explored pragmatic approaches to high dimensional data visualisation in light of
-the rise of popular non-linear embedding methods. 
-To finish, I will discuss the implications of this work and provide 
-suggestions for further research.
+A significant amount of my work has been devoted to the development of 
+open source R packages and workflows: **plyranges**, **fluentGenomics**, **superintronic** and **liminal**.
+I have emphasised how coherent software packages are tools for thought; they enable analysts to reason about their data and  models through the composition of workflows.
+To finish, I will discuss the
+implications of this work and provide suggestions for further research.
 
-## Findings and Impact
+## Software Development and Impact
 
-The early chapters of this thesis explored the *tidy data* semantic and its
-extension to range based genomics data. This culminated in the development
-of the **plyranges** package and consequent explorations of the applicability
-of its interface and use of the *tidy data* concept in chapters 
-\@ref(ch:fluentGenomics) and \@ref(ch:intron). The package has been relatively 
-successful: it has been downloaded `r plyranges_stats$no_downloads` times from 
-`r plyranges_stats$no_ip` unique IP addresses since its availability on 
-Bioconductor. I have also had the privilege of teaching workshops on **plyranges**
-at Bioconductor conferences which also led to the development of the work in 
-chapter\@ref(ch:fluentGenomics) and refined the development of the API. A 
-broader impact of the work, has been the discussions around the
+The **plyranges** package develops a suite of verbs for interacting
+with genomicd data as *GRanges* object. Since its release on Bioconductor,
+it  has been relatively  successful: it has been downloaded `r plyranges_stats$no_downloads` times from  `r plyranges_stats$no_ip` unique IP addresses.
+I have also had the privilege of teaching workshops on **plyranges**
+at Bioconductor conferences which also led to the development of the 
+**fluentGenomics** workflow package, outlined inchapter \@ref(ch:fluentGenomics). A broader impact of the work, has been the discussions around the
 concepts of fluent interfaces and tidy data within the Bioconductor community,
@@
 different approaches for fluent interfaces for other types of omics data.
+The **plyranges** package is available to download from 
+https://bioconductor.org/packages/release/bioc/html/plyranges.html and
+the **fluentGenomics** workflow is available to download from 
+https://bioconductor.org/packages/release/workflows/html/fluentGenomics.html.
 
@@
 cell and long-read based transcriptomics data, where scalability and much
-larger design matrices would be come an issue.
+larger design matrices would be come an issue. The **superintronic** package
+is available to download from https://github.com/sa-lee/superintronic.
 
@@
 used in **liminal** could be broadly applicable many high dimensional datasets.
+The **liminal** package is available to download from https://github.com/sa-lee/liminal.
 
-## Further Research
-
-### Scalable grammars for non-rectangular data structures
+## Further Work
 
@@
 
-### Extending tour displays for large $n$ data sets
-
-We showed in chapter \@ref(ch:liminal) that tours provide a global overview
+We showed in chapter \@ref(ch:tsne) that tours provide a global overview
 that can be used as tool for exploring model fits. An issue that arises is

### 2020-07-16 Rmd/06-conclusion.Rmd
fix spelling

@@
 I have emphasised how coherent software packages are tools for thought; they enable analysts to reason about their data and  models through the composition of workflows.
-To finish, I will discuss the
-implications of this work and provide suggestions for further research.
+To finish, I will discuss the implications of this work and provide suggestions for further research.
 
-## Software Development and Impact
+## Software Development
 
 The **plyranges** package develops a suite of verbs for interacting
-with genomicd data as *GRanges* object. Since its release on Bioconductor,
+with genomic data as a *GRanges* object. Since its release on Bioconductor,
 it  has been relatively  successful: it has been downloaded `r plyranges_stats$no_downloads` times from  `r plyranges_stats$no_ip` unique IP addresses.
@@
 at Bioconductor conferences which also led to the development of the 
-**fluentGenomics** workflow package, outlined inchapter \@ref(ch:fluentGenomics). A broader impact of the work, has been the discussions around the
+**fluentGenomics** workflow package, outlined in chapter \@ref(ch:fluentGenomics). A broader impact of the work, has been the discussions around the
 concepts of fluent interfaces and tidy data within the Bioconductor community,

### 2020-07-16 Rmd/04-intron.Rmd
pdf compiles with lualatex huzzah!

@@
 
-(ref:si-fig): An overview of the **superintronic** and **plyranges** workflow.
+(ref:si-fig) An overview of the **superintronic** and **plyranges** workflow.
 Coverage is estimated directly using the design matrix that contains a source column pointing to the locations of BAM files.
@@
 
-```{r si-fig, fig.cap = "(ref:si-fig)", out.width = "80%", echo = FALSE}
+```{r si-fig, fig.cap = "(ref:si-fig)", echo = FALSE}
 knitr::include_graphics(here::here("img", "si-workflow.png"))
@@
 
-(ref:si-splom): A hexbin scatter plot matrix of the data descriptors estimated for the *cal* rnpc3 knock out zebrafish line.
+(ref:si-splom) A hexbin scatter plot matrix of the data descriptors estimated for the *cal* rnpc3 knock out zebrafish line.
 To identify coverage traces with IR like events, we want a set of descriptors that will find genes with the following characteristics: the gene is "expressed" that has a large number of intron bases relative to the coverage of other intron features, and has  relatively stable coverage within exon features.
@@
 
-(ref:si-upset): Gene overlaps found between each combination of genotype and line using the thresholds defined in the text.
+(ref:si-upset) Gene overlaps found between each combination of genotype and line using the thresholds defined in the text.
 Our procedure mostly finds genes with IR like profiles across all groups (28 shared between all four) or that is unique to a single group, since we do not consider looking at differential IR and run our thresholding separately for each group of replicates.
@@
 
-(ref:si-cvg-ccdc43): Coverage traces faceted by biological group for the zebrafish knockout experiment, oriented from the 5' to 3' end of the gene.
+(ref:si-cvg-ccdc43) Coverage traces faceted by biological group for the zebrafish knockout experiment, oriented from the 5' to 3' end of the gene.
 Orange areas refer to intronic parts of the gene, while dark green areas refer to the exonic part.
@@
 
-```{r si-cvg-ccdc43, echo=FALSE, fig.cap = "(ref:si-cvg-ccdc43)", out.width="50%"}
+```{r si-cvg-ccdc43, echo=FALSE, fig.cap = "(ref:si-cvg-ccdc43)", out.width="80%"}
 knitr::include_graphics(here::here("img", "ccdc43.pdf"))
@@
 
-(ref:si-cvg-nat15): The gene nat15 exhibits another example of U12 intron retention (located at around 27.93Mb) in the rnpc3 knockout groups.
+(ref:si-cvg-nat15) The gene nat15 exhibits another example of U12 intron retention (located at around 27.93Mb) in the rnpc3 knockout groups.
 
-```{r si-cvg-nat15,  echo=FALSE, fig.cap = "(ref:si-cvg-nat15)", out.width="50%"}
+```{r si-cvg-nat15,  echo=FALSE, fig.cap = "(ref:si-cvg-nat15)", out.width="80%"}
 knitr::include_graphics(here::here("img", "nat15.pdf"))
@@
 
-(ref:si-cvg-mapk3): The gene mapk3 appears to have intron retention close to the start of the gene that have different impacts downstream, which result in different IR profiles between the knockout and wild-type groups.
+(ref:si-cvg-mapk3) The gene mapk3 appears to have intron retention close to the start of the gene that have different impacts downstream, which result in different IR profiles between the knockout and wild-type groups.
 
-```{r si-cvg-mapk3,  echo=FALSE, fig.cap = "(ref:si-cvg-mapk3)", out.width="50%"}
+```{r si-cvg-mapk3,  echo=FALSE, fig.cap = "(ref:si-cvg-mapk3)", out.width="80%"}
 knitr::include_graphics(here::here("img", "mapk3.pdf"))
@@
 
-(ref:si-cvg-tspan31): The gene tspan31 has a potential cryptic splice site
-within the intron retained in the knockout groups.
+(ref:si-cvg-tspan31) The gene tspan31 has a potential cryptic splice site within the intron retained in the knockout groups.
 
-```{r si-cvg-tspan31, echo=FALSE, fig.cap = "(ref:si-cvg-tspan31)", out.width="50%"}
+```{r si-cvg-tspan31, echo=FALSE, fig.cap = "(ref:si-cvg-tspan31)", out.width="80%"}
 knitr::include_graphics(here::here("img", "tspan31.pdf"))

### 2020-07-17 Rmd/00-b-abstract.Rmd
first pass at abstract

@@
 
-Exploratory data analysis  (EDA) is a vital element of the modern 
-statistical workflow - it is an analyst’s first pass at 
-understanding their data; revealing all it’s messes and (possibly) uncovering hidden insights. It is an iterative process that can 
-lead to new hypotheses that can be tested and formalised using 
-modelling. However, as data grows in complexity and there is an 
-increase in the number of observations, and variables EDA becomes 
-increasingly difficult.  This is certainly true in bioinformatic 
-data analysis, where there are constraints computation time and 
-memory, as well as a lack of guidelines or tooling for performing 
-EDA.
-
-This thesis presents a set of tools for performing exploratory 
-data analysis in the context of genomics and transcriptomics.  The first tool, “plyranges” provides a domain specific language for 
-reasoning about range based genomics data that simplifies common 
-operations such as overlaps and aggregations.  Next I extend 
-“plyranges”  to develop “superintronic” which presents an 
-approach for discovering interesting genomic regions within an 
-experimental design. We apply this tool to explore intron signal 
-in transcriptomic data. Finally, I turn my attention to focus on 
-embedding methods currently used for visualisation of large data 
-sets…
+Exploratory data analysis is vital to modern science workflows; it allows
+scientists to grasp problems with their data and generate new hypotheses.
+This work explores three facets of exploratory data analysis workflows as 
+applied to biological data science: data wrangling, integration and 
+visualisation.  It contributes new statistical computing interfaces and 
+frameworks with the explicit aim of enabling scientists to understand
+their data and models in their biological context. In chapter 
+\@ref(ch:plyranges) we show that genomics data can be represented using
+tidy data semantics, and consequently the process of wrangling it can
+be simplified via our grammar of genomic data transformation. 
+The next contribution is exploring the implications
+of our grammar on the integration and representation of genomics data.
+In chapter \@ref(ch:fluentGenomics), we provide a framework for integrating
+genomics data from multiple assays, via combining model estimates over
+their genomic regions. Next we extend our grammar to represent single
+variable measurements along the genome in multiple ways; 
+in chapter \@ref(ch:intron) we present a software tool that allows coverage
+scores to be aggregated and visualised over an experimental design and 
+genomic features and use this tool to uncover intron signal in RNA-seq data.
+Finally, in chapter \@ref(ch:tsne) we contribute a new visualisation 
+interface that provides scientists with a toolkit for discovering structure in 
+their high dimensional data, and assist them in understanding when non-linear 
+dimension reduction has worked appropriately.
 

### 2020-07-19 Rmd/05-embeddings.Rmd
update build
white space
update embeddings chapter

--- 2020-03-24 Rmd/05-embeddings.Rmd
@@
+```{r setup-videos, cache = FALSE}
+vid_lookup <- tibble::tribble(
+  ~key, ~url, ~path,
+  "gaussian",  "https://player.vimeo.com/video/439635921", here::here("video", "liminal-gaussian.mov"),
+  "hierachical", "https://player.vimeo.com/video/439635905", here::here("video", "liminal-hierachical.mov"),
+  "trees-01", "https://player.vimeo.com/video/439635892", here::here("video", "liminal-trees-01.mov"),
+  "trees-02", "https://player.vimeo.com/video/439635863", here::here("video", "liminal-trees-02.mov"),
+  "mouse-01", "https://player.vimeo.com/video/439635812", here::here("video", "liminal-mouse-01.mov"),
+  "mouse-02", "https://player.vimeo.com/video/439635843", here::here("video", "liminal-mouse-02.mov")
+)
+
+display_vid_html <- function(key, lookup) {
+  video <- lookup[lookup$key == key, ]
+  knitr::include_url(video[["url"]])
+}
+
+disply_vid_latex <- function(key, lookup) {
+    video <- lookup[lookup$key == key, ]
+    glue::glue(
+    "\\begin{{center}}
+   \\includemedia[
+    activate=pageopen,
+    width=640pt,
+    height=400pt,
+    keepaspectratio=true,
+    addresource={video},
+    flashvars={{%
+      src={video}
+      &scaleMode=stretch}
+  ]{{}}{{StrobeMediaPlayback.swf}}
+  \\end{{center}}",
+  video = video[["path"]]
+  )
+}
+
+```
+
 # Casting multiple shadows: high-dimensional interactive data visualisation with tours and embeddings {#ch:tsne}
 
-<!--
-- review of embedding algorithms, interactive graphics, tours
-- technology
-- design
-- examples/benchmarks
-  - Kevin moon
-  - single cell
-https://jlmelville.github.io/smallvis/init.html#other_initialization_options
--->
-
-There has been a rapid uptake in the use of non-linear dimensionality reduction (NLDR) methods such as t-distributed stochastic neighbour embedding (t-SNE) in the natural sciences as part of cluster orientation and dimension reduction workflows.
-The appropriate use of these methods is made difficult by their complex parameterisations and the multitude of decisions required to balance the preservation of
-local and global structure in the resulting visualisation.
-We present a visual analytics framework for the pragmatic usage of NLDR methods by combining them with a technique called the tour.
-A tour is a sequence of interpolated linear projections of multivariate data onto a lower dimensional space.
-The sequence is displayed as a dynamic visualisation, allowing a user to see the shadows the high-dimensional data casts in a lower dimensional view.
-By linking the tour to a view
-obtained from an NLDR method, we can preserve global structure and through user interactions like spatial linked brushing observe where
-the NLDR view may be misleading. We display several case 
-studies from single cell genomics, that shows our approach is useful for
-cluster orientation tasks. 
-The implementation of our framework is available as an R package called `liminal` available at https://github.com/sa-lee/liminal.
+There has been a rapid uptake in the use of non-linear dimensionality reduction
+(NLDR) methods such as t-distributed stochastic neighbour embedding (t-SNE) in
+the natural sciences as part of cluster orientation and dimension reduction
+workflows. The appropriate use of these methods is made difficult by their
+complex parameterisations and the multitude of decisions required to balance
+the preservation of local and global structure in the resulting visualisation.
+We present a visual analytics framework for the pragmatic usage of NLDR methods
+by combining them with a technique called the tour. A tour is a sequence of
+interpolated linear projections of multivariate data onto a lower dimensional
+space. The sequence is displayed as a dynamic visualisation, allowing a user to
+see the shadows the high-dimensional data casts in a lower dimensional view. By
+linking the tour to a view obtained from an NLDR method, we can preserve global
+structure and through user interactions like spatial linked brushing observe
+where the NLDR view may be misleading. We display several case studies from
+single cell transcriptomics and image processing, that shows our approach is
+useful for cluster orientation tasks.  The implementation of our framework is
+available as an R package called **liminal** available at
+https://github.com/sa-lee/liminal.
 
@@
 space into a low-dimensional one now form a core step of the data analysis
-workflow where they are used to ascertain hidden structure and de-noise data for
-downstream analysis (thereby nullifying the 'curse of dimensionality').
+workflow where they are used to ascertain hidden structure and de-noise data
+for downstream analysis (thereby nullifying the 'curse of dimensionality').
 
@@
 sequencing experiment, or identifying a gradient or trajectory along the
-embedded form @Nguyen2019-yh. The second relates to using embeddings as part of
+embedded form [@Nguyen2019-yh]. The second relates to using embeddings as part of
 a clustering workflow. Here analysts are interested in identifying and naming
@@
 variables that are a-priori known to distinguish clusters. Both of these
-workflow steps rely on the embedding being 'faithful' or the original high
-dimensional dataset, and become much more difficult when there is no underlying
+workflow steps rely on the embedding being representative the original high
+dimensional dataset, and becomes much more difficult when there is no underlying
 ground truth.
-
-Embedding methods can be classified into two broad groups: linear and
-non-linear methods. Linear methods perform a linear transformation of the data;
-one example is principal components analysis (PCA) which performs an
-eigendecomposition of the estimated sample covariance matrix. The eigenvalues
-are sorted in decreasing order and represent the variance explained by each
-component (eigenvector). A common approach to deciding on the number of
-principal components to retain is to plot the proportion of variance explained
-by each component and choose a cut-off.
-
-Non-linear methods generally perform pre-processing on the high-dimensional data
-such as generating a neighborhood graph and perform transformations on the
-pre-processed form. We restrict our attention to three methods that are commonly
-used in high-throughput biology: t-distributed stochastic neighbor embedding
-(t-SNE), uniform manifold alignment and projection (UMAP), and potential of
-heat-diffusion for affinity-based transition embedding (PHATE). The t-SNE
-algorithm estimates the similarity of (Euclidean) distances of points in a high
-dimensional space using a Gaussian distribution and then estimates a
-configuration in the low dimensional embedding space by modelling similarities
-using a t-distribution with 1 degree of freedom. The resulting configuration is
-the one that minimizes the Kullback-Leibler divergence between the two
-distributions. A recent theoretical contribution by @Linderman2019-dq proved
-that t-SNE can recover spherical and well separated cluster shapes, and
-proposed new approaches for tuning the optimisation parameters. It is a known
-problem that t-SNE can have trouble recovering global structure and that
-configurations can be highly dependent on how the algorithm is initialised and
-parameterized [@Wattenberg2016-ji; @Kobak2019-lm]. UMAP is a method that is
-related to LargeVis [@Tang2016-oz], and like t-SNE acts on the k-nearest
-neighbor graph. Its main differences are that it uses a different cost function
-(cross entropy) which is optimized using stochastic gradient descent and
-defines a different kernel for similarities in the low dimensional space. Due
-to it's computational speed it's possible to generate UMAP embeddings in more
-than three dimensions. Finally, PHATE, inspired by diffusion maps, is based on
-estimating an affinity matrix via a distance matrix and k-nearest neighbors
-graph. The algorithm de-noises estimated distances in high dimensional space via
-transforming the affinity matrix into a Markov transition probability matrix and
-diffusing this matrix over a fixed number of time steps. Then the diffused
-probabilities are transformed once more to construct a distance matrix, and
-multidimensional scaling is performed to construct a 2d embedding for
-visualization.
 
@@
 2D scatter plots were mostly sufficient for detecting class separation, however
-they noted that often multiple embeddings were required. For the task of
+they also noted that often multiple embeddings were required. For the task of
 cluster identification, @Lewis2012-ai showed experimentally that novice users
@@
 the tour. A tour is a sequence of projections of a high dimensional dataset
-onto a low-dimensional orthonormal basis matrix, that is represented as a
-dynamic visualization. The sequence of generated bases are interpolated to form
-the tour path, allowing a user to explore the subspace of projections. A grand
-tour corresponds to choosing new bases at random, and can give an overview of
-the structure in the data. Instead of picking projections at random, a guided
-tour can be used to generate a sequence 'interesting' projections as quantified
-by an index function. Given the dynamic nature of the tour, user interaction is
+onto a low-dimensional basis matrix, and is represented as a
+animated visualization.Given the dynamic nature of the tour, user interaction is 
 important for controlling and exploring the visualisation: the tour has been
-used previously by @Wickham2015-cx as tool for exploring statistical model fits
-and by @Buja1996-fk for exploring factorial experimental designs.
+used previously by @Wickham2015-cx and @Li2020-rh as a tool for 
+exploring statistical model fits and by @Buja1996-fk for exploring the space of 
+factorial experimental designs.
 
@@
 techniques: the Sleepwalk interface provides a click and highlight
-visualisation for colouring points in an embedding according to their distance
-in the original high-dimensional space [@Ovchinnikova2019-gf]. The work by
-@Pezzotti2017-cz provides a user guided and modified form of the t-SNE
-algorithm, that allows users to modified optimisation parameters in real-time.
-Similarly, the embedding projector is a web interface to running UMAP, t-SNE or
-PCA live in the browser and provides interactions to color points, and
-highlights nearest neighbors [@Smilkov2016-hp].
+visualisation for colouring points in an embedding according to an estimated
+pair-wise distance in the original high-dimensional space [@Ovchinnikova2020-sy].
+The idea behind this is that non-linear embedding methods should preserve 
+locality and that points that are 'close' to each other on the embedded view
+should be close according to the estimated pair-wise distance. 
+Similarly, the TensorFlow embedding projector is a web interface to 
+running some non-liner embedding methods live in the browser and 
+provides interactions  to colour points, and select nearest neighbours
+[@Smilkov2016-hp]. Finally, the work by @Pezzotti2017-cz provides a user 
+guided and modified form of the  t-SNE algorithm, that allows users to modified 
+optimisation parameters in real-time.
 
 There is no one-size fits all: finding an appropriate embedding for a given
-dataset is a difficult and somewhat poorly defined problem. For non-linear
+dataset is a difficult and a somewhat poorly defined problem. For non-linear
 methods, there are a lot of parameters to explore that can have an effect on
@@
 cluster orientation tasks. This workflow is incorporated into an R package
-called `liminal` (Available: https://github.com/sa-lee/liminal).
+called **liminal** (Available: https://github.com/sa-lee/liminal).
 
-## Design
+## Dimensionality Reduction Background
 
-We propose using tours as part of an analyst's workflow in in performing
-dimensionality reduction tasks. We have made extensive use of ensemble
-graphics, that is aligning related plots alongside each other to provide
-context. As we will see in the case studies, this allows analysts to quickly
-compare views from embedding methods and allows them to see how the embedding
-method alters the global structure of their data. Using ensembles allows the
-use of interaction techniques, that allow analysts to perform cluster
-orientation tasks via linking multiple views. This approach allows our
-interface, to achieve the three principles for interactive high-dimensional
-data visualisation outlined by @Buja1996-fk: finding gestalt, posing queries,
-and making comparisons.
+In the following section, we provide a brief overview of the goals of DR
+methods with respect to the data analyst, in addition to their high level the
+mathematical details. Here we restrict our attention to two recent methods that
+are commonly used in the literature: t-distributed stochastic neighbour embedding 
+(t-SNE) and uniform manifold alignment and projection (UMAP); however we do 
+mention several other techniques.
+
+To begin we suppose the data is in the form rectangular
+matrix of real numbers, $X = [\mathbf{x_1}, \dots, \mathbf{x_n}]$, where $n$ is
+the number of observations in $p$ dimensions.  The purpose of any DR algorithm
+is to find a low-dimensional representation of the data, 
+$Y = [\mathbf{y_1}, \dots, \mathbf{y_n}]$, such that $Y$ is an $n \times d$
+matrix where $d \ll p$. The hope of the analyst is that the DR procedure to 
+produces $Y$ will remove noise in the original dataset while retaining any 
+latent structure.
+
+DR methods can be classified into two broad classes: linear and non-linear
+methods. Linear methods perform a linear transformation of the data, that is,
+$Y$ is a linear transformation of $X$; one example is principal components
+analysis (PCA) which performs an eigendecomposition of the estimated sample
+covariance matrix. The eigenvalues are sorted in decreasing order and represent
+the variance explained by each component (eigenvector). 
+A common approach to deciding on the number of principal components to retain is
+to plot the proportion of variance explained by each component and choose a 
+cut-off.
+
+For non-linear methods $Y$ is generated via a pre-processed form of the input
+$X$ such as the $k$-nearest neighbours graph or via a kernel transformation.
+Multidimensional scaling (MDS) is a class of DR methods that aims to construct
+an embedding  $Y$ such that the pair-wise distances (inner products) in $Y$ 
+approximate the pair-wise distances (inner products) in $X$ 
+[@Torgerson1952-am; @Kruskal1964-cz]. There are many 
+variants of MDS, such as non-metric scaling which amounts to replacing
+distances with ranks instead [@Kruskal1964-cw]. 
+A related technique is Isomap which uses a $k$-nearest neighbour graph
+to estimate the pair-wise geodesic distance of points in $X$ then uses classical 
+MDS to construct $Y$ [@Silva2003-xw]. Other approaches are based on diffusion
+processes such as diffusion maps [@Coifman2005-ak]. A recent example of this 
+approach is the PHATE algorithm [@Moon2019-ce]. Here an affinity matrix is 
+estimated via the pair-wise distance matrix and k-nearest neighbours graph of $X$. 
+The algorithm de-noises estimated distances in high-dimensional space via 
+transforming the affinity matrix into a Markov transition probability matrix 
+and diffusing this matrix over a fixed number of time steps. Then the diffused 
+probabilities are  transformed once more to construct a distance matrix, 
+and classical MDS is  used to generate $Y$. A general difficulty with using 
+non-linear DR methods for exploratory data analysis is selecting and tuning 
+appropriate parameters. To see the extent of these choices 
+we now examine the underpinnings of t-SNE and UMAP.
+
+The t-SNE algorithm estimates
+the pair-wise similarity of (Euclidean) distances of points in a high
+dimensional space using a Gaussian distribution and then estimates a
+configuration in the low dimensional embedding space by modelling similarities
+using a t-distribution with 1 degree of freedom. There are several subtleties
+to the to use of the algorithm that are revealed by stepping through its
+machinery.
+
+To begin, t-SNE transforms pair-wise distances between $\mathbf{x_i}$ and
+$\mathbf{x_j}$ to similarities using a Gaussian kernel:
+
+$$ p_{i|j} = \frac{\exp(-\lVert \mathbf{x_i - x_j} \rVert ^ 2 /
+2\sigma_i^2)}{\sum_{k \ne i}\exp(-\lVert \mathbf{x_j - x_k} \rVert ^ 2 /
+2\sigma_i^2)} $$
+
+The conditional probabilities are then normalised and symmetrised to form a
+joint probability distribution via averaging:
+
+$$ p_{ij} = \frac{p_{i|j} + p_{j|i}}{2n} $$
+
+The variance parameter of the Gaussian kernel is controlled by the analyst
+using a fixed value of perplexity for all observations:
+
+$$ \text{perplexity}_i = \exp(-\log(2) \sum_{i \ne j}p_{j|i}\log_2(p_{j|i})) $$
+As the perplexity increases, $\sigma^2_{i}$ increases, until its bounded above
+by the number of observations , $n-1$, in the data, corresponding to
+$\sigma^2_{i} \rightarrow \infty$. This essentially turns $t-SNE$ into a
+nearest neighbours algorithm, $p_{i|j}$ will be close to zero for all
+observations that are not in the $\mathcal{O}(\text{perplexity}_i)$
+neighbourhood graph of the $i$th observation [@Van_Der_Maaten2014-zn].
+
+Next, the target low-dimensional space, $Y$, pair-wise distances between
+$\mathbf{y_i}$ and $\mathbf{y_j}$ are modelled as a symmetric probability
+distribution using a Cauchy kernel:
+
+$$ q_{ij} = \frac{w_{ij}}{Z} \\ \text{where } w_{ij} = \frac{1}{ 1 + \lVert
+\mathbf{y_i - y_j} \rVert ^ 2} \text{ and } Z = \sum_{k \ne l} w_{kl}. $$
+
+The resulting embedding $Y$ is the one that minimizes the Kullback-Leibler
+divergence between the probability distributions formed via similarities of
+observations in $X$, $\mathcal{P}$ and similarities of observations in $Y$,
+$\mathcal{Q}$:
+
+$$ \mathcal{L(\mathcal{P}, \mathcal{Q})} = \sum_{i \ne j} p_{ij} \log
+\frac{p_{ij}}{q_{ij}}$$
+
+Re-writing the loss function in terms of attractive (right) and repulsive
+(left) forces we obtain:
+
+$$ \mathcal{L(\mathcal{P}, \mathcal{Q})} = -\sum_{i \ne j} p_{ij}\log w_{ij} +
+\log\sum_{i \ne j} w_{ij} $$
+
+So when the the loss function is minimised this corresponds to large attractive
+forces, that is, the pair-wise distances in $Y$ are small when there are
+non-zero $p_{ij}$, i.e. $x_i$ and $x_j$ are close together. The repulsive force
+should also be small when the loss function is minimised, that is, when
+pair-wise distances in $Y$ are large regardless of the magnitude of the
+corresponding distances in $X$.
+
+Taken together, these details reveal the sheer number of decisions that an
+analyst must make. How does one choose the perplexity? How should the
+parameters that control the optimisation of the loss function (done with
+stochastic gradient descent), like the number of iterations, or early
+exaggeration ( a multiplier of the attractive force at the beginning of the
+optimisation), or the learning rate be selected?  It is a known problem that
+t-SNE can have trouble recovering topology and that configurations can be
+highly dependent on how the algorithm is initialised and parameterized
+[@Wattenberg2016-ji; @Kobak2019-lm; @Melville2020]. If the goal is cluster
+orientation a recent theoretical contribution by @Linderman2019-dq proved that
+t-SNE can recover spherical and well separated cluster shapes, and proposed new
+approaches for tuning the optimisation parameters. However, the cluster sizes
+and their relative orientation from a $t-SNE$ view can be misleading
+perceptually, due to the algorithms emphasis on locality.
+
+Another recent method UMAP has seen a large rise in popularity (at least in
+single cell transcriptomics) [@McInnes2018-co]. It is a method that is related to
+LargeVis [@Tang2016-oz], and like t-SNE acts on the k-nearest neighbour graph.
+Its main differences are that it uses a different cost function (cross entropy)
+which is optimized using stochastic gradient descent and defines a different
+kernel for similarities in the low dimensional space. Due to it's computational
+speed it's possible to generate UMAP embeddings in more than three dimensions.
+It appears to suffer from the same perceptual issues as t-SNE, however it
+supposedly preserves global structure better than t-SNE [@Coenen2019-to].
+
+## Visual Design
+
+We propose using tours as part of an analyst's workflow when using embedding
+methods like t-SNE. We have made extensive use of ensemble graphics, that is
+aligning related plots alongside each other to provide context [@Unwin2018-zb]. 
+As we will see in the case studies, this allows analysts to quickly compare 
+views from embedding methods and allows them to see how the embedding method 
+alters the global structure of their data. Using ensembles allows the use of 
+interaction techniques, that allow analysts to perform cluster orientation 
+tasks via linking multiple views. This approach allows our interface, to achieve 
+the three principles for interactive high-dimensional data visualisation outlined
+by @Buja1996-fk: finding gestalt, posing queries, and making comparisons.
 
@@
 
-We have combined the tour view in a side by side layout with 
-a scatter plot view that represents the output of an embedding algorithm. 
-These views are linked; analysts can brush regions or
-highlight collections of points in either view. Linked highlighting can
-be performed when points have been previously labeled according to some
-discrete structure, i.e. cluster labels are available. This is achieved
-via the analyst clicking on groups in the legend, which causes unselected groupings to have their points become less opaque. Consequently, 
-simple linked highlighting can alleviate a known downfall of methods such as 
-UMAP or t-SNE: that is distances between clusters are misleading. By
-highlighting corresponding clusters in the tour view, the analyst can
-see the relationship between clusters, and therefore obtain a more accurate 
-representation of the topology of their data.
+We have combined the tour view in a side by side layout with a scatter plot
+view that represents the output of an embedding algorithm.  These views are
+linked; analysts can brush regions or highlight collections of points in either
+view. Linked highlighting can be performed when points have been previously
+labelled according to some discrete structure, i.e. cluster labels are
+available. This is achieved via the analyst clicking on groups in the legend,
+which causes unselected groupings to have their points become less opaque.
+Consequently, simple linked highlighting can alleviate a known downfall of
+methods such as UMAP or t-SNE: that is distances between clusters are
+misleading. By highlighting corresponding clusters in the tour view, the
+analyst can see the relationship between clusters, and therefore obtain a more
+accurate representation of the topology of their data.
 
@@
 paused and corresponding points in the embedding view are highlighted.
-Likewise, when brushing occurs in the embedding view, corresponding points
-in the tour view are highlighted. In this case, an analyst can use brushing
-for manually identifying clusters and verifying cluster locations and shapes:
+Likewise, when brushing occurs in the embedding view, corresponding points in
+the tour view are highlighted. In this case, an analyst can use brushing for
+manually identifying clusters and verifying cluster locations and shapes:
 brushing in the embedding view gives analysts a sense of the shape and
@@
 
-As mentioned previously, when using any DR method, we are assuming
-the embedding is representative of the high-dimensional dataset it was
-computed from. Defining what it means for embedding to be 'representative` or
-'faithful' to high-dimensional data is ill-posed and depends on the underlying
-task an analyst is trying to achieve. At the very minimum, we are interested
-in distortions and diffusions of the high-dimensional data. Distortions
-occur when points that are near each other in the embedding view are 
-far from each other in the original dataset. This implies that the embedding 
-is not continuous. Diffusions occur when points are far from each other in the embedding view are near in the original data.  Whether, points are near or far 
-is reliant on the distance metric used; distortions and diffusions can
-be thought of as the preservation of distances or the nearest neighbors
-graphs between the high-dimensional space and the embedding space. As distances
-can be noisy in high-dimensions, ranks can be used instead as has been 
-proposed by ... Identifying distortions and diffusions allows an analyst
-to investigate the quality of their embedding and revise them iteratively. We
-propose this can be done visually using our side-by-side tour and embedding
-views. Some quality checks, such as finding distortions can be identified using simple linked brushing, however we can also interrogate them via spatial
-brushes and brush composition.
+As mentioned previously, when using any DR method, we are assuming the
+embedding is representative of the high-dimensional dataset it was computed
+from. Defining what it means for embedding to be 'representative` or 'faithful'
+to high-dimensional data is ill-posed and depends on the underlying task an
+analyst is trying to achieve. At the very minimum, we are interested in
+distortions and diffusions of the high-dimensional data. Distortions occur when
+points that are near each other in the embedding view are far from each other
+in the original dataset. This implies that the embedding is not continuous.
+Diffusions occur when points are far from each other in the embedding view are
+near in the original data.  Whether, points are near or far is reliant on the
+distance metric used; distortions and diffusions can be thought of as the
+preservation of distances or the nearest neighbours graphs between the
+high-dimensional space and the embedding space. As distances can be noisy in
+high-dimensions, ranks can be used instead as has been proposed by @Lee2009-zb.
+Identifying distortions and diffusions allows an analyst to investigate the
+quality of their embedding and revise them iteratively.
 
- Look up nearest neighbours graph from points that lie in a brushing region.
+We propose this can be
+done visually using our side-by-side tour and embedding views.
+In the simplest case, a local continuity check can be assessed via one to
+one linked brushing from the embedding to the tour view. This can be generalised further with one to many brushing, via spatial brushing. This form
+of brushing acts directly on $k$-nearest neighbours ($k$-nn) graph computed from a
+reference dataset, when a user brushes over a region in the embedding,
+all the points that match the graphs edges are selected on the 
+corresponding tour view. In place of highlighting, one could use opacity
+or binned colour scales to encode distances or ranks instead of the 
+neighbouring points, or by highlighting.
+Brush composition, such as 'and', 'or', or 'not' brushes,  could also be used to interrogate mismatches between the $k$-nn graphs estimated from both the embedding and tour views.
+This is left for
+future implementations of **liminal**.
 
-Highlight the corresponding neighbours using colour or transparency in the linked view.
+## Software Infrastructure
 
-- The $k$ nearest neighbours graph can be pre-computed quickly, for either $X$, $Y$ or both.
-- Instead of using the neighbour indices, we could use the neighbour distances instead.
-- Composition of multiple brushes could be used to show where there are matches/mismatches between nearest neighbour graphs.
+We have implemented the above design as an open source R package called
+**liminal** [@baseR]. The package allows analysts to construct concatenated
+visualisations, via the Vega-Lite grammar of interactive graphics via
+the **vegawidget** package [@Satyanarayan2017-gs; @Lyttle2020-hp]. It provides an interface for constructing linked
+and stand alone interfaces for manipulating tour paths via the **shiny** and **tourr** [@Chang2020-bq; @Wickham2011-st]. Below we provide some background
+on our implementation of the tour and user interaction.
 
-Using a linked neighbourhood brush, we can 
-visually investigate the nearest neighbour relationships in the 
-high-dimensional space via brushing in the embedding view. The user can select
-the number of nearest neighbours directly, and modify the distance metric 
-used for determining the neighbours. This allows users to interrogate the stability of clusters generated in the embedding view. Multiple brushes
-can be used to pose queries in either the tour view or the embedding view;
-interface controls allow these brushes to combine using logical operators
-such as 'and', 'or', or 'not'. The use of linked brushing goes beyond
-simple color highlights, allowing analysts to get a more holistic
-view of the effect of an embedding algorithm.
+### Tours explore the subspace of $d$-dimensional projections
 
-## Implementation
+Here we provide an overview of the mathematical aspects of the tour for
+readers who are unfamiliar with the technique before discussing the 
+computational aspects of how **liminal** interfaces with them. The tour
+is emphatically not a dimension reduction technique but rather a visualisation 
+technique that is grounded in mathematical theory, that is able to ascertain 
+the shape and global structure of a dataset via inspection of the subspace 
+generated by the set of low-dimensional projections. In fact, in some
+of the case studies below, we  first sphere our data via principal 
+components to reduce dimensionality of our input.
 
-We have implemnted the above design as an open source R package 
-called `liminal` [@baseR]. The package allows analysts to constuct
-concatenated visualisations, via the
-Vega-Lite grammar of interactive graphics (using `vegawidget` package)
-and provides an interface for constructing brushes and manipulating tour paths
-using the `shiny` and `tourr` packages. `liminal` also provides a stand-alone
-interface to the tour, in addition to the linked scatterplot interfaces 
-discussed above. Below we display the `liminal` API, and our approach to
-generating tour paths and user interactions.
+Like, when using DR techniques, the tour assumes we have a real data matrix
+$X$ consisting of $n$ observations in $p$ dimensions. First, the tour generates 
+a  sequence of $p \times d$ orthonormal projection matrices (bases) 
+$A_{t \in \mathbb{N}}$, where $d$ is typically $1$ or $2$. For each pair of 
+orthonormal bases $A_{t}$ and $A_{t+1}$ that are generated, the geodesic
+path between them is interpolated to form intermediate frames, and giving the 
+sense of continuous movement from one basis to another. The tour is then the 
+continuous visualisation of the  projections $Y_{t} = XA_{t}$, that is the 
+projection of $X$ onto $A_{t}$ as the tour path is interpolated between 
+successive bases. A *grand tour* corresponds to choosing new orthonormal
+bases at random; allowing a user to ascertain structure via exploring the
+subspace of $d$-dimensional projections. Instead of picking projections at 
+random, a *guided tour* can be used to generate a sequence 'interesting' 
+projections as quantified by an index function. While **liminal** is able to 
+visualise guided tours, our focus in the case studies uses the grand tour to
+see global structure in the data.
 
@@
 
-- don't need to realise the entire sequence
-- instead generate new bases according to a fixed frame rate
-- allows user interactions to play/pause on the current view
+The process of generating successive bases and interpolating between them to  
+construct intermediary frames, means the tour is a dynamic visualisation 
+technique. Generally, the user would set $d=2$ the tour is visualised as
+a animated scatter plot. This process of constructing bases and intermediate 
+frames and visualising the resulting projection is akin to making a "flip book" 
+animation. Like with a flip book, an interface to the tour requires the ability
+to interact and modify it in real time. The user interface generated in **liminal**
+allows a user to play, pause, and reset the tour animation,
+panning and zooming to modify the scales of the plot to provide context and
+click events to highlight groups of points if a labelling variable has
+been placed on the legend.
 
-### Linking views via brushes
+These interactions are enabled by treating the basis generation as a reactive
+stream. Instead of realising the entire sequence, which limits the animation
+to have a discrete number of frames, new bases and their intermediate frames
+are generated dynamically via pushing the current projection to the 
+visualisation interface. The interface listens to events like pressing a button
+or mouse-dragging and reacts by pausing the stream. This process allows the
+user to manipulate the tour in real time rather than having to fix the number
+of bases ahead of time. Additionally, once the user has identified an 
+interesting projection or is done with the tour, the interface will return the 
+current basis for use downstream.
+
+### Linking and highlighting views via interactions
+
+The embedding and tour views are linked together via rectangular brushes;
+when a brush is active, points will be highlighted in the adjacent view. 
+Because the tour is dynamic, brush events that become active will pause the
+animation, so that a user can interrogate the current view. By default,
+brushing on the embedding view will produce a one-to-one linking with the tour
+view (the special case of $k=1$ nearest neighbours brush). In future
+implementations it will be possible to shift from one-to-many linking by increasing $k$ via the 'controls' panel of the interface.
+If desired, the distance metric used to estimate nearest neighbours can by chosen by the user.
+The user may provide a reference data set for computing
+nearest neighbours (for example a distance matrix, or the complete data matrix)
+independently of the tour or embedding views. The estimation of neighbours
+is done with the `FNN` package for datasets with less than 4096 observations
+and `RcppAnnoy` for larger datasets or if a user selects a non-Euclidean
+distance metric [@fnn-pkg; @rcpp-annoy-pkg]. Once the user is finished 
+with the interface, the brush regions are returned for downstream analysis.
 
@@
 
-### Case Study 1: Exploring tree structured data with tours and t-SNE
+The next section steps through case studies cases of our approach using simulated
+and real data.  The videos presented show the visual 
+appearance of the **liminal** interface, and how we can interact with the tour
+via the controls previously described.
+In the first case study, we look at three different examples
+where t-SNE is applied to three simulations where cluster structure and geometry
+of the underlying data is known.
 
-(ref:fake-trees) Example from PHATE
+In the second case study, we apply our approach to clustering 
+the mouse retina data from @Macosko2015-ot, and use apply the tour
+to the process of verifying marker genes.
 
-```{r fake-trees, echo = FALSE, out.width = "50%", fig.cap="(ref:fake-trees)"}
-knitr::include_graphics(here::here("img", "tree.png"))
+<!--
+Finally, in the last case study we
+explore a subset of the QuickDraw dataset, we use **liminal** to identify
+instances of images from different labels that appear similar.
+-->
+
+### Case Study 1: Exploring simulation data with tours and t-SNE
+
+To begin we look at simulated datasets that reproduce known facts
+about the  t-SNE algorithm. Our first data set consists of five spherical 5-$d$
+Gaussian clusters embedded in 10-$d$ space, each cluster has the same 
+covariance matrix. We then computed a t-SNE layout with default settings 
+using the **Rtsne** package [@Rtsne],
+and set up the **liminal** linked interface with grand tour on the 10-$d$
+observations:
+
+```{r gaussian, results="asis", eval = knitr::is_latex_output()}
+cat(display_vid_latex("gaussian", vid_lookup))
 ```
 
-#### Using t-SNE
+```{r gaussian-h, eval = knitr::is_html_output()}
+display_vid_html("gaussian", vid_lookup)
+```
 
-- there's a lot of parameters to tweak
-- emphasises locality, distance between clusters of points can be misleading
-- size of clusters can be misleading 
-- may require a few different runs to capture topology of data
-- if there are clusters it will find them
+We see in the video that t-SNE has correctly split out each cluster and
+laid them out in a star like formation. This is agrees with the tour view,
+where once we start the animation, the five clusters begin to appear but
+generally points are more concentrated in the projection view compared to 
+the t-SNE layout. This can be seen from our brushing action on the t-SNE side.
 
-### Case Study 2: Clustering PBMC 10x single cell RNA-seq data
+We repeat the process of simulating Gaussian clusters in the next video, 
+but this time we have a cluster that has more granular structure. This time we have two
+Gaussian clusters with equal covariance embedded in $10-d$, and a third cluster
+with hierarchical structure. This cluster has two 3-$d$ clusters embedded in 
+10-$d$, where the second cluster is subdivided into three smaller clusters, that are each equidistant from each other and have the same covariance structure.
+Again t-SNE has correctly identified the sub-clusters but their relative 
+locations to each other is distorted, with the orange and blue groups being
+far from each other in the tour view. We see in this case that is difficult
+to see the sub-clusters in the tour view, however, once we zoom and highlight
+they become more apparent. When we brush the sub-clusters in the t-SNE, their
+relative placement is again exaggerated, with the tour showing that they
+are indeed much closer than the impression the t-SNE view gives.
 
-### Case Study 3: Single cell mouse retina data
+```{r hierachical, results='asis', eval = knitr::is_latex_output()}
+cat(display_vid_latex("hierachical", vid_lookup))
+```
+
+```{r hierachical-h, eval = knitr::is_html_output()}
+display_vid_html("hierachical", vid_lookup)
+```
+
+Next explore some simulated noisy tree structured data (figure
+\@ref(fig:fake-trees)), our interest here is how t-SNE visualisations break
+topology of the data, and then seeing if we can resolve this by tweaking the
+default parameters with reference to the global view of the data set.
+
+(ref:fake-trees) Example high-dimensional tree shaped data, $n = 3000$ and $p = 100$.
+*(a)* The true data lies 2-d tree like structure consisting of ten branches.
+This data is available in the **phateR** package and is simulated via diffusion-limited aggregation (a random walk along the branches of the tree) with Gaussian noise added [@Moon2019-ce].
+*(b)* The first two principal components, which form the initial projection for the tour, note that the backbone of the three is obscured by this view.
+*(c)* The default t-SNE view breaks the global structure of the tree.
+*(d)* Altering t-SNE using the first two principal components as the starting coordinates for the embedding, results in clustering the tree at its branching points.
+
+```{r fake-trees, echo = FALSE, fig.align = "center", fig.cap="(ref:fake-trees)"}
+knitr::include_graphics(here::here("img", "tree-view.png"))
+```
+
+First, we apply principal components and restrict the results down to the first
+twelve principal components (which makes up approximately 70% of the variance
+explained in the data) to use with the grand tour.
+
+Moreover, we run t-SNE using the default arguments on the complete data (this
+keeps the first 50 PCs, sets the perplexity to equal 30 and performs random
+initialisation). We then create a linked
+tour with t-SNE layout with **liminal** as shown in the following video:
+
+```{r trees-01, results='asis', eval = knitr::is_latex_output()}
+cat(display_vid_latex("trees-01", vid_lookup))
+```
+
+```{r trees01-h, eval = knitr::is_html_output()}
+display_vid_html("trees-01", vid_lookup)
+```
+
+From the video, we see that the t-SNE view has been unable to recapitulate the
+topology of the tree - the backbone (blue) branch has been split into three
+fragments. We can see this immediately via the linked highlighting over both
+plots. If we click on the legend for the zero branch, the blue coloured points
+on each view are highlighted and the remaining points are made transparent.
+From, here it becomes apparent from the tour view that the blue branch forms
+the backbone of the tree and is connected to all other branches. We can also
+again see that cluster sizes formed via t-SNE can be misleading; from the tour
+view there is a lot of noise along the branches, while this does not appear to
+be the case for the t-SNE result.
+
+From the first view, we modify the inputs to the t-SNE view, to try and produce
+a better trade-off between local structure and retain the topology of the data.
+We keep every parameter the same except that we initialise $Y$ with the first
+two PCs (scaled to have standard deviation 1e-4) instead of the default random
+initialisation and increase the perplexity from 30 to 100. We then combine
+these results with our tour view as displayed in the video:
+
+```{r trees-02, results='asis', eval = knitr::is_latex_output()}
+cat(display_vid_latex("trees-02", vid_lookup))
+```
+
+```{r trees-02-h, eval = knitr::is_html_output()}
+display_vid_html("trees-02", vid_lookup)
+```
+
+The video shows that this selection of parameters results in the the tips of
+the branches (the three black dots in figure \@ref(fig:fake-trees) a) being split
+into three clusters representing the terminal branches of the tree. However, there
+are perceptual issues following the placement of the three groupings on the
+t-SNE view that become apparent via simple linked brushing. If we brush the
+tips of the yellow and brown branches (which appear to be close to each other
+on the t-SNE view), we immediately see the placement is distorted in the t-SNE
+view, and in the tour view these tips are at opposite ends of the tree.
+Although, this is a known issue of the t-SNE algorithm, we can easily identify
+it via simple interactivity without knowing the inner workings of the method.
+
+### Case Study 2: Clustering single cell RNA-seq data
+
+A common analysis task in single cell studies is performing clustering to
+identify groupings  of cells with similar expression profiles. Analysts
+in this area generally use non linear DR methods for verification and identification
+of clusters and developmental trajectories (i.e., case study 1). For clustering,
+workflows the primary task is verify the existence of clusters and then begin to 
+identify the clusters as cell types using the expression of "known" 
+marker genes. Here a 'faithful' a embedding should ideally preserve the topology
+of the data; cells that correspond to a cell type should lie in the same
+neighbourhood in high-dimensional space. In this case study we use our 
+linked brushing approaches to look at neighbourhood preservation
+and look at marker genes through the lens of the tour.
+
+First, we downloaded the raw mouse retinal single cell RNA-seq data from
+@Macosko2015-ot using the **scRNAseq** Bioconductor package [@scRNAseq-d]. We
+have followed a standard workflow for pre-processing and normalizing this data 
+(described by @Amezquita2020-at): we performed QC using the the `scater`
+by removing cells with high proportion of mitochondrial gene expression and low
+numbers of genes detected, we log-transformed and normalised the expression 
+values and  finally selected highly variable genes (HVGs) using `scran` 
+[@McCarthy2017; @Lun2016]. The top ten percent of HVGs were used to subset
+the normalised expression matrix and compute PCA using the first 25 components.
+Using the PCs we built a shared nearest neighbours graph (with $k = 10$) and 
+used Louvain clustering to generate clusters [@Blondel2008-bx].
+
+To check and verify the clustering we construct a **liminal** view. 
+We tour the first five PCs (approximately 20% of the variance in expression),
+along side the t-SNE view which was computed from all 25 PCs. 
+Due to latency of the interface we do weighted sample of rows based on
+cluster membership, leaving us with approximately 10 per cent of the original
+data size - 4,590 cells. Although this is not ideal, it still allows us to
+get a sense of the shape of the clusters as seen in the video below:
+
+```{r mouse-01, results='asis', eval = knitr::is_latex_output()}
+cat(display_vid_latex("mouse-01", vid_lookup))
+```
+
+```{r mouse-01-h, eval = knitr::is_html_output()}
+display_vid_html("mouse-01", vid_lookup)
+```
+
+In the following video we can observe that the embedding has mostly captured the
+clusters relative location to each other to their location in high dimensional space, with a notable exception of points in cluster 3 and 10 as shown with
+linked brushing. As expected avoids t-SNE mitigates the crowding problem that is
+an issue for tour in this case, where points in clusters 2,4,6, and 11 are
+clumped together in tour view, but are blown up in the embedding view. 
+The tour appears to form a tetrahedron like shape, with points lying on
+surface and along the vertices of the the tetrahedron in 5-$d$ PCA space -
+a phenomena that has also been observed in @Korem2015-af. Brushing on
+the tour view, reveals points in cluster 9 that are diffuse in the embedding 
+view, points in cluster 9 are relatively far away and spread apart from other clusters in the tour view, but has points placed in cluster 3 and 9 in the embedding.
+
+Next, we identify marker genes for clusters using one sided Welch t-tests with
+a minimum log fold change of one as recommended by @Amezquita2020-at, which
+uses the testing framework from @McCarthy2009-qx. We select the top 10 marker genes that are upregulated in cluster 2, which was one of the clumped clusters
+when we toured on principal components. Here the tour becomes an alternative
+to a standard heatmap view for assessing shared markers; the basis
+generation (shown as the biplot on the left view) reflects the relative 
+weighting of each gene. We run the tour directly on the 
+log normalised expression values using the same subset as before:
+
+```{r mouse-02, results='asis', eval = knitr::is_latex_output()}
+cat(display_vid_latex("mouse-02", vid_lookup))
+```
+
+```{r mouse-02-h, eval = knitr::is_html_output()}
+display_vid_html("mouse-02", vid_lookup)
+```
+
+We see that the expression of the marker genes using t, appear to separate
+the previously clumped clusters 2,4,6, and 11 from the other clusters, 
+indicating that these genes are expressed in all four clusters, and the 
+magnitude of expression in these genes is similar across these groups, 
+consistent with the results of marker gene analysis.
+
+<!--
+### Case Study 3: How do people draw animals?
+
+In the final case study, we explore a subset of the Google QuickDraw dataset 
+[@quickdraw-paper]. The data are sketches drawn by people of common concepts
+like 'airplane' or 'shoe', and corresponding metadata about the humans who
+created the sketches, such as country, the word category, whether the sketch
+was recognised by Google's neural network, and the number of strokes taken.
+We downloaded the metadata and normalised 28 by 28 pixel bitmaps via the 
+`quickdraw` package in R; unravelling the bitmaps unravelled the bitmaps into 
+a 784 dimension image vector. We sampled 500 images from each of the following
+categories: "broccoli", "flying saucer", "ice cream", "octopus", "pineapple",
+and "whale". We are interested in discovering images that are similiar but
+belong to different classifications 
+-->
 
@@
 
+We have shown that the use of tours as a tool for interacting with
+high dimensional data provides an additional insight for interrogating views
+generated from embeddings. The interface we have designed in the **liminal**
+package, allows a user to gain a deeper understanding of an embedding 
+algorithm, and rectifies perceptual issues associated with NLDR methods via 
+linked interactions with the tour. As we have shown in the simulation case
+studies, the t-SNE method can produce misleading embeddings but can be
+detected through the use linked brushing and highlighting. In the case when
+the data has a piecewise linear geometry, like the tree simulation, the tour
+preserves the shape of the data which can be obscured by the embedding method.
+
+Our holistic framework is also useful in real data situations, as displayed
+in our second case study. The tour when combined with t-SNE allowed us
+to identify clusters, while giving us an idea of their orientation to 
+each other. Moreover, we could visually inspect the separation of clusters
+using a tour on marker gene sets. We see our approach as being valuable
+to the single cell analyst who wants to make their embeddings more 
+interpretable.
+
+There are some limitations
+in using the **liminal** interface for larger datasets. First,
+t-SNE avoids the crowding problem, points are separated into distinct regions
+on the display canvas, while for the tour, points are concentrated in the
+centre of the projection and become difficult to see. 
+Second, as $n$ increases the animation becomes slower as there is more data
+passing from the server to the client. Both of these issues could be mitigated
+through alternate tour paths and displays. The recent work
+by @Laa2020-wr is a promising area for further investigation, as well as
+work from topological statistics [@Rieck2017-kk; @Genovese2017-iq].
+
 ## Acknowledgements
+
+We would like to thank David Frazier, Ursula Laa and Paul Harrison for their
+feedback on this work.

### 2020-07-19 Rmd/05-embeddings.Rmd
update pdf

@@
 
-disply_vid_latex <- function(key, lookup) {
+display_vid_latex <- function(key, lookup) {
     video <- lookup[lookup$key == key, ]
@@
    \\includemedia[
-    activate=pageopen,
-    width=640pt,
-    height=400pt,
+    width=0.6\\linewidth,
+    height=0.4\\linewidth,
     keepaspectratio=true,
@@
     flashvars={{%
-      src={video}
-      &scaleMode=stretch}
-  ]{{}}{{StrobeMediaPlayback.swf}}
+      source={video}
+      &autohide=1
+      &autoPlay=true
+      &loop=true}
+  ]{{}}{{VPlayer.swf}}
   \\end{{center}}",

### 2020-07-19 Rmd/05-embeddings.Rmd
add video, reduce PDF size with mp4

@@
   ~key, ~url, ~path,
-  "gaussian",  "https://player.vimeo.com/video/439635921", here::here("video", "liminal-gaussian.mov"),
-  "hierachical", "https://player.vimeo.com/video/439635905", here::here("video", "liminal-hierachical.mov"),
-  "trees-01", "https://player.vimeo.com/video/439635892", here::here("video", "liminal-trees-01.mov"),
-  "trees-02", "https://player.vimeo.com/video/439635863", here::here("video", "liminal-trees-02.mov"),
-  "mouse-01", "https://player.vimeo.com/video/439635812", here::here("video", "liminal-mouse-01.mov"),
-  "mouse-02", "https://player.vimeo.com/video/439635843", here::here("video", "liminal-mouse-02.mov")
+  "gaussian",  "https://player.vimeo.com/video/439635921", here::here("video", "liminal-gaussian.mp4"),
+  "hierachical", "https://player.vimeo.com/video/439635905", here::here("video", "liminal-hierachical.mp4"),
+  "trees-01", "https://player.vimeo.com/video/439635892", here::here("video", "liminal-trees-01.mp4"),
+  "trees-02", "https://player.vimeo.com/video/439635863", here::here("video", "liminal-trees-02.mp4"),
+  "mouse-01", "https://player.vimeo.com/video/439635812", here::here("video", "liminal-mouse-01.mp4"),
+  "mouse-02", "https://player.vimeo.com/video/439635843", here::here("video", "liminal-mouse-02.mp4")
 )
@@
     width=0.6\\linewidth,
-    height=0.4\\linewidth,
+    height=0.45\\linewidth,
     keepaspectratio=true,
+    activate=pageopen
     addresource={video},
@@
       source={video}
-      &autohide=1
       &autoPlay=true

### 2020-07-23 Rmd/05-embeddings.Rmd
minor adjustments per comments in #6

@@
 variables that are a-priori known to distinguish clusters. Both of these
-workflow steps rely on the embedding being representative the original high
+workflow steps rely on the embedding being representative of the original high
 dimensional dataset, and becomes much more difficult when there is no underlying
@@
 onto a low-dimensional basis matrix, and is represented as a
-animated visualization.Given the dynamic nature of the tour, user interaction is 
+animated visualization. Given the dynamic nature of the tour, user interaction is 
 important for controlling and exploring the visualisation: the tour has been
@@
 
-While there has been much work on the algorithmic details of the aforementioned
+While there has been much work on the algorithmic details of
 embedding methods, there has been relatively few tools designed to assist users
 to interact with these techniques and assist them in making comparisons between
-embeddings and performing the aforementioned cluster orientation tasks. Several
+embeddings and performing cluster orientation tasks. Several
 interactive interfaces have been proposed for evaluating or using embedding
@@
 embeddings require interaction but should also be able to be incorporated into
-an analysts workflow. We propose a more pragmatic workflow inspired by
+an analysts workflow. Instead, we implement a more pragmatic workflow by
 incorporating interactive graphics and tours with embeddings that allows users
@@
 
-We propose using tours as part of an analyst's workflow when using embedding
-methods like t-SNE. We have made extensive use of ensemble graphics, that is
+Tours provide a supportive visualisation to NLDR graphics, and 
+can be easily incorporated into an analysts workflow with our interface in 
+**liminal**. 
+We have made extensive use of ensemble graphics, that is
 aligning related plots alongside each other to provide context [@Unwin2018-zb]. 
@@
 
-We propose this can be
-done visually using our side-by-side tour and embedding views.
+These checks are done visually using our side-by-side tour and embedding views.
 In the simplest case, a local continuity check can be assessed via one to
-one linked brushing from the embedding to the tour view. This can be generalised further with one to many brushing, via spatial brushing. This form
-of brushing acts directly on $k$-nearest neighbours ($k$-nn) graph computed from a
-reference dataset, when a user brushes over a region in the embedding,
-all the points that match the graphs edges are selected on the 
-corresponding tour view. In place of highlighting, one could use opacity
-or binned colour scales to encode distances or ranks instead of the 
-neighbouring points, or by highlighting.
-Brush composition, such as 'and', 'or', or 'not' brushes,  could also be used to interrogate mismatches between the $k$-nn graphs estimated from both the embedding and tour views.
-This is left for
-future implementations of **liminal**.
+one linked brushing from the embedding to the tour view. Similarly, diffusions
+are identified from linked brushing on the tour view, higlighting points
+in the embedding view. Section <> details
 
@@
 We have implemented the above design as an open source R package called
-**liminal** [@baseR]. The package allows analysts to construct concatenated
+**liminal**. The package allows analysts to construct concatenated
 visualisations, via the Vega-Lite grammar of interactive graphics via
@@
 brushing on the embedding view will produce a one-to-one linking with the tour
-view (the special case of $k=1$ nearest neighbours brush). In future
-implementations it will be possible to shift from one-to-many linking by increasing $k$ via the 'controls' panel of the interface.
-If desired, the distance metric used to estimate nearest neighbours can by chosen by the user.
-The user may provide a reference data set for computing
-nearest neighbours (for example a distance matrix, or the complete data matrix)
-independently of the tour or embedding views. The estimation of neighbours
-is done with the `FNN` package for datasets with less than 4096 observations
-and `RcppAnnoy` for larger datasets or if a user selects a non-Euclidean
-distance metric [@fnn-pkg; @rcpp-annoy-pkg]. Once the user is finished 
-with the interface, the brush regions are returned for downstream analysis.
+view. For interpreting specific combinations of clusters, the legend
 
@@
 In the first case study, we look at three different examples
-where t-SNE is applied to three simulations where cluster structure and geometry
-of the underlying data is known.
+where t-SNE is applied to  simulations where cluster structure and geometry of the underlying data is known.
 
 In the second case study, we apply our approach to clustering 
-the mouse retina data from @Macosko2015-ot, and use apply the tour
-to the process of verifying marker genes.
+the mouse retina data from @Macosko2015-ot, and apply the tour
+to the process of verifying marker genes that separate clusters.
 
@@
 This data is available in the **phateR** package and is simulated via diffusion-limited aggregation (a random walk along the branches of the tree) with Gaussian noise added [@Moon2019-ce].
-*(b)* The first two principal components, which form the initial projection for the tour, note that the backbone of the three is obscured by this view.
+*(b)* The first two principal components, which form the initial projection for the tour, note that the backbone of the tree is obscured by this view.
 *(c)* The default t-SNE view breaks the global structure of the tree.
@@
 a minimum log fold change of one as recommended by @Amezquita2020-at, which
-uses the testing framework from @McCarthy2009-qx. We select the top 10 marker genes that are upregulated in cluster 2, which was one of the clumped clusters
+uses the testing framework from @McCarthy2009-qx. We select the top 10 marker 
+genes that are upregulated in cluster 2, which was one of the clumped clusters
 when we toured on principal components. Here the tour becomes an alternative
@@
 to the single cell analyst who wants to make their embeddings more 
-interpretable.
+interpretable
+
+In future implementations it will be possible to shift from one to one linked
+brushing to one to many linked brushing. This form of brushing acts directly on
+a $k$-nearest neighbours ($k$-nn) graph computed from a
+reference dataset: when a user brushes over a region in the embedding,
+all the points that match the graphs edges are selected on the 
+corresponding tour view.  The reference data set for computing
+nearest neighbours (for example a distance matrix, or the complete data matrix)
+can be independent of the tour or embedding views. The estimation of neighbours
+can be computed effeciently with the `FNN` or `RcppAnnoy` packages
+on the server side [@fnn-pkg; @rcpp-annoy-pkg].. In place of highlighting, one 
+could use opacity or binned colour scales to encode distances or ranks instead 
+of the neighbouring points, Brush composition, such as 'and', 'or',
+or 'not' brushes, could also be used to interrogate mismatches between the 
+$k$-nn graphs estimated from both the embedding and tour views.
 

### 2020-07-23 Rmd/05-embeddings.Rmd
some notes and rejigging

@@
 the preservation of local and global structure in the resulting visualisation.
-We present a visual analytics framework for the pragmatic usage of NLDR methods
+We present a visual <change word here> framework for the pragmatic usage of NLDR methods
 by combining them with a technique called the tour. A tour is a sequence of
@@
 important for controlling and exploring the visualisation: the tour has been
-used previously by @Wickham2015-cx and @Li2020-rh as a tool for 
+used previously by @Wickham2015-cx and  
 exploring statistical model fits and by @Buja1996-fk for exploring the space of 
 factorial experimental designs.
+
+<!--- add tour reference; linked brushing -->
 
@@
 guided and modified form of the  t-SNE algorithm, that allows users to modified 
-optimisation parameters in real-time.
+optimisation parameters in real-time. <!-- add visualising with mds reference -->
 
@@
 
-## Dimensionality Reduction Background
+## Overview of Dimension Reduction
 
@@
 **liminal**. 
-We have made extensive use of ensemble graphics, that is
-aligning related plots alongside each other to provide context [@Unwin2018-zb]. 
 As we will see in the case studies, this allows analysts to quickly compare 
@@
 
+#### Spherical Gaussian Clusters
+
 To begin we look at simulated datasets that reproduce known facts
@@
 
+#### Spherical Gaussian clusters with hierarchical structure
+
 We repeat the process of simulating Gaussian clusters in the next video, 
@@
 ```
+
+####
 
@@
 the tour view, reveals points in cluster 9 that are diffuse in the embedding 
-view, points in cluster 9 are relatively far away and spread apart from other clusters in the tour view, but has points placed in cluster 3 and 9 in the embedding.
+view, points in cluster 9 are relatively far away and spread apart from other 
+clusters in the tour view, but has points placed in cluster 3 and 9 in the embedding.
 
@@
 feedback on this work.
+
+<!-- add supplementary materials -->

### 2020-07-23 Rmd/05-embeddings.Rmd
note to self
update and rejig section on DR

@@
 
-Another recent method UMAP has seen a large rise in popularity (at least in
+Another recent method, UMAP, has seen a large rise in popularity (at least in
 single cell transcriptomics) [@McInnes2018-co]. It is a method that is related to
@@
 which is optimized using stochastic gradient descent and defines a different
-kernel for similarities in the low dimensional space. Due to it's computational
-speed it's possible to generate UMAP embeddings in more than three dimensions.
-It appears to suffer from the same perceptual issues as t-SNE, however it
-supposedly preserves global structure better than t-SNE [@Coenen2019-to].
+kernel for similarities in the low dimensional space. Due to its computational
+speed it is possible to generate UMAP embeddings in more than three dimensions.
+It appears to suffer from the same perceptual issues as t-SNE, however it supposedly preserves global structure better than t-SNE [@Coenen2019-to].
+
+### Tours explore the subspace of $d$-dimensional projections
+
+The tour is a visualisation technique that is grounded in 
+mathematical theory, and is able to ascertain 
+the shape and global structure of a dataset via inspection of the subspace generated by the set of low-dimensional projections. In fact, in some
+of the case studies below, we first sphere our data via principal 
+components to reduce dimensionality of our input.
+
+Like, when using other DR techniques, the tour assumes we have a real data matrix
+$X$ consisting of $n$ observations in $p$ dimensions. First, the tour generates 
+a  sequence of $p \times d$ orthonormal projection matrices (bases) 
+$A_{t \in \mathbb{N}}$, where $d$ is typically $1$ or $2$. For each pair of 
+orthonormal bases $A_{t}$ and $A_{t+1}$ that are generated, the geodesic
+path between them is interpolated to form intermediate frames, and giving the 
+sense of continuous movement from one basis to another. The tour is then the 
+continuous visualisation of the  projections $Y_{t} = XA_{t}$, that is the 
+projection of $X$ onto $A_{t}$ as the tour path is interpolated between 
+successive bases. A *grand tour* corresponds to choosing new orthonormal
+bases at random; allowing a user to ascertain structure via exploring the
+subspace of $d$-dimensional projections. Instead of picking projections at 
+random, a *guided tour* can be used to generate a sequence 'interesting' 
+projections as quantified by an index function. While **liminal** is able to 
+visualise guided tours, our focus in the case studies uses the grand tour to see global structure in the data.
 
@@
 the **vegawidget** package [@Satyanarayan2017-gs; @Lyttle2020-hp]. It provides an interface for constructing linked
-and stand alone interfaces for manipulating tour paths via the **shiny** and **tourr** [@Chang2020-bq; @Wickham2011-st]. Below we provide some background
-on our implementation of the tour and user interaction.
-
-### Tours explore the subspace of $d$-dimensional projections
-
-Here we provide an overview of the mathematical aspects of the tour for
-readers who are unfamiliar with the technique before discussing the 
-computational aspects of how **liminal** interfaces with them. The tour
-is emphatically not a dimension reduction technique but rather a visualisation 
-technique that is grounded in mathematical theory, that is able to ascertain 
-the shape and global structure of a dataset via inspection of the subspace 
-generated by the set of low-dimensional projections. In fact, in some
-of the case studies below, we  first sphere our data via principal 
-components to reduce dimensionality of our input.
-
-Like, when using DR techniques, the tour assumes we have a real data matrix
-$X$ consisting of $n$ observations in $p$ dimensions. First, the tour generates 
-a  sequence of $p \times d$ orthonormal projection matrices (bases) 
-$A_{t \in \mathbb{N}}$, where $d$ is typically $1$ or $2$. For each pair of 
-orthonormal bases $A_{t}$ and $A_{t+1}$ that are generated, the geodesic
-path between them is interpolated to form intermediate frames, and giving the 
-sense of continuous movement from one basis to another. The tour is then the 
-continuous visualisation of the  projections $Y_{t} = XA_{t}$, that is the 
-projection of $X$ onto $A_{t}$ as the tour path is interpolated between 
-successive bases. A *grand tour* corresponds to choosing new orthonormal
-bases at random; allowing a user to ascertain structure via exploring the
-subspace of $d$-dimensional projections. Instead of picking projections at 
-random, a *guided tour* can be used to generate a sequence 'interesting' 
-projections as quantified by an index function. While **liminal** is able to 
-visualise guided tours, our focus in the case studies uses the grand tour to
-see global structure in the data.
+and stand alone interfaces for manipulating tour paths via the **shiny** and **tourr** packages [@Chang2020-bq; @Wickham2011-st].
 
@@
 can be computed effeciently with the `FNN` or `RcppAnnoy` packages
-on the server side [@fnn-pkg; @rcpp-annoy-pkg].. In place of highlighting, one 
+on the server side [@fnn-pkg; @rcpp-annoy-pkg]. In place of highlighting, one 
 could use opacity or binned colour scales to encode distances or ranks instead 
-of the neighbouring points, Brush composition, such as 'and', 'or',
-or 'not' brushes, could also be used to interrogate mismatches between the 
+of the neighbouring points. Brush composition, such as 'and', 'or',
+or 'not' brushes, could be used to interrogate mismatches between the 
 $k$-nn graphs estimated from both the embedding and tour views.
@@
 
-<!-- add supplementary materials -->
+<!-- add supplementary materials here -->

### 2020-07-24 Rmd/01-intro.Rmd
typo spotting and reference adds

@@
 [@Cook1995-bi] alongside an NLDR visualisation to highlight potential pitfalls 
-and distortions of obtained from an NLDR method. This approach acknowledges that
+and distortions obtained from an NLDR method. This approach acknowledges that
 there is no 'one' best visualisation or dimension reduction for a 
@@
 constructing these views and a user interaction framework for identifying 
-distortions. We present several case studies using data sets from both 
-single-cell transcriptomics and  machine learning  for using our approach to diagnose the quality of  results obtained via popular NLDR methods
-like t-distributed stochastic neighbour embeddings (t-SNE).
+distortions. We present several case studies using data sets  both 
+single-cell transcriptomics and  machine learning  for using our approach to diagnose the quality   results obtained via popular NLDR methods
+like t-distributed stochastic neighbour embeddings (t-SNE) [@Maaten2008-sk].

### 2020-07-24 Rmd/05-embeddings.Rmd
add reference to tSNE!

@@
 (t-SNE) and uniform manifold alignment and projection (UMAP); however we do 
-mention several other techniques.
+mention several other techniques [@Maaten2008-sk; @McInnes2018-co].
 
@@
 configuration in the low dimensional embedding space by modelling similarities
-using a t-distribution with 1 degree of freedom. There are several subtleties
+using a t-distribution with 1 degree of freedom [@Maaten2008-sk]. There are several subtleties
 to the to use of the algorithm that are revealed by stepping through its

### 2020-07-26 Rmd/00-c-declaration.Rmd
refix declaration
updates based on Matt's comments

@@
----
-output:
-  pdf_document: default
-  html_document: default
----
 `r if (knitr::is_html_output()) '<!--'`
@@
 Division at the Walter and Eliza Hall Institute. The following table details
-the publications and my and my fellow co-authors contributions:
+the publications, including my and my fellow co-authors contributions:
 
@@
 \hline
-3 & Fluent genomics with \textit{plyranges} and \textit{tximeta} & Published (awaiting peer review) & 60\%.
+3 & Fluent genomics with \textbf{plyranges} and \textbf{tximeta} & Published & 60\%.
 Concept, software development, data analysis and manuscript writing  & (1) Michael Lawrence, manuscript feedback and editing 5\% (2) Michael I Love, Concept, data analysis and manuscript writing 35\% & \multicolumn{1}{c|}{N} \\
@@
 
-Note that chapter \@ref(ch:intron), reflects my  contribution (in the form of a 
-software package and data analysis) to the submitted manuscript entitled 
+Chapter \@ref(ch:intron), reflects my contribution (in the form of a 
+software package) to the submitted manuscript entitled 
 _Covering all your bases: incorporating intron signal from RNA-seq data_
-[@Lee2019-mf] and is joint work under the supervision of Dr Charity Law
+[@Lee2019-mf], and includes new data analyses not found in the paper.  It is joint work under the supervision of Dr Charity Law
 and Associate Professor Matthew Ritchie.

### 2020-07-26 Rmd/02-plyranges.Rmd
update toc; rebuild
update text
fix spelling
consistency of formatting

@@
 There is a cognitive load placed on users in learning a data abstraction from the Bioconductor project and understanding its appropriate use.
-Users must navigate these abstractions to perform a genomic analysis task, when a single data abstraction, a GRanges object will suffice.
-By recognizing that the GRanges class follows 'tidy' data principles, we create a grammar of genomic data transformation, defining verbs for performing actions on and between genomic interval data and providing a way of performing common data analysis tasks through a coherent interface to existing Bioconductor infrastructure.
-We implement this grammar as a Bioconductor/R package called `plyranges`.
+Users must navigate these abstractions to perform a genomic analysis task, when a single data abstraction, a _GRanges_ object will suffice.
+By recognizing that the _GRanges_ class follows 'tidy' data principles, we create a grammar of genomic data transformation, defining verbs for performing actions on and between genomic interval data and providing a way of performing common data analysis tasks through a coherent interface to existing Bioconductor infrastructure.
+We implement this grammar as a Bioconductor/R package called **plyranges**.
 
@@
 computations to a database, or operate over collections of files in
-standard formats like BED. An example of the former is the Genome Query 
-Language (GQL) and its distributed implementation GenAp which use a 
-SQL-like syntax for fast retrieval of information of unprocessed 
-sequencing data [@Kozanitis2014-va; @Kozanitis2016-bm]. Similarly, the Genometric 
-Query Language (GMQL) implements a DSL for combining genomic datasets 
-[@Kaitoua2017-pw].  The command line application BEDtools develops an extensive 
-algebra for performing arithmetic between two or more sets of genomic regions 
-[@Quinlan2010-gc]. All of the aforementioned DSLs are designed to be evaluated
-either at the command line or embedded in scripts for batch processing. They 
-exist in a sparse ecosystem, mostly consisting of UNIX and database tools that
-lack biological semantics and operate at the level of files and
-database tables.
+standard formats like BED. An example of the former is the 
+**Genome Query Language (GQL)** and its distributed implementation **GenAp** 
+which use a  SQL-like syntax for fast retrieval of information of unprocessed 
+sequencing data [@Kozanitis2014-va; @Kozanitis2016-bm]. Similarly, the 
+**Genometric Query Language (GMQL)** implements a DSL for combining genomic 
+datasets [@Kaitoua2017-pw].  The command line application **BEDtools** develops 
+an extensive algebra for performing arithmetic between two or more sets of 
+genomic regions [@Quinlan2010-gc]. All of the aforementioned DSLs are 
+designed to be evaluated either at the command line or embedded in scripts 
+for batch processing. They exist in a sparse ecosystem, mostly consisting of 
+UNIX and database tools that lack biological semantics and operate at the 
+level of files and database tables.
 
-The Bioconductor/R packages `IRanges` and `GenomicRanges` 
+The Bioconductor/R packages **IRanges** and **GenomicRanges** 
 [@r-core; @Lawrence2013-wg; @Huber2015-ei] define a DSL for 
@@
 abstractions in their own code and packages. Other genomic DSLs 
-that are embedded in programming languages include pybedtools and valr 
+that are embedded in programming languages include **pybedtools** and **valr** 
 [@Dale2011-js; @Kent2017], however these packages lack the interoperability
@@
 us to see how the data relate to the design of an experiment
-and the variables measured. The `dplyr` package [@Wickham2017-dplyr] 
+and the variables measured. The **dplyr** package [@Wickham2017-dplyr] 
 defines an application programming interface (API) that maps notions from 
@@
 can be composed together on one or more tidy datasets with the pipe operator 
-from the `magrittr` package [@R-magrittr]. Taken together these features 
+from the **magrittr** package [@R-magrittr]. Taken together these features 
 enable a user to write human readable analysis workflows.
 
-We have created a genomic DSL called `plyranges` that reformulates
+We have created a genomic DSL called **plyranges** that reformulates
 notions from existing genomic algebras and embeds them in R as a
-genomic extension of `dplyr`. By analogy, `plyranges` is to the
-genomic algebra, as `dplyr` is to the relational algebra.  The
-`plyranges` Bioconductor package implements the language on top of a
+genomic extension of **dplyr**. By analogy, **plyranges** is to the
+genomic algebra, as **dply**` is to the relational algebra.  The
+**plyranges`**Bioconductor package implements the language on top of a
 key subset of Bioconductor data structures and thus fully integrates
@@
 
-(ref:GRanges-cap) An illustration of the GRanges data model for a sample from an RNA-seq experiment.
+(ref:GRanges-cap) An illustration of the _GRanges_ data model for a sample from an RNA-seq experiment.
 The core components of the data model include a seqname column (representing the chromosome), a ranges column which consists of start and end coordinates for a genomic region, and a strand identifier (either positive, negative, or unstranded).
@@
 
-The `plyranges` DSL is built on the core Bioconductor data structure
-GRanges, which is a constrained table, with fixed columns
+The **plyranges** DSL is built on the core Bioconductor data structure
+_GRanges_, which is a constrained table, with fixed columns
 for the chromosome, start and end coordinates, and the strand, along
@@
 measurements or metadata specific to the data type or experiment
-(figure \@ref(fig:GRanges)).  GRanges balances flexibility with formal
+(figure \@ref(fig:GRanges)).  _GRanges_ balances flexibility with formal
 constraints, so that it is applicable to virtually any genomic
@@
 high-level operations on genomic ranges. As a core data structure,
-GRanges enables interoperability between `plyranges` and the rest of
+_GRanges_ enables interoperability between **plyranges** and the rest of
 Bioconductor. Adhering to a single data structure simplifies the API
@@
 
-GRanges follow the intuitive tidy data pattern: it is a rectangular
+_GRanges_ follow the intuitive tidy data pattern: it is a rectangular
 table corresponding to a single biological context. Each row contains
 a single observation and each column is a variable describing the
-observations.  GRanges specializes the tidy pattern in that the
+observations.  _GRanges_ specializes the tidy pattern in that the
 observations always pertain to some genomic feature, but it largely
 remains compatible with the general relational operations defined by
-`dplyr`. Thus, we define our algebra as an extension of the `dplyr`
+**dplyr**. Thus, we define our algebra as an extension of the **dplyr**
 algebra, and borrow its syntax conventions and design principles.
@@
    & \textbf{\emph{summarize()}} & aggregate over column(s) \\ 
-   Aggregate & \emph{disjoin\_ranges()} & aggregate column(s) over the union of end coordinates \\
-   &  \emph{reduce\_ranges()} & aggregate column(s) by merging overlapping and neighboring ranges \\
+   Aggregate & \textt{disjoin\_ranges()} & aggregate column(s) over the union of end coordinates \\
+   &  \textt{reduce\_ranges()} & aggregate column(s) by merging overlapping and neighboring ranges \\
    \hline
-   &  \textbf{\emph{mutate()}} & modifies any column \\
-   & \textbf{\emph{select()}} & select columns \\
-  Modify (Unary) & \textbf{\emph{arrange()}} & sort by columns \\
-   & \emph{stretch()} & extend range by fixed amount \\
-   &  \emph{shift\_(direction)} & shift coordinates \\
-   & \emph{flank\_(direction)} & generate flanking regions \\
-   & \emph{\%intersection\% } & row-wise intersection \\
-   & \emph{\%union\%} & row-wise union \\
-   & \emph{compute\_coverage} & coverage over all ranges \\
-  Modify (Binary) &  \emph{\%setdiff\%} & row-wise set difference \\
-   & \emph{between()} & row-wise gap range \\
-   & \emph{span()} & row-wise spanning range \\
+   &  \textbf{\textt{mutate()}} & modifies any column \\
+   & \textbf{\textt{select()}} & select columns \\
+  Modify (Unary) & \textbf{\textt{arrange()}} & sort by columns \\
+   & \textt{stretch()} & extend range by fixed amount \\
+   &  \textt{shift\_(direction)} & shift coordinates \\
+   & \textt{flank\_(direction)} & generate flanking regions \\
+   & \textt{\%intersection\% } & row-wise intersection \\
+   & \textt{\%union\%} & row-wise union \\
+   & \textt{compute\_coverage} & coverage over all ranges \\
+  Modify (Binary) &  \textt{\%setdiff\%} & row-wise set difference \\
+   & \textt{between()} & row-wise gap range \\
+   & \textt{span()} & row-wise spanning range \\
    \hline
-    & \emph{join\_overlap\_*()} & merge by overlapping ranges \\
-    & \emph{join\_nearest} & merge by nearest neighbor ranges \\
-    & \emph{join\_follow} & merge by following ranges \\
-    Merge & \emph{join\_precedes} & merge by preceding ranges \\
-    & \emph{union\_ranges} & range-wise union \\
-    & \emph{intersect\_ranges} & range-wise intersect \\
-    & \emph{setdiff\_ranges} & range-wise set difference \\
-    & \emph{complement\_ranges} & range-wise set complement \\
+    & \textt{join\_overlap\_*()} & merge by overlapping ranges \\
+    & \textt{join\_nearest} & merge by nearest neighbor ranges \\
+    & \textt{join\_follow} & merge by following ranges \\
+    Merge & \textt{join\_precedes} & merge by preceding ranges \\
+    & \textt{union\_ranges} & range-wise union \\
+    & \textt{intersect\_ranges} & range-wise intersect \\
+    & \textt{setdiff\_ranges} & range-wise set difference \\
+    & \textt{complement\_ranges} & range-wise set complement \\
   \hline
-   & \emph{anchor\_direction()} & fix coordinates at direction \\
-  Operate & \textbf{\emph{group\_by()}} & partition by column(s)  \\ 
-   & \emph{group\_by\_overlaps()} & partition by overlaps \\
+   & \textt{anchor\_direction()} & fix coordinates at direction \\
+  Operate & \textbf{\textt{group\_by()}} & partition by column(s)  \\ 
+   & \textt{group\_by\_overlaps()} & partition by overlaps \\
    \hline
-   & \textbf{\emph{filter()}} & subset rows \\
-  Restrict & \emph{filter\_by\_overlaps()} & subset by overlap \\
-    & \emph{filter\_by\_non\_overlaps()} & subset by no overlap \\
+   & \textbf{\textt{filter()}} & subset rows \\
+  Restrict & \textt{filter\_by\_overlaps()} & subset by overlap \\
+    & \textt{filter\_by\_non\_overlaps()} & subset by no overlap \\
    \hline
 \end{tabular}
-\caption{Overview of the \texttt{plyranges} grammar. The core verbs are
+\caption{Overview of the \textbf{plyranges} grammar. The core verbs are
 briefly described and categorized into one of the following higher level 
 categories: aggregate, modify, merge, operate, or restrict. A verb is given bold text if
-its origin is from the \texttt{dplyr} grammar.}\label{tab:grammar}
+its origin is from the \textbf{dplyr} grammar.}\label{tab:grammar}
 \end{table}
@@
 
-The `plyranges` DSL defines an expressive algebra for performing
-genomic operations with and between GRanges objects (see table
+The **plyranges** DSL defines an expressive algebra for performing
+genomic operations with and between _GRanges_ objects (see table
 \ref{tab:grammar}). The grammar includes several classes of operation
@@
 indicating the direction: left/right or, for stranded features,
-upstream/downstream. For example, _shift\_right_ shifts a range to the
+upstream/downstream. For example, `shift_right()` shifts a range to the
 right.
@@
 _difference_, and _complement_. There are two sets of these: parallel
-and merging. For example, the parallel intersection (_x %intersect% y_) finds the
+and merging. For example, the parallel intersection (`x %intersect% y`) finds the
 intersecting range between _xi_ and _yi_ for _i_ in _1...n_, where _n_
 is the length of both _x_ and _y_. In contrast, the merging
-intersection (_intersect\_ranges(x, y)_) returns a new set of disjoint
+intersection ( `intersect_ranges(x, y)`) returns a new set of disjoint
 ranges representing wherever there was overlap between a range in _x_
 and a range in _y_. Finding the parallel union will fail when two ranges
-have a gap, so we introduce a _span_ operator that takes the union
-while filling any gap. The _complement_ operation is unique in that it
+have a gap, so we introduce a `span()` operator that takes the union
+while filling any gap. The `complement()` operation is unique in that it
 is unary. It finds the regions not covered by any of the ranges in a
-single set. Closely related is the _between_ parallel operation, which
+single set. Closely related is the `between()` parallel operation, which
 finds the gap separating _xi_ and _yi_. The binary operations are
@@
 (ref:olaps-cap) Illustration of the three overlap join operators.
-Each join takes two  GRanges objects, \textit{x} and \textit{y} as input.
+Each join takes two  _GRanges_ objects, \textit{x} and \textit{y} as input.
 A 'Hits' object for the  join is computed which consists of two components.
@@
 The resulting 'Hits' object is used to modify \textit{x} by where it was 'hit' by \textit{y} and merge all metadata columns from \textit{x} and \textit{y} based on the indices contained in the 'Hits' object.
-This procedure is applied generally in the \texttt{plyranges} DSL for both overlap and nearest neighbor operations.
+This procedure is applied generally in the \textbf{plyranges} DSL for both overlap and nearest neighbor operations.
 The join semantics alter what is returned: \textbf{A}: for an \textbf{inner} join the \textit{x} ranges that are overlapped by \textit{y} are returned.
@@
 genomic regions as variants of the relational join operator. A join
-acts on two GRanges objects: _x_ and _y_. The join operator
+acts on two _GRanges_ objects: _x_ and _y_. The join operator
 is relational in the sense that metadata from the _x_ and _y_
@@
 
-(ref:code-comparison-cap) Idiomatic code examples for \texttt{plyranges} (\textbf{A}) and \texttt{GenomicRanges} (\textbf{B}) illustrating an overlap and aggregate operation that returns the same result.
+(ref:code-comparison-cap) Idiomatic code examples for \textbf{plyranges} (\textbf{A}) and \textbf{GenomicRanges} (\textbf{B}) illustrating an overlap and aggregate operation that returns the same result.
 In each example, we have two BED files consisting of SNPs that are genome-wide association study (GWAS) hits and reference exons.
 Each code block counts for each SNP the number of distinct exons it overlaps.
-The \texttt{plyranges} code achieves this with an overlap join followed by partitioning and aggregation.
+The \textbf{plyranges} code achieves this with an overlap join followed by partitioning and aggregation.
 Strand is ignored by default here.
-The \texttt{GenomicRanges} code achieves this using the 'Hits' and 'List' classes and their methods.
+The \textbf{GenomicRanges} code achieves this using the _Hits_ and _List_ classes and their methods.
 
-Since the GRanges object is a tabular data structure, our grammar includes
-operators to filter, sort and aggregate by columns in a GRanges. These 
+Since the _GRanges_ object is a tabular data structure, our grammar includes
+operators to filter, sort and aggregate by columns in a _GRanges_. These 
 operations can be performed over partitions formed
-using the *group_by* modifier. Together with our algebra for arithmetic and
+using the `group_by()` modifier. Together with our algebra for arithmetic and
 merging, these operations conform to the semantics and syntax of the 
-`dplyr` grammar. Consequently, `plyranges` code is generally more compact
-than the equivalent `GenomicRanges` code (figure \@ref(fig:code-comparison)).
+**dplyr** grammar. Consequently, **plyranges** code is generally more compact
+than the equivalent **GenomicRanges** code (figure \@ref(fig:code-comparison)).
 
-### Developing workflows with `plyranges`
+### Developing workflows with **plyranges**
 
-Here we provide illustrative examples of using the `plyranges` DSL to show
+Here we provide illustrative examples of using the **plyranges** DSL to show
 how our grammar could be integrated into genomic data workflows. As we
@@
 In the workflow of ChIP-seq data analysis, we are interested in finding peaks 
-from islands of coverage over chromosome. Here we will use `plyranges`
+from islands of coverage over chromosome. Here we will use **plyranges**
 to call peaks from islands of coverage above 8 then plot the region
@@
 
-Using `plyranges` and the the Bioconductor package `AnnotationHub` [@R-ahub]
+Using **plyranges** and the the Bioconductor package **AnnotationHub** [@R-ahub]
 we can download and read BigWig files from ChIP-Seq experiments from 
@@
 
-```{r load-bw, echo = TRUE}
+```{r load-bw, echo = TRUE }
 ```
@@
 
-```{r chr10-scores-read, echo = TRUE}
+```{r chr10-scores-read, echo = TRUE }
 ```
@@
 
-```{r chr10-scores-reduce, echo = TRUE}
+```{r chr10-scores-reduce, echo = TRUE }
 ```
 
-Returning to the GRanges object containing normalized coverage scores, 
+Returning to the _GRanges_ object containing normalized coverage scores, 
 we filter to find the coordinates of the peak containing the maximum coverage 
@@
 
-```{r max-score-region}
+```{r max-score-region }
 ```
@@
 
-```{r peak_region}
+```{r peak_region }
 ```
 
-(ref:peak-viz-cap) The final result of the `plyranges` operations to find a 5000nt region surrounding the peak of normalised coverage scores over chromosome 10, displayed as a density plot.
+(ref:peak-viz-cap) The final result of the **plyranges** operations to find a 5000nt region surrounding the peak of normalised coverage scores over chromosome 10, displayed as a density plot.
 
@@
 Another common operation in genomics data analysis is to compute data
-summaries over genomic windows. In `plyranges` this can be achieved
+summaries over genomic windows. In **plyranges** this can be achieved
 via the `group_by_overlaps()` operator.  We bin and count and
@@
 are loaded into memory. We can add another column representing the GC proportion 
-for each alignment using the `letterFrequency()` function from the `Biostrings`
+for each alignment using the `letterFrequency()` function from the **Biostrings**
 package [@R-biostrings]. After computing the GC proportion as the score
-column, we drop all other columns in the GRanges object.
+column, we drop all other columns in the _GRanges_ object.
 
@@
 
-We have created a GRanges object from genotyping performed on the
+We have created a _GRanges_ object from genotyping performed on the
 H1 cell line, consisting of approximately two million single nucleotide
-polymorphisms (SNPs) and short insertion/deletions (indels). The GRanges object consists
+polymorphisms (SNPs) and short insertion/deletions (indels). The _GRanges_ object consists
 of 7 columns, relating to the alleles of a SNP or indel, the B-allele frequency,
@@
 name of the probe. We can use this information to compute the transition-transversion
-ratio, a quality control metric, within each chromosome in GRanges object.
+ratio, a quality control metric, within each chromosome in _GRanges_ object.
 
@@
 
-(ref:titv-viz-cap) The final result of computing quality control metrics over the SNP array data with `plyranges`, displayed as a dot plot.
+(ref:titv-viz-cap) The final result of computing quality control metrics over the SNP array data with **plyranges**, displayed as a dot plot.
 Chromosomes are ordered by their estimated transition-transversion ratio.
@@
 
-The design of `plyranges` adheres to well understood principles of
+The design of **plyranges** adheres to well understood principles of
 language and API design: cognitive consistency, cohesion,
@@
 To varying degrees, these principles also underlie the
-design of `dplyr` and the Bioconductor infrastructure.
+design of **dplyr** and the Bioconductor infrastructure.
 
-We have aimed for `plyranges` to have a simple and direct mapping to the
+We have aimed for **plyranges** to have a simple and direct mapping to the
 user's cognitive model, i.e., how the user thinks about the data. This
@@
 user can express workflows in the language of genomics. This motivates
-the adoption of the tidy GRanges object as our central data
-structure. The basic data.frame and `dplyr` tibble lack any notion of
+the adoption of the tidy _GRanges_ object as our central data
+structure. The basic _data.frame_ and **dplyr** _tibble_ lack any notion of
 genomic ranges and so could not easily support our genomic grammar,
 with its specific verbs for range-oriented data manipulation. Another
-example of cognitive consistency is how `plyranges` is insensitive to
+example of cognitive consistency is how **plyranges** is insensitive to
 direction/strand by default when, e.g., detecting
-overlaps. `GenomicRanges` has the opposite behavior. We believe that
+overlaps. **GenomicRanges** has the opposite behavior. We believe that
 defaulting to purely spatial overlap is most intuitive to most users.
 
-To further enable cognitive consistency, `plyranges` functions are cohesive. 
+To further enable cognitive consistency, **plyranges** functions are cohesive. 
 A function is defined to be cohesive if it performs a singular task without
@@
 position (start, end, midpoint) should be invariant over the
-transformation. The `resize()` function from the `GenomicRanges` package
+transformation. The `resize()` function from the **GenomicRanges** package
 has a `fix` argument that sets the anchor, so calling `resize()`
@@
 function call level is justified since the effect of setting the width
-depends on the anchor. However, `plyranges` increases cohesion and
+depends on the anchor. However, **plyranges** increases cohesion and
 decouples the anchoring into its own function call.
@@
 of function and argument names. This results in the
-user being able to conceptualize the `plyranges` DSL as a flat catalog of 
+user being able to conceptualize the **plyranges** DSL as a flat catalog of 
 functions, without having to descend further into documentation to understand
@@
 
-Like `dplyr`, `plyranges` verbs are functional: they are free of side
-effects and are generally endomorphic, meaning that when the input is a GRanges
-object they return a GRanges object. This enables chaining of verbs
-through syntax like the forward pipe operator from the `magrittr` package. 
+Like **dplyr**, **plyranges** verbs are functional: they are free of side
+effects and are generally endomorphic, meaning that when the input is a _GRanges_
+object they return a _GRanges_ object. This enables chaining of verbs
+through syntax like the forward pipe operator from the **magrittr** package. 
 This syntax has a direct cognitive mapping to natural language and the intuitive notion of
@@
 concise than the multiple steps necessary to achieve the same effect
-in the original `GenomicRanges` API. In other cases, the `plyranges` API
+in the original **GenomicRanges** API. In other cases, the **plyranges** API
 increases verbosity for the sake of clarity and cohesion. Explicitly
-calling _anchor()_ can require more typing, but the code is easier to
+calling `anchor()` can require more typing, but the code is easier to
 comprehend. Another example is the set of routines for importing
-genomic annotations, including _read\_gff()_, _read\_bed()_, and
-_read\_bam()_. Compared to the generic _import()_ in `rtracklayer`, the
-explicit format-based naming in `plyranges` clarifies intent and the
-type of data being returned. Similarly, every `plyranges` function that
+genomic annotations, including `read_gff()`, `read_bed()`, and
+`read_bam()`. Compared to the generic `import()` in **rtracklayer**, the
+explicit format-based naming in **plyranges** clarifies intent and the
+type of data being returned. Similarly, every **plyranges** function that
 computes with strand information indicates its intentions by including
 suffixes such as _directed_, _upstream_ or _downstream_ in its name,
-otherwise strand is ignored. The `GenomicRanges` API does not make this
+otherwise strand is ignored. The **GenomicRanges** API does not make this
 distinction explicit in its function naming, instead relying on a
@@
 
-The implementation of `plyranges` is built on top of 
+The implementation of **plyranges** is built on top of 
 Bioconductor infrastructure, meaning most functions are constructed
 by composing generic functions from core Bioconductor packages. As a result, any 
-Bioconductor packages that uses data structures that inherit from GRanges will
-be able to use `plyranges`  for free. Another consequence of building on top of
-Bioconductor generics is that the speed and memory usage of `plyranges` 
+Bioconductor packages that uses data structures that inherit from _GRanges_ will
+be able to use **plyranges**  for free. Another consequence of building on top of
+Bioconductor generics is that the speed and memory usage of **plyranges** 
 functions are similar to the highly optimized methods implemented in Bioconductor
-for GRanges objects.
+for _GRanges_ objects.
 
-A caveat to constructing a compatible interface with `dplyr` is that
-`plyranges` makes extensive use of non-standard evaluation in R via
-the `rlang` package [@R-rlang]. Simply, this means that computations are
-evaluated in the context of the GRanges objects. Both `dplyr` and
-`plyranges` are based on the `rlang` language, because it allows for more
+A caveat to constructing a compatible interface with **dplyr** is that
+**plyranges** makes extensive use of non-standard evaluation in R via
+the **rlang** package [@R-rlang]. Simply, this means that computations are
+evaluated in the context of the _GRanges_ objects. Both **dplyr** and
+**plyranges** are based on the **rlang** language, because it allows for more
 expressive code that is free of repeated references to the
@@
 convenient when programming interactively.  Consequently, when
-programming with `plyranges`, a user needs to generally understand
-the `rlang` language and how to adapt their code accordingly. Users
+programming with **plyranges**, a user needs to generally understand
+the **rlang** language and how to adapt their code accordingly. Users
 familiar with the tidyverse should already have such knowledge.
@@
 We have shown how to create expressive and 
-reproducible genomic workflows using the `plyranges` DSL. By
-realising that the GRanges data model is tidy we have highlighted how to 
+reproducible genomic workflows using the **plyranges** DSL. By
+realising that the _GRanges_ data model is tidy we have highlighted how to 
 implement a grammar for performing genomic arithmetic, aggregation, restriction 
-and merging. Our examples show that `plyranges` code is succinct, human
+and merging. Our examples show that **plyranges** code is succinct, human
 readable and can take advantage of the interoperability provided by
@@
 structure. We chose R because it is what we are familiar with and because
-the aforementioned Bioconductor packages have implemented the GRanges data
+the aforementioned Bioconductor packages have implemented the _GRanges_ data
 structure.
 
-We aim to continue developing the `plyranges` package and to extend it
+We aim to continue developing the **plyranges** package and to extend it
 for use with more complex data structures, such as the
-SummarizedExperiment class, the core Bioconductor data structure for
+_SummarizedExperiment_ class, the core Bioconductor data structure for
 representing experimental results (e.g., counts) from multiple sample
 experiments in conjunction with feature and sample metadata. Although,
-the SummarizedExperiment is not strictly tidy, it does consist of 
+the _SummarizedExperiment_ is not strictly tidy, it does consist of 
 three tidy data structures that are related by feature and sample identifiers. 
-Therefore, the grammar and design of the `plryanges` DSL is naturally 
-extensible to the SummarizedExperiment.
+Therefore, the grammar and design of the **plryanges** DSL is naturally 
+extensible to the _SummarizedExperiment_.
 
-As the `plyranges` interface encourages tidy data practices, it
+As the **plyranges** interface encourages tidy data practices, it
 integrates well with the grammar of graphics [@Wickham2016-gz].  To
@@
 access and computing patterns, so the deferred mechanisms within
-`plyranges` should help support interactive genomics applications.
-
-## Acknowledgements
-
-We would like to thank Dr Matthew Ritchie at the Walter and Eliza Hall 
-Institute and Dr Paul Harrison for their feedback on earlier drafts of
-this work. We would also like to thank Lori Shepherd and Hèrve Pages
-for the code review they performed and users who have submitted feedback
-and pull requests.
+**plyranges** should help support interactive genomics applications.
 
@@
 blood ChIP-seq data from the Human Roadmap Epigenomics project  
-was downloaded from the `AnnotationHub` package (2.13.1)
+was downloaded from the **AnnotationHub** package (2.13.1)
 under accession AH33458 [@R-ahub; @Roadmap-Epigenomics-Consortium2015-pr]. 
@@
 
-The `plyranges` package is open source under an Artistic 2.0 license 
+The **plyranges** package is open source under an Artistic 2.0 license 
 [@R-plyranges-z]. The software 
@@
 [https://github.com/sa-lee/plyranges](https://github.com/sa-lee/plyranges).
+
+## Acknowledgements {-}
+
+We would like to thank Dr Matthew Ritchie at the Walter and Eliza Hall 
+Institute and Dr Paul Harrison for their feedback on earlier drafts of
+this work. We would also like to thank Lori Shepherd and Hèrve Pages
+for the code review they performed and users who have submitted feedback
+and pull requests.

### 2020-07-26 Rmd/04-intron.Rmd
tidy up chapter 4

@@
 
-In this paper, we describe methodology for establishing data descriptors by turning coverage vectors into long form tidy data using **superintronic**
+In this chapter, we describe methodology for establishing data descriptors by turning coverage vectors into long form tidy data using **superintronic**
 and **plyranges**. We provide a workflow below
@@
 
-**superintronic** is an R package available on GitHub at https://github.com/sa-lee/superintronic and used for estimating, representing,
+**superintronic** is an R package used for estimating, representing,
 and visualising per-base coverage scores, that can then be 
@@
 along with an optional experimental design table that is returns 
-a long-form tidy **GRanges** data structure. The coverage estimation is
+a long-form tidy _GRanges_ data structure. The coverage estimation is
 computed via the **Rsamtools** package, and
@@
 in the genome where there is no coverage [@rsamtools; @granges]. This
-representation is tidy, since each row of the resulting **GRanges** data
+representation is tidy, since each row of the resulting _GRanges_ data
 structure corresponds to position(s) within a given sample with a given 
 coverage score alongside any variables such as biological group. While the long-form representation repeats the same information
-for a sample within the design, the size of the resulting **GRanges** in
+for a sample within the design, the size of the resulting _GRanges_ in
 memory can be compressed using run-length encoding for any categorical 
@@
 External reference annotations, perhaps transcripts or exons, 
-can be coerced to **GRanges** objects, are  incorporated into the coverage **GRanges** by taking the intersection of
-the annotation with the **GRanges** using an overlap intersect join from the **plyranges** software. The resulting intersection will now contain the
+can be coerced to _GRanges_ objects, are  incorporated into the coverage _GRanges_ by taking the intersection of
+the annotation with the _GRanges_ using an overlap intersect join from the **plyranges** software. The resulting intersection will now contain the
 per base coverage that are overlapped the genomic features in the annotation,
@@
 provides some syntactic sugar for unravelling gene annotations into their exonic
-and intronic parts, and intersecting them with a coverage **GRanges**.
+and intronic parts, and intersecting them with a coverage _GRanges_.
 
@@
 
-Once the coverage **GRanges** has reference genomic features, data 
+Once the coverage _GRanges_ has reference genomic features, data 
 descriptors can be computed via collecting summary statistics across
@@
 thresholds can be applied to filter the genomic features that had extreme
-descriptors on the coverage **GRanges**, and the traces can be visualised.
+descriptors on the coverage _GRanges_, and the traces can be visualised. 
+By default **superintronic** displays coverage traces oriented from the 5' to 3' end of the gene, with the `view_coverage()` function.
+Traces can be highlighted according to a genomic feature of interest, figures in this chapter
+have orange areas corresponding to intronic parts 
+of the gene, while dark green areas referring to exonic parts.
 
@@
 Minor class splicing is a regulatory process where a class of 
-introns (called U12 introns in this case but there exist other classes in eukaryotic organisms such as U2 introns) are removed from pre-mRNA prior to gene expression.
+introns (called U12 introns in this case but there exist other classes in eukaryotic organisms such as U2 introns) are removed from pre-mRNA prior to gene expression [@Turunen2013-qu].
 The removal of these introns is catalysed by small ribonucleoproteins (snRNPs)
@@
 Here we describe an exploratory workflow for finding coverage traces
-with evidence for intron retention using RNA-seq data from a knockout experiment in zebrafish obtained from the Heath Lab at WEHI.
+with evidence for intron retention using RNA-seq data from a knockout experiment in zebrafish obtained from the Heath Lab at Walter and Eliza Hall Institute. Code
+for this analysis is available at https://github.com/sa-lee/thesis/tree/master/scripts/superintronic.R.
 
@@
 sample [@Liao2013-eb; @Liao2019-qn]. The coverage was then estimated directly from the set of BAM files
-using **superintronic** into the long form *GRanges* representation we described above.
+using **superintronic** into the long form _GRanges_ representation we described above.
 
@@
 obtained as GFF files from RefSeq and used that to construct the exonic
-and intronic parts of each gene as **GRanges** object using *superintronic*. 
+and intronic parts of each gene as _GRanges_ object using *superintronic*. 
 We further filtered genes that had a single exon or genes that were overlapping
@@
 computing data descriptors on in order to detecting IR-like coverage traces.
-Across each combination of genotype and line, we first log$_2$ transformed the coverage score with an offset of one, and then intersected the coverage *GRanges* with exonic and intronic features of each gene.
+Across each combination of genotype and line, we first $log_2$ transformed the coverage score with an offset of one, and then intersected the coverage _GRanges_ with exonic and intronic features of each gene.
 For each gene,
@@
 
-We decided to select genes with the following thresholds:  genes have an average exon log-coverage greater than the mean of average exon log-coverage values across all genes, have an average intron log-coverage greater than the average standard deviation of exon log-coverage values over all genes, and
+We selected genes with the following thresholds:  genes have an average exon log-coverage greater than the mean of average exon log-coverage values across all genes, have an average intron log-coverage greater than the average standard deviation of exon log-coverage values over all genes, and
 the standard deviation of intron log-coverage values is twice the standard
@@
 The rnpc3 knockout lines
+share the largest overlap in results.
 
@@
 
-(ref:si-cvg-ccdc43) Coverage traces faceted by biological group for the zebrafish knockout experiment, oriented from the 5' to 3' end of the gene.
-Orange areas refer to intronic parts of the gene, while dark green areas refer to the exonic part.
-The ccdc43 gene is known to have enhanced U12 intron retention in the caliban phenotype and increase retention when rnpc3 is knocked out as can be seen directly from the intron located at around 31.603Mb.
+(ref:si-cvg-ccdc43) The ccdc43 gene is known to have enhanced U12 intron retention in the caliban phenotype and increase retention when rnpc3 is knocked out as can be seen directly from the intron located at around 31.603Mb.
 
@@
 or combine our data descriptors with external estimates using **limma** 
-[@Ritchie2015-mj], like our proposed index method in @Lee2019-mf.
+[@Ritchie2015-mj], like our proposed index method in @Lee2019-mf. The gene
+candiates obtain by our thresholds have been validated by the Heath lab
+using qPCR.
 
@@
 data descriptors could be generalised to other types of omics analyses,
-and to use more sophisticated methods for identifying 
-thresholds of 'interesting' traces. Our approach would also greatly benefit
-from interactive graphics that dynamically link say a gene description to
-it's underlying coverage trace, for rapid exploration. This is left for 
+such as peaking finding in ChIP-seq, and to use more sophisticated methods for identifying  thresholds of 'interesting' traces.
+Our approach would also greatly benefit from interactive graphics that dynamically link say a gene description to
+an underlying coverage trace, for rapid exploration. This is left for 
 future work.
 
-## Acknowledgements
+## Acknowledgements {-}
 
-We would like to thank Alexandra Garnham, Stephen Mieruszynski and Joan Heath
+We would like to thank Dr Alexandra Garnham, Dr Stephen Mieruszynski and 
+Associate Professor Joan Heath
 for providing the zebrafish data and helping interpret the results from the workflow.

### 2020-07-26 scripts/superintronic.R

@@ -1,0 +1,350 @@
+## ----setup, include=FALSE-------------------
+# pre load libraries
+suppressPackageStartupMessages(library(plyranges))
+suppressPackageStartupMessages(library(superintronic))
+library(ggplot2)
+
+# use hexbinning for pairs plot
+hex <- function(data, mapping, ...) {
+  ggplot2::ggplot(data = data, mapping = mapping) +
+    ggplot2::geom_hex(...) +
+    ggplot2::scale_fill_viridis_c(direction = -1)
+}
+
+# plot function for pretty coverage traces
+pretty_cov_plot <- function(cvg, parts, target, design = NULL, alpha = FALSE, ...) {
+
+  if(is(target, "data.frame")) {
+    target <- plyranges::filter(parts, gene_id == !!unique(target$gene_id))
+  }
+  cvg_rng <- set_plot_rng(cvg, target, design, alpha)
+
+  if (alpha) {
+    p <- ggplot(as.data.frame(cvg_rng),
+                aes(x = start, xend = end, y = 0, yend = log_score)) +
+      geom_segment(aes(colour = feature_type, alpha = alpha)) +
+      superintronic:::rescale_by_width(cvg_rng) +
+      guides(alpha = FALSE)
+  } else {
+    p <- superintronic::view_coverage(cvg_rng,
+                                      score = log_score,
+                                      colour = feature_type,
+                                      facets = dplyr::vars(var))
+  }
+
+  p <- p +
+    ylim(0, NA) +
+    guides(colour = FALSE) +
+    scale_color_brewer(palette = "Dark2") +
+    labs(subtitle = paste("Coverage over", target$gene_id),
+         y = "Log coverage"
+    ) +
+    theme_bw() +
+    theme(axis.text.x = element_blank(),
+          axis.ticks.x = element_blank(),
+          strip.text = element_text(hjust = 0, size = 8),
+          strip.background = element_blank()) +
+    expand_limits(y = 0)
+
+  segments <- superintronic::unnest_parts(target)
+  track <- view_segments(segments, colour = feature_type) +
+    theme_bw() +
+    theme(axis.text.x = element_text(size = 8), axis.title.x = element_blank(),
+          axis.title.y = element_blank(), axis.text.y = element_blank())
+
+  patchwork::wrap_plots(p,
[…]

### 2020-07-26 Rmd/03-fluentGenomics.Rmd
tidy chapter 3

@@
-# Fluent genomics with _plyranges_ and _tximeta_ {#ch:fluentGenomics}
+# Fluent genomics with **plyranges** and **tximeta** {#ch:fluentGenomics}
 
@@
 Here we show how to implement these steps to integrate published RNA-seq and ATAC-seq experiments on macrophage cell lines.
-Using tximeta, we import RNA-seq transcript quantifications into an analysis-ready data structure, called the SummarizedExperiment, that contains the ranges of the reference transcripts and metadata on their provenance.
-Using SummarizedExperiments to represent the ATAC-seq and RNA-seq data, we model differentially accessible (DA) chromatin peaks and differentially expressed (DE) genes with existing Bioconductor packages.
-Using plyranges we then integrate the results to see if there is an enrichment of DA peaks near DE genes by finding overlaps and aggregating over log-fold change thresholds.
+Using **tximeta**, we import RNA-seq transcript quantifications into an analysis-ready data structure, called the _SummarizedExperiment_, that contains the ranges of the reference transcripts and metadata on their provenance.
+Using _SummarizedExperiments_ to represent the ATAC-seq and RNA-seq data, we model differentially accessible (DA) chromatin peaks and differentially expressed (DE) genes with existing Bioconductor packages.
+Using **plyranges** we then integrate the results to see if there is an enrichment of DA peaks near DE genes by finding overlaps and aggregating over log-fold change thresholds.
 The combination of these packages and their integration with the Bioconductor ecosystem provide a coherent framework for analysts to iteratively and reproducibly explore their biological data.
@@
 @alasoo, a study that involved treatment of macrophage cell lines from a number
-of human donors with interferon gamma (IFNg), *Salmonella* infection, or both
+of human donors with interferon gamma (IFNg), Salmonella infection, or both
 treatments combined. @alasoo examined gene expression and chromatin
@@
 (ref:workflow) An overview of the fluent genomics workflow.
-First, we *import* data as a *SummarizedExperiment* object, which enables interoperability with downstream analysis packages.
-Then we *model* our assay data, using the existing Bioconductor packages `DESeq2` and `limma`.
+First, we *import* data as a _SummarizedExperiment_ object, which enables interoperability with downstream analysis packages.
+Then we *model* our assay data, using the existing Bioconductor packages **DESeq2** and **limma**.
 We take the results of our models for each assay with respect to their genomic coordinates, and *integrate* them.
@@
 Bioconductor infrastructure to understand these datasets. In particular, we
-will emphasize the use of the Bioconductor packages `plyranges` and `tximeta`.
-The `plyranges` package fluently transforms data tied to genomic ranges using
+will emphasize the use of the Bioconductor packages **plyranges** and **tximeta**.
+The **plyranges** package fluently transforms data tied to genomic ranges using
 operations like shifting, window construction, overlap detection, etc. It is
 described by @Lee2019 and leverages underlying core Bioconductor infrastructure
-[@granges; @bioc] and the `tidyverse` design principles @tidyverse.
+[@granges; @bioc] and the tidyverse design principles @tidyverse.
 
-The `tximeta` package described by @Love2019-tximeta is used to read RNA-seq
+The **tximeta** package described by @Love2019-tximeta is used to read RNA-seq
 quantification data into R/Bioconductor, such that the transcript ranges and
@@
 
-The data used in this workflow is available from two packages: the `macrophage`
+The data used in this workflow is available from two packages: the **macrophage**
 Bioconductor ExperimentData package and from the workflow package 
-`fluentGenomics`.
+**fluentGenomics**.
 
-The `macrophage` package contains RNA-seq quantification from 24 RNA-seq
+The **macrophage** package contains RNA-seq quantification from 24 RNA-seq
 samples, a subset of the RNA-seq samples generated and analyzed by @alasoo. The
-paired-end reads were quantified using *Salmon* [@salmon], using the Gencode 29
+paired-end reads were quantified using **Salmon** [@salmon], using the Gencode 29
 human reference transcripts [@gencode]. For more details on quantification, and
 the exact code used, consult the vignette of the
-[macrophage](http://bioconductor.org/packages/macrophage) package. The package
-also contains the `Snakemake` file that was used to distribute the *Salmon*
+**macrophage** package. The package
+also contains the **Snakemake** file that was used to distribute the **Salmon**
 quantification jobs on a cluster [@snakemake].
 
-The `fluentGenomics` package contains functionality to download and generate a
-cached *SummarizedExperiment* object from the normalized ATAC-seq data provided
+The **fluentGenomics** package contains functionality to download and generate a
+cached _SummarizedExperiment_ object from the normalized ATAC-seq data provided
 by @alasooZenodo. This object contains all 145 ATAC-seq samples across all
@@
 The following code loads the path to the cached data file, or if it is not
-present, will create the cache and generate a *SummarizedExperiment* using the
-the `BiocFileCache` package [@bcfilecache].
+present, will create the cache and generate a _SummarizedExperiment_ using the
+the **BiocFileCache** package [@bcfilecache].
 
@@
 
-A precise description of how we obtained this *SummarizedExperiment* object can be found in section \@ref(atac).
+A precise description of how we obtained this _SummarizedExperiment_ object 
+can be found in section \@ref(atac).
 
@@
 
-### Using `tximeta` to import RNA-seq quantification data
+### Using **tximeta** to import RNA-seq quantification data
 
@@
 where the path is relative to your current R session. However, in this case we
-have distributed the files in the `macrophage` package. The relevant directory
-and associated files can be located using `system.file`.
+have distributed the files in the **macrophage** package. The relevant directory
+and associated files can be located using `system.file()`.
 
@@
 Information about the experiment is contained in the `coldata.csv` file. We
-leverage the `dplyr` and `readr` packages (as part of the `tidyverse`) to read
-this file into R [@tidyverse]. We will see later that `plyranges` extends these
+leverage the **dplyr** and **readr** packages (as part of the **tidyverse**) to read
+this file into R [@tidyverse]. We will see later that **plyranges** extends these
 packages to accommodate genomic ranges.
@@
 coldata <- read_csv(colfile) %>%
-  dplyr::select(
+  select(
     names,
@@
   ) %>%
-  dplyr::mutate(
+  mutate(
     files = file.path(dir, "quants", names, "quant.sf.gz"),
@@
 quantifications for each observation -- these files have been gzipped, but
-would typically not have the 'gz' ending if used from *Salmon* directly. One
+would typically not have the 'gz' ending if used from **Salmon** directly. One
 other thing to note is the use of the pipe operator,`%>%`, which can be read as
@@
 the quantifications. The following lines of code do a lot of work for the
-analyst: importing the RNA-seq quantification (dropping *inferential
-replicates* in this case), locating the relevant reference transcriptome,
+analyst: importing the RNA-seq quantification (dropping *inferential replicates* 
+in this case), locating the relevant reference transcriptome,
 attaching the transcript ranges to the data, and fetching genome information.
@@
 
-The result is a *SummarizedExperiment* object.
+The result is a _SummarizedExperiment_ object.
 
@@
 On a machine with a working internet connection, the above command works
-without any extra steps, as the `tximeta` function obtains any necessary
-metadata via FTP, unless it is already cached locally. The `tximeta` package
+without any extra steps, as the **tximeta** function obtains any necessary
+metadata via FTP, unless it is already cached locally. The **tximeta** package
 can also be used without an internet connection, in this case the linked
@@
 
-Because `tximeta` knows the correct reference transcriptome, we can ask
-`tximeta` to summarize the transcript-level data to the gene level using the
+Because **tximeta** knows the correct reference transcriptome, we can ask
+**tximeta** to summarize the transcript-level data to the gene level using the
 methods of @Soneson2015.
@@
 
-One final note is that the `start` of positive strand genes and the `end` of
+One final note is that the _start_ of positive strand genes and the _end_ of
 negative strand genes is now dictated by the genomic extent of the isoforms of
-the gene (so the `start` and `end` of the reduced *GRanges*). Another
+the gene (so the _start_ and _end_ of the reduced *GRanges*). Another
 alternative would be to either operate on transcript abundance, and perform
@@
 
-The *SummarizedExperiment* object containing ATAC-seq peaks can be created from
+The _SummarizedExperiment_ object containing ATAC-seq peaks can be created from
 the following tab-delimited files from @alasooZenodo:
@@
 
-The ATAC-seq counts have already been normalized with *cqn* [@Hansen2012] and
-log2 transformed. Loading the *cqn*-normalized matrix of log2 transformed read
+The ATAC-seq counts have already been normalized with **cqn** [@Hansen2012] and
+$log_2$ transformed. Loading the **cqn**-normalized matrix of $log_2$ transformed read
 counts takes ~30 seconds and loads an object of ~370 Mb. We set the column
@@
 
-We read in the peak metadata (locations in the genome), and convert it to a
-*GRanges* object. The `as_granges()` function automatically converts the
-*data.frame* into a *GRanges* object. From that result, we extract the peak_id
+We read in the peak metadata (locations in the genome), and convert it to a _GRanges_ object. The `as_granges()` function automatically converts the
+_data.frame_ into a _GRanges_ object. From that result, we extract the peak_id
 column and set the genome information to the build "GRCh38". We know this from
@@
 
-Finally, we construct a *SummarizedExperiment* object.  We place the matrix
+Finally, we construct a _SummarizedExperiment_ object.  We place the matrix
 into the assays slot as a named list, the annotated peaks into the row-wise
@@
 
-We can easily run a differential expression analysis with `DESeq2` using the
+We can easily run a differential expression analysis with **DESeq2** using the
 following code chunks [@Love2014]. The design formula indicates that we want to
-control for the donor baselines (`line`) and test for differences in gene
+control for the donor baselines (_line_) and test for differences in gene
 expression on the condition. For a more comprehensive discussion of DE
@@
 
-We now output the results as a *GRanges* object, and due to the conventions of
-`plyranges`, we construct a new column called `gene_id` from the row names of
-the results. Each row now contains the genomic region (`seqnames`, `start`,
-`end`, `strand`) along with corresponding metadata columns (the `gene_id` and
-the results of the test). Note that `tximeta` has correctly identified the
-reference genome as "hg38", and this has also been added to the *GRanges* along
+We now output the results as a _GRanges_ object, and due to the conventions of
+**plyranges**, we construct a new column called _gene\_id_ from the row names of
+the results. Each row now contains the genomic region (_seqnames_, _start_,
+_end_, _strand_) along with corresponding metadata columns (the _gene_id_ and
+the results of the test). Note that **tximeta** has correctly identified the
+reference genome as "hg38", and this has also been added to the **GRanges** along
 the results columns. This kind of book-keeping is vital once overlap operations
-are performed to ensure that `plyranges` is not comparing across incompatible
+are performed to ensure that **plyranges** is not comparing across incompatible
 genomes.
@@
 absolute value. To visualize the result of this test, you can run `results`
-without `format="GRanges"`, and pass this object to `plotMA` as before. We
+without `format="GRanges"`, and pass this object to `plotMA()` as before. We
 label these genes as `other_genes` and later as "non-DE genes", for comparison
@@
 The following section describes the process we have used for generating a
-*GRanges* object of differential peaks from the ATAC-seq data in @alasoo.
+_GRanges_ object of differential peaks from the ATAC-seq data in @alasoo. The 
+code chunks for the remainder of this section are not run.
 
-The code chunks for the remainder of this section are not run.
-
-For assessing differential accessibility, we run `limma` [@Smyth2004], and
+For assessing differential accessibility, we run **limma** [@Smyth2004], and
 generate the a summary of LFCs and adjusted p-values for the peaks:
@@
 
-We now take the `rowRanges` of the *SummarizedExperiment* and attach the LFCs
-and adjusted p-values from `limma`, so that we can consider the overlap with
+We now take the `rowRanges()` of the _SummarizedExperiment_ and attach the LFCs
+and adjusted p-values from **limma**, so that we can consider the overlap with
 differential expression. Note that we set the genome build to "hg38" and
@@
 
-The final *GRanges* object containing the DA peaks is included in the workflow
-package and can be loaded as follows:
+The final _GRanges_ object containing the DA peaks is included in 
+the **fluentGenomics** and can be loaded as follows:
 
@@
 
-### Finding overlaps with `plyranges`
+### Finding overlaps with **plyranges**
 
-We have already used `plyranges` a number of times above, to `filter`,
-`mutate`, and `select` on *GRanges* objects, as well as ensuring the correct
-genome annotation and style has been used. The `plyranges` package provides a
+We have already used **plyrange**  a number of times above, to `filter()`,
+`mutate()`, and `select()` on _GRanges_ objects, as well as ensuring the correct
+genome annotation and style has been used. The **plyranges** package provides a
 grammar for performing transformations of genomic data [@Lee2019]. Computations
-resulting from compositions of `plyranges` "verbs" are performed using
-underlying, highly optimized range operations in the `GenomicRanges` package
+resulting from compositions of **plyranges** "verbs" are performed using
+underlying, highly optimized range operations in the **GenomicRanges** package
 [@granges].
@@
 
-We now have *GRanges* objects that contain DE genes, genes without strong
+We now have _GRanges_ objects that contain DE genes, genes without strong
 signal of DE, and DA peaks. We are ready to answer the question: is there an
@@
 
-As `plyranges` is built on top of *dplyr*, it implements methods for many of
-its verbs for *GRanges* objects. Here we can use `slice` to randomly sample the
-rows of the `other_genes`. The `sample.int` function will generate random
+As **plyranges** is built on top of **dplyr**, it implements methods for many of
+its verbs for *GRanges* objects. Here we can use `slice()` to randomly sample the
+rows of the `other_genes`. The `sample.int()` function will generate random
 samples of size equal to the number of DE-genes from the number of rows in
@@
 
-We can repeat this many times to create many samples via `replicate`. By
+We can repeat this many times to create many samples via `replicate()`. By
 replicating the sub-sampling multiple times, we minimize the variance on the
@@
 
-This creates a list of *GRanges* objects as a list, and we can bind these
-together using the `bind_ranges` function. This function creates a new column
-called "resample" on the result that identifies each of the input *GRanges*
+This creates a list of _GRanges_ objects as a list, and we can bind these
+together using `bind_ranges()`.  This function creates a new column
+called _resample_ on the result that identifies each of the input _GRanges_
 objects:
@@
 
-Similarly, we can then combine the `boot_genes` *GRanges*, with the DE
-*GRanges* object. As the resample column was not present on the DE *GRanges*
+Similarly, we can then combine the `boot_genes` _GRanges_, with the DE
+_GRanges_ object. As the resample column was not present on the DE *GRanges*
 object, this is given a missing value which we recode to a 0 using `mutate()`
@@
 could do this, but we prefer an approach via the anchoring methods in
-`plyranges`. Because there is a mutual dependence between the start, end,
+**plyranges**. Because there is a mutual dependence between the start, end,
 width, and strand of a *GRanges* object, we define anchors to fix one of
-`start` and `end`, while modifying the `width`. As an example, to extract just
+_start_ and _end_, while modifying the _width_. As an example, to extract just
 the TSS, we can anchor by the 5' end of the range and modify the width of the
@@
 
-Anchoring by the 5' end of a range will fix the `end` of negatively stranded
-ranges, and fix the `start` of positively stranded ranges.
+Anchoring by the 5' end of a range will fix the _end_ of negatively stranded
+ranges, and fix the _start_ of positively stranded ranges.
 
 We can then repeat the same pattern but this time using `anchor_center()` to
-tell `plyranges` that we are making the TSS the midpoint of a range that has
+tell **plyranges** that we are making the TSS the midpoint of a range that has
 total width of 20kb, or 10kb both upstream and downstream of the TSS.
@@
 We are now ready to compute overlaps between RNA-seq genes (our DE set and
-bootstrap sets) and the ATAC-seq peaks. In `plyranges`, overlaps are defined as
+bootstrap sets) and the ATAC-seq peaks. In **plyranges**, overlaps are defined as
 joins between two *GRanges* objects: a _left_ and a _right_ *GRanges* object.
@@
 
-```{r olap-join, echo = FALSE}
+```{r olap-join, echo = TRUE}
 ```
@@
 peaks for each gene and the maximum peak based on LFC. This is achieved via
-`reduce_ranges_directed`, which allows an aggregation to result in a *GRanges*
+`reduce_ranges_directed()`, which allows an aggregation to result in a _GRanges_
 object via merging neighboring genomic regions. The use of the directed suffix
@@
 
-```{r reduce-ex01, echo = FALSE}
+```{r reduce-ex01, echo = TRUE}
 ```
@@
 
-For all variables except for the `origin` column we divide the first row's
+For all variables except for the _origin_ column we divide the first row's
 values by the second row, which will be the enrichment of peaks in DE genes
 compared to other genes. This requires us to reshape the summary table from
-long form back to wide form using the *tidyr* package. First we pivot the
-results of the `peak_count` columns into name-value pairs, then pivot again to
-place values into the `origin` column. Then we create a new column with the
+long form back to wide form using the **tidyr** package. First we pivot the
+results of the _peak\_count_ columns into name-value pairs, then pivot again to
+place values into the _origin_ column. Then we create a new column with the
 relative enrichment:
@@
 duplicate code. Instead we create a single function called
-`count_above_threshold` that accepts a variable and a vector of thresholds, and
+`count_above_threshold()` that accepts a variable and a vector of thresholds, and
 computes the sum of the absolute value of the variable for each element in the
@@
 hundred thresholds based on the range of absolute LFC values in the `da_peaks`
-*GRanges* object:
+_GRanges_ object:
 
@@
 The peak counts for each threshold are computed as a new list-column called
-`value`. First, the *GRanges* object has been grouped by the gene, origin, and
+_value_. First, the _GRanges_ object has been grouped by the gene, origin, and
 the number of resamples columns. Then we aggregate over those columns, so each
@@
 
-Now we can expand these list-columns into a long *GRanges* object using the
-`expand_ranges()` function. This function will unlist the `value` and
-`threshold` columns and lengthen the resulting *GRanges* object.  To compute
+Now we can expand these list-columns into a long _GRanges_ object using 
+`expand_ranges()`. This function will unlist the _value_ and
+_threshold_ columns and lengthen the resulting _GRanges_ object. To compute
 the peak and gene counts for each threshold, we apply the same summarization as
@@
 before, by pivoting the results to long form then back to wide form to compute
-the enrichment. We visualize the peak enrichment changes of DE genes relative
+the enrichment.
+
+```{r line-chart-data, echo = TRUE}
+```
+
+We visualize the peak enrichment changes of DE genes relative
 to other genes as a line chart:
@@
 
-We have shown that by using `plyranges` and `tximeta` (with support of
-Bioconductor and `tidyverse` ecosystems) we can fluently iterate through the
+We have shown that by using **plyranges** and **tximeta** (with the support of 
+the Bioconductor and **tidyverse** ecosystems) we can iterate through the
 biological data science workflow: from import, through to modeling, and data
@@
 Finally, our workflow illustrates the benefits of using appropriate data
-abstractions provided by Bioconductor such as the *SummarizedExperiment* and
-*GRanges*. These abstractions provide users with a mental model of their
+abstractions provided by Bioconductor such as the _SummarizedExperiment_ and
+_GRanges_. These abstractions provide users with a mental model of their
 experimental data and are the building blocks for constructing the modular and
@@
 interoperate many decoupled R packages (from both Bioconductor and the
-tidyverse) to construct a seamless end-to-end workflow that is far too
+**tidyverse**) to construct a seamless end-to-end workflow that is far too
 specialized for a single monolithic tool.
@@
 the development version of the workflow and all downstream dependencies can be
-installed using the `BiocManager` package by running:
+installed using the **BiocManager** package by running:
 
@@
 We would like to thank all participants of the Bioconductor 2019 and
-BiocAsia 2019 conferences who attended and provided feedback on early versions of this workflow paper.
+BiocAsia 2019 conferences who attended and provided feedback on early versions 
+of this workflow paper.

### 2020-07-26 scripts/fluentGenomics.R

@@ -1,8 +1,12 @@
 ## ----setup, include = FALSE------------------------
 library(fluentGenomics)
 dir <- system.file("extdata", package="macrophage")
-
+library(readr)
+library(dplyr)
+library(SummarizedExperiment)
 library(tximeta)
+library(DESeq2)
+library(plyranges)
 makeLinkedTxome(
   indexDir=file.path(dir, "gencode.v29_salmon_0.12.0"),
   source="Gencode",
@@ -13,7 +17,9 @@ makeLinkedTxome(
   gtf=file.path(dir, "gencode.v29.annotation.gtf.gz"), # local version
   write=FALSE
 )
-
+# alias to plyranges
+n_distinct <- plyranges::n_distinct
+n <- plyranges::n
 
 ## ----setdir----------------------------------------
 dir <- system.file("extdata", package="macrophage")
@@ -24,13 +30,13 @@ library(readr)
 library(dplyr)
 colfile <- file.path(dir, "coldata.csv")
 coldata <- read_csv(colfile) %>%
-  dplyr::select(
+  select(
     names,
     id = sample_id,
     line = line_id,
     condition = condition_name
   ) %>%
-  dplyr::mutate(
+  mutate(
     files = file.path(dir, "quants", names, "quant.sf.gz"),
     line = factor(line),
     condition = relevel(factor(condition), "naive")
@@ -39,7 +45,7 @@ coldata
 
 
 ## ----tximeta-run-----------------------------------
-suppressPackageStartupMessages(library(SummarizedExperiment))
+library(SummarizedExperiment)
 library(tximeta)
 se <- tximeta(coldata, dropInfReps=TRUE)
 se
@@ -74,7 +80,7 @@ DESeq2::plotMA(res, ylim=c(-10,10))
 
 
 ## ----results-GRanges-------------------------------
-suppressPackageStartupMessages(library(plyranges))
+library(plyranges)
 de_genes <- results(dds,
                     contrast=c("condition","IFNg","naive"),
                     lfcThreshold=1,
@@ -120,25 +126,25 @@ da_peaks <- peaks %>%
 
 ## ----slice-example---------------------------------
 size <- length(de_genes)
-slice(other_genes, sample.int(plyranges::n(), size))
+slice(other_genes, sample.int(n(), size))
 
 
 ## ----boot-set-01-----------------------------------
 # set a seed for the results
[…]

### 2020-07-26 Rmd/06-conclusion.Rmd
resolve Matt comments

@@
   )
+
+nice <- function(x) prettyNum(x, big.mark = ",", scientific = FALSE)
 
@@
 with genomic data as a *GRanges* object. Since its release on Bioconductor,
-it  has been relatively  successful: it has been downloaded `r plyranges_stats$no_downloads` times from  `r plyranges_stats$no_ip` unique IP addresses.
+it  has been relatively  successful: it has been downloaded `r nice(plyranges_stats$no_downloads)` times from  `r nice(plyranges_stats$no_ip)` unique IP addresses.
 I have also had the privilege of teaching workshops on **plyranges**
@@
 with embeddings. Based on the case studies provided I believe that the methods
-used in **liminal** could be broadly applicable many high dimensional datasets.
+used in **liminal** could be broadly applicable to many high dimensional datasets
+and NLDR methods.
 The **liminal** package is available to download from https://github.com/sa-lee/liminal.
@@
 like HDF5 that leverage existing Bioconductor frameworks would be useful.
-We showed in chapters \@ref(ch:fluentGenomics) ad \@ref(ch:intron) that an a
+We showed in chapters \@ref(ch:fluentGenomics) ad \@ref(ch:intron) that an
 analyst is able to do some very complex data transformations and re-sampling 
-procedures via casting results into *GRanges* object. However, it is unclear 
+procedures via casting results into _GRanges_ object. However, it is unclear 
 whether the semantics of our grammar can be extended to data that can not be 
@@
 to avoid crowding would be valuable. An added complexity to changes in visual
-displays are thinking about the design of user interactions, several promising
-avenues based on section tours should be explored [@Laa2020-wr; @Laa2020-fy].
+displays are thinking about the design of user interactions, and several promising
+avenues based on section tours could be explored [@Laa2020-wr; @Laa2020-fy].

### 2020-07-26 Rmd/05-embeddings.Rmd
Matt comments
typo

@@
 of non-linear embedding techniques were more likely to consider clusters of
-points on a 2d scatter plot to be the result of a spurious embedding compared
+points on a 2D scatter plot to be the result of a spurious embedding compared
 to advanced users who were aware of the inner workings of the embedding
@@
 
-## Acknowledgements
+## Acknowledgements {-}
 
-We would like to thank David Frazier, Ursula Laa and Paul Harrison for their
-feedback on this work.
+We would like to thank Dr David Frazier, Dr Ursula Laa and Dr Paul Harrison for their feedback on this work.
 

### 2020-07-26 Rmd/02-plyranges.Rmd
typos

@@
 genomic extension of **dplyr**. By analogy, **plyranges** is to the
-genomic algebra, as **dply**` is to the relational algebra.  The
+genomic algebra, as **dplyr**` is to the relational algebra.  The
 **plyranges`**Bioconductor package implements the language on top of a
@@
   \hline
-   & \textbf{\emph{summarize()}} & aggregate over column(s) \\ 
-   Aggregate & \textt{disjoin\_ranges()} & aggregate column(s) over the union of end coordinates \\
-   &  \textt{reduce\_ranges()} & aggregate column(s) by merging overlapping and neighboring ranges \\
+   & \textbf{\texttt{summarize()}} & aggregate over column(s) \\ 
+   Aggregate & \texttt{disjoin\_ranges()} & aggregate column(s) over the union of end coordinates \\
+   &  \texttt{reduce\_ranges()} & aggregate column(s) by merging overlapping and neighboring ranges \\
    \hline
-   &  \textbf{\textt{mutate()}} & modifies any column \\
-   & \textbf{\textt{select()}} & select columns \\
-  Modify (Unary) & \textbf{\textt{arrange()}} & sort by columns \\
-   & \textt{stretch()} & extend range by fixed amount \\
-   &  \textt{shift\_(direction)} & shift coordinates \\
-   & \textt{flank\_(direction)} & generate flanking regions \\
-   & \textt{\%intersection\% } & row-wise intersection \\
-   & \textt{\%union\%} & row-wise union \\
-   & \textt{compute\_coverage} & coverage over all ranges \\
-  Modify (Binary) &  \textt{\%setdiff\%} & row-wise set difference \\
-   & \textt{between()} & row-wise gap range \\
-   & \textt{span()} & row-wise spanning range \\
+   &  \textbf{\texttt{mutate()}} & modifies any column \\
+   & \textbf{\texttt{select()}} & select columns \\
+  Modify (Unary) & \textbf{\texttt{arrange()}} & sort by columns \\
+   & \texttt{stretch()} & extend range by fixed amount \\
+   &  \texttt{shift\_(direction)} & shift coordinates \\
+   & \texttt{flank\_(direction)} & generate flanking regions \\
+   & \texttt{\%intersection\% } & row-wise intersection \\
+   & \texttt{\%union\%} & row-wise union \\
+   & \texttt{compute\_coverage} & coverage over all ranges \\
+  Modify (Binary) &  \texttt{\%setdiff\%} & row-wise set difference \\
+   & \texttt{between()} & row-wise gap range \\
+   & \texttt{span()} & row-wise spanning range \\
    \hline
-    & \textt{join\_overlap\_*()} & merge by overlapping ranges \\
-    & \textt{join\_nearest} & merge by nearest neighbor ranges \\
-    & \textt{join\_follow} & merge by following ranges \\
-    Merge & \textt{join\_precedes} & merge by preceding ranges \\
-    & \textt{union\_ranges} & range-wise union \\
-    & \textt{intersect\_ranges} & range-wise intersect \\
-    & \textt{setdiff\_ranges} & range-wise set difference \\
-    & \textt{complement\_ranges} & range-wise set complement \\
+    & \texttt{join\_overlap\_*()} & merge by overlapping ranges \\
+    & \texttt{join\_nearest} & merge by nearest neighbor ranges \\
+    & \texttt{join\_follow} & merge by following ranges \\
+    Merge & \texttt{join\_precedes} & merge by preceding ranges \\
+    & \texttt{union\_ranges} & range-wise union \\
+    & \texttt{intersect\_ranges} & range-wise intersect \\
+    & \texttt{setdiff\_ranges} & range-wise set difference \\
+    & \texttt{complement\_ranges} & range-wise set complement \\
   \hline
-   & \textt{anchor\_direction()} & fix coordinates at direction \\
-  Operate & \textbf{\textt{group\_by()}} & partition by column(s)  \\ 
-   & \textt{group\_by\_overlaps()} & partition by overlaps \\
+   & \texttt{anchor\_direction()} & fix coordinates at direction \\
+  Operate & \textbf{\texttt{group\_by()}} & partition by column(s)  \\ 
+   & \texttt{group\_by\_overlaps()} & partition by overlaps \\
    \hline
-   & \textbf{\textt{filter()}} & subset rows \\
-  Restrict & \textt{filter\_by\_overlaps()} & subset by overlap \\
-    & \textt{filter\_by\_non\_overlaps()} & subset by no overlap \\
+   & \textbf{\texttt{filter()}} & subset rows \\
+  Restrict & \texttt{filter\_by\_overlaps()} & subset by overlap \\
+    & \texttt{filter\_by\_non\_overlaps()} & subset by no overlap \\
    \hline
@@
 three tidy data structures that are related by feature and sample identifiers. 
-Therefore, the grammar and design of the **plryanges** DSL is naturally 
+Therefore, the grammar and design of the **plyranges** DSL is naturally 
 extensible to the _SummarizedExperiment_.

### 2020-07-26 Rmd/03-fluentGenomics.Rmd
typos

@@
 
-We have already used **plyrange**  a number of times above, to `filter()`,
+We have already used **plyranges**  a number of times above, to `filter()`,
 `mutate()`, and `select()` on _GRanges_ objects, as well as ensuring the correct

### 2020-07-26 Rmd/05-embeddings.Rmd
typos

@@
 one linked brushing from the embedding to the tour view. Similarly, diffusions
-are identified from linked brushing on the tour view, higlighting points
+are identified from linked brushing on the tour view, highlighting points
 in the embedding view. Section <> details
@@
 can be independent of the tour or embedding views. The estimation of neighbours
-can be computed effeciently with the `FNN` or `RcppAnnoy` packages
+can be computed efficiently with the `FNN` or `RcppAnnoy` packages
 on the server side [@fnn-pkg; @rcpp-annoy-pkg]. In place of highlighting, one 

### 2020-07-27 Rmd/03-fluentGenomics.Rmd
refactor and tidy up scripts for chapter 3

@@
 ```
-<!-- should we describe the data structure in more detail, or use a figure -->
 
@@
 
-Similarly, we can then combine the `boot_genes` _GRanges_, with the DE
+Similarly, we can then combine the `subsamp_genes` _GRanges_, with the DE
 _GRanges_ object. As the resample column was not present on the DE *GRanges*
@@
 We are now ready to compute overlaps between RNA-seq genes (our DE set and
-bootstrap sets) and the ATAC-seq peaks. In **plyranges**, overlaps are defined as
+resampled sets) and the ATAC-seq peaks. In **plyranges**, overlaps are defined as
 joins between two *GRanges* objects: a _left_ and a _right_ *GRanges* object.
@@
 sets. We could also have computed variance in addition to the mean of the
-bootstrap set, and so drawn an interval around the enrichment line.
+resampled set, and so drawn an interval around the enrichment line.
 

### 2020-07-27 scripts/fluentGenomics.R

@@ -1,12 +1,16 @@
 ## ----setup, include = FALSE------------------------
+# due to bookdown loading everything at once, we
+# first detach and reattach to avoid conflicts
+detach("package:plyranges", unload = TRUE)
 library(fluentGenomics)
-dir <- system.file("extdata", package="macrophage")
-library(readr)
-library(dplyr)
 library(SummarizedExperiment)
 library(tximeta)
 library(DESeq2)
-library(plyranges)
+library(readr)
+library(dplyr)
+
+dir <- system.file("extdata", package="macrophage")
+
 makeLinkedTxome(
   indexDir=file.path(dir, "gencode.v29_salmon_0.12.0"),
   source="Gencode",
@@ -17,9 +21,9 @@ makeLinkedTxome(
   gtf=file.path(dir, "gencode.v29.annotation.gtf.gz"), # local version
   write=FALSE
 )
-# alias to plyranges
-n_distinct <- plyranges::n_distinct
-n <- plyranges::n
+# # alias to plyranges
+# n_distinct <- plyranges::n_distinct
+# n <- plyranges::n
 
 ## ----setdir----------------------------------------
 dir <- system.file("extdata", package="macrophage")
@@ -41,7 +45,7 @@ coldata <- read_csv(colfile) %>%
     line = factor(line),
     condition = relevel(factor(condition), "naive")
   )
-coldata
+glimpse(coldata)
 
 
 ## ----tximeta-run-----------------------------------
@@ -80,7 +84,7 @@ DESeq2::plotMA(res, ylim=c(-10,10))
 
 
 ## ----results-GRanges-------------------------------
-library(plyranges)
+suppressPackageStartupMessages(library(plyranges))
 de_genes <- results(dds,
                     contrast=c("condition","IFNg","naive"),
                     lfcThreshold=1,
@@ -92,7 +96,7 @@ de_genes
 ## ----de-genes--------------------------------------
 de_genes <- de_genes %>%
   filter(padj < 0.01) %>%
-  dplyr::select(
+  select(
     gene_id,
     de_log2FC = log2FoldChange,
     de_padj = padj
@@ -107,7 +111,7 @@ other_genes <- results(dds,
                        format="GRanges") %>%
   filter(padj < 0.01) %>%
   names_to_column("gene_id") %>%
-  dplyr::select(
+  select(
     gene_id,
     de_log2FC = log2FoldChange,
     de_padj = padj
[…]

### 2020-07-27 Rmd/01-intro.Rmd
update intro, closes #7

@@
 leading to new hypotheses that can be tested and formalised using 
-statistical modelling. As datasets grow in complexity and become
+statistical modelling (figures \@ref(fig:wflow-cycle)). As datasets grow in complexity and become
 increasingly heterogeneous and multidimensional, 
@@
 
+(ref:wflow-cycle): An idealised model of the biological data science workflow (adapted from  @Grolemund2017-uc).
+We begin with data generated from one or more biological assay(s) corresponding to a research hypothesis or question.
+Our primary focus is on data generated from bulk assays that measure gene expression (RNA-seq), genetic variation (DNA-seq), and gene regulation (ChIP-seq, ATAC-seq).
+Throughout this process, we need computational tools to gain insight into the biology under study and communicate our analysis in a reproducible manner.
+
+(ref:wflow-wrangle): In the wrangle phase of the workflow, data from an assay is imported into a programming language.
+It is then tidied into a new representation that should capture the biological semantics of the measurements.
+Following this, the representation can be transformed to generate new summaries.
+
+(ref:wflow-integrate): Data from multiple assays can be combined together using modelling and transformations.
+Data and model results can combined, tidied and transformed to have a common representation across different granularities of the genome.
+
+(ref:wflow-integrate): Visualisation is a tool for making sense of models. Here we explore, how well a model has captured the structure within the data using interactive graphics.
+
+```{r wflow-cycle, fig.cap="(ref:wflow-cycle)"}
+knitr::include_graphics(here::here("img", "wflow-thesis-cycle.png"))
+```
+
 This thesis focuses on core aspects of EDA as part of a biological data
@@
 on applications to genomics and transcriptomics. To begin we discuss wrangling
-biological data using a coherent representation and programming interface; 
-section \@ref(sec:tidygranges)
+biological data using a coherent representation and programming interface 
+(figure \@ref(fig:wflow-wrangle)). Section \@ref(sec:tidygranges)
 introduces a grammar-based framework for transforming genomics data that is 
 described in chapter \@ref(ch:plyranges). We then look at integrating data and 
-model outputs over genomic regions to gain biological insight; 
-\@ref(sec:integrate) introduces a framework for incorporating genomic regions 
-over multiple assays, described in chapter \@ref(ch:fluentGenomics), while 
-\@ref(sec:coverage) discusses finding 'interesting' genomic regions via 
-multiple summaries of a single assay, described in \@ref(ch:intron). Next, we 
-consider the challenges in  visualising  high dimensional data; section 
-\@ref(sec:va) introduces an  interactive visualisation approach for 
-understanding non-linear dimension reduction  techniques described in chapter 
-\@ref(ch:tsne). Lastly, in chapter \@ref(ch:conclusion) I describe the outputs 
-of my thesis and plans for future developments.
+model outputs over genomic regions to gain biological insight (\@ref(fig:wflow-integrate)). Section \@ref(sec:integrate) introduces a 
+framework for incorporating genomic regions  over multiple assays, described in 
+chapter \@ref(ch:fluentGenomics), while  \@ref(sec:coverage) discusses finding 'interesting' genomic regions via combining multiple summaries of a single assay, described in \@ref(ch:intron).
+Next, we
+consider the challenges in  visualising  high dimensional data (figure \@ref(fig:wflow-explore)). Section  \@ref(sec:va) introduces an interactive 
+visualisation approach for  understanding non-linear dimension reduction 
+techniques described in chapter  \@ref(ch:tsne). Lastly, in 
+chapter \@ref(ch:conclusion) describes the outputs of the thesis and plans for future developments.
 
 ## A grammar for genomic data analysis {#sec:tidygranges}
+
+```{r wflow-wrangle, fig.cap="(ref:wflow-wrangle)"}
+knitr::include_graphics(here::here("img", "wflow-thesis-wrangle.png"))
+```
 
@@
 
+```{r wflow-integrate, fig.cap="(ref:wflow-integrate)"}
+knitr::include_graphics(here::here("img", "wflow-thesis-integrate.png"))
+```
+
 It is rare that a biological data analysis will involve
@@
 
+```{r wflow-explore, fig.cap="(ref:wflow-explore)"}
+knitr::include_graphics(here::here("img", "wflow-thesis-explore.png"))
+```
+
 In the final chapter, we move away from data wrangling and towards the 

### 2020-07-27 Rmd/05-embeddings.Rmd
remove media9 for preparation of screenshots.

@@
   knitr::include_url(video[["url"]])
-}
-
-display_vid_latex <- function(key, lookup) {
-    video <- lookup[lookup$key == key, ]
-    glue::glue(
-    "\\begin{{center}}
-   \\includemedia[
-    width=0.6\\linewidth,
-    height=0.45\\linewidth,
-    keepaspectratio=true,
-    activate=pageopen
-    addresource={video},
-    flashvars={{%
-      source={video}
-      &autoPlay=true
-      &loop=true}
-  ]{{}}{{VPlayer.swf}}
-  \\end{{center}}",
-  video = video[["path"]]
-  )
 }
@@
 
-```{r gaussian, results="asis", eval = knitr::is_latex_output()}
-cat(display_vid_latex("gaussian", vid_lookup))
+```{r gaussian, eval = knitr::is_latex_output()}
 ```
@@
 ```{r hierachical, results='asis', eval = knitr::is_latex_output()}
-cat(display_vid_latex("hierachical", vid_lookup))
 ```
@@
 
-####
+#### Piecewise linear structure
 
@@
 
-```{r trees-01, results='asis', eval = knitr::is_latex_output()}
-cat(display_vid_latex("trees-01", vid_lookup))
+```{r trees-01, eval = knitr::is_latex_output()}
 ```
@@
 
-```{r trees-02, results='asis', eval = knitr::is_latex_output()}
-cat(display_vid_latex("trees-02", vid_lookup))
+```{r trees-02, eval = knitr::is_latex_output()}
 ```
@@
 
-```{r mouse-01, results='asis', eval = knitr::is_latex_output()}
-cat(display_vid_latex("mouse-01", vid_lookup))
+```{r mouse-01, eval = knitr::is_latex_output()}
 ```
@@
 
-```{r mouse-02, results='asis', eval = knitr::is_latex_output()}
-cat(display_vid_latex("mouse-02", vid_lookup))
+```{r mouse-02, eval = knitr::is_latex_output()}
 ```

### 2020-07-27 Rmd/01-intro.Rmd
always forgot you don't need the colon

@@
 
-(ref:wflow-cycle): An idealised model of the biological data science workflow (adapted from  @Grolemund2017-uc).
+(ref:wflow-cycle) An idealised model of the biological data science workflow (adapted from  @Grolemund2017-uc).
 We begin with data generated from one or more biological assay(s) corresponding to a research hypothesis or question.
@@
 
-(ref:wflow-wrangle): In the wrangle phase of the workflow, data from an assay is imported into a programming language.
+(ref:wflow-wrangle) In the wrangle phase of the workflow, data from an assay is imported into a programming language.
 It is then tidied into a new representation that should capture the biological semantics of the measurements.
@@
 
-(ref:wflow-integrate): Data from multiple assays can be combined together using modelling and transformations.
+(ref:wflow-integrate) Data from multiple assays can be combined together using modelling and transformations.
 Data and model results can combined, tidied and transformed to have a common representation across different granularities of the genome.
 
-(ref:wflow-integrate): Visualisation is a tool for making sense of models. Here we explore, how well a model has captured the structure within the data using interactive graphics.
+(ref:wflow-integrate) Visualisation is a tool for making sense of models. Here we explore, how well a model has captured the structure within the data using interactive graphics.
 

### 2020-07-27 scripts/fluentGenomics.R
tidy up code output so it doesn't bleed over lines

@@ -199,8 +199,10 @@ origin_peak_lfc <- genes_olap_peaks %>%
   group_by(origin) %>%
   summarize(
     peak_count = sum(!is.na(da_padj)) / n_distinct(resample),
-    lfc1_peak_count =sum(abs(da_log2FC) > 1, na.rm=TRUE)/ n_distinct(resample),
-    lfc2_peak_count = sum(abs(da_log2FC) > 2, na.rm=TRUE)/ n_distinct(resample)
+    lfc1_peak_count =
+      sum(abs(da_log2FC) > 1, na.rm=TRUE) / n_distinct(resample),
+    lfc2_peak_count =
+      sum(abs(da_log2FC) > 2, na.rm=TRUE) / n_distinct(resample)
   )
 origin_peak_lfc
 

### 2020-07-28 Rmd/05-embeddings.Rmd
updating references and sections for #6

@@
 the preservation of local and global structure in the resulting visualisation.
-We present a visual <change word here> framework for the pragmatic usage of NLDR methods
+We present visual diagnostics for the pragmatic usage of NLDR methods
 by combining them with a technique called the tour. A tour is a sequence of
@@
 linking the tour to a view obtained from an NLDR method, we can preserve global
-structure and through user interactions like spatial linked brushing observe
+structure and through user interactions like linked brushing observe
 where the NLDR view may be misleading. We display several case studies from
-single cell transcriptomics and image processing, that shows our approach is
-useful for cluster orientation tasks.  The implementation of our framework is
-available as an R package called **liminal** available at
+both simulated and real data from single cell transcriptomics, that shows our
+approach is useful for cluster orientation tasks.  The implementation of our 
+framework is available as an R package called **liminal** available at
 https://github.com/sa-lee/liminal.
@@
 High dimensional data is increasingly prevalent in the natural sciences and
-beyond but presents a challenge to the analyst in terms of both data cleaning /
+beyond but presents a challenge to the analyst in terms of data cleaning,
 pre-processing and visualisation. Methods to embed data from a high-dimensional
@@
 workflow where they are used to ascertain hidden structure and de-noise data
-for downstream analysis (thereby nullifying the 'curse of dimensionality').
+for downstream analysis .
 
@@
 onto a low-dimensional basis matrix, and is represented as a
-animated visualization. Given the dynamic nature of the tour, user interaction is 
-important for controlling and exploring the visualisation: the tour has been
-used previously by @Wickham2015-cx and  
+animated visualization [@Asimov1985-vp; @Buja1986-zr]. Given the dynamic nature
+of the tour, user interaction is  important for controlling and exploring the 
+visualisation: the tour has been used previously by @Wickham2015-cx and  
 exploring statistical model fits and by @Buja1996-fk for exploring the space of 
@@
 
-<!--- add tour reference; linked brushing -->
-
 While there has been much work on the algorithmic details of
-embedding methods, there has been relatively few tools designed to assist users
-to interact with these techniques and assist them in making comparisons between
-embeddings and performing cluster orientation tasks. Several
+embedding methods, there are relatively few tools designed to assist users
+to interact with these techniques: when is an embedding suffucient for the
+task at hand? Several
 interactive interfaces have been proposed for evaluating or using embedding
-techniques: the Sleepwalk interface provides a click and highlight
-visualisation for colouring points in an embedding according to an estimated
-pair-wise distance in the original high-dimensional space [@Ovchinnikova2020-sy].
-The idea behind this is that non-linear embedding methods should preserve 
-locality and that points that are 'close' to each other on the embedded view
-should be close according to the estimated pair-wise distance. 
-Similarly, the TensorFlow embedding projector is a web interface to 
+techniques. @Buja2008-fn used tours to guide analysts during the optimisation
+of multidimensional scaling methods by extending their interactive visualisation
+toolkits, **XGobi** and **GGobi** into a new tool called **GGvis** 
+[@Swayne1998-uq; @Swayne2003-qd; @Swayne2004-ua].
+Their interface allows the analyst to dynamically modify and check whether
+an MDS configuration has preserved the locality and closeness
+of points between the configuration and the original data.
+@Ovchinnikova2020-sy created the **Sleepwalk**  interface for checking
+non-linear embeddings in single cell RNA-seq data. It provides a click and 
+highlight visualisation for colouring points in an embedding according to an 
+estimated pairwise distance in the original high-dimensional space.
+Similarly, the **TensorFlow**  embedding projector is a web interface to 
 running some non-liner embedding methods live in the browser and 
@@
 [@Smilkov2016-hp]. Finally, the work by @Pezzotti2017-cz provides a user 
-guided and modified form of the  t-SNE algorithm, that allows users to modified 
-optimisation parameters in real-time. <!-- add visualising with mds reference -->
+guided and modified form of the  t-SNE algorithm, that allows users to modify 
+optimisation parameters in real-time.
 
@@
 to see a global overview of their high dimensional data and assists them with
-cluster orientation tasks. This workflow is incorporated into an R package
-called **liminal** (Available: https://github.com/sa-lee/liminal).
+cluster orientation tasks.
+
+The rest of the paper is organised as follows. The next section provides
+background on dimension reduction methods, including an overview of the tour.
+Then we describe the visual design of **liminal**, followed by implementation
+details. Next we provide case studies that show how our interface assists
+in using embedding methods. Finally, we describe the insights gained by using
+**liminal** and plans for extensions to the software.
 
@@
 mathematical theory, and is able to ascertain 
-the shape and global structure of a dataset via inspection of the subspace generated by the set of low-dimensional projections. In fact, in some
-of the case studies below, we first sphere our data via principal 
-components to reduce dimensionality of our input.
+the shape and global structure of a dataset via inspection of the subspace generated by the set of low-dimensional projections [@Asimov1985-vp; @Buja1986-zr].
 
@@
 bases at random; allowing a user to ascertain structure via exploring the
-subspace of $d$-dimensional projections. Instead of picking projections at 
-random, a *guided tour* can be used to generate a sequence 'interesting' 
-projections as quantified by an index function. While **liminal** is able to 
-visualise guided tours, our focus in the case studies uses the grand tour to see global structure in the data.
+subspace of $d$-dimensional projections.  In practice, we first sphere 
+our data via principal components to reduce dimensionality of $X$ prior to 
+running the tour. Instead of picking projections at 
+random, a *guided tour* can be used to generate a sequence of 'interesting' 
+projections as quantified by an index function [@Cook1995-bi]. While our software, 
+**liminal** is able to  visualise guided tours, our focus in the case studies 
+uses the grand tour to see global structure in the data.
 
@@
 Tours provide a supportive visualisation to NLDR graphics, and 
-can be easily incorporated into an analysts workflow with our interface in 
-**liminal**. 
-As we will see in the case studies, this allows analysts to quickly compare 
-views from embedding methods and allows them to see how the embedding method 
-alters the global structure of their data. Using ensembles allows the use of 
-interaction techniques, that allow analysts to perform cluster orientation 
-tasks via linking multiple views. This approach allows our interface, to achieve 
+can be easily incorporated into an analysts workflow with our software package, 
+**liminal**.  Our interface allows analysts to quickly compare 
+views from embedding methods and see how an embedding method 
+preserves or alters the geometry of their data. Using multiple concatenated and
+linked views with the tour enchances interaction techniques, and allows 
+analysts to perform  cluster orientation tasks via linked highlighting and 
+brushing [@McDonald1982-ew; @Becker1987-gd]. 
+This approach allows our interface to achieve 
 the three principles for interactive high-dimensional data visualisation outlined
@@
 tour is dynamic, it is often useful to be able to pause and highlight a
-particular view. In our interface, brushing will pause the tour path, allowing
-users to identify 'interesting' projections. The domain of the axis scales from
-running a tour is called the half range, and is computed by rescaling the input
-data onto hyper-dimensional unit cube. We bind the half range to a mouse wheel
-event, allowing a user to pan and zoom on the tour view dynamically. This is
-useful for peeling back dense clumps of points to reveal structure.
+particular view. In addition to pause, play and reset buttons, brushing will 
+pause the tour path, allowing users to identify 'interesting' projections. The
+domain of  the axis scales from running a tour is called the half range, and is 
+computed by rescaling the input data onto hyper-dimensional unit cube. We bind 
+the half range to a mouse wheel event, allowing a user to pan and zoom on the 
+tour view dynamically. This is useful for peeling back dense clumps of points 
+to reveal structure.
 
@@
 We have combined the tour view in a side by side layout with a scatter plot
-view that represents the output of an embedding algorithm.  These views are
+view as has been done in previous tour interfaces **XGobi** and **DataViwer** 
+[@Buja1986-ku; @Swayne1998-uq]. These views are
 linked; analysts can brush regions or highlight collections of points in either
@@
 are identified from linked brushing on the tour view, highlighting points
-in the embedding view. Section <> details
+in the embedding view.
 
@@
 We have implemented the above design as an open source R package called
-**liminal**. The package allows analysts to construct concatenated
-visualisations, via the Vega-Lite grammar of interactive graphics via
-the **vegawidget** package [@Satyanarayan2017-gs; @Lyttle2020-hp]. It provides an interface for constructing linked
-and stand alone interfaces for manipulating tour paths via the **shiny** and **tourr** packages [@Chang2020-bq; @Wickham2011-st].
+**liminal** [@r-liminal]. The package allows analysts to construct concatenated
+visualisations, drawn with the **Vega-Lite** grammar of interactive graphics via
+the **vegawidget** package [@Satyanarayan2017-gs; @Lyttle2020-hp]. It provides 
+an interface for constructing linked and stand alone interfaces for 
+manipulating tour paths via  the **shiny** and **tourr** packages 
+[@Chang2020-bq; @Wickham2011-st].
 
@@
 brushing on the embedding view will produce a one-to-one linking with the tour
-view. For interpreting specific combinations of clusters, the legend
+view. For interpreting specific combinations of clusters, the multiple
+guides on the legend can be selected in order to see their relative orientations.
+The interface is constructed as a **shiny** gadget specifically designed for
+interactive data analysis. Selections such as brushing regions and the current
+tour path are returned after the user clicks done on the interface and become
+available for further investigation.
 
@@
 The next section steps through case studies cases of our approach using simulated
-and real data.  The videos presented show the visual 
-appearance of the **liminal** interface, and how we can interact with the tour
-via the controls previously described.
-In the first case study, we look at three different examples
-where t-SNE is applied to  simulations where cluster structure and geometry of the underlying data is known.
+and real data.
 
-In the second case study, we apply our approach to clustering 
+The first three case studies qw use simulations where cluster structure and
+geometery of the underlying data is known. We start with a simple example
+where we generated spherical clusters that are embedded well by t-SNE. Then
+we move onto more complex examples where the tour provides insight, such
+as clusters that have substructure and where there is more complex geometry
+in the data.
+
+In the final case study, we apply our approach to clustering 
 the mouse retina data from @Macosko2015-ot, and apply the tour
@@
 
-<!--
-Finally, in the last case study we
-explore a subset of the QuickDraw dataset, we use **liminal** to identify
-instances of images from different labels that appear similar.
--->
+The videos presented show the visual 
+appearance of the **liminal** interface, and how we can interact with the tour
+via the controls previously described. Both code and videos available for 
+reproducing this paper are available at https://github.com/sa-lee/paper-liminal.
 
-### Case Study 1: Exploring simulation data with tours and t-SNE
-
-#### Spherical Gaussian Clusters
+### Case Study 1: Exploring spherical Gaussian clusters
 
@@
 
-#### Spherical Gaussian clusters with hierarchical structure
+### Case Study 2: Exploring spherical Gaussian clusters with hierarchical structure
 
@@
 
-#### Piecewise linear structure
+### Case Study 3: Exploring data with piecewise linear structure
 
@@
 
-### Case Study 2: Clustering single cell RNA-seq data
+### Case Study 4: Clustering single cell RNA-seq data
 
@@
 
-<!--
-### Case Study 3: How do people draw animals?
-
-In the final case study, we explore a subset of the Google QuickDraw dataset 
-[@quickdraw-paper]. The data are sketches drawn by people of common concepts
-like 'airplane' or 'shoe', and corresponding metadata about the humans who
-created the sketches, such as country, the word category, whether the sketch
-was recognised by Google's neural network, and the number of strokes taken.
-We downloaded the metadata and normalised 28 by 28 pixel bitmaps via the 
-`quickdraw` package in R; unravelling the bitmaps unravelled the bitmaps into 
-a 784 dimension image vector. We sampled 500 images from each of the following
-categories: "broccoli", "flying saucer", "ice cream", "octopus", "pineapple",
-and "whale". We are interested in discovering images that are similiar but
-belong to different classifications 
--->
-
 ## Discussion

### 2020-07-28 Rmd/05-embeddings.Rmd
update to include figure references.

@@
 and set up the **liminal** linked interface with grand tour on the 10-$d$
-observations:
+observations.
 
-```{r gaussian, eval = knitr::is_latex_output()}
+(ref:liminal-gaussian) Screenshots of the **liminal** interface applied to well clustered data, a video of the tour animation is available at https://player.vimeo.com/video/439635921. *(a)*:
+
+```{r gaussian, fig.cap = "(ref:liminal-gaussian)"}
 ```
@@
 
-We see in the video that t-SNE has correctly split out each cluster and
-laid them out in a star like formation. This is agrees with the tour view,
+We see in figure \@ref(fig:gaussian) and linked video that t-SNE has 
+correctly split out each cluster and laid them out in a star like formation. 
+This is agrees with the tour view,
 where once we start the animation, the five clusters begin to appear but
 generally points are more concentrated in the projection view compared to 
-the t-SNE layout. This can be seen from our brushing action on the t-SNE side.
+the t-SNE layout (figure \@ref(fig:gaussian)a). 
+This can be confirmed via brushing the t-SNE view (figure \@ref(fig:gaussian)b).
 
@@
 
-We repeat the process of simulating Gaussian clusters in the next video, 
-but this time we have a cluster that has more granular structure. This time we have two
-Gaussian clusters with equal covariance embedded in $10-d$, and a third cluster
+We repeat the process of viewing Gaussian clusters from the 
+_Multi Challenge Dataset_, a benchmark simulation data set for clustering tasks
+[@Rauber2009-vh]. This dataset has two Gaussian clusters with equal covariance 
+embedded in $10-d$, and a third cluster
 with hierarchical structure. This cluster has two 3-$d$ clusters embedded in 
-10-$d$, where the second cluster is subdivided into three smaller clusters, that are each equidistant from each other and have the same covariance structure.
+10-$d$, where the second cluster is subdivided into three smaller clusters, 
+that are each equidistant from each other and have the same covariance structure.
 Again t-SNE has correctly identified the sub-clusters but their relative 
 locations to each other is distorted, with the orange and blue groups being
-far from each other in the tour view. We see in this case that is difficult
+far from each other in the tour view (figure \@ref(fig:hierachical)a). We see 
+in this case that is difficult
 to see the sub-clusters in the tour view, however, once we zoom and highlight
-they become more apparent. When we brush the sub-clusters in the t-SNE, their
+they become more apparent (figure \@ref(fig:hierachical)b). When we brush 
+the sub-clusters in the t-SNE, their
 relative placement is again exaggerated, with the tour showing that they
@@
 
-```{r hierachical, results='asis', eval = knitr::is_latex_output()}
+(ref:liminal-hierachical) Screenshots of the **liminal** interface applied to subclustered data, a video of the tour animation is .available at https://player.vimeo.com/video/43963590
+
+```{r hierachical, fig.cap="(ref:liminal-hierachical)"}
 ```
@@
 initialisation). We then create a linked
-tour with t-SNE layout with **liminal** as shown in the following video:
+tour with t-SNE layout with **liminal** as shown in figure \@ref(fig:trees-01).
 
-```{r trees-01, eval = knitr::is_latex_output()}
+(ref:trees-01) Screenshots of the **liminal** interface applied to tree structured data, a video of the tour animation is available at https://player.vimeo.com/video/439635892.
+
+```{r trees-01, fig.cap = "(ref:trees-01)"}
 ```
@@
 
-From the video, we see that the t-SNE view has been unable to recapitulate the
+We see that the t-SNE view has been unable to recapitulate the
 topology of the tree - the backbone (blue) branch has been split into three
-fragments. We can see this immediately via the linked highlighting over both
+fragments (figure \@ref(fig:trees-01)a). We can see this immediately via the linked highlighting over both
 plots. If we click on the legend for the zero branch, the blue coloured points
@@
 view there is a lot of noise along the branches, while this does not appear to
-be the case for the t-SNE result.
+be the case for the t-SNE result (figure \@ref(fig:trees-01)b).
 
@@
 initialisation and increase the perplexity from 30 to 100. We then combine
-these results with our tour view as displayed in the video:
+these results with our tour view as displayed in figure \@ref(fig:trees-02).
 
-```{r trees-02, eval = knitr::is_latex_output()}
+(ref:trees-02) Screenshots of the **liminal** interface applied to tree structured data, a video of the tour animation is available at https://player.vimeo.com/video/439635863.
+
+```{r trees-02, fig.cap = "(ref:trees-02)"}
 ```
@@
 
-The video shows that this selection of parameters results in the the tips of
-the branches (the three black dots in figure \@ref(fig:fake-trees) a) being split
-into three clusters representing the terminal branches of the tree. However, there
-are perceptual issues following the placement of the three groupings on the
-t-SNE view that become apparent via simple linked brushing. If we brush the
+Figure \@(ref:trees-02) shows that this selection of parameters results in the 
+tips of the branches (the three black dots in figure \@ref(fig:fake-trees)a) 
+being split into three clusters representing the terminal branches of the tree. 
+However, there are perceptual issues following the placement of the three 
+groupings on the t-SNE view that become apparent via simple linked brushing. 
+If we brush the
 tips of the yellow and brown branches (which appear to be close to each other
 on the t-SNE view), we immediately see the placement is distorted in the t-SNE
-view, and in the tour view these tips are at opposite ends of the tree.
+view, and in the tour view these tips are at opposite ends of the tree 
+(figure \@ref(fig:trees-02)b).
 Although, this is a known issue of the t-SNE algorithm, we can easily identify
@@
 data size - 4,590 cells. Although this is not ideal, it still allows us to
-get a sense of the shape of the clusters as seen in the video below:
+get a sense of the shape of the clusters as seen in figure \@ref(fig:mouse-01).
 
-```{r mouse-01, eval = knitr::is_latex_output()}
+(ref:mouse-01) Screenshots of the **liminal** interface applied to single cell data. A video of the tour animation is available at https://player.vimeo.com/video/439635812.
+
+```{r mouse-01, fig.cap = "(ref:mouse-01)" }
 ```
@@
 
-In the following video we can observe that the embedding has mostly captured the
+We observe that the embedding has mostly captured the
 clusters relative location to each other to their location in high dimensional space, with a notable exception of points in cluster 3 and 10 as shown with
-linked brushing. As expected avoids t-SNE mitigates the crowding problem that is
+linked brushing (figure \@ref(fig:mouse-01)a). As expected, t-SNE mitigates the crowding problem that is
 an issue for tour in this case, where points in clusters 2,4,6, and 11 are
-clumped together in tour view, but are blown up in the embedding view. 
+clumped together in tour view, but are blown up in the embedding view (figure \@ref(fig:mouse-01)b). 
 The tour appears to form a tetrahedron like shape, with points lying on
 surface and along the vertices of the the tetrahedron in 5-$d$ PCA space -
-a phenomena that has also been observed in @Korem2015-af. Brushing on
+a phenomena that has also been observed in @Korem2015-af 
+(figure \@ref(fig:mouse-01)c). Brushing on
 the tour view, reveals points in cluster 9 that are diffuse in the embedding 
 view, points in cluster 9 are relatively far away and spread apart from other 
-clusters in the tour view, but has points placed in cluster 3 and 9 in the embedding.
+clusters in the tour view, but has points placed in cluster 3 and 9 in the embedding
+(figure \@ref(fig:mouse-01)d).
 
@@
 weighting of each gene. We run the tour directly on the 
-log normalised expression values using the same subset as before:
+log normalised expression values using the same subset as before figure \@ref(fig:mouse-02).
 
-```{r mouse-02, eval = knitr::is_latex_output()}
+(ref:mouse-01) Screenshots of the **liminal** tour applied to a marker gene set. A video of the tour animation is available at https://player.vimeo.com/video/439635843.
+
+```{r mouse-02, fig.cap = "(ref:mouse-02)"}
 ```
@@
 
-We see that the expression of the marker genes using t, appear to separate
+We see that the expression of the marker genes, appear to separate
 the previously clumped clusters 2,4,6, and 11 from the other clusters, 

### 2020-07-29 Rmd/05-embeddings.Rmd
update chapter 5

@@
   ~key, ~url, ~path,
-  "gaussian",  "https://player.vimeo.com/video/439635921", here::here("video", "liminal-gaussian.mp4"),
-  "hierachical", "https://player.vimeo.com/video/439635905", here::here("video", "liminal-hierachical.mp4"),
-  "trees-01", "https://player.vimeo.com/video/439635892", here::here("video", "liminal-trees-01.mp4"),
-  "trees-02", "https://player.vimeo.com/video/439635863", here::here("video", "liminal-trees-02.mp4"),
-  "mouse-01", "https://player.vimeo.com/video/439635812", here::here("video", "liminal-mouse-01.mp4"),
-  "mouse-02", "https://player.vimeo.com/video/439635843", here::here("video", "liminal-mouse-02.mp4")
+  "gaussian",  "https://player.vimeo.com/video/439635921", here::here("img", "liminal-screenshot-gaussian.png"),
+  "hierarchical", "https://player.vimeo.com/video/439635905", here::here("img", "liminal-screenshot-hierarchical.png"),
+  "trees-01", "https://player.vimeo.com/video/439635892", here::here("img", "liminal-screenshot-trees-01.png"),
+  "trees-02", "https://player.vimeo.com/video/439635863", here::here("img", "liminal-screenshot-trees-02.png"),
+  "mouse-01", "https://player.vimeo.com/video/439635812", here::here("img", "liminal-screenshot-mouse-01.png"),
+  "mouse-02", "https://player.vimeo.com/video/439635843", here::here("img", "liminal-screenshot-mouse-02.png")
 )
@@
   knitr::include_url(video[["url"]])
+}
+
+display_screenshot <- function(key, lookup) {
+  img <- lookup[lookup$key == key, ]
+  knitr::include_graphics(img[["path"]])
 }
@@
 embedding methods, there are relatively few tools designed to assist users
-to interact with these techniques: when is an embedding suffucient for the
+to interact with these techniques: when is an embedding sufficient for the
 task at hand? Several
@@
 of multidimensional scaling methods by extending their interactive visualisation
-toolkits, **XGobi** and **GGobi** into a new tool called **GGvis** 
+software called **XGobi** and **GGobi** into a new tool called **GGvis** 
 [@Swayne1998-uq; @Swayne2003-qd; @Swayne2004-ua].
@@
 preserves or alters the geometry of their data. Using multiple concatenated and
-linked views with the tour enchances interaction techniques, and allows 
+linked views with the tour enhances interaction techniques, and allows 
 analysts to perform  cluster orientation tasks via linked highlighting and 
@@
 We have combined the tour view in a side by side layout with a scatter plot
-view as has been done in previous tour interfaces **XGobi** and **DataViwer** 
+view as has been done in previous tour interfaces **XGobi** and **DataViewer** 
 [@Buja1986-ku; @Swayne1998-uq]. These views are
@@
 
-The first three case studies qw use simulations where cluster structure and
-geometery of the underlying data is known. We start with a simple example
+The first three case studies use simulations where the cluster structure and
+geometry of the underlying data is known. We start with a simple example
 where we generated spherical clusters that are embedded well by t-SNE. Then
@@
 
-(ref:liminal-gaussian) Screenshots of the **liminal** interface applied to well clustered data, a video of the tour animation is available at https://player.vimeo.com/video/439635921. *(a)*:
+(ref:liminal-gaussian) Screenshots of the **liminal** interface applied to well clustered data, a video of the tour animation is available at https://player.vimeo.com/video/439635921.
+*(a)*: t-SNE (left) and tour (right) in side by side layout coloured by cluster.
+*(b)*: brushing the red cluster on the t-SNE view, shows the corresponding points on the tour.
 
 ```{r gaussian, fig.cap = "(ref:liminal-gaussian)"}
+display_screenshot("gaussian", vid_lookup)
 ```
@@
 locations to each other is distorted, with the orange and blue groups being
-far from each other in the tour view (figure \@ref(fig:hierachical)a). We see 
+far from each other in the tour view (figure \@ref(fig:hierarchical)a). We see 
 in this case that is difficult
 to see the sub-clusters in the tour view, however, once we zoom and highlight
-they become more apparent (figure \@ref(fig:hierachical)b). When we brush 
+they become more apparent (figure \@ref(fig:hierarchical)b). When we brush 
 the sub-clusters in the t-SNE, their
@@
 
-(ref:liminal-hierachical) Screenshots of the **liminal** interface applied to subclustered data, a video of the tour animation is .available at https://player.vimeo.com/video/43963590
+(ref:liminal-hierarchical) Screenshots of the **liminal** interface applied to sub-clustered data, a video of the tour animation is available at https://player.vimeo.com/video/43963590.
+*(a)* t-SNE has identified and spread out the clusters with substructure.
+*(b)* Using linked highlighting and brushing, the tour view shows that sub-clusters are much closer together in high dimensional space than the t-SNE view implies.
 
-```{r hierachical, fig.cap="(ref:liminal-hierachical)"}
+```{r hierarchical, fig.cap="(ref:liminal-hierarchical)"}
+display_screenshot("hierarchical", vid_lookup)
 ```
 
-```{r hierachical-h, eval = knitr::is_html_output()}
-display_vid_html("hierachical", vid_lookup)
+```{r hierarchicall-h, eval = knitr::is_html_output()}
+display_vid_html("hierarchical", vid_lookup)
 ```
@@
 (ref:trees-01) Screenshots of the **liminal** interface applied to tree structured data, a video of the tour animation is available at https://player.vimeo.com/video/439635892.
+*(a)*: Default t-SNE has difficulties embedding the tree.
+*(b)* linked highlighting shows that the backbone of the tree has been broken by the embedding.
 
 ```{r trees-01, fig.cap = "(ref:trees-01)"}
+display_screenshot("trees-01", vid_lookup)
 ```
@@
 ```{r trees-02, fig.cap = "(ref:trees-02)"}
+display_screenshot("trees-02", vid_lookup)
 ```
@@
 
-(ref:mouse-01) Screenshots of the **liminal** interface applied to single cell data. A video of the tour animation is available at https://player.vimeo.com/video/439635812.
+(ref:mouse-01) Screenshots of the **liminal** interface applied to single cell data.
+A video of the tour animation is available at https://player.vimeo.com/video/439635812.
+*(a)*: linked brushing identifies distortions in cluster placement by the embedding.
+*(b)*: t-SNE spreads out larger clusters like cluster 2, 4, 6, and 11 which are concentrated in the tour view.
+*(c)*: the tour view gives a sense of the geometry of the data; the PCs appear to form a tetrahedron like shape.
+*(d)*: brushing on the tour view highlights diffusions, points in cluster 9 are close together in the tour view are far away in the t-SNE view.
 
 ```{r mouse-01, fig.cap = "(ref:mouse-01)" }
+display_screenshot("mouse-01", vid_lookup)
 ```
@@
 
-(ref:mouse-01) Screenshots of the **liminal** tour applied to a marker gene set. A video of the tour animation is available at https://player.vimeo.com/video/439635843.
+(ref:mouse-01) Screenshots of the **liminal** tour applied to a marker gene set.
+A video of the tour animation is available at https://player.vimeo.com/video/439635843.
+*(a)*: the axis (left) and tour (right) view of 10 marker genes, using linked highlighting we see that the marker genes separate clusters 2, 4, 6, and 11 from the other groups.
+*(b)*: after zooming, it appears that the marker genes give rise to a gradient along the clusters.
 
 ```{r mouse-02, fig.cap = "(ref:mouse-02)"}
+display_screenshot("mouse-02", vid_lookup)
 ```
@@
 the previously clumped clusters 2,4,6, and 11 from the other clusters, 
-indicating that these genes are expressed in all four clusters, and the 
-magnitude of expression in these genes is similar across these groups, 
-consistent with the results of marker gene analysis.
+indicating that these genes are expressed in all four clusters (figure \@ref(fig:mouse-02)a). Using zooming, we can see a trajectory forming along
+the clusters, while the axis view shows that magnitude of expression in 
+the marker genes is similar across these separated clusters which is 
+consistent with the results of marker gene analysis (figure \@ref(fig:mouse-02)b).
 
@@
 
-We would like to thank Dr David Frazier, Dr Ursula Laa and Dr Paul Harrison for their feedback on this work.
-
-<!-- add supplementary materials here -->
+We would like to thank Dr David Frazier, Dr Ursula Laa and Dr Paul Harrison for 
+their feedback on this work. The screenshots and images were compiled using the 
+**cowplot** and **ggplot2** packages [@r-cowplot; @Wickham2016-gz].

### 2020-07-29 scripts/liminal-mouse.R

@@ -1,0 +1,100 @@
+## ----init-----------------------------------------------------------
+suppressPackageStartupMessages(library(scater))
+suppressPackageStartupMessages(library(scran))
+library(liminal)
+library(ggplot2)
+source(here::here("scripts", "helpers.R"))
+
+sce <- scRNAseq::MacoskoRetinaData(ensembl = TRUE)
+# annotation ensembl database
+anno_db <- AnnotationHub::AnnotationHub()[["AH73905"]] 
+anno <- AnnotationDbi::select(anno_db, 
+               keys=rownames(sce), 
+               keytype="GENEID", columns=c("SYMBOL", "SEQNAME", "GENEBIOTYPE"))
+rowData(sce) <- anno[match(rownames(sce), anno$GENEID), ]
+
+
+## ----qc-------------------------------------------------------------
+is.mito <- grepl("^mt-", rowData(sce)$SYMBOL)
+qcstats <- perCellQCMetrics(sce, subsets=list(Mito=is.mito))
+filtered <- quickPerCellQC(qcstats, percent_subsets="subsets_Mito_percent")
+sce <- sce[, !filtered$discard]
+
+## ----normalise-counts-----------------------------------------------
+sce <- logNormCounts(sce)
+
+## ----gene-selection-------------------------------------------------
+dec <- modelGeneVar(sce) # fits mean-variance trend using `nls`.
+hvg <- getTopHVGs(dec, prop=0.1) # selects ~10% of nrow(sce) that are HVGs
+
+
+
+## ----pca------------------------------------------------------------
+sce <- runPCA(sce, ncomponents = 25, subset_row = hvg)
+# looking at elbow-plot looks like we get the first five PCs explain most of the 
+# variation
+var_explained <- tidy_var_explained(sce)
+ggplot(var_explained, aes(x = component, y = var_explained)) +
+  geom_point()
+
+## ----cluster--------------------------------------------------------
+# the recommended workflow is to use graph based clustering on the principal 
+# components
+g <- buildSNNGraph(sce, use.dimred = 'PCA')
+colData(sce)$cluster_membership <- factor(igraph::cluster_louvain(g)$membership)
+
+# extract PCs to tour and perform tSNE on, keep cluster labels too
+pc_df <- reducedDim(sce, "PCA")
+pc_df <- dplyr::as_tibble(pc_df)
[…]

### 2020-07-29 scripts/liminal-trees.R

@@ -1,0 +1,99 @@
+## ----init-----------------------------------------------------------
+library(liminal)
+dim(fake_trees)
+library(cowplot)
+library(ggplot2)
+
+tree_img <- ggdraw() + 
+  draw_image(here::here("img", "tree.png"), scale = 0.8) +
+  draw_figure_label("Tree Structure", position = "bottom.right") +
+  theme_void()
+
+
+## ----pc-view--------------------------------------------------------
+pcs  <- prcomp(fake_trees[, -ncol(fake_trees)])
+# var explained
+summary(pcs)$importance[3,1:12]
+
+## ----pca-xy---------------------------------------------------------
+fake_trees <- dplyr::bind_cols(fake_trees, as.data.frame(pcs$x))
+
+pca_xy <- ggplot(fake_trees, aes(x = PC1, y = PC2, color = branches)) +
+  geom_point() +
+  scale_color_manual(values = limn_pal_tableau10()) +
+  labs(caption = "PCA") +
+  theme(aspect.ratio = 1) +
+  theme_void() 
+
+## ----tsne-xy--------------------------------------------------------
+set.seed(2099)
+tsne <- Rtsne::Rtsne(dplyr::select(fake_trees, dplyr::starts_with("dim")))
+
+tsne_df <- data.frame(tsneX = tsne$Y[,1],
+                      tsneY = tsne$Y[,2],
+                      branches = fake_trees$branches)
+
+tsne_xy  <- ggplot(tsne_df, aes(x = tsneX, y = tsneY, color = branches)) +
+  geom_point() +
+  scale_color_manual(values = limn_pal_tableau10()) + 
+  guides(color = FALSE) +
+  labs(caption = "tSNE") +
+  theme(aspect.ratio = 1) +
+  theme_void() 
+
+## ----tour----------------------------------------------------------
+
+# now together
+limn_tour_link(
+  tour_data = dplyr::select(fake_trees, PC1:PC12, branches),
+  embed_data = tsne_df,
+  embed_color = branches,
+  tour_color = branches
+)
+
+
+
+## ----pca-init--------------------------------------------------------
+tsne_v2 <- Rtsne::Rtsne(
+  dplyr::select(fake_trees, dplyr::starts_with("dim")),
[…]

### 2020-07-29 Rmd/05-embeddings.Rmd
fix bib

@@
 linked interactions with the tour. As we have shown in the simulation case
-studies, the t-SNE method can produce misleading embeddings but can be
+studies, the t-SNE method can produce misleading embeddings which can be
 detected through the use linked brushing and highlighting. In the case when
@@
 to the single cell analyst who wants to make their embeddings more 
-interpretable
+interpretable.
 

### 2020-07-29 Rmd/04-intron.Rmd
typos
paper

@@
 
-In this chapter, we describe methodology for establishing data descriptors by turning coverage vectors into long form tidy data using **superintronic**
+In this paper, we describe methodology for establishing data descriptors by turning coverage vectors into long form tidy data using **superintronic**
 and **plyranges**. We provide a workflow below
@@
 [@Ritchie2015-mj], like our proposed index method in @Lee2019-mf. The gene
-candiates obtain by our thresholds have been validated by the Heath lab
+candidates obtain by our thresholds have been validated by the Heath lab
 using qPCR.

### 2020-07-29 index.Rmd
update toc; rebuild
add lay summary

@@
 degrees: 'B.MatComSci., University of Adelaide; MSc.Bioinformatics. University of Melbourne'
-bibliography: [bib/plyranges.bib, bib/fluentGenomics.bib, bib/thesisrefs.bib]
+bibliography: [bib/plyranges.bib, bib/fluentGenomics.bib, bib/superintronic.bib, bib/liminal.bib, bib/thesisrefs.bib]
 link-citations: yes
@@
 
-This is the website for my PhD thesis at Monash University (Australia), titled "Fluent statistical computing interfaces for biological data analysis". 
-It is a work in progress.
+This work contributes new computing interfaces for the analysis of data 
+collected from technologies that measure changes in biology, like gene 
+expression. The first interface is a new tool that unites data measured along
+the genome under a common language of transformation. This interface enables
+analysts to interpret and combine multiple results from different technologies
+in a clear and simple way. The second interface provides a framework for 
+identifying interesting regions along the genome. The third interface empowers
+analysts to interrogate complex models using visual diagnostics. All 
+are provided as open source software packages.
 

### 2020-07-30 Rmd/01-intro.Rmd
typo
bold package names

@@
 
-(ref:wflow-integrate) Visualisation is a tool for making sense of models. Here we explore, how well a model has captured the structure within the data using interactive graphics.
+(ref:wflow-explore) Visualisation is a tool for making sense of models. Here we explore, how well a model has captured the structure within the data using interactive graphics.
 
@@
 The approach taken by the suite of software packages collectively known as the 
-*tidyverse*  is an attempt to formalise aspects of the EDA process in the R
+**tidyverse**  is an attempt to formalise aspects of the EDA process in the R
 programming language under a single semantic known as *tidy data* 
@@
 mapped to graphical aesthetics of plots.
-This paradigm enables the grammar of graphics as implemented by *ggplot2* [@Wickham2016-gz; @Wilkinson2005-kq].
-User interfaces as implemented by *tidyverse*, and in particular the *dplyr* package, are *fluent*;  they form a domain specific language (DSL) that gives users a mental model for performing and composing common data transformation tasks
+This paradigm enables the grammar of graphics as implemented by **ggplot2** [@Wickham2016-gz; @Wilkinson2005-kq].
+User interfaces as implemented by **tidyverse**, and in particular the **dplyr** package, are *fluent*;  they form a domain specific language (DSL) that gives users a mental model for performing and composing common data transformation tasks
 [@Wickham2017-dplyr; @FowlerFluent].
@@
 in fields such as high-throughput biology where domain specific semantics are 
-required. This is particularly true in the *Bioconductor* ecosystem, where much 
+required. This is particularly true in the Bioconductor ecosystem, where much 
 thought has gone into the design of data structures that enable interoperability
@@
 to range-based genomics data and develops a *fluent* interface to transforming
-it called *plyranges*. The software provides a framework to an assist
+it called **plyranges**. The software provides a framework to an assist
 an analyst to compose queries on genomics datasets. This chapter has been 
@@
 we describe a simple end-to-end workflow for integrating 
-results along the genome using *plyranges*. This workflow also shows that
+results along the genome using **plyranges**. This workflow also shows that
 our grammar based approach does not impair interoperability between
-the *tidyverse* and *Bioconductor* approaches, and in fact they work seamlessly
+the **tidyverse** and Bioconductor approaches, and in fact they work seamlessly
 together. This chapter has been published as @Lee2020-ie.
@@
 In chapter \@ref(ch:intron) we explore the limits of the *tidy data*
-semantic by extending *plyranges* to analyse coverage estimated on RNA-seq
-data by developing a new software tool called *superintronic*. We show that the
+semantic by extending **plyranges** to analyse coverage estimated on RNA-seq
+data by developing a new software tool called **superintronic**. We show that the
 long-form tidy representation is an effective way of combining the experimental
 design and reference annotations into a single genomic data structure for
-exploration. We use *superintronic* to develop a framework for discovering
+exploration. We use **superintronic** to develop a framework for discovering
 interesting regions of coverage and apply our approach to integrating intron
@@
 
-Chapter \@ref(ch:tsne) introduces a software package called *liminal* for 
+Chapter \@ref(ch:tsne) introduces a software package called **liminal** for 
 constructing these views and a user interaction framework for identifying 

### 2020-07-30 Rmd/00-c-declaration.Rmd
formatting

@@
 This thesis includes 3 publications, two of which have been published and one 
-which has not yet been submitted [@Lee2019; @Lee2020-ie] . As the core theme of my thesis is the development of software
+which has not yet been submitted [@Lee2019; @Lee2020-ie]. As the core theme of my thesis is the development of software
 interfaces for biological data analysis, and given the collaborative nature
@@
 papers in this thesis reflect and acknowledge the contributions of my 
-co-authors, including my  primary supervisors Professor Dianne Cook (DC) in the 
+co-authors, including my  primary supervisors Professor Dianne Cook in the 
 Department of Econometrics and Business Statistics at Monash University
@@
 \hline
-5 & Casting multiple shadows: high-dimensional interactive data visualisation with tours and embeddings & To be submitted & 80\%.
+5 & Casting multiple shadows: high-dimensional interactive data visualisation with tours and embeddings & To be submitted & 85\%.
 Concept, software development, and manuscript writing & (1) Dianne Cook, Concept and manuscript writing 15\% & \multicolumn{1}{c|}{N} \\

### 2020-07-30 Rmd/04-intron.Rmd
formatting

@@
 FASTQ files were aligned to the GRCz11 reference genome using subjunc with
-the default parameters called from Rsubread to produce BAM files for each
+the default parameters called from **Rsubread** to produce BAM files for each
 sample [@Liao2013-eb; @Liao2019-qn]. The coverage was then estimated directly from the set of BAM files
@@
 obtained as GFF files from RefSeq and used that to construct the exonic
-and intronic parts of each gene as _GRanges_ object using *superintronic*. 
+and intronic parts of each gene as _GRanges_ object using **superintronic**. 
 We further filtered genes that had a single exon or genes that were overlapping

### 2020-07-30 Rmd/05-embeddings.Rmd
update caption
formatting

@@
 have followed a standard workflow for pre-processing and normalizing this data 
-(described by @Amezquita2020-at): we performed QC using the the `scater`
+(described by @Amezquita2020-at): we performed QC using the the **scater**
 by removing cells with high proportion of mitochondrial gene expression and low
 numbers of genes detected, we log-transformed and normalised the expression 
-values and  finally selected highly variable genes (HVGs) using `scran` 
+values and  finally selected highly variable genes (HVGs) using **scran** 
 [@McCarthy2017; @Lun2016]. The top ten percent of HVGs were used to subset
@@
 
-(ref:mouse-01) Screenshots of the **liminal** tour applied to a marker gene set.
+(ref:mouse-02) Screenshots of the **liminal** tour applied to a marker gene set.
 A video of the tour animation is available at https://player.vimeo.com/video/439635843.
@@
 can be independent of the tour or embedding views. The estimation of neighbours
-can be computed efficiently with the `FNN` or `RcppAnnoy` packages
+can be computed efficiently with the **FNN** or ***RcppAnnoy** packages
 on the server side [@fnn-pkg; @rcpp-annoy-pkg]. In place of highlighting, one 

### 2020-07-30 Rmd/01-intro.Rmd
clarification

@@
 data science, where constraints on computation time and memory, in addition
-to the analyst's time, results in EDA becoming increasingly difficult and 
-neglected.
+to the analyst's time, makes EDA difficult and neglected, which impacts the robustness and reliability of any downstream analysis.
 
@@
 in fields such as high-throughput biology where domain specific semantics are 
-required. This is particularly true in the Bioconductor ecosystem, where much 
+required (figure \@ref(fig:wflow-wrangle)). This is particularly true in the Bioconductor ecosystem, where much 
 thought has gone into the design of data structures that enable interoperability
@@
 a single measurement assay or that only one aspect of a measurement assay
-will be of interest to the biological question under study. While there are 
+will be of interest to the biological question under study (figure \@ref(fig:wflow-integrate)). While there are 
 many approaches to integrating data sets from multiple assays using multivariate
@@
 integration of visualisation with model-based summaries of high-dimensional
-data sets. We focus on a common tool for EDA (especially applied to single-cell
+data sets (figure \@ref(fig:wflow-explore)). We focus on a common tool for EDA (especially applied to single-cell
 transcriptomics): non-linear dimension reduction (NLDR). We consider the incorporation of 

### 2020-07-30 Rmd/01-intro.Rmd
formatting

@@
 leading to new hypotheses that can be tested and formalised using 
-statistical modelling (figures \@ref(fig:wflow-cycle)). As datasets grow in complexity and become
+statistical modelling (figure \@ref(fig:wflow-cycle)). As datasets grow in complexity and become
 increasingly heterogeneous and multidimensional, 
@@
 (ref:wflow-integrate) Data from multiple assays can be combined together using modelling and transformations.
-Data and model results can combined, tidied and transformed to have a common representation across different granularities of the genome.
+Data and model results can be combined, tidied and transformed to have a common representation across different granularities of the genome.
 

### 2020-07-30 Rmd/05-embeddings.Rmd
formatting
add supplementary

@@
 and interpretation of embedding methods as well. @Sedlmair2013-pn showed that
-2D scatter plots were mostly sufficient for detecting class separation, however
+scatter plots were mostly sufficient for detecting class separation, however
 they also noted that often multiple embeddings were required. For the task of
@@
 of non-linear embedding techniques were more likely to consider clusters of
-points on a 2D scatter plot to be the result of a spurious embedding compared
+points on a scatter plot to be the result of a spurious embedding compared
 to advanced users who were aware of the inner workings of the embedding
@@
 onto a low-dimensional basis matrix, and is represented as a
-animated visualization [@Asimov1985-vp; @Buja1986-zr]. Given the dynamic nature
-of the tour, user interaction is  important for controlling and exploring the 
-visualisation: the tour has been used previously by @Wickham2015-cx and  
-exploring statistical model fits and by @Buja1996-fk for exploring the space of 
-factorial experimental designs.
+animated visualization [@Asimov1985-vp; @Buja1986-zr]. Given the dynamic 
+nature of the tour, user interaction is  important for controlling and 
+exploring the visualisation: the tour has been used previously by 
+@Wickham2015-cx and exploring statistical model fits and by @Buja1996-fk for 
+exploring the space of factorial experimental designs.
 
@@
 
-The videos presented show the visual 
+We *strongly* recommend viewing the linked videos for each case study
+while reading. The videos presented show the visual 
 appearance of the **liminal** interface, and how we can interact with the tour
-via the controls previously described. Both code and videos available for 
-reproducing this paper are available at https://github.com/sa-lee/paper-liminal.
+via the controls previously described.
 
@@
 
-We repeat the process of viewing Gaussian clusters from the 
+Next we view Gaussian clusters from the 
 _Multi Challenge Dataset_, a benchmark simulation data set for clustering tasks
 [@Rauber2009-vh]. This dataset has two Gaussian clusters with equal covariance 
-embedded in $10-d$, and a third cluster
+embedded in 10-$d$, and a third cluster
 with hierarchical structure. This cluster has two 3-$d$ clusters embedded in 
@@
 
-Our holistic framework is also useful in real data situations, as displayed
-in our second case study. The tour when combined with t-SNE allowed us
+Our framework can also be useful in practice, as displayed
+in the fourth case study. The tour when combined with t-SNE allowed us
 to identify clusters, while giving us an idea of their orientation to 
@@
 
-In future implementations it will be possible to shift from one to one linked
-brushing to one to many linked brushing. This form of brushing acts directly on
-a $k$-nearest neighbours ($k$-nn) graph computed from a
+We have shown in the case studies, that one to one linked brushing can
+be used to identify distortions in the embedding, however we would
+like extend this to one to many linked brushing, which would allow us to directly interrogate neighbourhood preservation.
+This form of brushing acts directly on a $k$-nearest neighbours ($k$-nn) graph computed from a
 reference dataset: when a user brushes over a region in the embedding,
@@
 nearest neighbours (for example a distance matrix, or the complete data matrix) 
-can be independent of the tour or embedding views. The estimation of neighbours
-can be computed efficiently with the **FNN** or ***RcppAnnoy** packages
-on the server side [@fnn-pkg; @rcpp-annoy-pkg]. In place of highlighting, one 
-could use opacity or binned colour scales to encode distances or ranks instead 
-of the neighbouring points. Brush composition, such as 'and', 'or',
-or 'not' brushes, could be used to interrogate mismatches between the 
-$k$-nn graphs estimated from both the embedding and tour views.
+can be independent of the tour or embedding views. In place of 
+highlighting, one  could use opacity or binned colour scales to encode 
+distances or ranks instead of the neighbouring points. We have begun 
+implementing this brush in **liminal**,  using the  **FNN** or **RcppAnnoy**
+packages for fast neighbourhood estimation on the server side, however there
+are still technicalaties that need be resolved [@fnn-pkg; @rcpp-annoy-pkg]. 
+Brush composition, such as 'and', 'or', or 'not' brushes, could be used to 
+further investigate mismatches between the $k$-nn graphs estimated from both the embedding and tour views.
 
@@
 **cowplot** and **ggplot2** packages [@r-cowplot; @Wickham2016-gz].
+
+## Supplementary Materials {-}
+
+Code, data, and video for reproducing this paper are available at https://github.com/sa-lee/paper-liminal. Direct links to videos for viewing
+online are available in table \@ref(tab:vimeo).
+
+```{r vimeo}
+knitr::kable(dplyr::select(vid_lookup, `Case Study` = key, `URL` = url), 
+             caption = "Case Study Videos",
+             booktabs = TRUE)
+```

### 2020-07-30 Rmd/00-c-declaration.Rmd
update declaration

@@
 
-This thesis includes 3 publications, two of which have been published and one 
-which has not yet been submitted [@Lee2019; @Lee2020-ie]. As the core theme of my thesis is the development of software
+This thesis includes 4 publications, two of which have been published and two 
+which have not been submitted yet [@Lee2019; @Lee2020-ie].
+As the core theme of my thesis is the development of software
 interfaces for biological data analysis, and given the collaborative nature
@@
 papers in this thesis reflect and acknowledge the contributions of my 
-co-authors, including my  primary supervisors Professor Dianne Cook in the 
-Department of Econometrics and Business Statistics at Monash University
-and Associate Professor Matthew Ritchie in the Epigenetics and Development 
-Division at the Walter and Eliza Hall Institute. The following table details
-the publications, including my and my fellow co-authors contributions:
+co-authors. Of special note is chapter \@ref(ch:intron), which reflects my 
+contribution (in the form of a software package) to the submitted manuscript entitled 
+_Covering all your bases: incorporating intron signal from RNA-seq data_
+[@Lee2019-mf], and includes new data analyses not found in the paper. 
+The following table details the publications, including my and my fellow co-authors contributions:
 
@@
 \RaggedRight\textbf{Thesis Chapter}  & \RaggedRight\textbf{Publication Title}  & \RaggedRight\textbf{Status (published, in press, accepted or returned for revision)}  & \RaggedRight\textbf{Nature and} {\%} \RaggedRight\textbf{of student contribution} & \RaggedRight\textbf{Co-author name(s) Nature and} {\%} \RaggedRight\textbf{of Co-author’s contribution} & \RaggedRight\textbf{Co-author(s), Monash student Y/N} \\ \hline
-2 & plyranges: a grammar of genomic data transformation & Published  & 70\%.
+2 & \textbf{plyranges}: a grammar of genomic data transformation & Published  & 70\%.
 Concept, software development, data analysis, and manuscript writing & (1) Dianne Cook, Concept and manuscript revision 10\% (2) Michael Lawrence, Concept and software development  20\% & \multicolumn{1}{c|}{N}  \\
@@
 Concept, software development, data analysis and manuscript writing  & (1) Michael Lawrence, manuscript feedback and editing 5\% (2) Michael I Love, Concept, data analysis and manuscript writing 35\% & \multicolumn{1}{c|}{N} \\
+\hline
+4 & Expoloratory coverage analysis with \textbf{superintronic} and \textbf{plyranges} & Published & 50\%.
+Concept, software development, data analysis and manuscript writing  & (1) Charity Law, Concept, manuscript feedback and editing 45\% (2) Matthew Ritchie, manuscript writing and editing 5\% & \multicolumn{1}{c|}{N} \\
 \hline
@@
 \FloatBarrier
-
-Chapter \@ref(ch:intron), reflects my contribution (in the form of a 
-software package) to the submitted manuscript entitled 
-_Covering all your bases: incorporating intron signal from RNA-seq data_
-[@Lee2019-mf], and includes new data analyses not found in the paper.  It is joint work under the supervision of Dr Charity Law
-and Associate Professor Matthew Ritchie.
 

### 2020-07-31 Rmd/05-embeddings.Rmd
spelling

@@
 packages for fast neighbourhood estimation on the server side, however there
-are still technicalaties that need be resolved [@fnn-pkg; @rcpp-annoy-pkg]. 
+are still technicalities that need be resolved [@fnn-pkg; @rcpp-annoy-pkg]. 
 Brush composition, such as 'and', 'or', or 'not' brushes, could be used to 

### 2020-08-03 Rmd/04-intron.Rmd
fix typos found in #10

@@
 The gene annotation files were
-obtained as GFF files from RefSeq and used that to construct the exonic
+obtained as GFF files from RefSeq and used to construct the exonic
 and intronic parts of each gene as _GRanges_ object using **superintronic**. 
@@
 computing data descriptors on in order to detecting IR-like coverage traces.
-Across each combination of genotype and line, we first $log_2$ transformed the coverage score with an offset of one, and then intersected the coverage _GRanges_ with exonic and intronic features of each gene.
+Across each combination of genotype and line, we first $\log_2$ transformed the coverage score with an offset of one, and then intersected the coverage _GRanges_ with exonic and intronic features of each gene.
 For each gene,
-the weighted mean and standard deviation of the log-transformed coverage score were computed over all intron and exon parts within the groupings of 
+the mean and standard deviation of the log-transformed coverage score weighted by the number of bases covered were computed over all intron and exon parts within the groupings of 
 genotype and line. We also computed the bases above descriptor for intron
@@
 
-We have shown how to coverage can be represented in the tidy data framework
+We have shown how coverage can be represented in the tidy data framework
 and integrated with experimental metadata and reference annotations. This

### 2020-08-03 _render.R
fix conversion to png

@@ -24,7 +24,7 @@ for (i in img_pdf) {
   file_pdf <- paste0("img/", i)
   dest_pdf <- paste0("img/", sub("pdf$", "png", i))
   if (file.exists(dest_pdf)) {
-    break
+    next
   }
   magick::image_write(
     magick::image_read(file_pdf, 300), dest_pdf, "png",

### 2020-08-03 Rmd/04-intron.Rmd
addresses comments and closes #10

@@
 Our procedure mostly finds genes with IR like profiles across all groups (28 shared between all four) or that is unique to a single group, since we do not consider looking at differential IR and run our thresholding separately for each group of replicates.
-The rnpc3 knockout lines
-share the largest overlap in results.
+The rnpc3 knockout lines share the largest overlap in results.
 
@@
 ```{r si-cvg-ccdc43, echo=FALSE, fig.cap = "(ref:si-cvg-ccdc43)", out.width="80%"}
-knitr::include_graphics(here::here("img", "ccdc43.pdf"))
+ext <- ifelse(knitr::is_latex_output(), ".pdf", ".png")
+
+knitr::include_graphics(here::here("img", paste0("ccdc43", ext)))
 ```
 
-(ref:si-cvg-nat15) The gene nat15 exhibits another example of U12 intron retention (located at around 27.93Mb) in the rnpc3 knockout groups.
+(ref:si-cvg-nat15) The gene nat15 exhibits another example of U12 intron retention, located at around 27.93Mb, in the rnpc3 knockout groups.
 
 ```{r si-cvg-nat15,  echo=FALSE, fig.cap = "(ref:si-cvg-nat15)", out.width="80%"}
-knitr::include_graphics(here::here("img", "nat15.pdf"))
+knitr::include_graphics(here::here("img", paste0("nat15", ext)))
 ```
@@
 ```{r si-cvg-mapk3,  echo=FALSE, fig.cap = "(ref:si-cvg-mapk3)", out.width="80%"}
-knitr::include_graphics(here::here("img", "mapk3.pdf"))
+knitr::include_graphics(here::here("img", paste0("mapk3", ext)))
 ```
@@
 ```{r si-cvg-tspan31, echo=FALSE, fig.cap = "(ref:si-cvg-tspan31)", out.width="80%"}
-knitr::include_graphics(here::here("img", "tspan31.pdf"))
+knitr::include_graphics(here::here("img", paste0("tspan31", ext)))
 ```

### 2020-08-03 Rmd/05-embeddings.Rmd
addressing typos in #9

@@
 the tour. A tour is a sequence of projections of a high dimensional dataset
-onto a low-dimensional basis matrix, and is represented as a
+onto a low-dimensional basis matrix, and is represented as an
 animated visualization [@Asimov1985-vp; @Buja1986-zr]. Given the dynamic 
@@
 exploring the visualisation: the tour has been used previously by 
-@Wickham2015-cx and exploring statistical model fits and by @Buja1996-fk for 
+@Wickham2015-cx for exploring statistical model fits and by @Buja1996-fk for 
 exploring the space of factorial experimental designs.
@@
 $X$ such as the $k$-nearest neighbours graph or via a kernel transformation.
-Multidimensional scaling (MDS) is a class of DR methods that aims to construct
+Multidimensional scaling (MDS) is a class of DR method that aims to construct
 an embedding  $Y$ such that the pair-wise distances (inner products) in $Y$ 
@@
 $\mathbf{y_i}$ and $\mathbf{y_j}$ are modelled as a symmetric probability
-distribution using a Cauchy kernel:
+distribution using a t-distribution with one degree of freedom (Cauchy kernel):
 
@@
 
-Like, when using other DR techniques, the tour assumes we have a real data matrix
+Like, when, using other DR techniques, the tour assumes we have a real data matrix
 $X$ consisting of $n$ observations in $p$ dimensions. First, the tour generates 
@@
 Diffusions occur when points are far from each other in the embedding view are
-near in the original data.  Whether, points are near or far is reliant on the
+near in the original data.  Whether, points are near, or far is reliant on the
 distance metric used; distortions and diffusions can be thought of as the
@@
 construct intermediary frames, means the tour is a dynamic visualisation 
-technique. Generally, the user would set $d=2$ the tour is visualised as
-a animated scatter plot. This process of constructing bases and intermediate 
+technique. Generally, the user would set $d=2$ and the tour is visualised as
+an animated scatter plot. This process of constructing bases and intermediate 
 frames and visualising the resulting projection is akin to making a "flip book" 
@@
 
-The next section steps through case studies cases of our approach using simulated
-and real data.
+The next section steps through case studies of our approach using simulations
+and an application to single cell RNA-seq data.
 
@@
 
-Next explore some simulated noisy tree structured data (figure
+Next we explore some simulated noisy tree structured data (figure
 \@ref(fig:fake-trees)), our interest here is how t-SNE visualisations break

### 2020-08-03 Rmd/06-conclusion.Rmd
addressing typos in #9

@@
 cell and long-read based transcriptomics data, where scalability and much
-larger design matrices would be come an issue. The **superintronic** package
+larger design matrices would become an issue. The **superintronic** package
 is available to download from https://github.com/sa-lee/superintronic.
@@
 whether the semantics of our grammar can be extended to data that can not be 
-easily reshaped into long form tidy representations. Moreover, further work
+effeciently reshaped into long form tidy representations. Moreover, further work
 is required to explore the design space of grammars for data 
-transformations and grammars for graphics when the data large, multifaceted
+transformations and grammars for graphics when the data are large, multifaceted
 and non-rectangular.

### 2020-08-03 Rmd/05-embeddings.Rmd
fixing typos in #9

@@
 default parameters with reference to the global view of the data set. 
+This simulation aims to mimick branching trajectories of cell differentiation:
+if there were only mature cells, we would just see the tips of the branches 
+which have a hierarchical pattern of clustering.
 
 (ref:fake-trees) Example high-dimensional tree shaped data, $n = 3000$ and $p = 100$.
-*(a)* The true data lies 2-d tree like structure consisting of ten branches.
+*(a)* The true data lies on 2-$d$ tree consisting of ten branches.
 This data is available in the **phateR** package and is simulated via diffusion-limited aggregation (a random walk along the branches of the tree) with Gaussian noise added [@Moon2019-ce].
@@
 
-Figure \@(ref:trees-02) shows that this selection of parameters results in the 
+The video linked in figure \@ref(fig:trees-02) shows that this selection of 
+parameters results in the 
 tips of the branches (the three black dots in figure \@ref(fig:fake-trees)a) 
@@
 in this area generally use non linear DR methods for verification and identification
-of clusters and developmental trajectories (i.e., case study 1). For clustering,
+of clusters and developmental trajectories (i.e., case study 1). For clustering
 workflows the primary task is verify the existence of clusters and then begin to 
@@
 linked brushing approaches to look at neighbourhood preservation
-and look at marker genes through the lens of the tour.
+and look at marker genes through the lens of the tour. The data we have
+selected for this case study has features similar to those found in case studies
+2 and 3.
 
@@
 We tour the first five PCs (approximately 20% of the variance in expression),
-along side the t-SNE view which was computed from all 25 PCs. 
-Due to latency of the interface we do weighted sample of rows based on
+alongside the t-SNE view which was computed from all 25 PCs. 
+Due to latency of the interface we do a weighted sample of  the rows based on
 cluster membership, leaving us with approximately 10 per cent of the original
@@
 weighting of each gene. We run the tour directly on the 
-log normalised expression values using the same subset as before figure \@ref(fig:mouse-02).
+log normalised expression values using the same subset as before (figure \@ref(fig:mouse-02)).
 

### 2020-08-03 Rmd/05-embeddings.Rmd
update screenshots, strongly prompt the viewer to watch the videos

@@
 We *strongly* recommend viewing the linked videos for each case study
-while reading. The videos presented show the visual 
+while reading. The videos are available to view online from the links
+in the captions of each sections figures. The videos presented show the visual 
 appearance of the **liminal** interface, and how we can interact with the tour
-via the controls previously described.
+via the controls previously described. If you are unable to view the videos,
+the figures in each case study consist of screenshots that summarise what is
+learned from combining the tour and an embedding view.
 
@@
 
-We see in figure \@ref(fig:gaussian) and linked video that t-SNE has 
+From the video linked inn figure \@ref(fig:gaussian), we learn that t-SNE has 
 correctly split out each cluster and laid them out in a star like formation. 
@@
 the t-SNE layout (figure \@ref(fig:gaussian)a). 
-This can be confirmed via brushing the t-SNE view (figure \@ref(fig:gaussian)b).
+This can be seen via brushing the t-SNE view (figure \@ref(fig:gaussian)b).
 
@@
 that are each equidistant from each other and have the same covariance structure.
-Again t-SNE has correctly identified the sub-clusters but their relative 
+From the video linked in figure \@ref(fig:hierarchical), we see 
+that t-SNE has correctly identified the sub-clusters. However, their relative 
 locations to each other is distorted, with the orange and blue groups being
@@
 default parameters with reference to the global view of the data set. 
-This simulation aims to mimick branching trajectories of cell differentiation:
+This simulation aims to mimic branching trajectories of cell differentiation:
 if there were only mature cells, we would just see the tips of the branches 
@@
 
-We see that the t-SNE view has been unable to recapitulate the
-topology of the tree - the backbone (blue) branch has been split into three
-fragments (figure \@ref(fig:trees-01)a). We can see this immediately via the linked highlighting over both
+From the linked video, we see that the t-SNE view has been unable to 
+recapitulate the topology of the tree - the backbone (blue) branch has been 
+split into three fragments (figure \@ref(fig:trees-01)a). We can see 
+this immediately via the linked highlighting over both
 plots. If we click on the legend for the zero branch, the blue coloured points
@@
 From, here it becomes apparent from the tour view that the blue branch forms
-the backbone of the tree and is connected to all other branches. We can also
-again see that cluster sizes formed via t-SNE can be misleading; from the tour
+the backbone of the tree and is connected to all other branches. From the video 
+it is easy to see that cluster sizes formed via t-SNE can be misleading; from the tour
 view there is a lot of noise along the branches, while this does not appear to
@@
 initialisation and increase the perplexity from 30 to 100. We then combine
-these results with our tour view as displayed in figure \@ref(fig:trees-02).
+these results with our tour view as displayed in the linked video in the caption
+of figure \@ref(fig:trees-02).
 
@@
 data size - 4,590 cells. Although this is not ideal, it still allows us to
-get a sense of the shape of the clusters as seen in figure \@ref(fig:mouse-01).
+get a sense of the shape of the clusters as seen from the linked video in figure \@ref(fig:mouse-01).
 
@@
 
-We observe that the embedding has mostly captured the
+From the video linked in figure \@ref(fig:mouse-01), we learn that the embedding has mostly captured the
 clusters relative location to each other to their location in high dimensional space, with a notable exception of points in cluster 3 and 10 as shown with
@@
 weighting of each gene. We run the tour directly on the 
-log normalised expression values using the same subset as before (figure \@ref(fig:mouse-02)).
+log normalised expression values using the same subset as before.
 
@@
 
-We see that the expression of the marker genes, appear to separate
+From the video linked in figure \@ref(fig:mouse-02), we see that the 
+expression of the marker genes, appear to separate
 the previously clumped clusters 2,4,6, and 11 from the other clusters, 
-indicating that these genes are expressed in all four clusters (figure \@ref(fig:mouse-02)a). Using zooming, we can see a trajectory forming along
+indicating that these genes are expressed in all four clusters (figure \@ref(fig:mouse-02)a). After zooming, we can see a trajectory forming along
 the clusters, while the axis view shows that magnitude of expression in 
@@
 ```{r vimeo}
-knitr::kable(dplyr::select(vid_lookup, `Case Study` = key, `URL` = url), 
+vid_lookup$`Case Study` <- c(1, 2, 3,3,4,4)
+knitr::kable(dplyr::select(vid_lookup, `Case Study`, 
+                           `Example` = key, `URL` = url), 
              caption = "Case Study Videos",

### 2020-08-03 Rmd/06-conclusion.Rmd
spelling

@@
 whether the semantics of our grammar can be extended to data that can not be 
-effeciently reshaped into long form tidy representations. Moreover, further work
+efficiently reshaped into long form tidy representations. Moreover, further work
 is required to explore the design space of grammars for data 

### 2020-08-03 Rmd/05-embeddings.Rmd
update reference
adjust fig sizes

@@
 We *strongly* recommend viewing the linked videos for each case study
-while reading. The videos are available to view online from the links
-in the captions of each sections figures. The videos presented show the visual 
+while reading. Links to the videos are available in table \@ref(tab:vimeo) 
+and in the figures for each case study. The videos presented show the visual 
 appearance of the **liminal** interface, and how we can interact with the tour
@@
 (ref:liminal-gaussian) Screenshots of the **liminal** interface applied to well clustered data, a video of the tour animation is available at https://player.vimeo.com/video/439635921.
-*(a)*: t-SNE (left) and tour (right) in side by side layout coloured by cluster.
-*(b)*: brushing the red cluster on the t-SNE view, shows the corresponding points on the tour.
 
-```{r gaussian, fig.cap = "(ref:liminal-gaussian)"}
+```{r gaussian, fig.cap = "(ref:liminal-gaussian)", out.height = "75%"}
 display_screenshot("gaussian", vid_lookup)
@@
 (ref:liminal-hierarchical) Screenshots of the **liminal** interface applied to sub-clustered data, a video of the tour animation is available at https://player.vimeo.com/video/43963590.
-*(a)* t-SNE has identified and spread out the clusters with substructure.
-*(b)* Using linked highlighting and brushing, the tour view shows that sub-clusters are much closer together in high dimensional space than the t-SNE view implies.
 
-```{r hierarchical, fig.cap="(ref:liminal-hierarchical)"}
+```{r hierarchical, fig.cap="(ref:liminal-hierarchical)", out.height = "75%"}
 display_screenshot("hierarchical", vid_lookup)
@@
 (ref:trees-01) Screenshots of the **liminal** interface applied to tree structured data, a video of the tour animation is available at https://player.vimeo.com/video/439635892.
-*(a)*: Default t-SNE has difficulties embedding the tree.
-*(b)* linked highlighting shows that the backbone of the tree has been broken by the embedding.
 
-```{r trees-01, fig.cap = "(ref:trees-01)"}
+```{r trees-01, fig.cap = "(ref:trees-01)", out.height = "75%" }
 display_screenshot("trees-01", vid_lookup)
@@
 
-(ref:mouse-01) Screenshots of the **liminal** interface applied to single cell data.
-A video of the tour animation is available at https://player.vimeo.com/video/439635812.
-*(a)*: linked brushing identifies distortions in cluster placement by the embedding.
-*(b)*: t-SNE spreads out larger clusters like cluster 2, 4, 6, and 11 which are concentrated in the tour view.
-*(c)*: the tour view gives a sense of the geometry of the data; the PCs appear to form a tetrahedron like shape.
-*(d)*: brushing on the tour view highlights diffusions, points in cluster 9 are close together in the tour view are far away in the t-SNE view.
+(ref:mouse-01) Screenshots of the **liminal** interface applied to single cell data, a video of the tour animation is available at https://player.vimeo.com/video/439635812.
 
-```{r mouse-01, fig.cap = "(ref:mouse-01)" }
+```{r mouse-01, fig.cap = "(ref:mouse-01)", out.height = "75%"}
 display_screenshot("mouse-01", vid_lookup)
@@
 
-(ref:mouse-02) Screenshots of the **liminal** tour applied to a marker gene set.
-A video of the tour animation is available at https://player.vimeo.com/video/439635843.
-*(a)*: the axis (left) and tour (right) view of 10 marker genes, using linked highlighting we see that the marker genes separate clusters 2, 4, 6, and 11 from the other groups.
-*(b)*: after zooming, it appears that the marker genes give rise to a gradient along the clusters.
+(ref:mouse-02) Screenshots of the **liminal** tour applied to a marker gene set, a video of the tour animation is available at https://player.vimeo.com/video/439635843.
 
-```{r mouse-02, fig.cap = "(ref:mouse-02)"}
+```{r mouse-02, fig.cap = "(ref:mouse-02)", out.height = "75%"}
 display_screenshot("mouse-02", vid_lookup)

### 2020-08-03 Rmd/00-d-acknowlegements.Rmd
add acknowledgements

# Acknowledgements {-}

I have had the great privilege to work with and be inspired by many people
during my PhD candidature. It has been a challenging and rewarding experience
and I would not have been able to get through it without the support of my
advisers, colleagues, friends and family.

I am deeply grateful to my supervisors: Dianne Cook, Matthew Ritchie, and Paul
Harrison. Di has been instrumental in helping me grow as a researcher
and teacher; she has been patient and always full of interesting insights and 
enthusiasm for the practice of doing data analysis. I am sorry that I have not named any of my software after Australian animals. Matt is one of the kindest and most generous scientists I know, thank you for allowing me to be a part of your lab and for all of the lab lunches! Paul has been a great advocate for me and my work, thank you for getting down in the trenches and always being available to help when I nag you about technical difficulties I'm having.

My PhD would have been very different if I did not have the chance to collaborate with and learn from Michael Lawrence. Thank you for hosting me at Genentech in San Francisco and pushing me to grow as a programmer and researcher. Likewise, I am thankful to have had the opportunity to work with Michael Love who has been an enthusiastic collaborator and supporter. Hopefully I can visit North Carolina soon!

Thank you to my PhD committee: David Frazier, David Powell, and Farshid Vahid. I am lucky to be in such a supportive environment at Monash EBS, a big thank you to Gael Martin and Rob Hyndman for their advice and guidance during the PhD.

I am thankful to be part of the broader R and Bioconductor communities. Special thanks to Ian Lyttle and Alicia Schep for answering all of my annoying questions and letting me contribute to the **vegawidget** package. I thank Earo Wang, Ursula Laa, Shian Su, and Nick Tierney for always being available for chats (but really most importantly coffee or snacks) about software development and visualisation.

I would like to thank my colleagues Charity Law, Peter Hickey, and Saskia Freytag for their friendship and guidance on all things statistical during my thesis. I look forward to having a drink with you in person.

I would not have been able to make it through this PhD if not for the support
of friends and family. Thank you to the members of FCC, Casey, Clare, Cyrus, Joy and Margs, for always being a delight and making my sides hurt from laughter and fried chicken. Thank you Michael, Rachel, Robbie, Anna, LP, Laura, Saskia, Lan and Josh for making life more fun. A big thanks to my mum, dad and sister for always being supportive of my choices and listening to my complaints. Finally, thank you to my partner Joy for encouraging me to pursue a PhD and always believing in me, even when I don't believe in myself. I am a better person because of you and I am so happy I get to spend my life with you.

### 2020-08-03 Rmd/00-d-acknowlegements.Rmd
typos and additional thanks

@@
 
-I have had the great privilege to work with and be inspired by many people
-during my PhD candidature. It has been a challenging and rewarding experience
-and I would not have been able to get through it without the support of my
-advisers, colleagues, friends and family.
+I have had the great privilege to work with and be inspired by many people during my PhD candidature.
+It has been a challenging and rewarding experience and I would not have been able to get through it without the support of my advisers, colleagues, friends and family.
 
-I am deeply grateful to my supervisors: Dianne Cook, Matthew Ritchie, and Paul
-Harrison. Di has been instrumental in helping me grow as a researcher
-and teacher; she has been patient and always full of interesting insights and 
-enthusiasm for the practice of doing data analysis.
+I am deeply grateful to my supervisors: Dianne Cook, Matthew Ritchie, and Paul Harrison.
+Di has been instrumental in helping me grow as a researcher and teacher; she has been patient and always full of interesting insights and enthusiasm for the practice of doing data analysis.
 I am sorry that I have not named any of my software after Australian animals.
@@
 
-I would like to thank my colleagues Charity Law, Peter Hickey, and Saskia Freytag for their friendship and guidance on all things statistical during my thesis.
+I would like to thank my colleagues Charity Law, Peter Hickey, Alexandra Garnham, Hannah Coughlan, and Saskia Freytag for their friendship and guidance on all things statistical during my thesis.
 I look forward to having a drink with you in person.
 
-I would not have been able to make it through this PhD if not for the support
-of friends and family.
+I would not have been able to make it through this PhD if not for the support of friends and family.
 Thank you to the members of FCC, Casey, Clare, Cyrus, Joy and Margs, for always being a delight and making my sides hurt from laughter and fried chicken.

### 2020-08-04 Rmd/00-e-preface.Rmd
update text
don't number preface
addresses comments and closes #10
update preface and rebuild

@@
 # Preface {-}
+
+This thesis has been written using R Markdown with the **bookdown**
+package [@Xie2016-oa; @Xie2017-ke]. The **renv** package
+has been used to create a reproducible environment to build the thesis from source (available at https://github.com/sa-lee/thesis ) [@renv].
+An online version of the thesis is available to read at https://sa-lee.github.io/thesis.
 
@@
 \@ref(ch:intron) is based on my software contributions to the submitted
-article @Lee2019-mf. Chapter \@ref(ch:embedding) has not been submitted yet.
-
-## Open Research
-
-This thesis has been written using R Markdown with the **bookdown**
-package [@Xie2016-oa; @Xie2017-ke]. All materials required to build the
-thesis from source can be obtained from https://github.com/sa-lee/thesis. An
-online version of the thesis is available at https://sa-lee.github.io/thesis.
+article @Lee2019-mf. Chapter \@ref(ch:tsne) has not been submitted.
 

### 2020-08-04 _render.R

@@ -7,7 +7,7 @@ start_app(theme = simple_theme())
 cli_h1("Preprocessing")
 # spelling check
 cli_h2("Spelling check")
-rmd_files <- list.files(pattern = "*.Rmd", full.names = TRUE, recursive = TRUE)
+rmd_files <- list.files(path = "Rmd/", pattern = "*.Rmd", full.names = TRUE, recursive = TRUE)
 wordlist <- readLines("WORDLIST")
 spell_gb <- spelling::spell_check_files(rmd_files, wordlist, "en_GB")
 spell_us <- spelling::spell_check_files(rmd_files, wordlist, "en_US")

### 2020-08-04 Rmd/00-e-preface.Rmd
style

@@
 package [@Xie2016-oa; @Xie2017-ke]. The **renv** package
-has been used to create a reproducible environment to build the thesis from source (available at https://github.com/sa-lee/thesis ) [@renv].
+has been used to create a reproducible environment to build the thesis from source [@renv].
+All materials required to compile the thesis are available at https://github.com/sa-lee/thesis.
 An online version of the thesis is available to read at https://sa-lee.github.io/thesis.

### 2020-08-04 Rmd/01-intro.Rmd
style

@@
 integration of visualisation with model-based summaries of high-dimensional
-data sets (figure \@ref(fig:wflow-explore)). We focus on a common tool for EDA (especially applied to single-cell
-transcriptomics): non-linear dimension reduction (NLDR). We consider the incorporation of 
-interactive and dynamic graphics to assist analysts in using NLDR techniques
-for cluster orientation tasks. In particular, we advocate for the use of tours
-[@Cook1995-bi] alongside an NLDR visualisation to highlight potential pitfalls 
-and distortions obtained from an NLDR method. This approach acknowledges that
-there is no 'one' best visualisation or dimension reduction for a 
-high-dimensional dataset, and we often want to have an understanding of both the global and local structure within our data.
+data sets (figure \@ref(fig:wflow-explore)). We focus on a common tool for EDA 
+(especially applied to single-cell transcriptomics): non-linear dimension 
+reduction (NLDR). We consider the incorporation of interactive and dynamic 
+graphics to assist analysts in using  NLDR techniques for cluster orientation 
+tasks. In particular, we advocate  for the use of tours [@Cook1995-bi] alongside
+an NLDR visualisation to  highlight potential pitfalls  and distortions obtained
+from an NLDR method. This approach acknowledges that there is no 'one' best 
+visualisation or dimension reduction for a high-dimensional dataset, and we 
+often want to  have an understanding of both the global and local structure 
+within our data.
 
@@
 constructing these views and a user interaction framework for identifying 
-distortions. We present several case studies using data sets  both 
-single-cell transcriptomics and  machine learning  for using our approach to diagnose the quality   results obtained via popular NLDR methods
-like t-distributed stochastic neighbour embeddings (t-SNE) [@Maaten2008-sk].
+distortions. We present several case studies using data that capture aspects of
+single cell transcriptomics workflows, and use our approach to diagnose the
+quality of results obtained  via popular NLDR methods like t-distributed 
+stochastic neighbour embeddings (t-SNE) [@Maaten2008-sk].

### 2020-08-04 Rmd/02-plyranges.Rmd
style

@@
 genomic extension of **dplyr**. By analogy, **plyranges** is to the
-genomic algebra, as **dplyr**` is to the relational algebra.  The
-**plyranges`**Bioconductor package implements the language on top of a
+genomic algebra, as **dplyr** is to the relational algebra.  The
+**plyranges** Bioconductor package implements the language on top of a
 key subset of Bioconductor data structures and thus fully integrates
@@
 ranges are retained in the joined range.  All join operators in the
-`plyranges` DSL generate a set of hits based on overlap or proximity
+**plyranges** DSL generate a set of hits based on overlap or proximity
 of ranges and use those hits to merge the two datasets in different
@@
 the **rlang** language and how to adapt their code accordingly. Users
-familiar with the tidyverse should already have such knowledge.
+familiar with the **tidyverse** should already have such knowledge.
 

### 2020-08-04 Rmd/01-intro.Rmd
reference fixes

@@
 programming language under a single semantic known as *tidy data* 
-[@baser; @tidyverse; @Wickham2014-jc]. Simply put, a *tidy data* set is a 
+[@r-core; @tidyverse; @Wickham2014-jc]. Simply put, a *tidy data* set is a 
 rectangular table where each row of the table corresponds to an observation, 

### 2020-08-04 Rmd/03-fluentGenomics.Rmd
reference fixes

@@
 described by @Lee2019 and leverages underlying core Bioconductor infrastructure
-[@granges; @bioc] and the tidyverse design principles @tidyverse.
+[@granges; @bioc] and the **tidyverse** design principles @tidyverse.
 
@@
 The data used in this workflow is available from two packages: the **macrophage**
-Bioconductor ExperimentData package and from the workflow package 
+Bioconductor **ExperimentData** package and from the workflow package 
 **fluentGenomics**.
@@
 would typically not have the 'gz' ending if used from **Salmon** directly. One
-other thing to note is the use of the pipe operator,`%>%`, which can be read as
+other thing to note is the use of the pipe operator, `%>%`, which can be read as
 "then", i.e. first read the data, *then* select columns, *then* mutate them.
@@
 analysis, but here we will use a point estimate for the per-gene counts and
-perform gene-level analysis.
-
-The result is a _SummarizedExperiment_ object.
+perform gene-level analysis. The result is a _SummarizedExperiment_ object.
 
@@
 can also be used without an internet connection, in this case the linked
-transcriptome can be created directly from a *Salmon* index and gtf.
+transcriptome can be created directly from a **Salmon** index and gtf.
 
@@
 alternative would be to either operate on transcript abundance, and perform
-differential analysis on transcript (and so avoid defining the TSS of a set of
-isoforms), or to use gene-level summarized expression but to pick the most
-representative TSS based on isoform expression.
+differential analysis on transcript, and so avoid defining the transcription start
+site (TSS) of a set of isoforms, or to use gene-level summarized expression 
+but to pick the most representative TSS based on isoform expression.
 
@@
 The ATAC-seq counts have already been normalized with **cqn** [@Hansen2012] and
-$log_2$ transformed. Loading the **cqn**-normalized matrix of $log_2$ transformed read
+$\log_2$ transformed. Loading the **cqn**-normalized matrix of $\log_2$ transformed read
 counts takes ~30 seconds and loads an object of ~370 Mb. We set the column
@@
 **plyranges**, we construct a new column called _gene\_id_ from the row names of
-the results. Each row now contains the genomic region (_seqnames_, _start_,
+the results. Each row now contains the genomic region ( _seqnames_, _start_,
 _end_, _strand_) along with corresponding metadata columns (the _gene_id_ and
@@
 From this, we can restrict the results to those that meet our FDR threshold and
-select (and rename) the metadata columns we're interested in:
+select (and rename) the metadata columns we are interested in:
 
@@
 _GRanges_ object of differential peaks from the ATAC-seq data in @alasoo. The 
-code chunks for the remainder of this section are not run.
+code chunks for the remainder of this section are optional.
 
@@
 Now we would like to modify our gene ranges so they contain the 10 kilobases on
-either side of their transcription start site (TSS). There are many ways one
+either side of their TSS. There are many ways one
 could do this, but we prefer an approach via the anchoring methods in
@@
 the metadata columns from those overlapping DA peaks.  For any gene that has no
-overlaps, the DA peak columns will have `NA`'s.
+overlaps, the DA peak columns will have `NA` values.
 
@@
 object via merging neighboring genomic regions. The use of the directed suffix
-indicates we're maintaining strand information. In this case, we are simply
+indicates we are maintaining strand information. In this case, we are simply
 merging ranges (genes) via the groups we mentioned above. We also have to
@@
 
-For all variables except for the _origin_ column we divide the first row's
+For all variables except for the _origin_ column we divide the first rows
 values by the second row, which will be the enrichment of peaks in DE genes

### 2020-08-04 Rmd/04-intron.Rmd
reference fixes

@@
 
-Although there are many flexible and powerful software tools, like BEDtools or deepTools, for exploring and estimating coverage from common genomic data formats [@Quinlan2010-gc; @Ramirez2014-sh], it is advantageous to have tooling
+Although there are many flexible and powerful software tools, like **BEDtools** or **deepTools**, for exploring and estimating coverage from common genomic data formats [@Quinlan2010-gc; @Ramirez2014-sh], it is advantageous to have tooling
 that is tightly coupled to a statistical computing language such as R.
@@
 Coverage is estimated directly using the design matrix that contains a source column pointing to the locations of BAM files.
-The long-form representation is output as a **GRanges** object, and contains columns that were part of the design.
-Additional annotations are add with join functions, here we show the particular case of expanding the coverage **GRanges** to include exonic and intronic parts of a gene.
-This object can be further analysed using **plyranges** and our data descriptors approach, and then descriptors can be visualised as a scatter plot matrix with the `GGally` and `ggplot2` packages [@ggally; @Wickham2016-gz].
+The long-form representation is output as a _GRanges_ object, and contains columns that were part of the design.
+Additional annotations are added with join functions, here we show the particular case of expanding the coverage _GRanges_ to include exonic and intronic parts of a gene.
+This object can be further analysed using **plyranges** and our data descriptors approach, and then descriptors can be visualised as a scatter plot matrix with the **GGally** and **ggplot2** packages [@ggally; @Wickham2016-gz].
 Coverage traces can be directly generated with **superintronic** and collapsed over parts of the design matrix to identify differences between groups.
@@
 [@Ritchie2015-mj], like our proposed index method in @Lee2019-mf. The gene
-candidates obtain by our thresholds have been validated by the Heath lab
+candidates obtained by our thresholds have been validated by the Heath lab
 using qPCR.

### 2020-08-04 Rmd/00-d-acknowlegements.Rmd
update acknowledgements

@@
 Di has been instrumental in helping me grow as a researcher and teacher; she has been patient and always full of interesting insights and enthusiasm for the practice of doing data analysis.
-I am sorry that I have not named any of my software after Australian animals.
+Thank you Di for your support and guidance throughout this process.
+I look forward to keep working with you in the future.
 Matt is one of the kindest and most generous scientists I know, thank you for allowing me to be a part of your lab and for all of the lab lunches!
@@
 Thank you to the members of FCC, Casey, Clare, Cyrus, Joy and Margs, for always being a delight and making my sides hurt from laughter and fried chicken.
-Thank you Michael, Rachel, Robbie, Anna, LP, Laura, Saskia, Lan and Josh for making life more fun.
+Thank you Michael, Wai Keen, Rachel, Robbie, Anna, LP, Laura, Saskia, Kathy, Lan and Josh for making life more fun.
 A big thanks to my mum, dad and sister for always being supportive of my choices and listening to my complaints.

### 2020-08-07 Rmd/00-c-declaration.Rmd
spelling
typos and additional thanks
update declaration

@@
 2 & \textbf{plyranges}: a grammar of genomic data transformation & Published  & 70\%.
-Concept, software development, data analysis, and manuscript writing & (1) Dianne Cook, Concept and manuscript revision 10\% (2) Michael Lawrence, Concept and software development  20\% & \multicolumn{1}{c|}{N}  \\
+Concept, software development, data analysis, and manuscript writing & (1) Dianne Cook, Concept and manuscript revision 10\% (2) Michael Lawrence, Concept and software development  20\% & \multicolumn{1}{c|}{(1) No (2) No}  \\
 \hline
 3 & Fluent genomics with \textbf{plyranges} and \textbf{tximeta} & Published & 60\%.
-Concept, software development, data analysis and manuscript writing  & (1) Michael Lawrence, manuscript feedback and editing 5\% (2) Michael I Love, Concept, data analysis and manuscript writing 35\% & \multicolumn{1}{c|}{N} \\
-\hline
-4 & Expoloratory coverage analysis with \textbf{superintronic} and \textbf{plyranges} & Published & 50\%.
-Concept, software development, data analysis and manuscript writing  & (1) Charity Law, Concept, manuscript feedback and editing 45\% (2) Matthew Ritchie, manuscript writing and editing 5\% & \multicolumn{1}{c|}{N} \\
-\hline
-5 & Casting multiple shadows: high-dimensional interactive data visualisation with tours and embeddings & To be submitted & 85\%.
-Concept, software development, and manuscript writing & (1) Dianne Cook, Concept and manuscript writing 15\% & \multicolumn{1}{c|}{N} \\
+Concept, software development, data analysis and manuscript writing  & (1) Michael Lawrence, manuscript feedback and editing 5\% (2) Michael I Love, Concept, data analysis and manuscript writing 35\% & \multicolumn{1}{c|}{(1) No (2) No} \\
 \hline
@@
 \FloatBarrier
+
+I have renumbered sections of submitted or published papers in order to generate a consistent presentation within the thesis.
 
@@
 
+\textbf{Date:} 05/08/2020
+
+I hereby certify that the above declaration correctly reflects the nature and extent of the student’s and co-authors’ contributions to this work.
+In instances where I am not the responsible author I have consulted with the responsible author to agree on the respective contributions of the authors.
+
+\textbf{Supervisor name:} Dianne Helen Cook
+
+\textbf{Supervisor signature:}
+
+```{r supervisor, out.width = "20%", fig.align = "left", eval = file.exists("img/di.png")}
+knitr::include_graphics("img/di.png")
+```
+
+\textbf{Date:} 05/08/2020
+
 `r if (knitr::is_html_output()) '-->'`

### 2020-09-18 data-raw/hub.R
refactor local caches into data/

@@ -1,0 +1,4 @@
+# download hub resources locally
+library(AnnotationHub)
+AnnotationHub(cache = here::here("data"))[["AH73905"]]
+AnnotationHub(cache = here::here("data"))[["AH33458"]]

### 2020-09-18 data-raw/plyranges-ex-bam.R

@@ -1,8 +1,6 @@
 # download and prepare BAM file example
-
 library(BiocFileCache)
-bfc <- BiocFileCache(cache = here::here("data"))
-
+bfc <- BiocFileCache(cache = here::here("data"), ask = FALSE)
 # ftp link on GEO
 url <- "ftp://ftp.ncbi.nlm.nih.gov/geo/samples/GSM433nnn/GSM433167/suppl/GSM433167%5FBI%2EH3K27me3%2Ebam"
 fpath <- "GSM433167_BI.H3K27me3.bam"

### 2020-09-18 data-raw/plyranges-ex-snparray.R

@@ -3,7 +3,7 @@ library(readr)
 library(BiocFileCache)
 library(plyranges)
 
-bfc <- BiocFileCache(cache = here::here("data"))
+bfc <- BiocFileCache(cache = here::here("data"), ask = FALSE)
 
 rname <- "h1_snp_array"
 res <- bfcquery(bfc, rname)

### 2020-09-18 scripts/plyranges.R

@@ -5,12 +5,12 @@ library(Biostrings)
 library(plyranges)
 library(AnnotationHub)
 # bam and snp array resources are stored in BiocFileCache
-bfc <- BiocFileCache(cache = here::here("data"))
+bfc <- BiocFileCache(cache = here::here("data"), ask = FALSE)
 h1_bam_sorted <- bfcrpath(bfc, "GSM433167_BI.H3K27me3.sorted.bam")
 h1_bam_sorted_index <- bfcrpath(bfc, "GSM433167_BI.H3K27me3.sorted.bam.bai")
 h1_snp_array_path <- bfcrpath(bfc, "h1_snp_array") # this is an rds file
 # AnnotationHub resource for T-cell BigWig file
-bw_file <- AnnotationHub()[["AH33458"]]
+bw_file <- AnnotationHub(cache = here::here("data"))[["AH33458"]]
 
 ## ----load-bw---------------------------------------
 library(plyranges)

### 2020-09-18 Rmd/05-embeddings.Rmd
fix screenshots layout, closes #6
fix typos

@@
 
-So when the the loss function is minimised this corresponds to large attractive
+So when the loss function is minimised this corresponds to large attractive
 forces, that is, the pair-wise distances in $Y$ are small when there are
@@
 have followed a standard workflow for pre-processing and normalizing this data 
-(described by @Amezquita2020-at): we performed QC using the the **scater**
+(described by @Amezquita2020-at): we performed QC using the **scater** package
 by removing cells with high proportion of mitochondrial gene expression and low
@@
 
-```{r mouse-01, fig.cap = "(ref:mouse-01)", out.height = "75%"}
+```{r mouse-01, fig.cap = "(ref:mouse-01)", out.height = "85%"}
 display_screenshot("mouse-01", vid_lookup)
@@
 The tour appears to form a tetrahedron like shape, with points lying on
-surface and along the vertices of the the tetrahedron in 5-$d$ PCA space -
+surface and along the vertices of the tetrahedron in 5-$d$ PCA space -
 a phenomena that has also been observed in @Korem2015-af 

### 2020-10-21 Rmd/05-embeddings.Rmd
typo corrections

@@
 Similarly, the **TensorFlow**  embedding projector is a web interface to 
-running some non-liner embedding methods live in the browser and 
+running some non-linear embedding methods live in the browser and 
 provides interactions to colour points, and select nearest neighbours
@@
 dataset is a difficult and a somewhat poorly defined problem. For non-linear
-methods, there are a lot of parameters to explore that can have an effect on
+methods, there are many parameters to explore that can have an effect on
 the resulting visualisation and interpretation. Interfaces for evaluating
@@
 matrix where $d \ll p$. The hope of the analyst is that the DR procedure to 
-produces $Y$ will remove noise in the original dataset while retaining any 
+produce $Y$ will remove noise in the original dataset while retaining any 
 latent structure.
@@
 $$ \text{perplexity}_i = \exp(-\log(2) \sum_{i \ne j}p_{j|i}\log_2(p_{j|i})) $$
-As the perplexity increases, $\sigma^2_{i}$ increases, until its bounded above
+As the perplexity increases, $\sigma^2_{i}$ increases, until it is bounded above
 by the number of observations , $n-1$, in the data, corresponding to
@@
 
-Like, when, using other DR techniques, the tour assumes we have a real data matrix
+Like when using other DR techniques, the tour assumes we have a real data matrix
 $X$ consisting of $n$ observations in $p$ dimensions. First, the tour generates 
@@
 Diffusions occur when points are far from each other in the embedding view are
-near in the original data.  Whether, points are near, or far is reliant on the
+near in the original data.  Whether points are near or far is reliant on the
 distance metric used; distortions and diffusions can be thought of as the
@@
 
-From the video linked inn figure \@ref(fig:gaussian), we learn that t-SNE has 
+From the video linked in figure \@ref(fig:gaussian), we learn that t-SNE has 
 correctly split out each cluster and laid them out in a star like formation. 
@@
 
-(ref:liminal-hierarchical) Screenshots of the **liminal** interface applied to sub-clustered data, a video of the tour animation is available at https://player.vimeo.com/video/43963590.
+(ref:liminal-hierarchical) Screenshots of the **liminal** interface applied to sub-clustered data, a video of the tour animation is available at https://player.vimeo.com/video/439635905.
 
@@
 Next we explore some simulated noisy tree structured data (figure
-\@ref(fig:fake-trees)), our interest here is how t-SNE visualisations break
+\@ref(fig:fake-trees)). Our interest here is how t-SNE visualisations break
 topology of the data, and then seeing if we can resolve this by tweaking the

### 2020-10-21 Rmd/01-intro.Rmd
typo corrections

@@
 statistical workflow - it is an analyst’s first pass at 
-understanding their data; revealing all it’s messes and 
+understanding their data; revealing all its messes and 
 uncovering hidden insights [@Tukey1977-pn; @Grolemund2017-uc]. 

### 2020-10-21 Rmd/02-plyranges.Rmd
typo corrections

@@
 
-Using **plyranges** and the the Bioconductor package **AnnotationHub** [@R-ahub]
+Using **plyranges** and the Bioconductor package **AnnotationHub** [@R-ahub]
 we can download and read BigWig files from ChIP-Seq experiments from 

### 2020-10-22 Rmd/05-embeddings.Rmd
correct typos

@@
 projections as quantified by an index function [@Cook1995-bi]. While our software, 
-**liminal** is able to  visualise guided tours, our focus in the case studies 
+**liminal**, is able to  visualise guided tours, our focus in the case studies 
 uses the grand tour to see global structure in the data.

### 2020-10-26 response/response.Rmd
start tracking response

---
title: Thesis Amendments
author: Stuart Lee
output:
  MonashEBSTemplates::memo:
    includes:
      in_header: preamble2.tex
fontfamily: mathpazo
---

First, I would like to thank my examiners for their constructive feedback and insights. I am excited to incorporate their ideas into my future research. The point by point description of changes are below: the examiners' comments are in red and my response is in black.

## Professor Levi Waldron {-}

Thank you for your comments and identification of typographic errors.

> p. 1 Introduction: it’s -> its

Amended.

> p. 16 “Using plyranges and the the”

Amended.

> p. 34: use of `dir` for a directory path string when `dir()` is a function returning contents of a directory path could be confusing. However, this chapter is already peer-reviewed and published so I would not change it.

Noted. I can change this in the workflow for next Bioconductor release.

> P. 43-44: Why is the genome (“hg38”) set using an endomorphic function, but style (“UCSC”) set using a replacement function (`seqlevelsStyle`)?

At the time of writing, `plyranges` does not have any genome restyling functions, hence why the replacement function was used.

> p. 77: Tensorflow “non-liner embedding” p. 77-78” “a lot of” -> many

Amended.

> p. 80: “its bounded above” ->it’s

Amended.

> p. 82: “Like, when, using” -> Like when using “While our software, liminal is able “ -> While our software, liminal, is able

Amended.

>Section 5.3.1 “Finding Gestalt: focus and context” - this is probably a clever metaphor, but I didn’t understand it.

This was a reference to Buja, Cook, and Swayne (1996) paper mentioned in the paragraph above. Here "Gestalt" refers to Gestalt theory in perception and psychology, it refers to identifying patterns in visual forms.

> p. 84: “Whether, points are near, or far” -> remove both commas

Amended.

>Figure 5.2 video https://player.vimeo.com/video/43963590 does not exist. The URL given in Table 5.1 is different and correct.

Amended.

> p. 98: “We have shown in the case studies, that” -> no comma

Amended.

## Assistant Professor Stephanie C. Hicks {-}

Thank you for your thorough review. I am looking forward to incorporating your suggestions into future work.

### General Comments

### 2020-10-27 Rmd/01-intro.Rmd
modify intro

@@
 
-(ref:wflow-integrate) Data from multiple assays can be combined together using modelling and transformations.
-Data and model results can be combined, tidied and transformed to have a common representation across different granularities of the genome.
+(ref:wflow-integrate) Data from multiple assays can be combined together using modelling and transformations to gain biological insight.
+This is achieved either via a statistical technique or more simply via joining data and model results so they have a common representation across different granularities of the genome.
 
@@
 Chapter \@ref(ch:plyranges) shows that the *tidy data* semantic is applicable
-to range-based genomics data and develops a *fluent* interface to transforming
+to in memory data measured along the genome and develops a *fluent* interface to transforming
 it called **plyranges**. The software provides a framework to an assist
-an analyst to compose queries on genomics datasets. This chapter has been 
-published as @Lee2019.
+an analyst to compose queries on genomics datasets.
+Our software is agnostic to how counts from bulk assays have been obtained.
+Indeed, we have used data obtained from both alignment and quantification based approaches throughout the thesis to perform useful analyses.
+This chapter has been  published as @Lee2019.
 
@@
 many approaches to integrating data sets from multiple assays using multivariate
-statistical techniques [@Meng2016-pz], in chapter \@ref(ch:fluentGenomics)
+statistical techniques [@Meng2016-pz; @Stein-OBrien2018-bk] and data structures to represent them [@Ramos2017-kn], there has been little thought given to the interoperability between these approaches and the **tidyverse**.
+In chapter \@ref(ch:fluentGenomics)
 we describe a simple end-to-end workflow for integrating 
-results along the genome using **plyranges**. This workflow also shows that
+results along the genome using **plyranges**. This workflow shows that
 our grammar based approach does not impair interoperability between

### 2020-10-27 Rmd/07-appendix-A.Rmd
add appendix

# (APPENDIX) Appendix {-}

# **plyranges** vignette

## Getting started with the **plyranges** package

### `Ranges` revisited

In Bioconductor there are two classes, `IRanges` and `GRanges`, that are
standard data structures for representing genomics data.  Throughout this
document I refer to either of these classes as `Ranges` if an operation can be
performed on either class, otherwise I explicitly mention if a function is
appropriate for an `IRanges` or `GRanges`.

`Ranges` objects can either represent sets of integers as `IRanges` (which have
start, end and width attributes) or represent genomic intervals (which have
additional attributes, sequence name, and strand) as `GRanges`. In addition,
both types of `Ranges` can store information about their intervals as metadata
columns (for example GC content over a genomic interval).

`Ranges` objects follow the tidy data principle: each row of a `Ranges` object
corresponds to an interval, while each column will represent a variable about
that interval, and generally each object will represent a single unit of
observation (like gene annotations).

Consequently, `Ranges` objects provide a powerful representation for reasoning
about genomic data. In this vignette, you will learn more about `Ranges`
objects and how via grouping, restriction and aggregation you can perform
common data tasks.

### Constructing `Ranges`

To construct an `IRanges` we require that there are at least two columns that
represent at either a starting coordinate, finishing coordinate or the width of
the interval.

```{r}
suppressPackageStartupMessages(library(plyranges))
set.seed(100)
df <- data.frame(start=c(2:-1, 13:15),
                 width=c(0:3, 2:0))

# produces IRanges
rng <- df %>% as_iranges()
rng

```

To construct a `GRanges` we require a column that represents that sequence name
( contig or chromosome id), and an optional column to represent the
strandedness of an interval.

```{r}
# seqname is required for GRanges, metadata is automatically kept
grng <- df %>%
  transform(seqnames = sample(c("chr1", "chr2"), 7, replace = TRUE),
         strand = sample(c("+", "-"), 7, replace = TRUE),
         gc = runif(7)) %>%
  as_granges()

grng
```

### Arithmetic on Ranges

Sometimes you want to modify a genomic interval by altering the width of the
interval while leaving the start, end or midpoint of the coordinates unaltered.
This is achieved with the `mutate` verb along with `anchor_*` adverbs.

The act of anchoring fixes either the start, end, center coordinates of the
`Range` object, as shown in the figure and code below and anchors are used in
combination with either `mutate` or `stretch`.

```{r anchor_fig, echo = FALSE, out.width="400px", fig.align="center"}
knitr::include_graphics(here::here("img", "anchors.png"), dpi = 150)
```

```{r}
rng <- as_iranges(data.frame(start=c(1, 2, 3), end=c(5, 2, 8)))
grng <- as_granges(data.frame(start=c(1, 2, 3), end=c(5, 2, 8),
                          seqnames = "seq1",
                          strand = c("+", "*", "-")))
mutate(rng, width = 10)
mutate(anchor_start(rng), width = 10)
mutate(anchor_end(rng), width = 10)
mutate(anchor_center(rng), width = 10)
mutate(anchor_3p(grng), width = 10) # leave negative strand fixed
mutate(anchor_5p(grng), width = 10) # leave positive strand fixed
```

Similarly, you can modify the width of an interval using the `stretch` verb.
Without anchoring, this function will extend the interval in either direction
by an integer amount. With anchoring, either the start, end or midpoint are
preserved.

```{r}
rng2 <- stretch(anchor_center(rng), 10)
rng2
stretch(anchor_end(rng2), 10)
stretch(anchor_start(rng2), 10)
stretch(anchor_3p(grng), 10)
stretch(anchor_5p(grng), 10)
```

`Ranges` can be shifted left or right. If strand information is available we
can also shift upstream or downstream.

```{r}
shift_left(rng, 10)
shift_right(rng, 10)
shift_upstream(grng, 10)
shift_downstream(grng, 10)
```

### Grouping `Ranges`

**plyranges** introduces a new class of `Ranges` called `RangesGrouped`, this is
a similar idea to the grouped `data.frame\tibble` in **dplyr**.

Grouping can act on either the core components or the metadata columns of a
`Ranges` object.

It is most effective when combined with other verbs such as `mutate()`,
`summarise()`, `filter()`, `reduce_ranges()` or `disjoin_ranges()`.
```{r}
grng <- data.frame(seqnames = sample(c("chr1", "chr2"), 7, replace = TRUE),
         strand = sample(c("+", "-"), 7, replace = TRUE),
         gc = runif(7),
         start = 1:7,
         width = 10) %>%
  as_granges()

grng_by_strand <- grng %>%
  group_by(strand)

grng_by_strand
```

### Restricting `Ranges`

The verb `filter` can be used to restrict rows in the `Ranges`. Note that
grouping will cause the `filter` to act within each group of the data.
```{r}
grng %>% filter(gc < 0.3)
# filtering by group
grng_by_strand %>% filter(gc == max(gc))
```

We also provide the convenience methods `filter_by_overlaps` and
`filter_by_non_overlaps` for restricting by any overlapping `Ranges`.

```{r}
ir0 <- data.frame(start = c(5,10, 15,20), width = 5) %>%
  as_iranges()
ir1 <- data.frame(start = 2:6, width = 3:7) %>%
  as_iranges()
ir0
ir1
ir0 %>% filter_by_overlaps(ir1)
ir0 %>% filter_by_non_overlaps(ir1)
```

### Summarising `Ranges`

The `summarise` function will return a `DataFrame` because the information
required to return a `Ranges` object is lost. It is often most useful to use
`summarise()` in combination with the `group_by()` family of functions.

```{r}
ir1 <- ir1 %>%
  mutate(gc = runif(length(.)))

ir0 %>%
  group_by_overlaps(ir1) %>%
  summarise(gc = mean(gc))
```

### Joins, or another way at looking at overlaps between `Ranges`

A join acts on two GRanges objects, a query and a subject.

```{r}
query <- data.frame(seqnames = "chr1",
               strand = c("+", "-"),
               start = c(1, 9),
               end =  c(7, 10),
               key.a = letters[1:2]) %>%
  as_granges()

subject <- data.frame(seqnames = "chr1",
               strand = c("-", "+"),
               start = c(2, 6),
               end = c(4, 8),
               key.b = LETTERS[1:2]) %>%
  as_granges()
```

```{r olap, echo = FALSE, fig.cap = "Query and Subject Ranges", message = FALSE}
library(ggplot2)
query_df <- as.data.frame(query)[, -6]
query_df$key <- "Query"
subject_df <- as.data.frame(subject)[, -6]
subject_df$key <- "Subject"
melted_ranges <- rbind(query_df, subject_df)
ggplot(melted_ranges, aes(xmin = start, xmax = end, ymin = 1, ymax = 3)) +
  geom_rect() +
  facet_grid(key ~ .) +
  scale_x_continuous(breaks = seq(1, 10, by = 1)) +
  xlab("Position") +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank())
```

The join operator is relational in the sense that metadata from the query and
subject ranges is retained in the joined range.  All join operators in the
**plyranges** DSL generate a set of hits based on overlap or proximity of ranges
and use those hits to merge the two datasets in different ways.  There are four
supported matching algorithms: _overlap_, _nearest_, _precede_, and _follow_.
We can further restrict the matching by whether the query is completely
_within_ the subject, and adding the _directed_ suffix ensures that matching
ranges have the same direction (strand).

```{r olaps-diagram, echo = FALSE, out.width="600px"}
knitr::include_graphics(here::here("img", "olap-fig.png"))
```

The first function, `join_overlap_intersect()` will return a `Ranges` object
where the start, end, and width coordinates correspond to the amount of any
overlap between the left and right input `Ranges`. It also returns any
metadata in the subject range if the subject overlaps the query.

```{r}
intersect_rng <- join_overlap_intersect(query, subject)
intersect_rng
```

```{r intersect-join, echo = FALSE, fig.cap="Intersect Join"}
intersect_df <- as.data.frame(intersect_rng)[, -c(6,7)]
intersect_df$key <- "Intersect Join"
melted_ranges <- rbind(query_df, subject_df, intersect_df)
melted_ranges$key <- factor(melted_ranges$key,
                              levels = c("Query", "Subject", "Intersect Join"))
ggplot(melted_ranges, aes(xmin = start, xmax = end, ymin = 1, ymax = 3)) +
  geom_rect() +
  facet_grid(key ~ .) +
  scale_x_continuous(breaks = seq(1, 10, by = 1)) +
  xlab("Position") +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank())
```

The `join_overlap_inner()` function will return the `Ranges` in the query that
overlap any `Ranges` in the subject. Like the `join_overlap_intersect()`
function metadata of the subject `Range` is returned if it overlaps the query.

```{r}
inner_rng <- join_overlap_inner(query, subject)
inner_rng
```

```{r inner-join, echo = FALSE, fig.cap = "Inner Join", message = FALSE}
inner_df <- as.data.frame(inner_rng)[, -c(6,7)]
inner_df$ymin <- c(1,4)
inner_df$ymax <- c(3,6)
inner_df$key <- "Inner Join"
melted_ranges <- rbind(query_df, subject_df)
melted_ranges$ymin <- 1
melted_ranges$ymax <- 3
melted_ranges <- rbind(melted_ranges, inner_df)
melted_ranges$key <- factor(melted_ranges$key,
                              levels = c("Query", "Subject", "Inner Join"))

ggplot(melted_ranges, aes(xmin = start, xmax = end, ymin = ymin, ymax = ymax)) +
  geom_rect() +
  facet_grid(key ~ ., scales = "free_y") +
  scale_x_continuous(breaks = seq(1, 10, by = 1)) +
  xlab("Position") +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank())
```

We also provide a convenience method called `find_overlaps` that computes the
same result as `join_overlap_inner()`.

```{r}
find_overlaps(query, subject)
```

The `join_overlap_left()` method will perform an outer left join.

First any overlaps that are found will be returned similar to
`join_overlap_inner()`. Then any non-overlapping ranges will be returned, with
missing values on the metadata columns.

```{r}
left_rng <- join_overlap_left(query, subject)
left_rng
```

```{r olap-left, echo = FALSE, fig.cap = "Left Join", message = FALSE}
left_df <- as.data.frame(left_rng)[, -c(6,7)]
left_df$ymin <- c(1,4, 1)
left_df$ymax <- c(3,6, 3)
left_df$key <- "Left Join"
melted_ranges <- rbind(query_df, subject_df)
melted_ranges$ymin <- 1
melted_ranges$ymax <- 3
melted_ranges <- rbind(melted_ranges, left_df)
melted_ranges$key <- factor(melted_ranges$key,
                              levels = c("Query", "Subject", "Left Join"))

ggplot(melted_ranges,
       aes(xmin = start, xmax = end, ymin = ymin, ymax = ymax)) +
  geom_rect() +
  facet_grid(key ~ ., scales = "free_y") +
  scale_x_continuous(breaks = seq(1, 10, by = 1)) +
  xlab("Position") +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank())
```

Compared with `filter_by_overlaps()` above, the overlap left join expands the
`Ranges` to give information about each interval on the query `Ranges` that
overlap those on the subject `Ranges` as well as the intervals on the left that
do not overlap any range on the right.

#### Finding your neighbours
We also provide methods for finding nearest, preceding or following `Ranges`.
Conceptually this is identical to our approach for finding overlaps, except the
semantics of the join are different.

```{r neigbours, echo = FALSE, out.width="600px"}
knitr::include_graphics(here::here("img", "neighbours.png"))
```

```{r}
join_nearest(ir0, ir1)
join_follow(ir0, ir1)
join_precede(ir0, ir1) # nothing precedes returns empty `Ranges`
join_precede(ir1, ir0)
```

### Example: dealing with multi-mapping
This example is taken from the Bioconductor support
[site](https://support.bioconductor.org/p/100046/).

We have two `Ranges` objects. The first contains single nucleotide positions
corresponding to an intensity measurement such as a ChiP-seq experiment, while
the other contains coordinates for two genes of interest.

We want to identify which positions in the `intensities` `Ranges` overlap the
genes, where each row corresponds to a position that overlaps a single gene.

First we create the two `Ranges` objects
```{r ex1}
intensities <- data.frame(seqnames = "VI",
                          start = c(3320:3321,3330:3331,3341:3342),
                          width = 1) %>%
  as_granges()

intensities

genes <- data.frame(seqnames = "VI",
                    start = c(3322, 3030),
                    end = c(3846, 3338),
                    gene_id=c("YFL064C", "YFL065C")) %>%
  as_granges()

genes
```

Now to find where the positions overlap each gene, we can perform an overlap
join. This will automatically carry over the gene_id information as well as
their coordinates (we can drop those by only selecting the gene_id).
```{r}
olap <- join_overlap_inner(intensities, genes) %>%
  select(gene_id)
olap
```

Several positions match to both genes. We can count them using `summarise` and
grouping by the `start` position:

```{r}
olap %>%
  group_by(start) %>%
  summarise(n = n())
```

### Grouping by overlaps

It's also possible to group by overlaps. Using this approach we can count the
number of overlaps that are greater than 0.

```{r}
grp_by_olap <- ir0 %>%
  group_by_overlaps(ir1)
grp_by_olap
grp_by_olap %>%
  mutate(n_overlaps = n())
```

Of course we can also add overlap counts via the `count_overlaps()` function.

```{r}
ir0 %>%
  mutate(n_overlaps = count_overlaps(., ir1))
```

### Data Import/Output

We provide convenience functions via **rtracklayer** and **GenomicAlignments** for
reading/writing the following data formats to/from `Ranges` objects.

| **plyranges** functions | File Format |
|-----------------------|-------------|
| `read_bam()` | BAM | 
| `read_bed()`/`write_bed()` | BED | 
| `read_bed_graph()`/ `write_bed_graph()` | BEDGraph | 
| `read_narrowpeaks()`/`write_narrowpeaks()` | narrowPeaks | 
| `read_gff()` / `write_gff()` | GFF(1-3)/ GTF | 
| `read_bigwig()` / `write_bigwig()` | BigWig | 
| `read_wig()` /`write_wig()` | Wig |

# Learning more

There are many other resources and workshops available to learn to use
**plyranges** and related Bioconductor packages, especially for more realistic
analyses than the ones covered here:

- The [fluentGenomics](https://bioconductor.org/packages/release/workflows/html/fluentGenomics.html) workflow package is an end-to-end workflow package for integrating differential
expression results with differential accessibility results.
- The [Bioc 2018 Workshop book](https://bioconductor.github.io/BiocWorkshops/fluent-genomic-data-analysis-with-plyranges.html) has worked examples of using **plyranges** to analyse publicly available genomics data.
- The [extended vignette in the plyrangesWorkshops package](https://github.com/sa-lee/plyrangesWorkshops) has a detailed
walk through of using plyranges for coverage analysis.
- The [case study](https://github.com/mikelove/plyrangesTximetaCaseStudy) by Michael Love using plyranges with [tximeta](https://bioconductor.org/packages/release/bioc/html/tximeta.html) to follow
up on interesting hits from a combined RNA-seq and ATAC-seq analysis.

### 2020-10-27 Rmd/06-conclusion.Rmd
update link in appendix

@@
 The **plyranges** package is available to download from 
-https://bioconductor.org/packages/release/bioc/html/plyranges.html and
+https://bioconductor.org/packages/plyranges and
 the **fluentGenomics** workflow is available to download from 

### 2020-10-27 response/response.Rmd
starting to work through comments

@@
 
-### General Comments
+### Addressing General Comments
+
+#### On Scalability
+
+I would like to comment that I have avoided performance benchmarking for two main reasons.
+First, I believe it is out of scope for the thesis, since my focus has been on the development of programming and user interfaces.
+Second, my implementation of **plyranges** is deeply entwined with the underlying core Bioconductor infrastructure, especially the data structures and methods found in the **IRanges** and **GenomicRanges** packages.
+This would result any benchmarking of **plyranges** turning into benchmarking of those packages which has been done elsewhere.
+
+Regarding out of memory data structures like `DelayedArray`, it is possible to incorporate them as columns to a `GRanges` object and then the full **plyranges** API would be available.
+Currently, there appear to be no plans in Bioconductor for out of memory `Ranges`, and indeed there are times where it is more appropriate to switch to a rectangular data layout for computation, as is reflected in the workflow described in chapter 3 of the thesis.
+I believe the development of a scalable out of memory interface for genomic data analysis will require the transformation between data measured along the genome (covered by the **plyranges** API) and the rectangular layouts that enable out of disk computing like `SummarizedExperiment`.
+This is an exciting research area that I would like to explore in the currently under development **plyexperiment**.
+
+#### On alignment vs. pseudoalignment
+
+I have amended the introduction to include a note about alignment. 
+I believe it is out of scope to go into the gritty details of when to use alignment or pseudo-alignment methods to obtain counts for RNA-seq data.
+I have included results from both approaches to highlight that **plyranges** is agnostic to the method used and is flexible for many types of data.
+
+### Chapter 1 comments
+
+> As you introduced a “biological data science workflow” in this chapter and did a great job of walking the read through the wrangling, combining and visualizing components of the workflow, I was hoping you would touch on tools for communication too.
+I understand this is outside the scope of the thesis, but I still think it would be nice to at least mention work others have done in this area.
+
+I agree that it would be nice but it is out of scope for the thesis. Communication is extremely broad and I feel that including it here would distract from the focus of my thesis.
+
+> Pg 1: I thought it was strange that the words “figure” in e.g. “figure 1.1” were not capitalized. Would recommend changing to “Figure 1.1” throughout.
+> Pg 1: Recommend capitalizing the word “chapter” when referring to specific chapters e.g. “chapter 2”.
+
+TODO
+
+> Pg 2: Figure 1.1 has been modified to include “combine”, but the caption does not reflect what is meant by that.
+Would recommend modifying caption to reflect importance of combining experimental assays together to gain biological insight.
+
+Amended.
+
+> Pg 3: you introduce “range-based genomics data”, but it is not clear what is meant by that. More explanation would be helpful.
+
+I have clarified what this refers to.
+
+>Pg 4: In Section 1.2, could you add a few more references on existing methods for integration of genomic data and then re-emphasize the point that however existing methods have not thought much about the interoperability between the tidyverse and Bioconductor approaches?
+
+I have expanded this section to reflect these comments.
+
+### Chapter 2 comments
+
+Most of the comments pertain to the use and learning of the **plyranges** package.
+I believe they are best addressed through the **plyranges** documentation and vignette.
+I have amended the thesis to include the package vignette as an appendix.
+
+### Chapter 6 comments
+
+> Pg 102: Instead of https://bioconductor.org/packages/release/bioc/html/plyranges.html, you can always use the short version https://bioconductor.org/packages/plyranges which points to the release version.
+
+Amended.

### 2020-10-27 Rmd/05-embeddings.Rmd
add PCA reference

@@
 analysis (PCA) which performs an eigendecomposition of the estimated sample
-covariance matrix. The eigenvalues are sorted in decreasing order and represent
+covariance matrix [@Hotelling1933-ck]. The eigenvalues are sorted in decreasing order and represent
 the variance explained by each component (eigenvector). 

### 2020-10-27 response/response.Rmd
more details for chapter 5

@@
 
+### Chapter 5 comments
+
+>  Pg 77: “non-liner” should be “non-linear”
+
+Amended.
+
+>  Pg 78: “produces” should be “produce”
+
+Amended.
+
+> Pg 78: Could you add a reference for PCA?
+
+Amended.
+> Pg 80 (also in 81): The way “tSNE” was written in “this essentially turns t-SNE” looks very different
+than previously written stylistically. Could this be converted to the way you were originally referring to
+it?
+
+Amended.
+
+> Pg 87: “inn” should be “in”
+
+Amended.
+
+> Pg 88: Inside Figure 5.1A and B, could you have liminal label which plot is which i.e. t-SNE vs tour
+view. Also, could axis labels be added? Maybe also add this information to the Figure legend? – Pg 89 – same comment
+
+I think this is not obvious in the static figure but is obvious in the video and when using the **liminal** package that the embedding view is always on the left, and the tour view is on the right (it will be the view that is animated).
+I think the suggestion of titling the views is a good one and will place that into the software in future.
+Regarding axis labels, I don't think they make sense for the tour since it's really a linear combination of multiple variables (there's an option to switch on a biplot view for the tour).
+I am also reluctant to include them for t-SNE because they can be misleading for the reasons mentioned in the chapter, however I do think this would be good to have as an option for **liminal** in future.
+
+> Pg 90: I believe there should be a period after “tree structure data (figure 5.3)” instead of a comma
+
+Amended.
+
+> Pg 94: Can you comment on the scRNA-seq example where you only pick first 5 PCs that explained only 20% of the variance.
+Or more generally, what is the trade-off on the liminal view by picking more or less PCs along with a NLDR method.
+Do you have recommendations for the user on how many
+PCs to pick for and optimal liminal view?
+
+I have amended section 5.5.3 to provide more details about why the first five components were selected, and how to choose variables to tour.
+
+> Pg 94: Can you comment on the limits of the liminal view in terms of number of observations that
+one can visualize in a given tour? Have you thought about strategies for making this scale with more observations?
+– Ah, I see you touched on this at the end of the discussion of the capture. I think this is a really important point that would be greatly appreciated going forward.
+– And it was talked about in Chapter 6 in Future Work.
+Incorporating this more in the middle of the chapter with tables specifying what is or is not reasonable to plot in terms of the number of observations, would be helpful.
+
+I have amended section 5.5.3 to discuss a strategy in the specific case of cluster analysis.
+I have also amended the discussion of the chapter with specific work that can mitigate the need to show all of the data.
+In the case of tasks performed for clustering I don't think it is useful or necessary to show all points in the scatter plot when we are most interested in density and shapes.
+There is interesting work to be done here in future for appropriate sampling strategies or aggregations that could be done prior to producing a visualisation.
+
 ### Chapter 6 comments

### 2020-10-27 Rmd/05-embeddings.Rmd
more details for chapter 5

@@
 alongside the t-SNE view which was computed from all 25 PCs.
-Due to latency of the interface we do a weighted sample of  the rows based on
+We have selected only the first five PCs because there is a large drop in the percentage of variance explained after the fifth component, with each component after contributing less than one percent of variance.
+Consequently, increasing the number of PCs to tour would increase the dimensionality and volume of the subspace we are touring but without adding any additional signal to the view.
+
+Due to latency of the **liminal** interface we do a weighted sample of  the rows based on
 cluster membership, leaving us with approximately 10 per cent of the original
@@
 get a sense of the shape of the clusters as seen from the linked video in figure \@ref(fig:mouse-01).
+If one was interested in performing more in-depth cluster analysis we recommend an iterative approach of removing large clusters and then re-running the **liminal** view as a way finding more granular cluster structure.
 
@@
 on the display canvas, while for the tour, points are concentrated in the
-centre of the projection and become difficult to see. 
-Second, as $n$ increases the animation becomes slower as there is more data
-passing from the server to the client. Both of these issues could be mitigated
-through alternate tour paths and displays. The recent work
+centre of the projection and become difficult to see. We have recently proposed a simple non-linear transformation for the tour called a sage tour that aims to fix this problem [@Laa2020-vk]. 
+Second, as $n$ increases both the embedding view and tour view become harder to read due to over-plotting, while the interactivity and animation become slower as there is more data passing from the server to the client.
+For the tasks we have looked at in this paper, where shape and density are important to the analyst, we think that better displays and sub-sampling strategies are more useful than being able to look at every single point on the canvas.
+We showed in our single cell clustering case study that doing a weighted sample based on cluster membership still allowed us to get a sense of relative cluster orientation, however there are alternative sampling approaches that could be applied, like selecting points close to the cluster centres.
+Alternative displays via statistical transformations could also mitigate the need to show all of the data.
+Recent work
 by @Laa2020-wr is a promising area for further investigation, as well as

### 2020-10-28 response/response.Rmd
chapter 4 comments

@@
       in_header: preamble2.tex
-fontfamily: mathpazo
 ---
@@
 
+### Chapter 3 comments
+
+### Chapter 4 comments
+
+> Much of this chapter depends on understanding what is meant by a coverage trace plot.
+You do have examples (e.g.
+Figures 4.4-4.7), but this is at the very end of the chapter.
+It would be good to provide a motivating figure be added at the start to explain what is meant by “coverage” in this context?
+> Pg 63: Introduce what you mean by coverage traces and provide more motiviation why you would want this over just numerical summaries of the data.
+
+Amended.
+
+> Pg 66: You state “resulting GRanges in memory can be compressed using run-length encoding for any categorical variable”.
+Could you give a numeric example of the amount of compression that can be obtained with RLEs?
+Also, maybe add a brief sentence explaining what are RLEs.
+
+Amended.
+
+> Pg 68: Can you comment on why the GFF files from RefSeq were used?
+
+The model organism is Zebrafish which has limited annotations available compared to human and mouse.
+I chose the RefSeq GFF build because that is what was recommended to me by our collaborators as the canonical reference.
+
+> Pg 68: How did you handle reads that map over boundaries for introns and exons? or do you only keep the ones that are entirely inside an intron or exon?
+
+Reads that are split are across exon / intron boundaries are counted towards both categories.
+
+> Pg 69: In the facets, you used a short descriptor, but could you write out what is each facet label is in the legend?
+
+I think the facet labels are sufficient in conjunction with the figure captions.
+
 ### Chapter 5 comments
+
+>  One question I had as I was reading throughout this chapter, was what is your recommendation for users who want to calculate distances?
+can the liminal framework be used to better approximate distances instead of users who perform clustering directly on top of the tSNE or UMAP spaces?
+While
+is not ideal, and it is recommended clustering should performed in the PC space, it happens all the time.
+Another way of framing this question is can your framework be used to improve distance estimates in the tour with a NLDR method as you note it preserves global structure?
+
+This is a really interesting question.
+To begin, I have amended the clustering case study to comment on a more realistic approach to perform cluster verification.
+I like the idea of using the tour to calibrate distances obtained from an NLDR method (or the other way around).
+As mentioned in the text if there are (well-behaved) clusters in the data, it appears that t-SNE will find them but often messes up their orientation and topology.
+I think you would get some of the way there using the spatial linking approach I described.
+Another approach would be to look at distance preservation between successive tour frames.
+One could then ask, how often are these groups of points close according to the tour?
+Of interest would also be the manual clustering approaches described in Cook and Swayne, 2007.
+This is left as future work.
 

### 2020-10-28 Rmd/04-intron.Rmd
chapter 4 comments

@@
 reads that overlap a single position of the reference genome, and is important for assessing sequencing data quality and used in many different aspects of omics analysis such peak-calling in ChIP-seq or variant calling in DNA-seq [@Sims2014-qk].
-Here we emphasise looking at coverage traces to find biological events of interest, rather than only relying on numerical summaries of the data.
+Here we emphasise looking at coverage traces to find biological events of interest, rather than only relying on numerical summaries of the data (Figure \@ref(fig:cov-trace).
 By faceting these traces over combinations of the experimental design and with their biological context such as gene annotations, we can gain an
@@
 searching for unusual time series via estimating descriptors such as seasonality or autocorrelation, and visualising those descriptors instead [@Hyndman2015-sc].
+
+(ref:example-trace): An illustrative coverage trace plot from Chapter \@ref(ch:plyranges).
+The coverage _score_ is defined as the number of reads that overlap a single base position within a reference genome.
+For example, at position 21.807 of chromosome 8 there are approximately 12 reads overlapping it.
+One could also think of a coverage trace as a histogram with the bin width varies as you move along the genome.
+
+```{r cov-trace, echo = FALSE, fig.cap="(ref:example-trace)}
+knitr::include_graphics(here::here("img", "peak-viz-1.png"))
+```
 
@@
 
+```{r rle-example, echo =FALSE}
+set.seed(1836)
+x <- sample(letters[1:2], 10, replace = TRUE)
+x_rle <- S4Vectors::Rle(x)
+```
 The per base coverage score is estimated directly from one or more BAM files
@@
 memory can be compressed using run-length encoding for any categorical 
-variable. This representation allows us to easily transform the coverage scores
+variable, which is a form of data compression where "runs" of a vector are stored rather than their values.
+For example, the character vector of letters "a", and "b" is shortened so successive values are stored as a single value with their lengths:
+
+```{r, echo = FALSE}
+x
+x_rle
+```
+
+This representation is memory efficient: at worst it will be the same size as the input, and at best reduce the size by a factor corresponding to the largest run.
+The allows us to easily transform the coverage scores
 and integrate annotations using the **plyranges** grammar, and visualise traces using **ggplot2** [@Wickham2016-gz].

### 2020-10-28 Rmd/03-fluentGenomics.Rmd
add reference for MA, increase figure res

@@
 ```{r workflow, fig.cap = "(ref:workflow)"}
-knitr::include_graphics(here::here("img", "workflow.png"))
+knitr::include_graphics(here::here("img", "workflow-fluentgenomics.pdf"))
 ```
@@
 Bioconductor **ExperimentData** package and from the workflow package 
-**fluentGenomics**.
+**fluentGenomics** [@fg-pkg].
 
@@
 
-To see the results of the expression analysis, we can generate a summary table
-and an MA plot:
+To see the results of the expression analysis, we can generate a summary table and a mean-abundance (MA) plot [@Dudoit2002-vz]:
 
@@
 
-(ref:maplot) Visualization of `DESeq2` results as an "MA plot". Genes that have an adjusted *p-value* below 0.01 are colored red.
+(ref:maplot) Visualization of `DESeq2` results as an "MA plot".
+Genes  that have an adjusted *p-value* below 0.01 are colored red.In this case the LFC between conditions is shown on the y-axis, while the average normalised gene counts accross all conditions are shown on the x-axis.
+The assumption of an RNA-seq analysis that most genes are not DE, so most genes are scattered about zero on the y-axis while genes that have evidence of DE are far from the zero baseline.
+The x-axis gives a sense of the total expression of the gene.
 

### 2020-10-28 response/response.Rmd
finishing touches

@@
 
-Most of the comments pertain to the use and learning of the **plyranges** package.
+This chapter has already been published and it appears most of the comments pertain to the use and learning of the **plyranges** package.
 I believe they are best addressed through the **plyranges** documentation and vignette.
@@
 ### Chapter 3 comments
+
+> Almost all of Chapter 2 discusses the use of GRanges objects.
+As most of Chapter 3 discusses the use of the SummarizedExperiment object, could you add a motivating figure at the start explaining the relationship between a SummarizedExperiment and GRanges object (i.e.
+rowRanges)?
+I do not think this is obvious to a reader who may not very familiar with these objects.
+However, I think it is an important aspect to explain in this chapter.
+
+I agree and have amended this by adding a figure which describes these features.
+
+> Pg 31: Figure 3.1 is fairly low resolution. Is it possible to make a higher resolution version?
+
+Amended.
+
+> Pg 31: Figure 3.1 demonstrates how you use plyranges to integrate the results from DE and DA (i.e.
+finding overlaps).
+Would it also make sense to the integration step prior to modeling?
+I am thinking
+about this as a way to reduce the amount of tests performed for DE and DA.
+
+I think it would make sense but it is out of scope for this paper.
+
+> Pg 32: The fluentGenomics package needs a reference
+
+Amended.
+
+>Pg 39: Could you add 1-2 sentences (and maybe a reference) about what is a MA plot and why it’s
+useful for visualizing DE results?
+
+Amended.
+
+> Pg 40: I think it would be useful if you could comment on the results() function a bit more to explain
+what is happening under the hood there.
+It’s a pretty critical step for everything downstream after identifying the DE genes in preparation for the integration step.
+As it stands, I felt like it was kind of glossed over.
+
+Amended.
+
+>Pg 45: More motivation could be provided on why you want to generate random samples of size equal to the number of DE-genes from the other_genes with your larger question in mind of “how many DA peaks are near DE genes relative to”other" non-DE genes¿‘.
+It is stated on pg 46”we minimize the variance on the enrichment statistics induced by the sampling process", but it might be good to explain this in greater detail at a higher level.
+
+I have expanded on this now.
 

### 2020-10-28 Rmd/03-fluentGenomics.Rmd
finishing touches

@@
 ```{r workflow, fig.cap = "(ref:workflow)"}
-knitr::include_graphics(here::here("img", "workflow-fluentgenomics.pdf"))
+knitr::include_graphics(here::here("img", "workflow-fluentgenomics.png"))
 ```
@@
 Bioconductor infrastructure to understand these datasets. In particular, we
-will emphasize the use of the Bioconductor packages **plyranges** and **tximeta**.
+will emphasize the use of the Bioconductor packages **plyranges** and **tximeta** and the _SummarizedExperiment_ class (Figure \@ref(fig:se).
 The **plyranges** package fluently transforms data tied to genomic ranges using
@@
 [@granges; @bioc] and the **tidyverse** design principles @tidyverse.
+
+```{r se, fig.cap = "(ref:se)"}
+knitr::include_graphics(here::here("img", "se-digram.png"))
+```
+
+(ref:se): A _SummarizedExperiment_ orients bulk assay measurements as matrices where rows correspond to features, and columns correspond to samples.
+The rows have their own accessor functions which contains additional data about the feature of interest, in this `rowRanges()` returns a _GRanges_  where rows are the genomic coordinates for each gene measured in the assay, while `colData()` returns a _DataFrame_  where rows contain information about the samples.
 
@@
 
-First, we specify a directory `dir`, where the quantification files are stored.
+First, we specify a directory `path`, where the quantification files are stored.
 You could simply specify this directory with:
@@
 ```{r dir, eval=FALSE, echo = TRUE}
-dir <- "/path/to/quant/files"
+path <- "/path/to/quant/files"
 ```
@@
 library(readr)
-colfile <- file.path(dir, "coldata.csv")
+colfile <- file.path(path, "coldata.csv")
 coldata <- read_csv(colfile) %>%
@@
   mutate(
-    files = file.path(dir, "quants", names, "quant.sf.gz"),
+    files = file.path(path, "quants", names, "quant.sf.gz"),
     line = factor(line),
@@
 makeLinkedTxome(
-  indexDir=file.path(dir, "gencode.v29_salmon_0.12.0"),
+  indexDir=file.path(path, "gencode.v29_salmon_0.12.0"),
   source="Gencode",
@@
   fasta="gencode.v29.transcripts.fa.gz", # ftp link to fasta file
-  gtf=file.path(dir, "gencode.v29.annotation.gtf.gz"), # local version
+  gtf=file.path(path, "gencode.v29.annotation.gtf.gz"), # local version
   write=FALSE
@@
 
+The `results()` function extracts a summary of the DE analysis: in this case for each gene we have the LFC comparing the two cell lines, the Wald test statistic of LFC from the fitted negative binomial GLM, and associated p-value and corrected p-value accounting for the FDR.
 To see the results of the expression analysis, we can generate a summary table and a mean-abundance (MA) plot [@Dudoit2002-vz]:
@@
 (ref:maplot) Visualization of `DESeq2` results as an "MA plot".
-Genes  that have an adjusted *p-value* below 0.01 are colored red.In this case the LFC between conditions is shown on the y-axis, while the average normalised gene counts accross all conditions are shown on the x-axis.
+Genes  that have an adjusted *p-value* below 0.01 are colored red.In this case the LFC between conditions is shown on the y-axis, while the average normalised gene counts across all conditions are shown on the x-axis.
 The assumption of an RNA-seq analysis that most genes are not DE, so most genes are scattered about zero on the y-axis while genes that have evidence of DE are far from the zero baseline.
@@
 hypothesis (`altHypothesis`) that the LFC is less than the threshold in
-absolute value. To visualize the result of this test, you can run `results`
+absolute value.
+In this case, the p-values are taken as maximum of the upper and lower Wald tests under the hypothesis absolute value of the estimated LFC is lower than the threshold.
+To visualize the result of this test, you can run `results`
 without `format="GRanges"`, and pass this object to `plotMA()` as before. We
@@
 
-We can repeat this many times to create many samples via `replicate()`. By
-replicating the sub-sampling multiple times, we minimize the variance on the
+We can repeat this many times to create many samples via `replicate()`.
+We are sub-sampling gene sets without evidence of DE to have the same size as DE gene set, because for each set we want to how different the ATAC peaks around these different sets assuming the number of DE genes is fixed.
+The sampling of the non-DE genes is done without replacement so each replication produces a different set to compare.
+By replicating the sub-sampling multiple times, we minimize the variance on the
 enrichment statistics induced by the sampling process.

### 2020-10-28 Rmd/07-appendix-A.Rmd
finishing touches

@@
 We have two `Ranges` objects. The first contains single nucleotide positions
-corresponding to an intensity measurement such as a ChiP-seq experiment, while
+corresponding to an intensity measurement such as a ChIP-seq experiment, while
 the other contains coordinates for two genes of interest.

### 2020-10-28 scripts/fluentGenomics.R

@@ -9,16 +9,16 @@ library(DESeq2)
 library(readr)
 library(dplyr)
 
-dir <- system.file("extdata", package="macrophage")
+path <- system.file("extdata", package="macrophage")
 
 makeLinkedTxome(
-  indexDir=file.path(dir, "gencode.v29_salmon_0.12.0"),
+  indexDir=file.path(path, "gencode.v29_salmon_0.12.0"),
   source="Gencode",
   organism="Homo sapiens",
   release="29",
   genome="GRCh38",
   fasta="ftp://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_human/release_29/gencode.v29.transcripts.fa.gz",
-  gtf=file.path(dir, "gencode.v29.annotation.gtf.gz"), # local version
+  gtf=file.path(path, "gencode.v29.annotation.gtf.gz"), # local version
   write=FALSE
 )
 # # alias to plyranges

### 2020-10-28 Rmd/01-intro.Rmd
capitalisation

@@
 leading to new hypotheses that can be tested and formalised using 
-statistical modelling (figure \@ref(fig:wflow-cycle)). As datasets grow in complexity and become
+statistical modelling (Figure \@ref(fig:wflow-cycle)). As datasets grow in complexity and become
 increasingly heterogeneous and multidimensional, 
@@
 biological data using a coherent representation and programming interface 
-(figure \@ref(fig:wflow-wrangle)). Section \@ref(sec:tidygranges)
+(Figure \@ref(fig:wflow-wrangle)). Section \@ref(sec:tidygranges)
 introduces a grammar-based framework for transforming genomics data that is 
-described in chapter \@ref(ch:plyranges). We then look at integrating data and 
+described in Chapter \@ref(ch:plyranges). We then look at integrating data and 
 model outputs over genomic regions to gain biological insight (\@ref(fig:wflow-integrate)). Section \@ref(sec:integrate) introduces a 
 framework for incorporating genomic regions  over multiple assays, described in 
-chapter \@ref(ch:fluentGenomics), while  \@ref(sec:coverage) discusses finding 'interesting' genomic regions via combining multiple summaries of a single assay, described in \@ref(ch:intron).
+Chapter \@ref(ch:fluentGenomics), while  \@ref(sec:coverage) discusses finding 'interesting' genomic regions via combining multiple summaries of a single assay, described in \@ref(ch:intron).
 Next, we
-consider the challenges in  visualising  high dimensional data (figure \@ref(fig:wflow-explore)). Section  \@ref(sec:va) introduces an interactive 
+consider the challenges in  visualising  high dimensional data (Figure \@ref(fig:wflow-explore)). Section  \@ref(sec:va) introduces an interactive 
 visualisation approach for  understanding non-linear dimension reduction 
-techniques described in chapter  \@ref(ch:tsne). Lastly, in 
-chapter \@ref(ch:conclusion) describes the outputs of the thesis and plans for future developments.
+techniques described in Chapter  \@ref(ch:tsne). Lastly, in 
+Chapter \@ref(ch:conclusion) describes the outputs of the thesis and plans for future developments.
 
@@
 in fields such as high-throughput biology where domain specific semantics are 
-required (figure \@ref(fig:wflow-wrangle)). This is particularly true in the Bioconductor ecosystem, where much 
+required (Figure \@ref(fig:wflow-wrangle)). This is particularly true in the Bioconductor ecosystem, where much 
 thought has gone into the design of data structures that enable interoperability
@@
 a single measurement assay or that only one aspect of a measurement assay
-will be of interest to the biological question under study (figure \@ref(fig:wflow-integrate)). While there are 
+will be of interest to the biological question under study (Figure \@ref(fig:wflow-integrate)). While there are 
 many approaches to integrating data sets from multiple assays using multivariate
 statistical techniques [@Meng2016-pz; @Stein-OBrien2018-bk] and data structures to represent them [@Ramos2017-kn], there has been little thought given to the interoperability between these approaches and the **tidyverse**.
-In chapter \@ref(ch:fluentGenomics)
+In Chapter \@ref(ch:fluentGenomics)
 we describe a simple end-to-end workflow for integrating 
@@
 
-In chapter \@ref(ch:intron) we explore the limits of the *tidy data*
+In Chapter \@ref(ch:intron) we explore the limits of the *tidy data*
 semantic by extending **plyranges** to analyse coverage estimated on RNA-seq
@@
 signal from RNA-seq data. This chapter is based on my software and analysis
-contributions to the preprint @Lee2019-mf.
+contributions to the  @Lee2019-mf.
 
@@
 
-In the final chapter, we move away from data wrangling and towards the 
+Finally, we move away from data wrangling and towards the 
 integration of visualisation with model-based summaries of high-dimensional
-data sets (figure \@ref(fig:wflow-explore)). We focus on a common tool for EDA 
+data sets (Figure \@ref(fig:wflow-explore)). We focus on a common tool for EDA 
 (especially applied to single-cell transcriptomics): non-linear dimension 

### 2020-10-28 Rmd/02-plyranges.Rmd
capitalisation

@@
 measurements or metadata specific to the data type or experiment
-(figure \@ref(fig:GRanges)).  _GRanges_ balances flexibility with formal
+(Figure \@ref(fig:GRanges)).  _GRanges_ balances flexibility with formal
 constraints, so that it is applicable to virtually any genomic
@@
 ways. There are four supported matching algorithms: _overlap_,
-_nearest_, _precede_, and _follow_ (figure 
+_nearest_, _precede_, and _follow_ (Figure 
 \@ref(fig:olaps-fig)). We can further restrict the
@@
 **dplyr** grammar. Consequently, **plyranges** code is generally more compact
-than the equivalent **GenomicRanges** code (figure \@ref(fig:code-comparison)).
+than the equivalent **GenomicRanges** code (Figure \@ref(fig:code-comparison)).
 
@@
 Finally, the overlap inner join is used to restrict the chromosome 10
-coverage islands, to the islands that are contained in the 5000nt region that surrounds the max peak (figure \@ref(fig:peak-viz)).
+coverage islands, to the islands that are contained in the 5000nt region that surrounds the max peak (Figure \@ref(fig:peak-viz)).
 
@@
 We then compute the transition-transversion ratio over each chromosome using
-`group_by()` in combination with `summarize()` (figure \@ref(fig:titv-viz)).
+`group_by()` in combination with `summarize()` (Figure \@ref(fig:titv-viz)).
 

### 2020-10-28 Rmd/04-intron.Rmd
finishing touches
capitalisation

@@
 
-```{r cov-trace, echo = FALSE, fig.cap="(ref:example-trace)}
+```{r cov-trace, echo = FALSE, fig.cap="(ref:example-trace)"}
 knitr::include_graphics(here::here("img", "peak-viz-1.png"))
@@
 over regions of the genome using our companion package **plyranges**. The
-aspects of this combined workflow are summarised in figure \@ref(fig:si-fig).
+aspects of this combined workflow are summarised in Figure \@ref(fig:si-fig).
 
@@
 features. It refers to the total number of bases within an intron that has a score above the overall average exon coverage score. The scatter plot 
-matrix view of these descriptors is shown in figure \@ref(fig:si-splom)
+matrix view of these descriptors is shown in Figure \@ref(fig:si-splom)
 for a single biological group in the experiment. Using these views and
@@
 to link back to their underlying coverage traces, with the overlaps shown
-in the UpSet plot in figure \@ref(fig:si-upset) [@Lex2014-bh].
+in the UpSet plot in Figure \@ref(fig:si-upset) [@Lex2014-bh].
 The procedure produces gene coverage traces with known minor class splice sites affected by the knockout procedure such as ccdc43 and nat15 (figures \@ref(fig:si-cvg-ccdc43) and \@ref(fig:si-cvg-nat15)), as well as some that appear to have U12 intron retention like events such as mapk3 or tspan31 (figures \@ref(fig:si-cvg-mapk3) and \@ref(fig:si-cvg-tspan31)) that affect other parts of the gene.
@@
 design to find differential IR like profiles. However, if that was of interest,
-one could look at the overlaps, like we did in figure \@ref(fig:si-upset),
+one could look at the overlaps, like we did in Figure \@ref(fig:si-upset),
 or combine our data descriptors with external estimates using **limma** 

### 2020-10-28 Rmd/05-embeddings.Rmd
capitalisation

@@
 
-From the video linked in figure \@ref(fig:gaussian), we learn that t-SNE has 
+From the video linked in Figure \@ref(fig:gaussian), we learn that t-SNE has 
 correctly split out each cluster and laid them out in a star like formation. 
@@
 generally points are more concentrated in the projection view compared to 
-the t-SNE layout (figure \@ref(fig:gaussian)a). 
-This can be seen via brushing the t-SNE view (figure \@ref(fig:gaussian)b).
+the t-SNE layout (Figure \@ref(fig:gaussian)a). 
+This can be seen via brushing the t-SNE view (Figure \@ref(fig:gaussian)b).
 
@@
 that are each equidistant from each other and have the same covariance structure.
-From the video linked in figure \@ref(fig:hierarchical), we see 
+From the video linked in Figure \@ref(fig:hierarchical), we see 
 that t-SNE has correctly identified the sub-clusters. However, their relative 
 locations to each other is distorted, with the orange and blue groups being
-far from each other in the tour view (figure \@ref(fig:hierarchical)a). We see 
+far from each other in the tour view (Figure \@ref(fig:hierarchical)a). We see 
 in this case that is difficult
 to see the sub-clusters in the tour view, however, once we zoom and highlight
-they become more apparent (figure \@ref(fig:hierarchical)b). When we brush 
+they become more apparent (Figure \@ref(fig:hierarchical)b). When we brush 
 the sub-clusters in the t-SNE, their
@@
 
-Next we explore some simulated noisy tree structured data (figure
+Next we explore some simulated noisy tree structured data (Figure
 \@ref(fig:fake-trees)). Our interest here is how t-SNE visualisations break
@@
 initialisation). We then create a linked
-tour with t-SNE layout with **liminal** as shown in figure \@ref(fig:trees-01).
+tour with t-SNE layout with **liminal** as shown in Figure \@ref(fig:trees-01).
 
@@
 recapitulate the topology of the tree - the backbone (blue) branch has been 
-split into three fragments (figure \@ref(fig:trees-01)a). We can see 
+split into three fragments (Figure \@ref(fig:trees-01)a). We can see 
 this immediately via the linked highlighting over both
@@
 view there is a lot of noise along the branches, while this does not appear to
-be the case for the t-SNE result (figure \@ref(fig:trees-01)b).
+be the case for the t-SNE result (Figure \@ref(fig:trees-01)b).
 
@@
 these results with our tour view as displayed in the linked video in the caption
-of figure \@ref(fig:trees-02).
+of Figure \@ref(fig:trees-02).
 
@@
 
-The video linked in figure \@ref(fig:trees-02) shows that this selection of 
+The video linked in Figure \@ref(fig:trees-02) shows that this selection of 
 parameters results in the 
-tips of the branches (the three black dots in figure \@ref(fig:fake-trees)a) 
+tips of the branches (the three black dots in Figure \@ref(fig:fake-trees)a) 
 being split into three clusters representing the terminal branches of the tree. 
@@
 view, and in the tour view these tips are at opposite ends of the tree 
-(figure \@ref(fig:trees-02)b).
+(Figure \@ref(fig:trees-02)b).
 Although, this is a known issue of the t-SNE algorithm, we can easily identify
@@
 data size - 4,590 cells. Although this is not ideal, it still allows us to
-get a sense of the shape of the clusters as seen from the linked video in figure \@ref(fig:mouse-01).
+get a sense of the shape of the clusters as seen from the linked video in Figure \@ref(fig:mouse-01).
 If one was interested in performing more in-depth cluster analysis we recommend an iterative approach of removing large clusters and then re-running the **liminal** view as a way finding more granular cluster structure.
@@
 
-From the video linked in figure \@ref(fig:mouse-01), we learn that the embedding has mostly captured the
+From the video linked in Figure \@ref(fig:mouse-01), we learn that the embedding has mostly captured the
 clusters relative location to each other to their location in high dimensional space, with a notable exception of points in cluster 3 and 10 as shown with
-linked brushing (figure \@ref(fig:mouse-01)a). As expected, t-SNE mitigates the crowding problem that is
+linked brushing (Figure \@ref(fig:mouse-01)a). As expected, t-SNE mitigates the crowding problem that is
 an issue for tour in this case, where points in clusters 2,4,6, and 11 are
-clumped together in tour view, but are blown up in the embedding view (figure \@ref(fig:mouse-01)b). 
+clumped together in tour view, but are blown up in the embedding view (Figure \@ref(fig:mouse-01)b). 
 The tour appears to form a tetrahedron like shape, with points lying on
@@
 a phenomena that has also been observed in @Korem2015-af 
-(figure \@ref(fig:mouse-01)c). Brushing on
+(Figure \@ref(fig:mouse-01)c). Brushing on
 the tour view, reveals points in cluster 9 that are diffuse in the embedding 
@@
 clusters in the tour view, but has points placed in cluster 3 and 9 in the embedding
-(figure \@ref(fig:mouse-01)d).
+(Figure \@ref(fig:mouse-01)d).
 
@@
 
-From the video linked in figure \@ref(fig:mouse-02), we see that the 
+From the video linked in Figure \@ref(fig:mouse-02), we see that the 
 expression of the marker genes, appear to separate
 the previously clumped clusters 2,4,6, and 11 from the other clusters, 
-indicating that these genes are expressed in all four clusters (figure \@ref(fig:mouse-02)a). After zooming, we can see a trajectory forming along
+indicating that these genes are expressed in all four clusters (Figure \@ref(fig:mouse-02)a). After zooming, we can see a trajectory forming along
 the clusters, while the axis view shows that magnitude of expression in 
 the marker genes is similar across these separated clusters which is 
-consistent with the results of marker gene analysis (figure \@ref(fig:mouse-02)b).
+consistent with the results of marker gene analysis (Figure \@ref(fig:mouse-02)b).
 

### 2020-10-28 Rmd/06-conclusion.Rmd
capitalisation

@@
 its extension to range based genomics data. This culminated in the development
-of "plyranges: a grammar of genomic data transformation" in chapter 
+of "plyranges: a grammar of genomic data transformation" in Chapter 
 \@ref(ch:plyranges), which developed a new domain specific language for genomics
 data analysis. The applicability of the **plyranges** interface and use of the 
-*tidy data* concept were further interrogated in chapter \@ref(ch:fluentGenomics),
+*tidy data* concept were further interrogated in Chapter \@ref(ch:fluentGenomics),
 "Fluent genomics with plyranges and tximeta", which described techniques 
 integrating data along the genome, and emphasised the importance of 
-interoperability between analysis tools. Similarly, chapter  \@ref(ch:intron), 
+interoperability between analysis tools. Similarly, Chapter  \@ref(ch:intron), 
 "Exploratory coverage analysis with superintronic and 
@@
 at Bioconductor conferences which also led to the development of the 
-**fluentGenomics** workflow package, outlined in chapter \@ref(ch:fluentGenomics). A broader impact of the work, has been the discussions around the
+**fluentGenomics** workflow package, outlined in Chapter \@ref(ch:fluentGenomics). A broader impact of the work, has been the discussions around the
 concepts of fluent interfaces and tidy data within the Bioconductor community,
@@
 
-The **superintronic** software described in chapter \@ref(ch:intron) has
+The **superintronic** software described in Chapter \@ref(ch:intron) has
 been used in @Lee2019-mf to disentangle and view intron signal in RNA-seq
@@
 
-We showed in chapter \@ref(ch:tsne) that tours provide a global overview
+We showed in Chapter \@ref(ch:tsne) that tours provide a global overview
 that can be used as tool for exploring model fits. An issue that arises is

### 2020-10-28 Rmd/07-appendix-A.Rmd
capitalisation

@@
 # (APPENDIX) Appendix {-}
+
+```{r detach, include = FALSE}
+detach("package:dplyr", unload = TRUE)
+detach("package:plyranges", unload = TRUE)
+
+```
 
@@
 
-## Getting started with the **plyranges** package
+# Getting started with the **plyranges** package
 
-### `Ranges` revisited
+## _Ranges_ revisited
 
@@
 standard data structures for representing genomics data.  Throughout this
-document I refer to either of these classes as `Ranges` if an operation can be
+document I refer to either of these classes as _Ranges_ if an operation can be
 performed on either class, otherwise I explicitly mention if a function is
-appropriate for an `IRanges` or `GRanges`.
+appropriate for an _IRanges_ or _GRanges_.
 
-`Ranges` objects can either represent sets of integers as `IRanges` (which have
+_Ranges_ objects can either represent sets of integers as _IRanges_ (which have
 start, end and width attributes) or represent genomic intervals (which have
-additional attributes, sequence name, and strand) as `GRanges`. In addition,
-both types of `Ranges` can store information about their intervals as metadata
+additional attributes, sequence name, and strand) as _GRanges_. In addition,
+both types of _Ranges_ can store information about their intervals as metadata
 columns (for example GC content over a genomic interval).
 
-`Ranges` objects follow the tidy data principle: each row of a `Ranges` object
+_Ranges_ objects follow the tidy data principle: each row of a _Ranges_ object
 corresponds to an interval, while each column will represent a variable about
@@
 
-Consequently, `Ranges` objects provide a powerful representation for reasoning
-about genomic data. In this vignette, you will learn more about `Ranges`
+Consequently, _Ranges_ objects provide a powerful representation for reasoning
+about genomic data. In this vignette, you will learn more about _Ranges_
 objects and how via grouping, restriction and aggregation you can perform
@@
 
-### Constructing `Ranges`
+## Constructing _Ranges_
 
-To construct an `IRanges` we require that there are at least two columns that
+To construct an _IRanges_ we require that there are at least two columns that
 represent at either a starting coordinate, finishing coordinate or the width of
@@
 
-To construct a `GRanges` we require a column that represents that sequence name
+To construct a _GRanges_ we require a column that represents that sequence name
 ( contig or chromosome id), and an optional column to represent the
@@
 
-### Arithmetic on Ranges
+## Arithmetic on Ranges
 
@@
 
-```{r anchor_fig, echo = FALSE, out.width="400px", fig.align="center"}
+```{r anchor_fig, echo = FALSE}
 knitr::include_graphics(here::here("img", "anchors.png"), dpi = 150)
@@
 
-`Ranges` can be shifted left or right. If strand information is available we
+_Ranges_ can be shifted left or right. If strand information is available we
 can also shift upstream or downstream.
@@
 
-### Grouping `Ranges`
+## Grouping _Ranges_
 
-**plyranges** introduces a new class of `Ranges` called `RangesGrouped`, this is
+**plyranges** introduces a new class of _Ranges_ called `RangesGrouped`, this is
 a similar idea to the grouped `data.frame\tibble` in **dplyr**.
@@
 Grouping can act on either the core components or the metadata columns of a
-`Ranges` object.
+_Ranges_ object.
 
@@
 
-### Restricting `Ranges`
+## Restricting _Ranges_
 
-The verb `filter` can be used to restrict rows in the `Ranges`. Note that
+The verb `filter` can be used to restrict rows in the _Ranges_. Note that
 grouping will cause the `filter` to act within each group of the data.
@@
 We also provide the convenience methods `filter_by_overlaps` and
-`filter_by_non_overlaps` for restricting by any overlapping `Ranges`.
+`filter_by_non_overlaps` for restricting by any overlapping _Ranges_.
 
@@
 
-### Summarising `Ranges`
+## Summarising _Ranges_
 
 The `summarise` function will return a `DataFrame` because the information
-required to return a `Ranges` object is lost. It is often most useful to use
+required to return a _Ranges_ object is lost. It is often most useful to use
 `summarise()` in combination with the `group_by()` family of functions.
@@
 
-### Joins, or another way at looking at overlaps between `Ranges`
+## Joins, or another way at looking at overlaps between _Ranges_
 
@@
 
-```{r olaps-diagram, echo = FALSE, out.width="600px"}
+```{r olaps-diagram, echo = FALSE}
 knitr::include_graphics(here::here("img", "olap-fig.png"))
@@
 
-The first function, `join_overlap_intersect()` will return a `Ranges` object
+The first function, `join_overlap_intersect()` will return a _Ranges_ object
 where the start, end, and width coordinates correspond to the amount of any
-overlap between the left and right input `Ranges`. It also returns any
+overlap between the left and right input _Ranges_. It also returns any
 metadata in the subject range if the subject overlaps the query.
@@
 
-The `join_overlap_inner()` function will return the `Ranges` in the query that
-overlap any `Ranges` in the subject. Like the `join_overlap_intersect()`
+The `join_overlap_inner()` function will return the _Ranges_ in the query that
+overlap any _Ranges_ in the subject. Like the `join_overlap_intersect()`
 function metadata of the subject `Range` is returned if it overlaps the query.
@@
 Compared with `filter_by_overlaps()` above, the overlap left join expands the
-`Ranges` to give information about each interval on the query `Ranges` that
-overlap those on the subject `Ranges` as well as the intervals on the left that
+_Ranges_ to give information about each interval on the query _Ranges_ that
+overlap those on the subject _Ranges_ as well as the intervals on the left that
 do not overlap any range on the right.
@@
 #### Finding your neighbours
-We also provide methods for finding nearest, preceding or following `Ranges`.
+We also provide methods for finding nearest, preceding or following _Ranges_.
 Conceptually this is identical to our approach for finding overlaps, except the
@@
 
-```{r neigbours, echo = FALSE, out.width="600px"}
+```{r neigbours, echo = FALSE}
 knitr::include_graphics(here::here("img", "neighbours.png"))
@@
 join_follow(ir0, ir1)
-join_precede(ir0, ir1) # nothing precedes returns empty `Ranges`
+join_precede(ir0, ir1) # nothing precedes returns empty _Ranges_
 join_precede(ir1, ir0)
@@
 
-### Example: dealing with multi-mapping
+#### Example: dealing with multi-mapping
 This example is taken from the Bioconductor support
@@
 
-We have two `Ranges` objects. The first contains single nucleotide positions
+We have two _Ranges_ objects. The first contains single nucleotide positions
 corresponding to an intensity measurement such as a ChIP-seq experiment, while
@@
 
-We want to identify which positions in the `intensities` `Ranges` overlap the
+We want to identify which positions in the `intensities` _Ranges_ overlap the
 genes, where each row corresponds to a position that overlaps a single gene.
 
-First we create the two `Ranges` objects
+First we create the two _Ranges_ objects
 ```{r ex1}
@@
 
-### Grouping by overlaps
+## Grouping by overlaps
 
@@
 
-### Data Import/Output
+## Reading Genomic Files
 
 We provide convenience functions via **rtracklayer** and **GenomicAlignments** for
-reading/writing the following data formats to/from `Ranges` objects.
+reading/writing the following data formats to/from _Ranges_ objects.
 
@@
 
-# Learning more
+## Learning more
 

### 2020-10-28 Rmd/00-c-declaration.Rmd
typo fixes

@@
 co-authors. Of special note is chapter \@ref(ch:intron), which reflects my 
-contribution (in the form of a software package) to the submitted manuscript entitled 
+contribution (in the form of a software package) to the published manuscript entitled 
 _Covering all your bases: incorporating intron signal from RNA-seq data_

### 2020-10-28 Rmd/03-fluentGenomics.Rmd
correct path
typo fixes

@@
 ```{r workflow, fig.cap = "(ref:workflow)"}
-knitr::include_graphics(here::here("img", "workflow-fluentgenomics.png"))
+knitr::include_graphics(here::here("img", "workflow-fluentgenomics.pdf"))
 ```
@@
 Bioconductor infrastructure to understand these datasets. In particular, we
-will emphasize the use of the Bioconductor packages **plyranges** and **tximeta** and the _SummarizedExperiment_ class (Figure \@ref(fig:se).
+will emphasize the use of the Bioconductor packages **plyranges** and **tximeta** and the _SummarizedExperiment_ class (Figure \@ref(fig:se)).
 The **plyranges** package fluently transforms data tied to genomic ranges using
@@
 ```{r se, fig.cap = "(ref:se)"}
-knitr::include_graphics(here::here("img", "se-digram.png"))
+knitr::include_graphics(here::here("img", "se-diagram.pdf"))
 ```
 
-(ref:se): A _SummarizedExperiment_ orients bulk assay measurements as matrices where rows correspond to features, and columns correspond to samples.
+(ref:se) A _SummarizedExperiment_ orients bulk assay measurements as matrices where rows correspond to features, and columns correspond to samples.
 The rows have their own accessor functions which contains additional data about the feature of interest, in this `rowRanges()` returns a _GRanges_  where rows are the genomic coordinates for each gene measured in the assay, while `colData()` returns a _DataFrame_  where rows contain information about the samples.
@@
 (ref:maplot) Visualization of `DESeq2` results as an "MA plot".
-Genes  that have an adjusted *p-value* below 0.01 are colored red.In this case the LFC between conditions is shown on the y-axis, while the average normalised gene counts across all conditions are shown on the x-axis.
+Genes  that have an adjusted *p-value* below 0.01 are colored blue.In this case the LFC between conditions is shown on the y-axis, while the average normalised gene counts across all conditions are shown on the x-axis.
 The assumption of an RNA-seq analysis that most genes are not DE, so most genes are scattered about zero on the y-axis while genes that have evidence of DE are far from the zero baseline.

### 2020-10-28 Rmd/05-embeddings.Rmd
typo fixes

@@
 t-SNE avoids the crowding problem, points are separated into distinct regions
-on the display canvas, while for the tour, points are concentrated in the
+on the display canvas. For the tour, points are concentrated in the
 centre of the projection and become difficult to see. We have recently proposed a simple non-linear transformation for the tour called a sage tour that aims to fix this problem [@Laa2020-vk]. 

### 2020-10-28 scripts/fluentGenomics.R

@@ -26,13 +26,13 @@ makeLinkedTxome(
 # n <- plyranges::n
 
 ## ----setdir----------------------------------------
-dir <- system.file("extdata", package="macrophage")
+path <- system.file("extdata", package="macrophage")
 
 
 ## ----coldata-rna-----------------------------------
 library(readr)
 library(dplyr)
-colfile <- file.path(dir, "coldata.csv")
+colfile <- file.path(path, "coldata.csv")
 coldata <- read_csv(colfile) %>%
   select(
     names,
@@ -41,7 +41,7 @@ coldata <- read_csv(colfile) %>%
     condition = condition_name
   ) %>%
   mutate(
-    files = file.path(dir, "quants", names, "quant.sf.gz"),
+    files = file.path(path, "quants", names, "quant.sf.gz"),
     line = factor(line),
     condition = relevel(factor(condition), "naive")
   )

### 2020-10-28 response/response.Rmd
finalise response

@@
       in_header: preamble2.tex
+bibliography: response.bib
+fontfamily: mathpazo
+citation_package: biblatex
 ---
@@
 
-Noted. I can change this in the workflow for next Bioconductor release.
+Amended - since I can update on F1000 at some point.
 
@@
 
-This was a reference to Buja, Cook, and Swayne (1996) paper mentioned in the paragraph above.
-Here "Gestalt" refers to Gestalt theory in perception and psychology, it refers to identifying patterns in visual forms.
+This was a reference to @Buja1996-fk paper mentioned in the paragraph above. Here "Gestalt" refers to Gestalt theory in perception and psychology, it refers to identifying patterns in visual forms.
 
@@
 
-Thank you for your thorough review. I am looking forward to incorporating your suggestions into future work.
+Thank you for your thorough review and thoughtful suggestions. This feedback is extremely helpful and I am looking forward to incorporating your suggestions into future work.
 
-### Addressing General Comments
+### Addressing General Comments {-}
 
-#### On Scalability
+**On Scalability**
 
@@
 
-#### On alignment vs. pseudoalignment
+**On alignment vs. pseudoalignment**
 
@@
 
-### Chapter 1 comments
+### Chapter 1 comments {-}
 
@@
 
-TODO
+Amended. I have made this consistent throughout the thesis now.
 
@@
 
-### Chapter 2 comments
+### Chapter 2 comments {-}
 
@@
 
-### Chapter 3 comments
+### Chapter 3 comments {-}
 
@@
 
-### Chapter 4 comments
+### Chapter 4 comments {-}
 
@@
 
-### Chapter 5 comments
+### Chapter 5 comments {-}
 
@@
 One could then ask, how often are these groups of points close according to the tour?
-Of interest would also be the manual clustering approaches described in Cook and Swayne, 2007.
+Of interest would also be the manual clustering approaches described in @Cook2007-ld.
 This is left as future work.
@@
 Amended.
+
 > Pg 80 (also in 81): The way “tSNE” was written in “this essentially turns t-SNE” looks very different
@@
 
-I think this is not obvious in the static figure but is obvious in the video and when using the **liminal** package that the embedding view is always on the left, and the tour view is on the right (it will be the view that is animated).
+I think this is not immediately obvious in the static figure but is obvious in the video and when using the **liminal** package that the embedding view is always on the left, and the tour view is on the right (it will be the view that is animated).
 I think the suggestion of titling the views is a good one and will place that into the software in future.
@@
 
-### Chapter 6 comments
+### Chapter 6 comments {-}
 
@@
 Amended.
+
+# References

### 2020-10-28 response/response.Rmd
fix typos

@@
 Second, my implementation of **plyranges** is deeply entwined with the underlying core Bioconductor infrastructure, especially the data structures and methods found in the **IRanges** and **GenomicRanges** packages.
-This would result any benchmarking of **plyranges** turning into benchmarking of those packages which has been done elsewhere.
+This would result in any benchmarking of **plyranges** turning into benchmarking of those packages which has been done elsewhere.
 
@@
 I have amended the introduction to include a note about alignment. 
-I believe it is out of scope to go into the gritty details of when to use alignment or pseudo-alignment methods to obtain counts for RNA-seq data.
+I believe it is out of scope to go into the finer details of when to use alignment or pseudo-alignment methods to obtain counts for RNA-seq data.
 I have included results from both approaches to highlight that **plyranges** is agnostic to the method used and is flexible for many types of data.
@@
 
-I have clarified what this refers to.
+Amended. It now refers to data measured along the genome.
 
@@
 
-I agree and have amended this by adding a figure which describes these features.
+I agree and have amended this by adding a figure which explains what a `SummarizedExperiment`  is and how it is related to a `GRanges`.
 
@@
 
-I have expanded on this now.
+I have added a few more sentences to this section which describes the motivation for selecting sets of the same size between non-DE and DE genes, and why we want to do sub-sampling.
 

### 2020-10-29 Rmd/04-intron.Rmd
typo fixes
fix typos
clarify labelling

@@
 
-(ref:example-trace): An illustrative coverage trace plot from Chapter \@ref(ch:plyranges).
+(ref:example-trace) An illustrative coverage trace plot from Chapter \@ref(ch:plyranges).
 The coverage _score_ is defined as the number of reads that overlap a single base position within a reference genome.
 For example, at position 21.807 of chromosome 8 there are approximately 12 reads overlapping it.
-One could also think of a coverage trace as a histogram with the bin width varies as you move along the genome.
 
@@
 provides some syntactic sugar for unravelling gene annotations into their exonic
-and intronic parts, and intersecting them with a coverage _GRanges_.
+and intronic parts, and intersecting them with a coverage _GRanges_. Split reads that cross the boundaries of exon and intron parts are counted towards both unless filtered beforehand.
 
@@
 The gene annotation files were
-obtained as GFF files from RefSeq and used to construct the exonic
+obtained as GFF files from RefSeq as is standard for Zebrafish analyses. The GFF was used to construct the exonic
 and intronic parts of each gene as _GRanges_ object using **superintronic**. 

### 2020-10-29 Rmd/05-embeddings.Rmd
clarify labelling

@@
 the three principles for interactive high-dimensional data visualisation outlined
-by @Buja1996-fk: finding gestalt, posing queries, and making comparisons.
+by @Buja1996-fk: finding gestalt (identifying patterns in visual forms), posing queries, and making comparisons.
 
@@
 
-```{r mouse-01, fig.cap = "(ref:mouse-01)", out.height = "85%"}
+```{r mouse-01, fig.cap = "(ref:mouse-01)", out.height = "90%"}
 display_screenshot("mouse-01", vid_lookup)

### 2020-10-29 response/response.Rmd
update response

@@
 First, I would like to thank my examiners for their constructive feedback and insights.
-I am excited to incorporate their ideas into my future research.
-The point by point description of changes are below: the examiners' comments are in red and my response is in black.
+I have incorporated most of the comments into thesis revisions.
+Where changes have not been made I have described my plans to incorporate them into future work.
+The point by point description of changes is below: the examiners' comments are in black and my response is in red.
 
@@
 
-Thank you for your comments and identification of typographic errors.
+p. 1 Introduction: it’s -> its
 
-> p. 1 Introduction: it’s -> its
+> Amended.
 
-Amended.
+p. 16 “Using plyranges and the the”
 
-> p. 16 “Using plyranges and the the”
+> Amended.
 
-Amended.
-
-> p.
+p.
 34: use of `dir` for a directory path string when `dir()` is a function returning contents of a directory path could be confusing.
@@
 
-Amended - since I can update on F1000 at some point.
+> Amended - since I can update on F1000 at some point.
 
-> P. 43-44: Why is the genome (“hg38”) set using an endomorphic function, but style (“UCSC”) set using a replacement function (`seqlevelsStyle`)?
+p. 43-44: Why is the genome (“hg38”) set using an endomorphic function, but style (“UCSC”) set using a replacement function (`seqlevelsStyle`)?
 
-At the time of writing, `plyranges` does not have any genome restyling functions, hence why the replacement function was used.
+> No change was made. **plyranges** does not have any genome restyling functions, hence why the replacement function was used.
 
-> p. 77: Tensorflow “non-liner embedding” p. 77-78” “a lot of” -> many
+p. 77: Tensorflow “non-liner embedding” p. 77-78” “a lot of” -> many
 
-Amended.
+> Amended.
 
-> p. 80: “its bounded above” ->it’s
+p. 80: “its bounded above” ->it’s
 
-Amended.
+> Amended.
 
-> p. 82: “Like, when, using” -> Like when using “While our software, liminal is able “ -> While our software, liminal, is able
+p. 82: “Like, when, using” -> Like when using “While our software, liminal is able “ -> While our software, liminal, is able
 
-Amended.
+> Amended.
 
->Section 5.3.1 “Finding Gestalt: focus and context” - this is probably a clever metaphor, but I didn’t understand it.
+Section 5.3.1 “Finding Gestalt: focus and context” - this is probably a clever metaphor, but I didn’t understand it.
 
-This was a reference to @Buja1996-fk paper mentioned in the paragraph above. Here "Gestalt" refers to Gestalt theory in perception and psychology, it refers to identifying patterns in visual forms.
+> This is a reference to @Buja1996-fk paper mentioned in the paragraph above. I have amended section 5.3 to add what I mean by "Find Gestalt" which refers to identifying patterns in visual forms.
 
-> p. 84: “Whether, points are near, or far” -> remove both commas
+p. 84: “Whether, points are near, or far” -> remove both commas
 
-Amended.
+> Amended.
 
->Figure 5.2 video https://player.vimeo.com/video/43963590 does not exist. The URL given in Table 5.1 is different and correct.
+Figure 5.2 video https://player.vimeo.com/video/43963590 does not exist. The URL given in Table 5.1 is different and correct.
 
-Amended.
+> Amended.
 
-> p. 98: “We have shown in the case studies, that” -> no comma
+p. 98: “We have shown in the case studies, that” -> no comma
 
-Amended.
+> Amended.
 
 ## Assistant Professor Stephanie C. Hicks {-}
-
-Thank you for your thorough review and thoughtful suggestions. This feedback is extremely helpful and I am looking forward to incorporating your suggestions into future work.
 
@@
 
-I would like to comment that I have avoided performance benchmarking for two main reasons.
-First, I believe it is out of scope for the thesis, since my focus has been on the development of programming and user interfaces.
+> There are two main reasons to not talk about performance benchmarking.
+First, the focus has been on the development of programming and user interfaces and performance benchmarking would distract from that.
 Second, my implementation of **plyranges** is deeply entwined with the underlying core Bioconductor infrastructure, especially the data structures and methods found in the **IRanges** and **GenomicRanges** packages.
@@
 
-Regarding out of memory data structures like `DelayedArray`, it is possible to incorporate them as columns to a `GRanges` object and then the full **plyranges** API would be available.
+> Regarding out of memory data structures like `DelayedArray`, it is possible to incorporate them as columns to a `GRanges` object and then the full **plyranges** API would be available.
 Currently, there appear to be no plans in Bioconductor for out of memory `Ranges`, and indeed there are times where it is more appropriate to switch to a rectangular data layout for computation, as is reflected in the workflow described in chapter 3 of the thesis.
-I believe the development of a scalable out of memory interface for genomic data analysis will require the transformation between data measured along the genome (covered by the **plyranges** API) and the rectangular layouts that enable out of disk computing like `SummarizedExperiment`.
+The development of a scalable out of memory interface for genomic data analysis will require the transformation between data measured along the genome (covered by the **plyranges** API) and the rectangular layouts that enable out of disk computing like `SummarizedExperiment`.
 This is an exciting research area that I would like to explore in the currently under development **plyexperiment**.
@@
 
-I have amended the introduction to include a note about alignment. 
-I believe it is out of scope to go into the finer details of when to use alignment or pseudo-alignment methods to obtain counts for RNA-seq data.
+>I have amended the introduction to include a note about alignment.
+It is out of scope to go into the finer details of when to use alignment or pseudo-alignment methods to obtain counts for RNA-seq data.
 I have included results from both approaches to highlight that **plyranges** is agnostic to the method used and is flexible for many types of data.
@@
 
-> As you introduced a “biological data science workflow” in this chapter and did a great job of walking the read through the wrangling, combining and visualizing components of the workflow, I was hoping you would touch on tools for communication too.
+As you introduced a “biological data science workflow” in this chapter and did a great job of walking the read through the wrangling, combining and visualizing components of the workflow, I was hoping you would touch on tools for communication too.
 I understand this is outside the scope of the thesis, but I still think it would be nice to at least mention work others have done in this area.
 
-I agree that it would be nice but it is out of scope for the thesis. Communication is extremely broad and I feel that including it here would distract from the focus of my thesis.
+> This work focuses on exploratory tools, sometimes people confuse graphics used for exploratory work and graphics used for communication.
+It is not in the scope of the thesis to include communication graphics, so I have not made changes.
+The distinction is already provided in Figure 1.1.
 
-> Pg 1: I thought it was strange that the words “figure” in e.g. “figure 1.1” were not capitalized. Would recommend changing to “Figure 1.1” throughout.
-> Pg 1: Recommend capitalizing the word “chapter” when referring to specific chapters e.g. “chapter 2”.
+Pg 1: I thought it was strange that the words “figure” in e.g. “figure 1.1” were not capitalized. Would recommend changing to “Figure 1.1” throughout.
+Pg 1: Recommend capitalizing the word “chapter” when referring to specific chapters e.g. “chapter 2”.
 
-Amended. I have made this consistent throughout the thesis now.
+> Amended. I have made this consistent throughout the thesis now.
 
-> Pg 2: Figure 1.1 has been modified to include “combine”, but the caption does not reflect what is meant by that.
+Pg 2: Figure 1.1 has been modified to include “combine”, but the caption does not reflect what is meant by that.
 Would recommend modifying caption to reflect importance of combining experimental assays together to gain biological insight.
 
-Amended.
+> Amended.
 
-> Pg 3: you introduce “range-based genomics data”, but it is not clear what is meant by that. More explanation would be helpful.
+Pg 3: you introduce “range-based genomics data”, but it is not clear what is meant by that. More explanation would be helpful.
 
-Amended. It now refers to data measured along the genome.
+> Amended. It now refers to data measured along the genome.
 
->Pg 4: In Section 1.2, could you add a few more references on existing methods for integration of genomic data and then re-emphasize the point that however existing methods have not thought much about the interoperability between the tidyverse and Bioconductor approaches?
+Pg 4: In Section 1.2, could you add a few more references on existing methods for integration of genomic data and then re-emphasize the point that however existing methods have not thought much about the interoperability between the tidyverse and Bioconductor approaches?
 
-I have expanded this section to reflect these comments.
+> I have expanded this section to reflect these comments.
 
@@
 
-This chapter has already been published and it appears most of the comments pertain to the use and learning of the **plyranges** package.
-I believe they are best addressed through the **plyranges** documentation and vignette.
-I have amended the thesis to include the package vignette as an appendix.
+This chapter has already been published and the comments pertain to the use and learning of the **plyranges** package.
+They are best addressed through the **plyranges** documentation and vignette.
+Consequently, I have amended the thesis to include the package vignette as an appendix.
 
@@
 
-> Almost all of Chapter 2 discusses the use of GRanges objects.
+Almost all of Chapter 2 discusses the use of GRanges objects.
 As most of Chapter 3 discusses the use of the SummarizedExperiment object, could you add a motivating figure at the start explaining the relationship between a SummarizedExperiment and GRanges object (i.e.
@@
 
-I agree and have amended this by adding a figure which explains what a `SummarizedExperiment`  is and how it is related to a `GRanges`.
+> I have amended this by adding a figure which explains what a `SummarizedExperiment`  is and how it is related to a `GRanges`.
 
-> Pg 31: Figure 3.1 is fairly low resolution. Is it possible to make a higher resolution version?
+Pg 31: Figure 3.1 is fairly low resolution. Is it possible to make a higher resolution version?
 
-Amended.
+> Amended.
 
-> Pg 31: Figure 3.1 demonstrates how you use plyranges to integrate the results from DE and DA (i.e.
-finding overlaps).
-Would it also make sense to the integration step prior to modeling?
-I am thinking
+Pg 31: Figure 3.1 demonstrates how you use plyranges to integrate the results from DE and DA (i.e. finding overlaps). Would it also make sense to the integration step prior to modelling? I am thinking
 about this as a way to reduce the amount of tests performed for DE and DA.
 
-I think it would make sense but it is out of scope for this paper.
+> No changes made. It would make sense but it is out of scope for this paper.
 
-> Pg 32: The fluentGenomics package needs a reference
+Pg 32: The fluentGenomics package needs a reference
 
-Amended.
+> Amended.
 
->Pg 39: Could you add 1-2 sentences (and maybe a reference) about what is a MA plot and why it’s
+Pg 39: Could you add 1-2 sentences (and maybe a reference) about what is a MA plot and why it’s
 useful for visualizing DE results?
 
-Amended.
+> Amended.
 
-> Pg 40: I think it would be useful if you could comment on the results() function a bit more to explain
+Pg 40: I think it would be useful if you could comment on the results() function a bit more to explain
 what is happening under the hood there.
@@
 
-Amended.
+> Amended.
 
->Pg 45: More motivation could be provided on why you want to generate random samples of size equal to the number of DE-genes from the other_genes with your larger question in mind of “how many DA peaks are near DE genes relative to”other" non-DE genes¿‘.
+Pg 45: More motivation could be provided on why you want to generate random samples of size equal to the number of DE-genes from the other_genes with your larger question in mind of “how many DA peaks are near DE genes relative to”other" non-DE genes¿‘.
 It is stated on pg 46”we minimize the variance on the enrichment statistics induced by the sampling process", but it might be good to explain this in greater detail at a higher level.
 
-I have added a few more sentences to this section which describes the motivation for selecting sets of the same size between non-DE and DE genes, and why we want to do sub-sampling.
+> I have added a few more sentences to this section which describes the motivation for selecting sets of the same size between non-DE and DE genes, and why we want to do sub-sampling.
 
@@
 
-> Much of this chapter depends on understanding what is meant by a coverage trace plot.
+Much of this chapter depends on understanding what is meant by a coverage trace plot.
 You do have examples (e.g.
@@
 It would be good to provide a motivating figure be added at the start to explain what is meant by “coverage” in this context?
-> Pg 63: Introduce what you mean by coverage traces and provide more motiviation why you would want this over just numerical summaries of the data.
 
-Amended.
+Pg 63: Introduce what you mean by coverage traces and provide more motivation why you would want this over just numerical summaries of the data.
 
-> Pg 66: You state “resulting GRanges in memory can be compressed using run-length encoding for any categorical variable”.
+> Amended. There is a now a motivating figure that explains how to interpret coverage scores.
+
+Pg 66: You state “resulting GRanges in memory can be compressed using run-length encoding for any categorical variable”.
 Could you give a numeric example of the amount of compression that can be obtained with RLEs?
@@
 
-Amended.
+> Amended.
 
-> Pg 68: Can you comment on why the GFF files from RefSeq were used?
+Pg 68: Can you comment on why the GFF files from RefSeq were used?
 
-The model organism is Zebrafish which has limited annotations available compared to human and mouse.
-I chose the RefSeq GFF build because that is what was recommended to me by our collaborators as the canonical reference.
+> I have amended this to comment that RefSeq is standard for zebrafish unlike mouse or human.
 
-> Pg 68: How did you handle reads that map over boundaries for introns and exons? or do you only keep the ones that are entirely inside an intron or exon?
+Pg 68: How did you handle reads that map over boundaries for introns and exons? or do you only keep the ones that are entirely inside an intron or exon?
 
-Reads that are split are across exon / intron boundaries are counted towards both categories.
+> I have amended this. Reads that are split are across exon / intron boundaries are counted towards both categories.
 
-> Pg 69: In the facets, you used a short descriptor, but could you write out what is each facet label is in the legend?
+Pg 69: In the facets, you used a short descriptor, but could you write out what is each facet label is in the legend?
 
-I think the facet labels are sufficient in conjunction with the figure captions.
+> No changes made. The facet labels need to be short and the details are provided in the figure captions and main text.
 
@@
 
->  One question I had as I was reading throughout this chapter, was what is your recommendation for users who want to calculate distances?
+One question I had as I was reading throughout this chapter, was what is your recommendation for users who want to calculate distances?
 can the liminal framework be used to better approximate distances instead of users who perform clustering directly on top of the tSNE or UMAP spaces?
@@
 
-This is a really interesting question.
-To begin, I have amended the clustering case study to comment on a more realistic approach to perform cluster verification.
-I like the idea of using the tour to calibrate distances obtained from an NLDR method (or the other way around).
-As mentioned in the text if there are (well-behaved) clusters in the data, it appears that t-SNE will find them but often messes up their orientation and topology.
-I think you would get some of the way there using the spatial linking approach I described.
-Another approach would be to look at distance preservation between successive tour frames.
-One could then ask, how often are these groups of points close according to the tour?
-Of interest would also be the manual clustering approaches described in @Cook2007-ld.
-This is left as future work.
+> This is a really interesting question.
+I have amended the clustering case study to provide more details on performing cluster verification.
+The suggestions for using the tour to calibrate distances obtained from an NLDR method (or the other way around) is something that will be incorporated into future work.
 
->  Pg 77: “non-liner” should be “non-linear”
+Pg 77: “non-liner” should be “non-linear”
 
-Amended.
+> Amended.
 
->  Pg 78: “produces” should be “produce”
+Pg 78: “produces” should be “produce”
 
-Amended.
+> Amended.
 
-> Pg 78: Could you add a reference for PCA?
+Pg 78: Could you add a reference for PCA?
 
-Amended.
+> Amended.
 
-> Pg 80 (also in 81): The way “tSNE” was written in “this essentially turns t-SNE” looks very different
+Pg 80 (also in 81): The way “tSNE” was written in “this essentially turns t-SNE” looks very different
 than previously written stylistically. Could this be converted to the way you were originally referring to
@@
 
-Amended.
+> Amended. I now use "t-SNE" throughout.
 
-> Pg 87: “inn” should be “in”
+Pg 87: “inn” should be “in”
 
-Amended.
+> Amended.
 
-> Pg 88: Inside Figure 5.1A and B, could you have liminal label which plot is which i.e. t-SNE vs tour
+Pg 88: Inside Figure 5.1A and B, could you have liminal label which plot is which i.e. t-SNE vs tour
 view. Also, could axis labels be added? Maybe also add this information to the Figure legend? – Pg 89 – same comment
 
-I think this is not immediately obvious in the static figure but is obvious in the video and when using the **liminal** package that the embedding view is always on the left, and the tour view is on the right (it will be the view that is animated).
-I think the suggestion of titling the views is a good one and will place that into the software in future.
-Regarding axis labels, I don't think they make sense for the tour since it's really a linear combination of multiple variables (there's an option to switch on a biplot view for the tour).
-I am also reluctant to include them for t-SNE because they can be misleading for the reasons mentioned in the chapter, however I do think this would be good to have as an option for **liminal** in future.
+> When using the **liminal** package it is obvious that the embedding view is always on the left, and the tour view is on the right (it will be the view that is animated).
+I have annotated the static images of the interface that appear in the paper to make this clear to the reader which view is which.
+Regarding axis labels, they don't make sense for the tour since it's really a linear combination of multiple variables (there's an option to switch on a biplot view for the tour).
+It is also not necessary to include them for t-SNE because they can be misleading for the reasons mentioned in the chapter.
 
-> Pg 90: I believe there should be a period after “tree structure data (figure 5.3)” instead of a comma
+Pg 90: I believe there should be a period after “tree structure data (figure 5.3)” instead of a comma
 
-Amended.
+> Amended.
 
-> Pg 94: Can you comment on the scRNA-seq example where you only pick first 5 PCs that explained only 20% of the variance.
+Pg 94: Can you comment on the scRNA-seq example where you only pick first 5 PCs that explained only 20% of the variance.
 Or more generally, what is the trade-off on the liminal view by picking more or less PCs along with a NLDR method.
@@
 
-I have amended section 5.5.3 to provide more details about why the first five components were selected, and how to choose variables to tour.
+> I have amended section 5.5.3 to provide more details about why the first five components were selected, and how to choose variables to tour.
 
-> Pg 94: Can you comment on the limits of the liminal view in terms of number of observations that
+Pg 94: Can you comment on the limits of the liminal view in terms of number of observations that
 one can visualize in a given tour? Have you thought about strategies for making this scale with more observations?
@@
 
-I have amended section 5.5.3 to discuss a strategy in the specific case of cluster analysis.
+> I have amended section 5.5.3 to discuss a strategy in the specific case of cluster analysis.
 I have also amended the discussion of the chapter with specific work that can mitigate the need to show all of the data.
-In the case of tasks performed for clustering I don't think it is useful or necessary to show all points in the scatter plot when we are most interested in density and shapes.
-There is interesting work to be done here in future for appropriate sampling strategies or aggregations that could be done prior to producing a visualisation.
+In the case of tasks performed for clustering it is not necessarily useful to show all points in the scatter plot when we are most interested in density and shapes.
+In future, I will incorporate appropriate sampling strategies or aggregations that could be done prior to producing a visualisation.
 
@@
 
-> Pg 102: Instead of https://bioconductor.org/packages/release/bioc/html/plyranges.html, you can always use the short version https://bioconductor.org/packages/plyranges which points to the release version.
+Pg 102: Instead of https://bioconductor.org/packages/release/bioc/html/plyranges.html, you can always use the short version https://bioconductor.org/packages/plyranges which points to the release version.
 
-Amended.
+> Amended.
 

### 2020-10-29 Rmd/00-c-declaration.Rmd
update copyright

@@
 
-```{r signature, out.width = "20%", fig.align = "left", eval = file.exists("img/signature.png")}
+```{r signature, out.width = "20%", fig.align = "left", eval = FALSE}
 knitr::include_graphics("img/signature.png")
@@
 
-\textbf{Date:} 05/08/2020
-
+\textbf{Date:} 29/10/2020
+<!--
 I hereby certify that the above declaration correctly reflects the nature and extent of the student’s and co-authors’ contributions to this work.
@@
 
+-->
+
 `r if (knitr::is_html_output()) '-->'`
